---
title: "Data Import and Quality Control"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction
This script imports, cleans and formats raw data. 
The output is saved for downstream scripts to use.


# Setup
`r what_this_does("Loaded required utility packages (tidyverse, data.table, readxl, matrixStats).", "Established dynamic output directory paths to safely save data and figures.", "Configured default knitr chunk options to suppress unnecessary warnings.")`

```{r setup, message=FALSE, warning=FALSE}
# Utility Packages
library(tidyverse)   # Data manipulation
library(readxl)      # Reading messy Excel files
library(data.table)  # High-performance data wrangling
library(here)        # Safe file paths
library(matrixStats) # Fast matrix math for QC and Imputation

# Get current file name to make folder
current_file <- "01_data_import_and_qc"

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Configuration

`r what_this_does("Defined strict quality control missingness thresholds for proteomics filtering (5% max missingness).", "Set the clinical variable completeness threshold (variables must be 80% complete in patients to survive).", "Established core reference dates for accurate feature engineering.", "Listed protected variables and explicit date columns to prevent accidental deletion during automated pruning.", "Defined regex patterns to globally purge procedural meta-variables.", "Established a global renaming dictionary.", "Parameterized all raw data file names to avoid hardcoding deep in the script.")`

```{r config}
# 1. Global File Paths ---------------------------------------------------------
file_paths <- list(
  clinical_raw  = "Patients__20022025_105625_modified.xlsx",
  id_map        = "Samples for OLink submitted_final_0226.xlsx",
  clinical_dict = "clinical_dictionary.txt",
  olink_raw     = "pre_quant.csv"
)

# 2. Global QC Thresholds & Column Names ---------------------------------------
qc_config <- list(
  max_missing_pct_prot = 0.05,
  max_missing_pct_pat  = 0.05,
  selected_col         = "SELECTED", 
  
  min_completeness_clin_var = 0.80, 
  
  exclude_from_clean = c("Subject_ID", "inCentraxx", "Entnahmedatum", 
                         "Einlagerungsdatum", "Entnahmedatum_Clean", "Visit Date"),
  
  date_cols            = c("Visit Date", "Entnahmedatum", "Einlagerungsdatum", 
                           "If yes, Date of HRCT", "onset_nonRP", "Date of PFT", 
                           "If yes, Date of assessment for ECG", "If yes, Date of assessment for ECHO",
                           "If, yes Date of assessment for Walk test"),
                           
  meta_drop_patterns   = c("performed", "measured", "done", "velocity\\?", "ratio\\?", "area\\?", "excursion\\?", "fraction %\\?", "done\\?"),
  
  olink_date           = as.Date("2026-03-30")
)

# 3. Global Variable Renaming Map ----------------------------------------------
clinical_rename_map <- c(
  "Sclerodactyly" = "of the fingers as sclerodactyly (distal to the metacarpophalangeal joints but proximal to the proximal interphalangeal joints)",
  "SSc Antibody Score" = "Score: Scleroderma related antibodies (any of anti-centromere, Anti-topoisomerase I [ anti-SCL 70], anti-RNA polymerase III",
  "SSc Antibodies" = "Scleroderma related antibodies (any of anti-centromere, Anti-topoisomerase I [ anti-SCL 70], anti-RNA polymerase III",
  "Skin Thickening (Ext. MCP)" = "of the fingers of both hands extending proximally to the metacarpophalangeal joints",
  "Skin Thickening (Prox. MCP)" = "Skin thickening of the fingers of both hands extending proximal to the MCP joints",
  "New PAH" = "Does the patient have pulmonary arterial hypertension since the last visit?",
  "PAH/ILD" = "Pulmonary arterial hypertension and/or Interstitial lung Disease",
  "Max Borg Dyspnea" = "Worst modified Borg dyspnea score during the test (from 0 to 10)",
  "New Cardiac Manifest." = "Does the patient have cardiac manifestation since last visit?",
  "Diastolic Dysf. (Echo)" = "Diastolic function abnormal(on echo, E/a less than 10cm/sec)",
  "MD Global Activity (NA)" = "Physicians global assessment of disease activity VAS Unknown",
  "Pt Global Activity (NA)" = "Patients global assessment of disease activity VAS Unknown",
  "mRSS Worsening (1yr)" = "Was there a worsening of mRSS over the last 12 months?",
  "Intestinal Symptoms" = "Intestinal symptoms (diarrhea, bloating, constipation)",
  "Arthritis Activity (NA)" = "Activity of arthritis during the past week Unknown",
  "EULAR/ACR 2013 Criteria" = "2013 EULAR/ACR classification criteria fulfilled",
  "TAPSE (cm)" = "Tricuspid annular plane systolic excursion (cm)",
  "Stomach Symptoms" = "Stomach symptoms (early satiety, vomiting)",
  "FVC %pred" = "Forced Vital Capacity (FVC- % predicted)",
  "Capillaroscopy Score" = "Score: Abnormal nailfold capillaroscopy",
  "TLC %pred" = "Total lung capacity (TLC- % predicted)",
  "GAVE" = "Gastric Antral Vascular Ectasia (GAVE)",
  "LVEF %" = "Left ventricular ejection fraction %",
  "Min SpO2 (Exercise)" = "Worst O2-saturation at exercise (%)",
  "Capillaroscopy Pattern" = "Capillaroscopy scleroderma pattern",
  "Joint Contractures" = "Joint contractures (at any joint)",
  "Abnormal Capillaroscopy" = "Abnormal nailfold capillaroscopy",
  "Non-Raynaud Symptom" = "Any non-Raynaud symptom present?",
  "Puffy Fingers" = "of the fingers as puffy fingers"
)
```

# Clinical Data Cleanup
`r analysis_decision("Dropped the follow-up sample 'Act_4_2' to strictly enforce a cross-sectional study design.", "Excluded patients who were missing from either the CentraXX output or the Biobank sample manifest.")`

`r what_this_does("Imported raw clinical and sample mapping Excel files and resolved ID discrepancies.", "Engineered core analytical features including Age, Sample_Age, cohort_group, and Sex.", "Calculated time deltas (in years) between clinical assessment dates and the blood draw date.", "Pruned clinical variables according to the defined 80% completeness filter.", "Enforced strict data typing and dropped manually excluded variables using the clinical_dictionary.", "Dynamically dropped procedural meta-variables based on parameterized config patterns.", "Standardized factor levels: converted 'UN' and 'Unknown' to NA.", "Finalized downstream features (renamed long columns, fixed illogical CRP values, locked factor levels).")`

```{r clin_cleanup}
audit_log <- c("Importing Clinical Data and ID Maps...")


# 1. Process Master Clinical File (Sheet 1 only) -------------------------------
raw_clin <- read_excel(here::here("data", file_paths$clinical_raw), sheet = 1)
setDT(raw_clin) 

# Force the clinical ID to be a character string for safe merging
raw_clin <- raw_clin[, inCentraxx := as.character(inCentraxx)]

# Strict Filter: Keep ONLY samples explicitly marked for this study
raw_clin <- raw_clin[get(qc_config$selected_col) == 1 | get(qc_config$selected_col) == "1"]


# 1.5 Fix Excel Numeric Dates (DIRECT PARSING) ---------------------------------
valid_date_cols <- intersect(qc_config$date_cols, names(raw_clin))

if(length(valid_date_cols) > 0) {
  raw_clin <- raw_clin[, (valid_date_cols) := lapply(.SD, function(x) {
    suppressWarnings(as.Date(as.numeric(x), origin = "1899-12-30"))
  }), .SDcols = valid_date_cols]
}


# 2. Process Mapping File & Coalesce IDs ---------------------------------------
id_map <- read_excel(here::here("data", file_paths$id_map))
setDT(id_map)

# Explicitly drop the follow-up sample to enforce cross-sectional design
id_map <- id_map[`Subject ID` != "Act_4_2"]

id_map <- id_map[, `Subject-ID (EUSTAR)`      := as.character(`Subject-ID (EUSTAR)`)]
id_map <- id_map[, `Subject-ID (Biobank RUZ)` := as.character(`Subject-ID (Biobank RUZ)`)]

# COALESCE: Create a master 'Merge_ID'
id_map <- id_map[, Merge_ID := fifelse(
  is.na(`Subject-ID (EUSTAR)`) | `Subject-ID (EUSTAR)` == "", 
  `Subject-ID (Biobank RUZ)`, 
  `Subject-ID (EUSTAR)`
)]

id_map <- id_map[!is.na(Merge_ID) & Merge_ID != ""]


# 3. Merge and Structure -------------------------------------------------------
master_spine <- merge(
  raw_clin, 
  id_map, 
  by.x = "inCentraxx", 
  by.y = "Merge_ID", 
  all = FALSE
)

# SANITY CHECK: Ensure the merge actually yielded data
stopifnot("FATAL: Inner join resulted in 0 rows. Check ID formatting." = nrow(master_spine) > 0)
audit_log <- c(audit_log, "- Sanity Check: Inner join successful.")

setnames(master_spine, "Right atrium area (cm²)", "Right atrium area (cm2)", skip_absent = TRUE)
setnames(master_spine, "Right ventricular area (cm²)", "Right ventricular area (cm2)", skip_absent = TRUE)


# 4. Feature Engineering (Groups, Dates, Age, Sex) -----------------------------
master_spine <- setnames(master_spine, old = "Subject ID", new = "Subject_ID", skip_absent = TRUE)

master_spine <- master_spine[, cohort_group := fcase(
  str_detect(Subject_ID, "^Nonact"), GRP_NON_ACTIVE,
  str_detect(Subject_ID, "^Act"),    GRP_ACTIVE,
  str_detect(Subject_ID, "^Ctrl"),   GRP_CONTROL,
  default = NA_character_
)]

master_spine <- master_spine[, Entnahmedatum_Clean := as.Date(Entnahmedatum)]

master_spine <- master_spine[, Birth_Year := as.numeric(trimws(Geburtsdatum))]
master_spine <- master_spine[, Age := year(Entnahmedatum_Clean) - Birth_Year]
master_spine <- master_spine[, Sample_Age := as.numeric(difftime(qc_config$olink_date, Entnahmedatum_Clean, units = "days")) / 365.25]

master_spine <- master_spine[, Sex := fcase(
  Geschlecht == "weiblich", SEX_FEMALE,
  Geschlecht == "männlich", SEX_MALE,
  default = NA_character_
)]


# 4.5 Feature Engineering (Clinical Date Deltas) -------------------------------
assessment_dates <- setdiff(qc_config$date_cols, c("Visit Date", "Entnahmedatum", "Einlagerungsdatum"))

for (d_col in assessment_dates) {
  if (d_col %in% names(master_spine)) {
    clean_prefix <- gsub(".*Date of assessment for |.*Date of ", "", d_col)
    clean_name <- paste0("Delta_", gsub("[^A-Za-z0-9]", "_", clean_prefix), "_Years")
    master_spine[, (clean_name) := as.numeric(difftime(get(d_col), Entnahmedatum_Clean, units = "days")) / 365.25]
  }
}


# 5. Completeness Pruning & Dictionary Enforcement -----------------------------
audit_log <- c(audit_log, "\nFiltering by completeness and applying TSV Data Dictionary...")

protected_vars <- c("Subject_ID", "inCentraxx", "cohort_group", "Age", "Sex", "Sample_Age", qc_config$date_cols)

completeness <- master_spine[cohort_group %in% c(GRP_ACTIVE, GRP_NON_ACTIVE), 
                             lapply(.SD, function(x) sum(!is.na(x)) / .N)]

cols_to_keep <- names(completeness)[completeness >= qc_config$min_completeness_clin_var | names(completeness) %in% protected_vars]

n_dropped <- ncol(master_spine) - length(cols_to_keep)
master_spine <- master_spine[, ..cols_to_keep]
audit_log <- c(audit_log, sprintf("- Pruning: Dropped %d columns for having < %s%% completeness in patients.", 
            n_dropped, qc_config$min_completeness_clin_var * 100))

dict_path <- here::here("data", file_paths$clinical_dict)
if (file.exists(dict_path)) {
  data_dict <- readr::read_tsv(dict_path, show_col_types = FALSE)
  
  vars_n <- intersect(data_dict %>% filter(Type == "n") %>% pull(Variable_Name), names(master_spine))
  vars_c <- intersect(data_dict %>% filter(Type == "c") %>% pull(Variable_Name), names(master_spine))
  vars_x <- intersect(data_dict %>% filter(Type == "x") %>% pull(Variable_Name), names(master_spine))
  
  if (length(vars_x) > 0) master_spine[, (vars_x) := NULL]
  
  safe_numeric <- function(x) {
    if (is.numeric(x)) return(x)
    clean_x <- gsub(",", ".", trimws(as.character(x)))
    suppressWarnings(as.numeric(clean_x))
  }
  if (length(vars_n) > 0) master_spine[, (vars_n) := lapply(.SD, safe_numeric), .SDcols = vars_n]
  if (length(vars_c) > 0) master_spine[, (vars_c) := lapply(.SD, function(x) as.factor(as.character(x))), .SDcols = vars_c]
  
  audit_log <- c(audit_log, sprintf("- Enforcement: Cleaned %d Numeric and %d Categorical variables based on dictionary.", length(vars_n), length(vars_c)))
  audit_log <- c(audit_log, sprintf("- Exclusion: Dropped %d manually blacklisted variables.", length(vars_x)))
} else {
  audit_log <- c(audit_log, "! WARNING: 'clinical_dictionary.txt' not found in data/. Skipping explicit type enforcement.")
}


# 5.5 Meta-Variable Pruning (Dropping Measurement Indicators) ------------------
audit_log <- c(audit_log, "\nPruning procedural meta-variables (measurement indicators)...")

regex_pattern <- paste(tolower(qc_config$meta_drop_patterns), collapse = "|")
cols_to_drop <- names(master_spine)[str_detect(tolower(names(master_spine)), regex_pattern)]

if (length(cols_to_drop) > 0) {
  master_spine <- master_spine %>% select(-all_of(cols_to_drop))
  audit_log <- c(audit_log, sprintf("- Dropped %d procedural variables to prevent analytical confounding.", length(cols_to_drop)))
}


# 6. Factor Standardization (Binarization & Cleaning) --------------------------
audit_log <- c(audit_log, "\nStandardizing factor variables (Yes/No -> 1/0, UN/Unknown -> NA)...")

master_spine <- master_spine %>%
  mutate(across(where(is.factor), ~ {
    char_vec <- as.character(.x)
    lower_vec <- tolower(char_vec)
    
    char_vec[lower_vec %in% c("un", "unknown", "not done")] <- NA
    char_vec[lower_vec == "yes"] <- "1"
    char_vec[lower_vec == "no"] <- "0"
    
    as.factor(char_vec)
  }))
setDT(master_spine)


# 7. Final Downstream Feature Engineering --------------------------------------
audit_log <- c(audit_log, "\nApplying final clinical feature engineering (Renaming, Factor levels)...")

setnames(master_spine, 
         old = as.character(clinical_rename_map), 
         new = names(clinical_rename_map), 
         skip_absent = TRUE)

setnames(master_spine, 
         old = c("OR serum creatinine (micromoles/L)", "CRP (mg/dl)"), 
         new = c("Creatinine", "CRP"), 
         skip_absent = TRUE)

if("CRP" %in% names(master_spine)) {
  master_spine[, CRP := as.numeric(gsub(",", ".", as.character(CRP)))]
  master_spine[CRP <= 0, CRP := NA_real_]
}

if("Creatinine" %in% names(master_spine)) {
  master_spine[, Creatinine := as.numeric(gsub(",", ".", as.character(Creatinine)))]
}

master_spine[, cohort_group := factor(cohort_group, levels = c(GRP_ACTIVE, GRP_NON_ACTIVE, GRP_CONTROL))]

core_cols <- intersect(c("Subject_ID", "cohort_group", "Age", "Sex", "Sample_Age"), names(master_spine))
setcolorder(master_spine, core_cols)

audit_log <- c(audit_log, sprintf("\n✓ Clinical Spine Ready: %d rows x %d columns.", nrow(master_spine), ncol(master_spine)))

rm(raw_clin, id_map)
invisible(gc())

# Print consolidated output to prevent HTML fragmentation
cat(paste(audit_log, collapse = "\n"))
```

# OLINK Data
## OLINK Import

`r what_this_does("Pre-quantified Olink data was pivoted from long to wide format (Samples as columns, Assays as rows).", "The Olink matrix was strictly subset to include only samples definitively present in the clinical master_spine.")`

```{r olink_import}
audit_log_olink <- c("Importing and Pivoting Olink Data...")

raw_olink <- fread(here::here("data", file_paths$olink_raw))

raw_olink <- raw_olink[SampleID %in% master_spine$Subject_ID]

wide_olink <- dcast(raw_olink, Assay ~ SampleID, value.var = "NPX")

olink_mat <- as.matrix(wide_olink[, -1, with = FALSE])
rownames(olink_mat) <- wide_olink$Assay

audit_log_olink <- c(audit_log_olink, sprintf("- Initial Matrix: %d Proteins | %d Samples", nrow(olink_mat), ncol(olink_mat)))

rm(raw_olink, wide_olink)
invisible(gc())

cat(paste(audit_log_olink, collapse = "\n"))
```


## OLINK QC, Imputation & Sync

`r analysis_decision("Dropped proteins and samples exceeding the 5% missingness threshold (max_missing_pct_prot and max_missing_pct_pat) to ensure robust downstream imputation.")`

`r what_this_does("Proteins and Samples exceeding the previously defined missingness threshold were entirely dropped.", "Remaining NA values were imputed using a row-wise (protein-wise) median approach.", "The master_spine clinical dataset and the Olink matrix were cross-synchronized to guarantee identical sample inclusion and ordering.")`

```{r qc_imputation_sync}
cat("Running Missingness QC and Imputation...\n")

audit_log_qc <- c("Running Missingness QC and Imputation...")

miss_prots <- rowMeans(is.na(olink_mat))
miss_samps <- colMeans(is.na(olink_mat))

valid_prots <- miss_prots < qc_config$max_missing_pct_prot
valid_samps <- miss_samps < qc_config$max_missing_pct_pat

failed_prots <- rownames(olink_mat)[!valid_prots]
failed_samps <- colnames(olink_mat)[!valid_samps]

if(length(failed_prots) > 0) {
  audit_log_qc <- c(audit_log_qc, sprintf("! Dropped %d Proteins (>= %s%% missing):\n  %s", 
              length(failed_prots), qc_config$max_missing_pct_prot * 100, paste(failed_prots, collapse = ", ")))
} else {
  audit_log_qc <- c(audit_log_qc, sprintf("✓ All proteins passed QC (< %s%% missing).", qc_config$max_missing_pct_prot * 100))
}

if(length(failed_samps) > 0) {
  audit_log_qc <- c(audit_log_qc, sprintf("! Dropped %d Patients (>= %s%% missing):\n  %s", 
              length(failed_samps), qc_config$max_missing_pct_pat * 100, paste(failed_samps, collapse = ", ")))
} else {
  audit_log_qc <- c(audit_log_qc, sprintf("✓ All patients passed QC (< %s%% missing).", qc_config$max_missing_pct_pat * 100))
}

qc_mat <- olink_mat[valid_prots, valid_samps, drop = FALSE]
audit_log_qc <- c(audit_log_qc, sprintf("\n- Post-QC Matrix: %d Proteins | %d Samples", nrow(qc_mat), ncol(qc_mat)))

row_meds <- matrixStats::rowMedians(qc_mat, na.rm = TRUE)
na_coords <- which(is.na(qc_mat), arr.ind = TRUE)
qc_mat[na_coords] <- row_meds[na_coords[, "row"]]

master_spine <- master_spine[Subject_ID %in% colnames(qc_mat)]

qc_mat <- qc_mat[, master_spine$Subject_ID, drop = FALSE]

cat(paste(audit_log_qc, collapse = "\n"))
```
# Export 

`r what_this_does("The finalized arrays (clinical_spine.rds and clean_olink_matrix.rds) were exported as serialized R objects for quick, standardized ingestion by downstream scripts.", "The clinical spine was exported as .csv")`

```{r export}
saveRDS(master_spine, file.path(output_dir_data, "clinical_spine.rds"))
saveRDS(qc_mat, file.path(output_dir_data, "clean_olink_matrix.rds"))

write_csv(master_spine, file.path(output_dir_data, "clinical_spine.csv"))

cat("Script 01 Complete! Data is ready for analysis.")
```


