---
title: "Data Import and Quality Control"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction
This script imports, cleans and formats raw data. 
The output is saved for downstream scripts to use.


# Setup
`r what_this_does("Loaded required utility packages (tidyverse, data.table, readxl, matrixStats).", "Established dynamic output directory paths to safely save data and figures.", "Sourced global project configurations, color palettes, and standard themes.", "Configured default knitr chunk options to suppress unnecessary warnings.")`

```{r setup, message=FALSE, warning=FALSE}
# Utility Packages
library(tidyverse)   # Data manipulation
library(readxl)      # Reading messy Excel files
library(data.table)  # High-performance data wrangling
library(here)        # Safe file paths
library(matrixStats) # Fast matrix math for QC and Imputation

# Get current file name to make folder
current_file <- "01_data_import_and_qc"

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

# Source global formatting (colors, themes, seed)
source(here::here("code", "00_config.R"))

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

```

# Configuration

`r what_this_does("Defined strict quality control missingness thresholds for proteomics filtering (5% max missingness).", "Set the clinical variable completeness threshold (variables must be 80% complete in patients to survive).", "Established core reference dates for accurate feature engineering (Olink run date, default fallback control date).", "Listed protected variables and explicit date columns to prevent accidental deletion during automated pruning.")`

```{r config}
# Define thresholds and critical column names
qc_config <- list(
  max_missing_pct_prot = 0.05,
  max_missing_pct_pat  = 0.05,
  selected_col         = "SELECTED", # Column flagging study inclusion
  
  # Completeness Threshold (Variables must be at least this complete to survive)
  min_completeness_clin_var = 0.80, 
  
  # Protection list (IDs and Dates that should never be dropped or mutated automatically)
  exclude_from_clean = c("Subject_ID", "inCentraxx", "Entnahmedatum", 
                         "Einlagerungsdatum", "Entnahmedatum_Clean", "Visit Date"),
  
  # Replace these with the EXACT column names from your Excel file that need date conversion
  date_cols            = c("Visit Date", "Entnahmedatum", "Einlagerungsdatum", 
                           "If yes, Date of HRCT", "onset_nonRP", "Date of PFT", 
                           "If yes, Date of assessment for ECG", "If yes, Date of assessment for ECHO",
                           "If, yes Date of assessment for Walk test"),
  
  # REFERENCE DATES (Required for feature engineering)
  olink_date           = as.Date("2026-03-30"), 
  default_control_date = as.Date("2024-01-01") 
)
  
```

## Clinical Data Cleanup
# Data Ingestion and Cleanup

`r what_this_does("Imported raw clinical and sample mapping Excel files and resolved ID discrepancies.", "Engineered core analytical features including Age, Sample_Age, cohort_group, and Sex.", "Pruned clinical variables according to the defined 80% completeness filter.", "Enforced strict data typing and dropped manually excluded variables using the clinical_dictionary.", "Standardized factor levels: converted 'UN' and 'Unknown' to NA.", "Converted 'Yes' to 1 and 'No' to 0 (retained explicitly as factors to ensure mathematical compatibility with downstream models).")`

```{r clin_cleanup}
cat("Importing Clinical Data and ID Maps...\n")

# ------------------------------------------------------------------------------
# 1. Process Master Clinical File (Sheet 1 only)
# ------------------------------------------------------------------------------
raw_clin <- read_excel(here::here("data", "Patients__20022025_105625_modified.xlsx"), sheet = 1)
setDT(raw_clin) 

# Force the clinical ID to be a character string for safe merging (Silenced)
raw_clin <- raw_clin[, inCentraxx := as.character(inCentraxx)]

# Strict Filter: Keep ONLY samples explicitly marked for this study
raw_clin <- raw_clin[get(qc_config$selected_col) == 1 | get(qc_config$selected_col) == "1"]

# ------------------------------------------------------------------------------
# 1.5 Fix Excel Numeric Dates (DIRECT PARSING)
# ------------------------------------------------------------------------------
# Safely target only the date columns that actually exist in the file
valid_date_cols <- intersect(qc_config$date_cols, names(raw_clin))

if(length(valid_date_cols) > 0) {
  raw_clin <- raw_clin[, (valid_date_cols) := lapply(.SD, function(x) {
    # The data is exactly "45601" (character). Convert to numeric, then safely to Date.
    suppressWarnings(as.Date(as.numeric(x), origin = "1899-12-30"))
  }), .SDcols = valid_date_cols]
}

# ------------------------------------------------------------------------------
# 2. Process Mapping File & Coalesce IDs
# ------------------------------------------------------------------------------
id_map <- read_excel(here::here("data", "Samples for OLink submitted_final_0226.xlsx"))
setDT(id_map)

# Explicitly drop the follow-up sample to enforce cross-sectional design
id_map <- id_map[`Subject ID` != "Act_4_2"]

# Force both ID columns to character (Silenced)
id_map <- id_map[, `Subject-ID (EUSTAR)`      := as.character(`Subject-ID (EUSTAR)`)]
id_map <- id_map[, `Subject-ID (Biobank RUZ)` := as.character(`Subject-ID (Biobank RUZ)`)]

# COALESCE: Create a master 'Merge_ID' (Silenced)
id_map <- id_map[, Merge_ID := fifelse(
  is.na(`Subject-ID (EUSTAR)`) | `Subject-ID (EUSTAR)` == "", 
  `Subject-ID (Biobank RUZ)`, 
  `Subject-ID (EUSTAR)`
)]

# Drop rows from the mapping file that have no ID in either column
id_map <- id_map[!is.na(Merge_ID) & Merge_ID != ""]

# ------------------------------------------------------------------------------
# 3. Merge and Structure
# ------------------------------------------------------------------------------
# Inner join: keeps only participants present in BOTH the clinical and mapping files
master_spine <- merge(
  raw_clin, 
  id_map, 
  by.x = "inCentraxx", 
  by.y = "Merge_ID", 
  all = FALSE
)

# Sanitize special characters in column names to prevent encoding mismatches
setnames(master_spine, "Right atrium area (cm²)", "Right atrium area (cm2)", skip_absent = TRUE)
setnames(master_spine, "Right ventricular area (cm²)", "Right ventricular area (cm2)", skip_absent = TRUE)

# ------------------------------------------------------------------------------
# 4. Feature Engineering (Groups, Dates, Age, Sex)
# ------------------------------------------------------------------------------
master_spine <- setnames(master_spine, old = "Subject ID", new = "Subject_ID", skip_absent = TRUE)

# Assign Groups
master_spine <- master_spine[, cohort_group := fcase(
  str_detect(Subject_ID, "^Nonact"), GRP_NON_ACTIVE,
  str_detect(Subject_ID, "^Act"),    GRP_ACTIVE,
  str_detect(Subject_ID, "^Ctrl"),   GRP_CONTROL,
  default = NA_character_
)]

# The Date Fix & Age Calculation
master_spine <- master_spine[, Entnahmedatum_Clean := as.Date(Entnahmedatum)]
master_spine <- master_spine[is.na(Entnahmedatum_Clean) & cohort_group == GRP_CONTROL, 
                             Entnahmedatum_Clean := as.Date("2024-01-01")]

master_spine <- master_spine[, Birth_Year := as.numeric(trimws(Geburtsdatum))]
master_spine <- master_spine[, Age := year(Entnahmedatum_Clean) - Birth_Year]
master_spine <- master_spine[, Sample_Age := as.numeric(difftime(qc_config$olink_date, Entnahmedatum_Clean, units = "days")) / 365.25]

# Assign Sex
master_spine <- master_spine[, Sex := fcase(
  Geschlecht == "weiblich", SEX_FEMALE,
  Geschlecht == "männlich", SEX_MALE,
  default = NA_character_
)]

# ------------------------------------------------------------------------------
# 4.5 Feature Engineering (Clinical Date Deltas)
# ------------------------------------------------------------------------------
# Isolate the specific assessment dates (ignoring the core visit/blood draw dates)
assessment_dates <- setdiff(qc_config$date_cols, c("Visit Date", "Entnahmedatum", "Einlagerungsdatum"))

for (d_col in assessment_dates) {
  if (d_col %in% names(master_spine)) {
    # Generate a clean name: Strips out the messy "If yes, Date of assessment for..." prefixes
    clean_prefix <- gsub(".*Date of assessment for |.*Date of ", "", d_col)
    clean_name <- paste0("Delta_", gsub("[^A-Za-z0-9]", "_", clean_prefix), "_Years")
    
    # Calculate Time Delta (Test Date - Blood Draw Date)
    # Negative means test happened BEFORE the blood draw. Positive means AFTER.
    master_spine[, (clean_name) := as.numeric(difftime(get(d_col), Entnahmedatum_Clean, units = "days")) / 365.25]
  }
}

# ------------------------------------------------------------------------------
# 5. Completeness Pruning & Dictionary Enforcement
# ------------------------------------------------------------------------------
cat("\nFiltering by completeness and applying TSV Data Dictionary...\n")

# A. COMPLETENESS PRUNING
# Protect core variables and dates from being dropped due to missingness
protected_vars <- c("Subject_ID", "inCentraxx", "cohort_group", "Age", "Sex", "Sample_Age", qc_config$date_cols)

# Calculate completeness ONLY on Patients (ignore Controls for this threshold)
completeness <- master_spine[cohort_group %in% c(GRP_ACTIVE, GRP_NON_ACTIVE), 
                             lapply(.SD, function(x) sum(!is.na(x)) / .N)]

cols_to_keep <- names(completeness)[completeness >= qc_config$min_completeness_clin_var | names(completeness) %in% protected_vars]

n_dropped <- ncol(master_spine) - length(cols_to_keep)
master_spine <- master_spine[, ..cols_to_keep]
cat(sprintf("- Pruning: Dropped %d columns for having < %s%% completeness in patients.\n", 
            n_dropped, qc_config$min_completeness_clin_var * 100))

# B. DICTIONARY ENFORCEMENT
dict_path <- here::here("data", "clinical_dictionary.txt")
if (file.exists(dict_path)) {
  data_dict <- readr::read_tsv(dict_path, show_col_types = FALSE)
  
  # Extract classification lists
  vars_n <- data_dict %>% filter(Type == "n") %>% pull(Variable_Name)
  vars_c <- data_dict %>% filter(Type == "c") %>% pull(Variable_Name)
  vars_x <- data_dict %>% filter(Type == "x") %>% pull(Variable_Name)
  
  # Only attempt to mutate columns that survived the completeness prune
  vars_n <- intersect(vars_n, names(master_spine))
  vars_c <- intersect(vars_c, names(master_spine))
  vars_x <- intersect(vars_x, names(master_spine))
  
  # 1. Drop explicitly excluded columns
  if (length(vars_x) > 0) {
    master_spine[, (vars_x) := NULL]
  }
  
  # 2. Force Numerics (Handles Swiss decimal commas, converts text to NA safely)
  safe_numeric <- function(x) {
    if (is.numeric(x)) return(x)
    clean_x <- gsub(",", ".", trimws(as.character(x)))
    suppressWarnings(as.numeric(clean_x))
  }
  if (length(vars_n) > 0) {
    master_spine[, (vars_n) := lapply(.SD, safe_numeric), .SDcols = vars_n]
  }
  
  # 3. Force Categoricals
  if (length(vars_c) > 0) {
    master_spine[, (vars_c) := lapply(.SD, function(x) as.factor(as.character(x))), .SDcols = vars_c]
  }
  
  cat(sprintf("- Enforcement: Cleaned %d Numeric and %d Categorical variables based on dictionary.\n", 
              length(vars_n), length(vars_c)))
  cat(sprintf("- Exclusion: Dropped %d manually blacklisted variables.\n", length(vars_x)))
  
} else {
  cat("! WARNING: 'clinical_dictionary.txt' not found in data/. Skipping explicit type enforcement.\n")
}

# Reorder core columns to the front of the dataframe for easy viewing
core_cols <- intersect(c("Subject_ID", "cohort_group", "Age", "Sex", "Sample_Age"), names(master_spine))
setcolorder(master_spine, core_cols)

cat(sprintf("\n✓ Clinical Spine Ready: %d rows x %d columns.\n", nrow(master_spine), ncol(master_spine)))
print(table(master_spine$cohort_group))


# ------------------------------------------------------------------------------
# 6. Factor Standardization (Binarization & Cleaning)
# ------------------------------------------------------------------------------
cat("\nStandardizing factor variables (Yes/No -> 1/0, UN/Unknown -> NA)...\n")

master_spine <- master_spine %>%
  mutate(across(where(is.factor), ~ {
    char_vec <- as.character(.x)
    lower_vec <- tolower(char_vec)
    
    # Apply standardizations (case insensitive)
    char_vec[lower_vec %in% c("un", "unknown", "not done")] <- NA
    char_vec[lower_vec == "yes"] <- "1"
    char_vec[lower_vec == "no"] <- "0"
    
    # BEST PRACTICE: Keep everything as a factor for the VPA engine.
    # We will temporarily cast to numeric later inside the correlation engine.
    as.factor(char_vec)
  }))
rm(raw_clin, id_map)
invisible(gc())
```

## OLINK import

`r what_this_does("Pre-quantified Olink data was pivoted from long to wide format (Samples as columns, Assays as rows).", "The Olink matrix was strictly subset to include only samples definitively present in the clinical master_spine.")`

```{r olink_import}
cat("Importing and Pivoting Olink Data...\n")

# Load Long-format Olink Data (Update filename if necessary)
raw_olink <- fread(here::here("data", "pre_quant.csv"))

# SubjectID is the identifier in the master_spine
# SampleID is the identifier in the OLINK file
# They both contain the same information
raw_olink <- raw_olink[SampleID %in% master_spine$Subject_ID]

# Pivot Long to Wide (Rows = Proteins, Cols = Patients)
# Assuming "Assay" is the protein name and "NPX" is the read value
wide_olink <- dcast(raw_olink, Assay ~ SampleID, value.var = "NPX")

# Convert to standard numeric matrix
olink_mat <- as.matrix(wide_olink[, -1, with = FALSE])
rownames(olink_mat) <- wide_olink$Assay

cat(sprintf("- Initial Matrix: %d Proteins | %d Samples\n", nrow(olink_mat), ncol(olink_mat)))

# Clean up RAM
rm(raw_olink, wide_olink)
invisible(gc())
```


## Proteomics QC, Imputation, Sync

`r what_this_does("Proteins and Samples exceeding the previously defined missingness threshold were entirely dropped.", "Remaining NA values were imputed using a row-wise (protein-wise) median approach.", "The master_spine clinical dataset and the Olink matrix were cross-synchronized to guarantee identical sample inclusion and ordering.")`

```{r qc_imputation_sync}
cat("Running Missingness QC and Imputation...\n")

# 1. Quality Control (Missingness Rules from qc_config)
miss_prots <- rowMeans(is.na(olink_mat))
miss_samps <- colMeans(is.na(olink_mat))

valid_prots <- miss_prots < qc_config$max_missing_pct_prot
valid_samps <- miss_samps < qc_config$max_missing_pct_pat

# Extract the names of the failures before we drop them
failed_prots <- rownames(olink_mat)[!valid_prots]
failed_samps <- colnames(olink_mat)[!valid_samps]

# --- Print Protein QC Results ---
if(length(failed_prots) > 0) {
  cat(sprintf("! Dropped %d Proteins (>= %s%% missing):\n  %s\n", 
              length(failed_prots), qc_config$max_missing_pct_prot * 100, paste(failed_prots, collapse = ", ")))
} else {
  cat(sprintf("✓ All proteins passed QC (< %s%% missing).\n", qc_config$max_missing_pct_prot * 100))
}

# --- Print Patient QC Results ---
if(length(failed_samps) > 0) {
  cat(sprintf("! Dropped %d Patients (>= %s%% missing):\n  %s\n", 
              length(failed_samps), qc_config$max_missing_pct_pat * 100, paste(failed_samps, collapse = ", ")))
} else {
  cat(sprintf("✓ All patients passed QC (< %s%% missing).\n", qc_config$max_missing_pct_pat * 100))
}

# Apply the filters
qc_mat <- olink_mat[valid_prots, valid_samps, drop = FALSE]
cat(sprintf("\n- Post-QC Matrix: %d Proteins | %d Samples\n", nrow(qc_mat), ncol(qc_mat)))

# 2. Impute remaining NAs (Zero-Copy Median Imputation)
row_meds <- matrixStats::rowMedians(qc_mat, na.rm = TRUE)
na_coords <- which(is.na(qc_mat), arr.ind = TRUE)
qc_mat[na_coords] <- row_meds[na_coords[, "row"]]

# 3. Synchronize Clinical Spine
# Ensure our clinical spine ONLY contains samples that passed Olink QC
master_spine <- master_spine[Subject_ID %in% colnames(qc_mat)]

# Ensure matrix columns perfectly align with the spine 
qc_mat <- qc_mat[, master_spine$Subject_ID, drop = FALSE]
```
## Export final Data

`r what_this_does("The finalized arrays (clinical_spine.rds and clean_olink_matrix.rds) were exported as serialized R objects for quick, standardized ingestion by downstream scripts.")`

```{r export}
# 4. Export Final Data
cat("Saving finalized datasets to output folder...\n")
saveRDS(master_spine, file.path(output_dir_data, "clinical_spine.rds"))
saveRDS(qc_mat, file.path(output_dir_data, "clean_olink_matrix.rds"))

cat("Script 01 Complete! Data is ready for analysis.\n")
```


