---
title: "Data Import"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction
This script represents Phase 1a of the pipeline. It strictly handles the ingestion of raw data files, resolution of patient identifiers, and technical quality control (missingness filtering and imputation). No biological feature engineering or statistical transformations are performed here.


# Setup
`r what_this_does("Loaded required utility packages (tidyverse, data.table, readxl, matrixStats).", "Established dynamic output directory paths to safely save synchronized data.", "Configured default knitr chunk options to suppress unnecessary warnings.")`

```{r setup, message=FALSE, warning=FALSE}
# Packages
library(tidyverse)   # Data manipulation
library(readxl)      # Reading messy Excel files
library(data.table)  # High-performance data wrangling
library(here)        # Safe file paths
library(matrixStats) # Fast matrix math for QC and Imputation

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "01a_data_import_and_qc"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)
```

# Configuration

`r what_this_does("Defined strict quality control missingness thresholds for proteomics filtering (5% max missingness).", "Set the clinical variable completeness threshold (variables must be 80% complete to survive).", "Listed protected raw variables to prevent the missingness filter from accidentally dropping foundational columns needed for downstream engineering in Script 01b.", "Parameterized all raw data file names to avoid hardcoding deep in the script.")`

```{r config}
# 1. Global File Paths ---------------------------------------------------------
file_paths <- list(
  clinical_raw  = "clinical_data.xlsx",
  id_map        = "Samples for OLink submitted_final_0226_modified.xlsx",
  olink_raw     = "pre_quant.csv",
  lipid_raw     = "Dr Gabriela Kaina and Dr Amela Hurka aLipid Mediator Profiling raw data.xlsx",
  lipid_map     = "Samples for SPM profiling_final export_081225.xlsx",
  lipid_lod     = "Lipidomics_LOD.xlsx"
)

# 2. Global QC Thresholds & Protected Columns ----------------------------------
qc_config <- list(
  max_missing_pct_prot = 0.05,
  max_missing_pct_pat  = 0.05,
  max_missing_pct_lipid = 0.60,
  selected_col         = "SELECTED", 
  max_missingness_clin_var = 0.20,
  tech_ids_to_remove = c("FS93796088"),
  
  # Protect these from the 20% NA filter because Script 01b needs them to engineer features
  protected_raw_vars = c(
    "Subject_ID", "inCentraxx", "Geburtsdatum", "Geschlecht", "Entnahmedatum", 
    "Visit Date", "Einlagerungsdatum"
  )
)
```

# Clinical Data
`r analysis_decision("Dropped the follow-up sample 'Act_4_2' to strictly enforce a cross-sectional study design.", "Excluded patients who were missing from either the CentraXX output or the Biobank sample manifest.", "Converted explicit missing-data strings ('UN', 'Unknown', 'Not done') to true NAs to prevent phantom data from surviving the completeness filter.", "Dropped any raw clinical column with >20% missingness in SSc patients (excluding controls to prevent artificial NA inflation). Protected foundational variables were retained.")`

`r what_this_does("Imported raw clinical and sample mapping Excel files and coalesced IDs.", "Standardized missing value strings into true NAs.", "Calculated missingness exclusively on patient samples (ignoring healthy controls) and applied the 20% missingness filter.", "Generated a scrollable HTML audit table detailing exactly which columns were dropped and their missingness percentages.")`
```{r clin_import, results='asis'}
audit_log <- c("Clinical Import Log\n")

# 1. Process Master Clinical File ----------------------------------------------
raw_clin <- read_excel(here::here("data", file_paths$clinical_raw), sheet = 1)
setDT(raw_clin) 

# Force the clinical ID to be a character string for safe merging
raw_clin <- raw_clin[, inCentraxx := as.character(inCentraxx)]

# Strict Filter: Keep ONLY samples explicitly marked for this study
raw_clin <- raw_clin[get(qc_config$selected_col) == 1 | get(qc_config$selected_col) == "1"]

# 2. Process Mapping File & Coalesce IDs ---------------------------------------
id_map <- read_excel(here::here("data", file_paths$id_map))
setDT(id_map)

# Explicitly drop the follow-up sample to enforce cross-sectional design
id_map <- id_map[`Subject ID` != "Act_4_2"]
id_map <- id_map[, `Subject-ID (EUSTAR)`      := as.character(`Subject-ID (EUSTAR)`)]
id_map <- id_map[, `Subject-ID (Biobank RUZ)` := as.character(`Subject-ID (Biobank RUZ)`)]

# COALESCE: Create a master 'Merge_ID'
id_map <- id_map[, Merge_ID := fifelse(
  is.na(`Subject-ID (EUSTAR)`) | `Subject-ID (EUSTAR)` == "", 
  `Subject-ID (Biobank RUZ)`, 
  `Subject-ID (EUSTAR)`
)]

id_map <- id_map[!is.na(Merge_ID) & Merge_ID != ""]

# 3. Merge and Structure -------------------------------------------------------
raw_spine <- merge(
  raw_clin, 
  id_map, 
  by.x = "inCentraxx", 
  by.y = "Merge_ID", 
  all = FALSE
)

# Rename the final subject ID column for global consistency
raw_spine <- setnames(raw_spine, old = "Subject ID", new = "Subject_ID", skip_absent = TRUE)

audit_log <- c(audit_log, sprintf("- Inner join successful. Matched %d unique patients.", nrow(raw_spine)))

# 3.5 Standardize Missing Values ("Phantom Data" Sweep) ------------------------
# Convert case-insensitive "un", "unknown", and "not done" to true NA 
raw_spine <- raw_spine %>%
  mutate(across(where(is.character), ~ if_else(tolower(.x) %in% c("un", "unknown", "not done"), NA_character_, .x)))
setDT(raw_spine) # Ensure it stays a high-performance data.table

audit_log <- c(audit_log, "- Standardized phantom missing data ('UN', 'Unknown') to true NA.")

# 4. Completeness Pruning (The Technical NA Filter) ----------------------------
# CRITICAL: Exclude Healthy Controls from the calculation so their natural NAs don't inflate the missingness!
missingness <- raw_spine[!str_detect(Subject_ID, "^Ctrl"), lapply(.SD, function(x) sum(is.na(x)) / .N)]

# Convert 1-row data.table to a named numeric vector for safe logical filtering
miss_vec <- unlist(missingness)

cols_to_keep <- names(miss_vec)[miss_vec <= qc_config$max_missingness_clin_var | names(miss_vec) %in% qc_config$protected_raw_vars]
cols_to_drop <- setdiff(names(miss_vec), cols_to_keep)

n_dropped <- length(cols_to_drop)
raw_spine <- raw_spine[, ..cols_to_keep]

audit_log <- c(audit_log, sprintf("- Pruning: Dropped %d columns for having > %s%% missingness in the SSc cohort.", 
            n_dropped, qc_config$max_missingness_clin_var * 100))
audit_log <- c(audit_log, sprintf("- Clinical Spine Dimensions: %d rows x %d columns.", nrow(raw_spine), ncol(raw_spine)))

# 5. Build the Dropped Variables Audit Table -----------------------------------
dropped_html <- ""
if (n_dropped > 0) {
  dropped_df <- data.frame(
    Variable = cols_to_drop,
    Missingness = miss_vec[cols_to_drop]
  ) %>% 
    arrange(desc(Missingness)) %>% 
    mutate(Missingness = sprintf("%.1f%%", Missingness * 100))
  
  dropped_tbl <- dropped_df %>%
    kableExtra::kbl(format = "html", row.names = FALSE, 
                    caption = sprintf("<b style='font-size:1.2em; color:#333;'>Excluded Clinical Variables (Missingness > %s%%)</b>", qc_config$max_missingness_clin_var * 100)) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE, position = "left") %>%
    kableExtra::scroll_box(width = "100%", height = "600px")
  
  if (interactive()) {
    print(dropped_tbl)
  } else {
    dropped_html <- as.character(dropped_tbl)
  }
}

# Cleanup Memory
rm(raw_clin, id_map, missingness, miss_vec, cols_to_keep, cols_to_drop)
if(exists("dropped_df")) rm(dropped_df)
invisible(gc())

# Print HTML output as a single block to prevent code fragmentation
cat(paste(audit_log, collapse = "\n\n"))
cat("\n\n")
if (dropped_html != "") cat(dropped_html)
```

# Proteomics Data
`r analysis_decision("Dropped proteins and samples exceeding the 5% missingness threshold to ensure robust downstream imputation.", "NA values were imputed using a row-wise (protein-wise) median approach.")`

`r what_this_does("Pivoted the Olink data to a wide matrix.", "Executed missingness filtering and median imputation.", "Cross-synchronized the clinical spine and the Olink matrix to guarantee identical sample inclusion and ordering.")`
```{r proteomics_data, results='asis'}
audit_log_olink <- c("Proteomics QC Log\n")

# 1. Import and Pivot ----------------------------------------------------------
raw_olink <- fread(here::here("data", file_paths$olink_raw))
raw_olink <- raw_olink[SampleID %in% raw_spine$Subject_ID]

wide_olink <- dcast(raw_olink, Assay ~ SampleID, value.var = "NPX")
olink_mat <- as.matrix(wide_olink[, -1, with = FALSE])
rownames(olink_mat) <- wide_olink$Assay

audit_log_olink <- c(audit_log_olink, sprintf("- Initial Matrix: %d Proteins | %d Samples", nrow(olink_mat), ncol(olink_mat)))

# 2. Missingness Filtering -----------------------------------------------------
miss_prots <- rowMeans(is.na(olink_mat))
miss_samps <- colMeans(is.na(olink_mat))

valid_prots <- miss_prots < qc_config$max_missing_pct_prot
valid_samps <- miss_samps < qc_config$max_missing_pct_pat

failed_prots <- rownames(olink_mat)[!valid_prots]
failed_samps <- colnames(olink_mat)[!valid_samps]

if(length(failed_prots) > 0) {
  audit_log_olink <- c(audit_log_olink, sprintf("- Dropped %d Proteins (>= %s%% missing).", length(failed_prots), qc_config$max_missing_pct_prot * 100))
}
if(length(failed_samps) > 0) {
  audit_log_olink <- c(audit_log_olink, sprintf("- Dropped %d Patients (>= %s%% missing).", length(failed_samps), qc_config$max_missing_pct_pat * 100))
}

qc_mat <- olink_mat[valid_prots, valid_samps, drop = FALSE]

# 3. Median Imputation Tracking & Execution ------------------------------------
na_coords <- which(is.na(qc_mat), arr.ind = TRUE)
row_meds <- matrixStats::rowMedians(qc_mat, na.rm = TRUE)

impute_html <- ""
if (nrow(na_coords) > 0) {
  # Count how many NAs belong to each protein row
  na_counts <- table(rownames(qc_mat)[na_coords[, "row"]])
  
  impute_df <- data.frame(
    Protein = names(na_counts),
    Imputed_Values = as.integer(na_counts)
  )
  
  # Append the median that was used
  impute_df$Imputed_Median_NPX <- row_meds[match(impute_df$Protein, rownames(qc_mat))]
  
  impute_df <- impute_df %>% 
    arrange(desc(Imputed_Values)) %>%
    mutate(Imputed_Median_NPX = sprintf("%.4f", Imputed_Median_NPX))

  impute_tbl <- impute_df %>%
    kableExtra::kbl(format = "html", row.names = FALSE, 
                    caption = "<b style='font-size:1.2em; color:#333;'>Proteomics Median Imputation Log</b>") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE, position = "left") %>%
    kableExtra::scroll_box(width = "100%", height = "300px")
  
  if (interactive()) {
    print(impute_tbl)
  } else {
    impute_html <- as.character(impute_tbl)
  }
  
  # Execute the actual imputation
  qc_mat[na_coords] <- row_meds[na_coords[, "row"]]
  audit_log_olink <- c(audit_log_olink, sprintf("- Median Imputation Complete: %d total values imputed. Post-QC Matrix: %d Proteins | %d Samples", nrow(na_coords), nrow(qc_mat), ncol(qc_mat)))

} else {
  audit_log_olink <- c(audit_log_olink, sprintf("- Median Imputation Complete: 0 missing values found. No imputation required. Post-QC Matrix: %d Proteins | %d Samples", nrow(qc_mat), ncol(qc_mat)))
}

# 4. Final Sync ----------------------------------------------------------------
# Ensure the clinical spine perfectly matches the surviving omics samples
raw_spine <- raw_spine[Subject_ID %in% colnames(qc_mat)]

# Ensure exact column ordering in the matrix
qc_mat <- qc_mat[, raw_spine$Subject_ID, drop = FALSE]

# Cleanup Memory
rm(raw_olink, wide_olink, olink_mat, miss_prots, miss_samps, valid_prots, valid_samps, failed_prots, failed_samps, row_meds, na_coords)
if(exists("impute_df")) rm(impute_df, na_counts)
invisible(gc())

# Print HTML output as a single block
cat(paste(audit_log_olink, collapse = "\n\n"))
cat("\n\n<br>\n\n")
if (impute_html != "") cat(impute_html)
```
# Lipidomics Data
```{r lipidomics_data}
# 1. Ingestion and ID Translation 
audit_log_lipid <- c("Lipidomics QC Log\n")

# Import the data
raw_lipid <- read_excel(here::here("data", file_paths$lipid_raw), sheet = "Clean", na = "*")
setDT(raw_lipid)

# Filter out configured tech_ids_to_exclude (e.g., orphaned follow-ups)
# Note: Using idiomatic data.table syntax (:= NULL) to delete by reference and save RAM
cols_to_drop <- intersect(qc_config$tech_ids_to_remove, names(raw_lipid))

if (length(cols_to_drop) > 0) {
  raw_lipid[, (cols_to_drop) := NULL]
  audit_log_lipid <- c(
    audit_log_lipid, 
    sprintf("- Dropped %d manually configured tech_ids (%s).", 
            length(cols_to_drop), paste(cols_to_drop, collapse = ", "))
  )
} else {
  audit_log_lipid <- c(audit_log_lipid, "- Manual QC: No configured tech_ids found to drop.")
}

# Extract lipid names and clean the matrix
lipid_names <- raw_lipid[[1]] 
raw_lipid_mat <- as.matrix(raw_lipid[, -1, with = FALSE])

# Convert all values to numeric
raw_lipid_mat <- apply(raw_lipid_mat, 2, as.numeric)

# Restore row names
rownames(raw_lipid_mat) <- lipid_names

# Import Mapping file (Links Sample ID to EUSTAR ID)
id_map_lipid <- read_excel(here::here("data", file_paths$lipid_map))
setDT(id_map_lipid)

# Translation Strategy:
# 1. Identify which Lipidomics columns map to which EUSTAR ID
# 2. Match those EUSTAR IDs against the 'raw_spine' to get the definitive Subject_ID
tech_ids <- colnames(raw_lipid_mat)

# Create a lookup for EUSTAR ID based on Sample ID
eustar_lookup <- setNames(id_map_lipid$`Subject ID (EUSTAR)`, id_map_lipid$`Sample ID (Plasma)`)

# Get the EUSTAR IDs for our current samples
corresponding_eustar <- eustar_lookup[tech_ids]

# Now, match the EUSTAR IDs to the Subject_IDs present in our pre-processed clinical spine
# We assume 'Subject-ID (EUSTAR)' is the column name in raw_spine
eustar_to_subject_lookup <- setNames(raw_spine$Subject_ID, raw_spine$`inCentraxx`)

# Perform the translation: Sample ID -> EUSTAR -> Subject_ID
final_subject_ids <- eustar_to_subject_lookup[corresponding_eustar]

# Identify columns that failed to map (either no match in ID map or no match in clinical spine)
unmatched <- is.na(final_subject_ids)

if(any(unmatched)) {
  audit_log_lipid <- c(audit_log_lipid, sprintf("- Dropping %d unmatched samples.", sum(unmatched)))
  raw_lipid_mat <- raw_lipid_mat[, !unmatched, drop = FALSE]
  final_subject_ids <- final_subject_ids[!unmatched]
}

# Rename matrix columns to the definitive Subject_IDs
colnames(raw_lipid_mat) <- final_subject_ids

# 2. Missingness Filtering 
# Applying the 50% threshold as requested
miss_lipids <- rowMeans(is.na(raw_lipid_mat))
valid_lipids <- miss_lipids < qc_config$max_missing_pct_lipid

audit_log_lipid <- c(audit_log_lipid, sprintf("- Initial Matrix: %d Lipids | %d Samples", nrow(raw_lipid_mat), ncol(raw_lipid_mat)))
audit_log_lipid <- c(audit_log_lipid, sprintf("- Dropped %d Lipids (>= 50%% missing). Which corresponds to %.1f%%", 
                                              sum(!valid_lipids), 100*sum(!valid_lipids)/nrow(raw_lipid_mat)))

qc_mat_lipid <- raw_lipid_mat[valid_lipids, , drop = FALSE]

# 3. LOD Imputation (LOD / 5)
# Load Raw LOD Data
lod_raw <- readxl::read_excel(here::here("data", file_paths$lipid_lod), sheet = "Master")

# Aggregate Technical Replicates & Preserve Family Mapping
mean_lod_df <- lod_raw %>%
  # FIX: By adding Family to the group_by, it is preserved in the output!
  dplyr::group_by(Family, Mediator) %>%
  dplyr::summarise(
    LOD_Matrix = mean(LOD_Matrix, na.rm = TRUE),
    n_replicates = dplyr::n(),
    .groups = "drop"
  )

# Export Mean LOD for Downstream Scripts (e.g., Script 04a)
audit_log_lipid <- c(audit_log_lipid, sprintf("- LOD Aggregation: %d unique lipids averaged from technical replicates.", nrow(mean_lod_df)))

#Create a Named Vector for Fast Matrix Lookup
lod_vec <- setNames(mean_lod_df$LOD_Matrix, mean_lod_df$Mediator)

# Execute Imputation
na_coords <- which(is.na(qc_mat_lipid), arr.ind = TRUE)

if(nrow(na_coords) > 0) {
  # Map missing rows to lipid names, fetch LOD/5
  lipid_row_names <- rownames(qc_mat_lipid)[na_coords[, "row"]]
  imputed_vals <- lod_vec[lipid_row_names] / 5
  
  # Inject imputed values
  qc_mat_lipid[na_coords] <- imputed_vals
  
  audit_log_lipid <- c(audit_log_lipid, sprintf("- LOD Imputation Complete: %d values imputed.", nrow(na_coords)))
} else {
  audit_log_lipid <- c(audit_log_lipid, "- No missing values found; imputation not required.")
}

# 4. Final Sync 
# Ensure Lipidomics samples are synced with the existing clinical spine
qc_mat_lipid <- qc_mat_lipid[, colnames(qc_mat_lipid) %in% raw_spine$Subject_ID, drop = FALSE]
raw_spine <- raw_spine[Subject_ID %in% colnames(qc_mat_lipid)]

# Cleanup
rm(raw_lipid, id_map_lipid, map_match, tech_ids, unmatched, miss_lipids, valid_lipids, lipid_names, 
   na_coords, lipid_row_names, imputed_vals)
invisible(gc())

# Print output
cat(paste(audit_log_lipid, collapse = "\n\n"))
```

# Export 

`r what_this_does("Exported the raw_filtered_spine.rds and clean_olink_matrix.rds for downstream biological engineering in Script 01b.")`

```{r export}
# Save the RAW but mathematically synced and filtered files
saveRDS(raw_spine, file.path(output_dir_data, "raw_filtered_spine.rds"))
saveRDS(qc_mat, file.path(output_dir_data, "clean_olink_matrix.rds"))
saveRDS(qc_mat_lipid, file.path(output_dir_data, "clean_lipid_matrix.rds"))
saveRDS(raw_lipid_mat, file.path(output_dir_data, "raw_lipid_matrix.rds"))
saveRDS(mean_lod_df, file.path(output_dir_data, "lipid_mean_lod.rds"))

cat("Script 01a Complete!\n")
```

```{r}
# Extract the lipid names from both sources
matrix_lipids <- rownames(raw_lipid_mat)
lod_lipids <- mean_lod_df$Mediator # (or lod_raw$Mediator depending on which object is in your environment)

# Find lipids in the matrix that are missing from the LOD file
missing_from_lod <- setdiff(matrix_lipids, lod_lipids)

# Print the results
cat(sprintf("There are %d lipids in your matrix missing from or named differently in the LOD file:\n", length(missing_from_lod)))
sort(missing_from_lod)

```

