---
title: "Data Import & QC"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction
This script represents Phase 1a of the pipeline. It strictly handles the ingestion of raw data files, resolution of patient identifiers, and technical quality control (missingness filtering and imputation). No biological feature engineering or statistical transformations are performed here.


# Setup
`r what_this_does("Loaded required utility packages (tidyverse, data.table, readxl, matrixStats).", "Established dynamic output directory paths to safely save synchronized data.", "Configured default knitr chunk options to suppress unnecessary warnings.")`

```{r setup, message=FALSE, warning=FALSE}
# Packages
library(tidyverse)   # Data manipulation
library(readxl)      # Reading messy Excel files
library(data.table)  # High-performance data wrangling
library(here)        # Safe file paths
library(matrixStats) # Fast matrix math for QC and Imputation
library(OlinkAnalyze)
library(UpSetR)
library(patchwork)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "01a_data_import_and_qc"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)
```

# Configuration

`r what_this_does("Defined strict quality control missingness thresholds for proteomics filtering (5% max missingness).", "Set the clinical variable completeness threshold (variables must be 80% complete to survive).", "Listed protected raw variables to prevent the missingness filter from accidentally dropping foundational columns needed for downstream engineering in Script 01b.", "Parameterized all raw data file names to avoid hardcoding deep in the script.")`

```{r config}
# 1. Global File Paths ---------------------------------------------------------
file_paths <- list(
  clinical_raw  = "clinical_data.xlsx",
  id_map        = "Samples for OLink submitted_final_0226_modified.xlsx",
  olink_raw     = "pre_quant.csv",
  lipid_raw     = "Dr Gabriela Kaina and Dr Amela Hurka aLipid Mediator Profiling raw data.xlsx",
  lipid_map     = "Samples for SPM profiling_final export_081225.xlsx",
  lipid_lod     = "Lipidomics_LOD.xlsx"
)

# 2. Global QC Thresholds & Protected Columns ----------------------------------
qc_config <- list(
  max_missing_pct_prot = 0.05,
  max_missing_pct_pat  = 0.05,
  min_detectability_pct = 25,
  apply_lod_filter     = TRUE,  # Toggle to exclude proteins below min_detectability_pct
  apply_outlier_filter = FALSE,  # Toggle to exclude Median/IQR sample outliers (3 SD)
  max_missing_pct_lipid = 0.60,
  selected_col         = "SELECTED", 
  max_missingness_clin_var = 0.20,
  tech_ids_to_remove = c("FS93796088"),
  
  # Protect these from the 20% NA filter because Script 01b needs them to engineer features
  protected_raw_vars = c(
    "Subject_ID", "inCentraxx", "Geburtsdatum", "Geschlecht", "Entnahmedatum", 
    "Visit Date", "Einlagerungsdatum"
  )
)
```

# Clinical Data
`r analysis_decision("Dropped the follow-up sample 'Act_4_2' to strictly enforce a cross-sectional study design.", "Excluded patients who were missing from either the CentraXX output or the Biobank sample manifest.", "Converted explicit missing-data strings ('UN', 'Unknown', 'Not done') to true NAs to prevent phantom data from surviving the completeness filter.", "Dropped any raw clinical column with >20% missingness in SSc patients (excluding controls to prevent artificial NA inflation). Protected foundational variables were retained.")`

`r what_this_does("Imported raw clinical and sample mapping Excel files and coalesced IDs.", "Standardized missing value strings into true NAs.", "Calculated missingness exclusively on patient samples (ignoring healthy controls) and applied the 20% missingness filter.", "Generated a scrollable HTML audit table detailing exactly which columns were dropped and their missingness percentages.")`
```{r clin_import, results='asis'}
audit_log <- c("Clinical Import Log\n")

# 1. Process Master Clinical File ----------------------------------------------
raw_clin <- read_excel(here::here("data", file_paths$clinical_raw), sheet = 1)
setDT(raw_clin) 

# Force the clinical ID to be a character string for safe merging
raw_clin <- raw_clin[, inCentraxx := as.character(inCentraxx)]

# Strict Filter: Keep ONLY samples explicitly marked for this study
raw_clin <- raw_clin[get(qc_config$selected_col) == 1 | get(qc_config$selected_col) == "1"]

# 2. Process Mapping File & Coalesce IDs ---------------------------------------
id_map <- read_excel(here::here("data", file_paths$id_map))
setDT(id_map)

# Explicitly drop the follow-up sample to enforce cross-sectional design
id_map <- id_map[`Subject ID` != "Act_4_2"]
id_map <- id_map[, `Subject-ID (EUSTAR)`      := as.character(`Subject-ID (EUSTAR)`)]
id_map <- id_map[, `Subject-ID (Biobank RUZ)` := as.character(`Subject-ID (Biobank RUZ)`)]

# COALESCE: Create a master 'Merge_ID'
id_map <- id_map[, Merge_ID := fifelse(
  is.na(`Subject-ID (EUSTAR)`) | `Subject-ID (EUSTAR)` == "", 
  `Subject-ID (Biobank RUZ)`, 
  `Subject-ID (EUSTAR)`
)]

id_map <- id_map[!is.na(Merge_ID) & Merge_ID != ""]

# 3. Merge and Structure -------------------------------------------------------
raw_spine <- merge(
  raw_clin, 
  id_map, 
  by.x = "inCentraxx", 
  by.y = "Merge_ID", 
  all = FALSE
)

# Rename the final subject ID column for global consistency
raw_spine <- setnames(raw_spine, old = "Subject ID", new = "Subject_ID", skip_absent = TRUE)

audit_log <- c(audit_log, sprintf("- Inner join successful. Matched %d unique patients.", nrow(raw_spine)))

# 3.5 Standardize Missing Values ("Phantom Data" Sweep) ------------------------
# Convert case-insensitive "un", "unknown", and "not done" to true NA 
raw_spine <- raw_spine %>%
  mutate(across(where(is.character), ~ if_else(tolower(.x) %in% c("un", "unknown", "not done"), NA_character_, .x)))
setDT(raw_spine) # Ensure it stays a high-performance data.table

audit_log <- c(audit_log, "- Standardized phantom missing data ('UN', 'Unknown') to true NA.")

# 4. Completeness Pruning (The Technical NA Filter) ----------------------------
# CRITICAL: Exclude Healthy Controls from the calculation so their natural NAs don't inflate the missingness!
missingness <- raw_spine[!str_detect(Subject_ID, "^Ctrl"), lapply(.SD, function(x) sum(is.na(x)) / .N)]

# Convert 1-row data.table to a named numeric vector for safe logical filtering
miss_vec <- unlist(missingness)

cols_to_keep <- names(miss_vec)[miss_vec <= qc_config$max_missingness_clin_var | names(miss_vec) %in% qc_config$protected_raw_vars]
cols_to_drop <- setdiff(names(miss_vec), cols_to_keep)

n_dropped <- length(cols_to_drop)
raw_spine <- raw_spine[, ..cols_to_keep]

audit_log <- c(audit_log, sprintf("- Pruning: Dropped %d columns for having > %s%% missingness in the SSc cohort.", 
            n_dropped, qc_config$max_missingness_clin_var * 100))
audit_log <- c(audit_log, sprintf("- Clinical Spine Dimensions: %d rows x %d columns.", nrow(raw_spine), ncol(raw_spine)))

# 5. Build the Dropped Variables Audit Table -----------------------------------
dropped_html <- ""
if (n_dropped > 0) {
  dropped_df <- data.frame(
    Variable = cols_to_drop,
    Missingness = miss_vec[cols_to_drop]
  ) %>% 
    arrange(desc(Missingness)) %>% 
    mutate(Missingness = sprintf("%.1f%%", Missingness * 100))
  
  dropped_tbl <- dropped_df %>%
    kableExtra::kbl(format = "html", row.names = FALSE, 
                    caption = sprintf("<b style='font-size:1.2em; color:#333;'>Excluded Clinical Variables (Missingness > %s%%)</b>", qc_config$max_missingness_clin_var * 100)) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE, position = "left") %>%
    kableExtra::scroll_box(width = "100%", height = "600px")
  
  # If we are in RStudio (interactive) AND knitr is NOT actively building the document
  if (interactive() && !isTRUE(getOption('knitr.in.progress'))) {
    print(dropped_tbl) # Prints to RStudio Viewer
  } else {
    dropped_html <- as.character(dropped_tbl) # Captures safely for rendering
  }
}

# Premature Feature Engineering of cohort_group
raw_spine[, cohort_group := fcase(
  str_detect(Subject_ID, "^Nonact"), GRP_NON_ACTIVE,
  str_detect(Subject_ID, "^Act"),    GRP_ACTIVE,
  str_detect(Subject_ID, "^Ctrl"),   GRP_CONTROL,
  default = NA_character_
)]
raw_spine[, cohort_group := factor(cohort_group, levels = c(GRP_ACTIVE, GRP_NON_ACTIVE, GRP_CONTROL))]

# Cleanup Memory
rm(raw_clin, id_map, missingness, miss_vec, cols_to_keep, cols_to_drop)
if(exists("dropped_df")) rm(dropped_df)
invisible(gc())

# Print HTML output as a single block to prevent code fragmentation
cat(paste(audit_log, collapse = "\n\n"))

if (dropped_html != "") {
  # Add structural space
  cat("\n\n")
  
  # Tell Pandoc to strictly ignore the following HTML
  cat("```{=html}\n")
  cat(dropped_html)
  cat("\n```\n\n")
}
```

# Proteomics Data

## Proteomics Import

`r what_this_does("Ingested raw Olink long-format data.", "Retained strictly the patient samples present in the clinical spine, while defensively preserving technical Negative Controls (NC/NEG) required for accurate LOD calculation.")`

```{r proteomics_import, results='asis'}
audit_log_olink <- c("Proteomics Import Log\n")

# 1. Import Raw Long Format ----------------------------------------------------
raw_olink <- fread(here::here("data", file_paths$olink_raw))

# CRITICAL: Keep valid patients AND technical negative controls for LOD math
valid_subjects <- raw_spine$Subject_ID
raw_olink <- raw_olink[SampleID %in% valid_subjects | str_detect(toupper(SampleType), "NEG|NC")]

audit_log_olink <- c(audit_log_olink, sprintf("- Initial Long Matrix: %d total rows imported (Patients + Negative Controls).", nrow(raw_olink)))

# Print Import Log
cat(paste(audit_log_olink, collapse = "\n\n"))
```

## Proteomics QC {.tabset}
`r what_this_does("Calculated the Limit of Detection (LOD) using negative controls.", "Evaluated cohort-specific detectability rates, generating a tiered master summary table to isolate true global failures from state-dependent signals.", "Generated global NPX distribution and Median-IQR scatter plots to map technical variance and flag 3-SD sample outliers.", "Mapped biological representation across cohorts using UpSet intersections.")`

```{r proteomics_qc, fig.width=10, fig.height=8, results='asis'}
cat("\n\n### LOD & Detectability\n\n")

# 1. LOD Calculation -----------------------------------------------------------
nc_data <- raw_olink %>% filter(str_detect(toupper(SampleType), "NEG|NC"))
n_nc_wells <- n_distinct(nc_data$SampleID)

cat("<b>Negative control wells available:</b>", n_nc_wells, "<br><br>")

npx_lod <- NULL
lod_col <- NULL
has_olink_analyze <- requireNamespace("OlinkAnalyze", quietly = TRUE)

if (has_olink_analyze && n_nc_wells >= 10) {
  npx_lod <- tryCatch(
    OlinkAnalyze::olink_lod(as.data.frame(raw_olink), lod_method = "NCLOD", 
                            preferred_names = "PCNormalizedNPX"),
    error = function(e) { NULL }
  )
  if (!is.null(npx_lod)) {
    lod_col <- if ("PCNormalizedLOD" %in% colnames(npx_lod)) "PCNormalizedLOD" else "LOD"
  }
}

if (is.null(npx_lod)) {
  lod_tbl <- nc_data %>%
    group_by(Assay) %>%
    summarise(
      mean_NPX = mean(NPX, na.rm = TRUE),
      sd_NPX = sd(NPX, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(LOD_manual = mean_NPX + (3 * sd_NPX))
  
  npx_lod <- raw_olink %>% left_join(lod_tbl, by = "Assay")
  lod_col <- "LOD_manual"
}

# 2. Master Detectability Calculation ------------------------------------------
samples_only <- npx_lod %>% 
  filter(toupper(SampleType) == "SAMPLE") %>%
  left_join(raw_spine %>% select(Subject_ID, cohort_group), by = c("SampleID" = "Subject_ID"))

# A. Global percentages
global_det <- samples_only %>%
  mutate(Target_NPX = if("PCNormalizedNPX" %in% names(.)) PCNormalizedNPX else NPX,
         above_lod = Target_NPX > .data[[lod_col]]) %>%
  group_by(Assay, OlinkID, UniProt) %>%
  summarise(Overall_Pct = 100 * mean(above_lod, na.rm = TRUE), .groups = "drop")

# B. Cohort-specific percentages
cohort_det <- samples_only %>%
  filter(!is.na(cohort_group)) %>%
  mutate(Target_NPX = if("PCNormalizedNPX" %in% names(.)) PCNormalizedNPX else NPX,
         above_lod = Target_NPX > .data[[lod_col]]) %>%
  group_by(Assay, cohort_group) %>%
  summarise(pct = 100 * mean(above_lod, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = cohort_group, values_from = pct, values_fill = 0)

# C. Merge and assign tiers
cohort_cols <- as.character(unique(samples_only$cohort_group[!is.na(samples_only$cohort_group)]))

master_det <- global_det %>%
  left_join(cohort_det, by = "Assay") %>%
  rowwise() %>%
  mutate(
    max_cohort_pct = max(c_across(all_of(cohort_cols)), na.rm = TRUE),
    min_cohort_pct = min(c_across(all_of(cohort_cols)), na.rm = TRUE),
    # Fails ONLY if below threshold in ALL groups (meaning max percentage across groups is < threshold)
    Pass = if_else(max_cohort_pct >= qc_config$min_detectability_pct, "Yes", "No"),
    Tier = case_when(
      Pass == "No" ~ 1,
      Pass == "Yes" & min_cohort_pct < qc_config$min_detectability_pct ~ 2,
      TRUE ~ 3
    )
  ) %>%
  ungroup() %>%
  arrange(Tier, Overall_Pct) # Sorted by Tier, then lowest detectability at the top

# Isolate the definitive failures for downstream exclusion in the next chunk
low_detect <- master_det %>% filter(Pass == "No")

# 3. Build Master Table --------------------------------------------------------
master_tbl_df <- master_det %>%
  select(Assay, Overall_Pct, all_of(cohort_cols), Pass) %>%
  mutate(across(c("Overall_Pct", all_of(cohort_cols)), 
                ~ kableExtra::cell_spec(sprintf("%.1f%%", .x), 
                                        color = ifelse(.x < qc_config$min_detectability_pct, "red", "black")))) %>%
  rename(Overall = Overall_Pct)

master_tbl <- master_tbl_df %>%
  kableExtra::kbl(format = "html", escape = FALSE, row.names = FALSE,
                  caption = sprintf("<b style='font-size:1.2em; color:#333;'>Master Detectability Summary (Threshold: %s%%)</b>", qc_config$min_detectability_pct)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE, position = "left") %>%
  kableExtra::scroll_box(width = "100%", height = "600px")

# 4. Global Distribution & QC Plots --------------------------------------------
cat("\n\n### Global Distribution\n\n")

if (has_olink_analyze) {
  npx_distribution_full <- OlinkAnalyze::olink_dist_plot(
    df = as.data.frame(samples_only),
    color_g = "SampleQC"
  ) + 
    ggplot2::theme(text = ggplot2::element_text(size = 12)) +
    ggplot2::labs(title = "Global NPX Distribution across all valid samples")
  
  print(npx_distribution_full)
  cat("<br><br>")
  
  # Calculate pass percentage for subtitle
  pct_pass_qc <- 100 * mean(toupper(samples_only$SampleQC) == "PASS", na.rm = TRUE)
  
  median_iqr_full <- OlinkAnalyze::olink_qc_plot(
    df = as.data.frame(samples_only),
    color_g = "cohort_group",
    label_outliers = TRUE,
    IQR_outlierDef = 3L,
    median_outlierDef = 3L,
    outlierLines = TRUE
  ) +
    ggplot2::labs(
      title = "Median–IQR QC plot for all valid samples",
      subtitle = sprintf("Outlier thresholds defined as ± 3 Standard Deviations from the mean Median/IQR.\nOverall Sample QC Pass Rate: %.1f%%", pct_pass_qc),
      color = "Cohort Group"
    ) +
    ggplot2::theme(
      text = ggplot2::element_text(size = 13),
      plot.title = ggplot2::element_text(size = 17, face = "bold"),
      plot.subtitle = ggplot2::element_text(size = 12, color = "#555555")
    )
  
  print(median_iqr_full)
}

# 5. Cohort Overlaps (UpSet Plot) ----------------------------------------------
cat("\n\n### Cohort Overlaps\n\n")

has_upsetr <- requireNamespace("UpSetR", quietly = TRUE)

if (has_upsetr) {
  
  upset_detection_data <- master_det %>%
    mutate(across(all_of(cohort_cols), ~ .x >= qc_config$min_detectability_pct)) %>%
    select(Assay, all_of(cohort_cols))
  
  upset_df <- as.data.frame(upset_detection_data)
  for(col in cohort_cols) {
    upset_df[[col]] <- as.integer(upset_df[[col]])
  }
  
  olink_upset_plot <- UpSetR::upset(
    upset_df,
    sets = cohort_cols,
    keep.order = TRUE,
    order.by = c("freq", "degree"),
    sets.bar.color = "#4C72B0",
    mainbar.y.label = "Number of Detected Proteins",
    sets.x.label = "Total Proteins per Cohort",
    text.scale = 1.3
  )
  
  print(olink_upset_plot)
  
} else {
  cat("<br><b>Skipped UpSet Plot because the 'UpSetR' package is missing.</b><br>")
}

# 6. Outputs -------------------------------------------------------------------
cat("\n\n### Assay Detectability Table\n\n")
cat(sprintf("<p><i>Proteins dropping below the %s%% threshold in ALL cohorts (Tier 1) are flagged 'No' and will be excluded. Proteins failing in at least one cohort but passing in others (Tier 2) are retained.</i></p>", qc_config$min_detectability_pct))

if (exists("master_tbl")) {
  print(master_tbl)
  cat("\n\n")
}
```

## Proteomics Filtering & Imputation
`r what_this_does("Pivoted the Olink data to a wide matrix for standard math operations.", "Executed strict missingness filtering (5% threshold) for proteins and samples.", "Applied cohort-aware LOD exclusions, safely retaining proteins that showed signal in at least one clinical group.", "Excluded sample outliers using the exact 3-SD Median/IQR math utilized in the Olink QC plots.", "Applied row-wise median imputation exclusively to surviving missing values.")`

`r analysis_decision("Enforced a cohort-aware LOD filter, dropping only proteins that fell below the parameterized detectability threshold simultaneously across all three clinical groups (Active, Non-Active, Control).", "Excluded technical sample outliers exceeding 3 standard deviations from the global mean Median or IQR to prevent skewed biological signals downstream.")`

```{r proteomics_filtering_and_imputation, results='asis'}
audit_log_impute <- c("Filtering & Imputation Log\n")

# 1. Pre-Processing & Pivot ----------------------------------------------------
target_val <- if("PCNormalizedNPX" %in% names(samples_only)) "PCNormalizedNPX" else "NPX"

# Perfect replica of OlinkAnalyze's internal QC math
has_panel <- "Panel" %in% names(samples_only)

if (has_panel) {
  qc_calc <- samples_only %>%
    group_by(SampleID, Panel) %>%
    summarise(
      S_Median = median(NPX, na.rm = TRUE), # Olink strictly calculates QC on NPX
      S_IQR = IQR(NPX, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    # 3-SD thresholds are calculated independently per Panel
    group_by(Panel) %>%
    mutate(
      median_outlier = abs(S_Median - mean(S_Median, na.rm = TRUE)) > (3 * sd(S_Median, na.rm = TRUE)),
      iqr_outlier = abs(S_IQR - mean(S_IQR, na.rm = TRUE)) > (3 * sd(S_IQR, na.rm = TRUE))
    ) %>%
    ungroup()
} else {
  qc_calc <- samples_only %>%
    group_by(SampleID) %>%
    summarise(
      S_Median = median(NPX, na.rm = TRUE),
      S_IQR = IQR(NPX, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      median_outlier = abs(S_Median - mean(S_Median, na.rm = TRUE)) > (3 * sd(S_Median, na.rm = TRUE)),
      iqr_outlier = abs(S_IQR - mean(S_IQR, na.rm = TRUE)) > (3 * sd(S_IQR, na.rm = TRUE))
    )
}

failed_outlier_samps <- qc_calc %>%
  filter(median_outlier | iqr_outlier) %>%
  pull(SampleID) %>%
  unique()

# Pivot to Wide Matrix
setDT(samples_only)
wide_olink <- dcast(samples_only, Assay ~ SampleID, value.var = target_val)

olink_mat <- as.matrix(wide_olink[, -1, with = FALSE])
rownames(olink_mat) <- wide_olink$Assay

audit_log_impute <- c(audit_log_impute, sprintf("- Pivot Complete: Wide Matrix configured with %d Proteins | %d Samples.", nrow(olink_mat), ncol(olink_mat)))

# 2. Missingness Filtering (NA-based) ------------------------------------------
miss_prots <- rowMeans(is.na(olink_mat))
miss_samps <- colMeans(is.na(olink_mat))

valid_prots <- miss_prots < qc_config$max_missing_pct_prot
valid_samps <- miss_samps < qc_config$max_missing_pct_pat

failed_prots <- rownames(olink_mat)[!valid_prots]
failed_samps <- colnames(olink_mat)[!valid_samps]

if(length(failed_prots) > 0) {
  audit_log_impute <- c(audit_log_impute, sprintf("- Dropped %d Proteins (>= %s%% missing NA).", length(failed_prots), qc_config$max_missing_pct_prot * 100))
}
if(length(failed_samps) > 0) {
  audit_log_impute <- c(audit_log_impute, sprintf("- Dropped %d Patients (>= %s%% missing NA).", length(failed_samps), qc_config$max_missing_pct_pat * 100))
}

qc_mat <- olink_mat[valid_prots, valid_samps, drop = FALSE]

# 3. Parameterized Exclusions (LOD & Outliers) ---------------------------------

# A. LOD Filter (Using cohort-aware logic from previous chunk)
assays_to_drop_lod <- low_detect$Assay[low_detect$Assay %in% rownames(qc_mat)]

if (qc_config$apply_lod_filter) {
  if(length(assays_to_drop_lod) > 0) {
    qc_mat <- qc_mat[!rownames(qc_mat) %in% assays_to_drop_lod, , drop = FALSE]
    audit_log_impute <- c(audit_log_impute, sprintf("- Strict LOD Filter Applied: Excluded %d proteins for falling below the %s%% detectability threshold across ALL cohort groups.", length(assays_to_drop_lod), qc_config$min_detectability_pct))
  } else {
    audit_log_impute <- c(audit_log_impute, "- Strict LOD Filter Applied: 0 proteins failed across all cohort groups. None excluded.")
  }
} else {
  audit_log_impute <- c(audit_log_impute, sprintf("- LOD Filter Bypassed (Config): %d proteins below the %s%% detectability threshold were retained.", 
                                                  length(assays_to_drop_lod), qc_config$min_detectability_pct))
}

# B. Median-IQR Outlier Filter
if (qc_config$apply_outlier_filter) {
  if (length(failed_outlier_samps) > 0) {
    qc_mat <- qc_mat[, !colnames(qc_mat) %in% failed_outlier_samps, drop = FALSE]
    audit_log_impute <- c(audit_log_impute, sprintf("- Outlier Filter Applied: Excluded %d samples that failed the 3 SD Median/IQR check (%s).", 
                                                    length(failed_outlier_samps), paste(failed_outlier_samps, collapse = ", ")))
  } else {
    audit_log_impute <- c(audit_log_impute, "- Outlier Filter Applied: 0 samples failed the Median/IQR check. None excluded.")
  }
} else {
  audit_log_impute <- c(audit_log_impute, sprintf("- Outlier Filter Bypassed (Config): %d flagged sample outliers were retained.", length(failed_outlier_samps)))
}

# 4. Median Imputation ---------------------------------------------------------
na_coords <- which(is.na(qc_mat), arr.ind = TRUE)
row_meds <- matrixStats::rowMedians(qc_mat, na.rm = TRUE)

if (nrow(na_coords) > 0) {
  na_counts <- table(rownames(qc_mat)[na_coords[, "row"]])
  
  impute_df <- data.frame(Protein = names(na_counts), Imputed_Values = as.integer(na_counts))
  impute_df$Imputed_Median_NPX <- row_meds[match(impute_df$Protein, rownames(qc_mat))]
  impute_df <- impute_df %>% arrange(desc(Imputed_Values)) %>% mutate(Imputed_Median_NPX = sprintf("%.4f", Imputed_Median_NPX))

  impute_tbl <- impute_df %>%
    kableExtra::kbl(format = "html", row.names = FALSE, caption = "<b style='font-size:1.2em; color:#333;'>Proteomics Median Imputation Log</b>") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE, position = "left") %>%
    kableExtra::scroll_box(width = "100%", height = "300px")
  
  qc_mat[na_coords] <- row_meds[na_coords[, "row"]]
  audit_log_impute <- c(audit_log_impute, sprintf("- Imputation Complete: %d total missing values replaced with row-wise medians.", nrow(na_coords)))
} else {
  audit_log_impute <- c(audit_log_impute, "- Imputation Bypassed: 0 missing values found in the surviving matrix.")
}

audit_log_impute <- c(audit_log_impute, sprintf("- Final QC Matrix Dimensions: %d Proteins | %d Samples.", nrow(qc_mat), ncol(qc_mat)))

# 5. Final Data Synchronization ------------------------------------------------
raw_spine <- raw_spine[Subject_ID %in% colnames(qc_mat)]
qc_mat <- qc_mat[, raw_spine$Subject_ID, drop = FALSE]

# Memory Cleanup
rm(raw_olink, wide_olink, olink_mat, miss_prots, miss_samps, valid_prots, valid_samps, 
   failed_prots, failed_samps, row_meds, na_coords, samples_only, npx_lod, nc_data,
   sample_medians, sample_iqrs, mean_median, sd_median, mean_iqr, sd_iqr, outlier_flags)
if(exists("impute_df")) rm(impute_df, na_counts)
invisible(gc())

# 6. Outputs -------------------------------------------------------------------
cat(paste(audit_log_impute, collapse = "\n\n"))
cat("\n\n<!-- Terminate List -->\n\n")

if (exists("impute_tbl")) {
  print(impute_tbl)
  cat("\n\n")
}
```


# Lipidomics Data
```{r lipidomics_data}
# 1. Ingestion and ID Translation 
audit_log_lipid <- c("Lipidomics QC Log\n")

# Import the data
raw_lipid <- read_excel(here::here("data", file_paths$lipid_raw), sheet = "Clean", na = "*")
setDT(raw_lipid)

# Filter out configured tech_ids_to_exclude (e.g., orphaned follow-ups)
# Note: Using idiomatic data.table syntax (:= NULL) to delete by reference and save RAM
cols_to_drop <- intersect(qc_config$tech_ids_to_remove, names(raw_lipid))

if (length(cols_to_drop) > 0) {
  raw_lipid[, (cols_to_drop) := NULL]
  audit_log_lipid <- c(
    audit_log_lipid, 
    sprintf("- Dropped %d manually configured tech_ids (%s).", 
            length(cols_to_drop), paste(cols_to_drop, collapse = ", "))
  )
} else {
  audit_log_lipid <- c(audit_log_lipid, "- Manual QC: No configured tech_ids found to drop.")
}

# Extract lipid names and clean the matrix
lipid_names <- raw_lipid[[1]] 
raw_lipid_mat <- as.matrix(raw_lipid[, -1, with = FALSE])

# Convert all values to numeric
raw_lipid_mat <- apply(raw_lipid_mat, 2, as.numeric)

# Restore row names
rownames(raw_lipid_mat) <- lipid_names

# Import Mapping file (Links Sample ID to EUSTAR ID)
id_map_lipid <- read_excel(here::here("data", file_paths$lipid_map))
setDT(id_map_lipid)

# Translation Strategy:
# 1. Identify which Lipidomics columns map to which EUSTAR ID
# 2. Match those EUSTAR IDs against the 'raw_spine' to get the definitive Subject_ID
tech_ids <- colnames(raw_lipid_mat)

# Create a lookup for EUSTAR ID based on Sample ID
eustar_lookup <- setNames(id_map_lipid$`Subject ID (EUSTAR)`, id_map_lipid$`Sample ID (Plasma)`)

# Get the EUSTAR IDs for our current samples
corresponding_eustar <- eustar_lookup[tech_ids]

# Now, match the EUSTAR IDs to the Subject_IDs present in our pre-processed clinical spine
# We assume 'Subject-ID (EUSTAR)' is the column name in raw_spine
eustar_to_subject_lookup <- setNames(raw_spine$Subject_ID, raw_spine$`inCentraxx`)

# Perform the translation: Sample ID -> EUSTAR -> Subject_ID
final_subject_ids <- eustar_to_subject_lookup[corresponding_eustar]

# Identify columns that failed to map (either no match in ID map or no match in clinical spine)
unmatched <- is.na(final_subject_ids)

if(any(unmatched)) {
  audit_log_lipid <- c(audit_log_lipid, sprintf("- Dropping %d unmatched samples.", sum(unmatched)))
  raw_lipid_mat <- raw_lipid_mat[, !unmatched, drop = FALSE]
  final_subject_ids <- final_subject_ids[!unmatched]
}

# Rename matrix columns to the definitive Subject_IDs
colnames(raw_lipid_mat) <- final_subject_ids

# 2. Missingness Filtering 
# Applying the 50% threshold as requested
miss_lipids <- rowMeans(is.na(raw_lipid_mat))
valid_lipids <- miss_lipids < qc_config$max_missing_pct_lipid

audit_log_lipid <- c(audit_log_lipid, sprintf("- Initial Matrix: %d Lipids | %d Samples", nrow(raw_lipid_mat), ncol(raw_lipid_mat)))
audit_log_lipid <- c(audit_log_lipid, sprintf("- Dropped %d Lipids (>= 50%% missing). Which corresponds to %.1f%%", 
                                              sum(!valid_lipids), 100*sum(!valid_lipids)/nrow(raw_lipid_mat)))

qc_mat_lipid <- raw_lipid_mat[valid_lipids, , drop = FALSE]

# 3. LOD Imputation (LOD / 5)
# Load Raw LOD Data
lod_raw <- readxl::read_excel(here::here("data", file_paths$lipid_lod), sheet = "Master")

# Aggregate Technical Replicates & Preserve Family Mapping
mean_lod_df <- lod_raw %>%
  # FIX: By adding Family to the group_by, it is preserved in the output!
  dplyr::group_by(Precursor, Class, Function, Mediator) %>%
  dplyr::summarise(
    LOD_Matrix = mean(LOD_Matrix, na.rm = TRUE),
    n_replicates = dplyr::n(),
    .groups = "drop"
  )

# Export Mean LOD for Downstream Scripts (e.g., Script 04a)
audit_log_lipid <- c(audit_log_lipid, sprintf("- LOD Aggregation: %d unique lipids averaged from technical replicates.", nrow(mean_lod_df)))

#Create a Named Vector for Fast Matrix Lookup
lod_vec <- setNames(mean_lod_df$LOD_Matrix, mean_lod_df$Mediator)

# Execute Imputation
na_coords <- which(is.na(qc_mat_lipid), arr.ind = TRUE)

if(nrow(na_coords) > 0) {
  # Map missing rows to lipid names, fetch LOD/5
  lipid_row_names <- rownames(qc_mat_lipid)[na_coords[, "row"]]
  imputed_vals <- lod_vec[lipid_row_names] / 5
  
  # Inject imputed values
  qc_mat_lipid[na_coords] <- imputed_vals
  
  audit_log_lipid <- c(audit_log_lipid, sprintf("- LOD Imputation Complete: %d values imputed.", nrow(na_coords)))
} else {
  audit_log_lipid <- c(audit_log_lipid, "- No missing values found; imputation not required.")
}

# 4. Final Sync 
# Ensure Lipidomics samples are synced with the existing clinical spine
qc_mat_lipid <- qc_mat_lipid[, colnames(qc_mat_lipid) %in% raw_spine$Subject_ID, drop = FALSE]
raw_spine <- raw_spine[Subject_ID %in% colnames(qc_mat_lipid)]

# Cleanup
rm(raw_lipid, id_map_lipid, map_match, tech_ids, unmatched, miss_lipids, valid_lipids, lipid_names, 
   na_coords, lipid_row_names, imputed_vals)
invisible(gc())

# Print output
cat(paste(audit_log_lipid, collapse = "\n\n"))

# Extract the lipid names from both sources
matrix_lipids <- rownames(raw_lipid_mat)
lod_lipids <- mean_lod_df$Mediator # (or lod_raw$Mediator depending on which object is in your environment)

# Find lipids in the matrix that are missing from the LOD file
missing_from_lod <- setdiff(matrix_lipids, lod_lipids)

# Print the results
cat(sprintf("There are %d lipids in your matrix missing from or named differently in the LOD file:\n", length(missing_from_lod)))
sort(missing_from_lod)
```

# Export 

`r what_this_does("Exported the raw_filtered_spine.rds and clean_olink_matrix.rds for downstream biological engineering in Script 01b.")`

```{r export}
# Save the RAW but mathematically synced and filtered files
saveRDS(raw_spine, file.path(output_dir_data, "raw_filtered_spine.rds"))
saveRDS(qc_mat, file.path(output_dir_data, "clean_olink_matrix.rds"))
saveRDS(qc_mat_lipid, file.path(output_dir_data, "clean_lipid_matrix.rds"))
saveRDS(raw_lipid_mat, file.path(output_dir_data, "raw_lipid_matrix.rds"))
saveRDS(mean_lod_df, file.path(output_dir_data, "lipid_mean_lod.rds"))

cat("Script 01a Complete!\n")
```



