---
title: "Cohort Characterization"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Data Import
`r what_this_does("Loaded required utility packages (tidyverse, gtsummary, corrplot, heatmaply, etc.).", "Created dynamic output directories to safely store Script 02 results and figures.", "Sourced global configurations and loaded the clinical spine generated previously.")`
```{r setup_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(lubridate)
library(gtsummary)
library(here)
library(patchwork)
library(ggpubr)
library(gt)
library(kableExtra)
library(corrplot)
library(heatmaply)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE # Helps prevent code chunk fragmentation
)

# Output paths
current_file <- "02_cohort_characterization"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Input paths (Parameterized)
input_paths <- list(
  clinical_spine = here::here("output", "01_data_import_and_qc", "clinical_spine.rds")
)

# Load the clinical spine from Script 01
master_spine <- readRDS(input_paths$clinical_spine)
```

# Configuration
`r what_this_does("Defined statistical thresholds for normality (Shapiro-Wilk) and variance (Bartlett) testing.", "Configured correlation matrix parameters including sparsity filters, significance limits, and targeted anchor variables.", "Constructed dynamic configuration lists to dictate Table 1 and Figure layouts using the pre-engineered master_spine.")`

```{r configuration}
# 1. Global Variables & Feature Engineering ------------------------------------
stat_config <- list(
  shapiro_alpha  = 0.05, 
  variance_alpha = 0.05, 
  p_cutoff       = 0.05
)

corr_config <- list(
  min_required_obs  = 10,   
  p_cutoff          = 0.05, 
  strong_rho_cutoff = 0.7,
  label_font_size   = 0.7  # Parameterized heatmap label size
)

# 2. Clinical Domains ----------------------------------------------------------
clinical_domains <- list(
  Demographics_and_Scores = c(
    "Age", "Sex", "Sample_Age", "cohort_group", "Total Score",
    "eustarAI", "ACTIVE_AI", "Active_our", "Inflamm_active", "fibrotic_active", "vascular_active"
  ),
  Cutaneous_and_Vascular = c(
    "Subset_Diffuse", "Subset_Limited", "Subset_Sine",
    "Modified rodnan skin score", "mRSS_a", "mRSS Worsening (1yr)",
    "Raynaud phenomenon", "Score : Raynaud phenomenon", "Raynaud's present", "Raynaud VAS",
    "Digital Tip Ulcers", "Digital ulcer", "Fingertip pitting scars", "Pitting scars on finger tips",
    "Telangiectasia", "Telangiectasia (any)", "Abnormal Capillaroscopy", "Capillaroscopy Score",
    "Capillaroscopy Pattern", "Giant capillaries", "Hemorrhages", "Capillary loss", "Ramified bushy capillaries",
    "Puffy Fingers", "Sclerodactyly", "Skin Thickening (Ext. MCP)", "Skin Thickening (Prox. MCP)",
    "Subcutaneous Calcinosis", "Gangrene"
  ),
  Musculoskeletal = c(
    "Joint involvement", "Joint Contractures", "Tendon friction rubs",
    "Proximal muscle weakness", "Myalgia", "Muscle atrophy", "Arthritis Activity (NA)"
  ),
  Pulmonary = c(
    "PAH/ILD", "ILD diagnosed via HRCT", "FVC %pred", "Forced Vital Capacity (ml)",
    "DLCO (SB) (% predicted)", "DLCO/VA (% predicted)", "TLC %pred",
    "O2-saturation at rest (%)", "Min SpO2 (Exercise)",
    "Dyspnea NYHA Stage", "Max Borg Dyspnea", "Distance in m", "New PAH"
  ),
  Cardiac = c(
    "New Cardiac Manifest.", "Does the patient have pulmonary arterial hypertension since the last visit?",
    "Right bundle branch block", "Right axis deviation", "Right ventricular hypertrophy",
    "Ventricular arrhythmias", "Auricular arrhythmias", "Conduction blocks", "Arrhythmias requiring therapy",
    "Right atrium area (cm2)", "Right ventricular area (cm2)", "TAPSE (cm)",
    "Pericardial effusion on echo", "PAPsys (on echo)?", "Diastolic Dysf. (Echo)", "LVEF %"
  ),
  Gastrointestinal = c(
    "Stomach Symptoms", "GAVE", "Intestinal Symptoms", "Malabsorption syndrome", "Proximal Dysphagia"
  ),
  Labs_and_Biomarkers = c(
    "CRP", "Creatinine", "NT-proBNP (pg/ml)", "Uric Acid (mg/dl)", 
    "Hemoglobin (g/dl)", "CK value in serum (U/U/L)", "Proteinuria (>300mg/d)",
    "SSc Antibodies", "SSc Antibody Score"
  )
)

# Helper function to determine parametric status (Kept local as requested)
check_parametric <- function(df, var_col, group_col, config, transform = NULL) { 
  test_vec <- df[[var_col]][!is.na(df[[var_col]])]
  group_vec <- df[[group_col]][!is.na(df[[var_col]])]
  
  if (length(unique(group_vec)) < 2 || length(test_vec) < 3) return(FALSE)
  
  if (!is.null(transform) && transform == "log10") {
    valid_idx <- test_vec > 0
    test_vec <- log10(test_vec[valid_idx])
    group_vec <- group_vec[valid_idx]
    
    if (length(unique(group_vec)) < 2 || length(test_vec) < 3) return(FALSE)
  }
  
  shapiro_p <- tryCatch(shapiro.test(test_vec)$p.value, error = function(e) 0)
  bartlett_p <- tryCatch(bartlett.test(test_vec ~ group_vec)$p.value, error = function(e) 0)
  
  return(shapiro_p > config$shapiro_alpha && bartlett_p > config$variance_alpha)
}

# 3. Table Configurations ------------------------------------------------------
table_configs <- list(
  table_1 = list(
    title = "Baseline Demographics and Clinical Characteristics",
    group_by = "cohort_group",
    pairwise = TRUE, 
    variables = c("Age", "Sex", "CRP", "Creatinine", "Sample_Age"),
    labels = list(
      Age = "Patient Age (Years)",
      Sex = "Biological Sex",
      CRP = "CRP (mg/dl)",
      Creatinine = "Serum Creatinine (µmol/L)",
      Sample_Age = "Sample Storage Time (Years)"
    )
  )
)

# 4. Figure Configurations -----------------------------------------------------
figure_configs <- list(
  figure_1 = list(
    figure_title = "Baseline Clinical Distributions", 
    plots = list(
      plot_1 = list(
        title = "Age and Sex Distribution", 
        type = "stacked_histogram",
        x_var = "Age",
        fill_var = "Sex",
        fill_palette = get_project_colors(c(SEX_FEMALE, SEX_MALE)), 
        facet_by = "cohort_group",
        bins = 10  
      ),
      plot_4 = list(
        title = "Sample Storage Age", 
        type = "violin",
        y_var = "Sample_Age",
        x_var = "cohort_group",
        y_transform = NULL,           
        y_label = "Sample Storage Time (Years)", 
        add_stat_test = TRUE,
        pairwise = TRUE
      )
    )
  ),
  
  figure_2 = list(
    figure_title = "Baseline Laboratory Values", 
    plots = list(
      plot_2 = list(
        title = "Baseline CRP", 
        type = "violin",
        y_var = "CRP",
        x_var = "cohort_group",
        y_transform = "log10",         
        y_label = "CRP (mg/dl) [Log10 Scale]",  
        add_stat_test = TRUE,
        pairwise = TRUE
      ),
      plot_3 = list(
        title = "Baseline Creatinine", 
        type = "violin",
        y_var = "Creatinine",
        x_var = "cohort_group",
        y_transform = "log10", 
        y_label = "Serum Creatinine (µmol/L) [Log10 Scale]", 
        add_stat_test = TRUE,
        pairwise = TRUE
      )
    )
  )
)
```

# Tables
`r what_this_does("Dynamically evaluated each clinical variable to select appropriate parametric (ANOVA/t-test) or non-parametric (Kruskal-Wallis/Wilcoxon) statistical tests.", "Generated comprehensive baseline demographic tables using the gtsummary package.", "Calculated and appended both global and pairwise group comparison p-values.", "Exported standard tables as CSV files and rendered interactive HTML equivalents.")`
```{r table_engine, results='asis'}
generated_tables <- list()

for (tbl_name in names(table_configs)) {
  conf <- table_configs[[tbl_name]]
  cat(sprintf("### %s\n\n", conf$title))
  
  tbl_data <- master_spine %>% select(all_of(c(conf$variables, conf$group_by)))
  
  gtsum_stats <- list(); gtsum_global_tests <- list(); gtsum_pairwise_tests <- list() 
  for (var in conf$variables) {
    if (is.numeric(tbl_data[[var]])) {
      is_parametric <- check_parametric(tbl_data, var, conf$group_by, stat_config, transform = NULL) 
      if (is_parametric) {
        gtsum_stats[[var]] <- "{mean} ({sd})"; gtsum_global_tests[[var]] <- "aov"; gtsum_pairwise_tests[[var]] <- "t.test" 
      } else {
        gtsum_stats[[var]] <- "{median} ({p25}, {p75})"; gtsum_global_tests[[var]] <- "kruskal.test"; gtsum_pairwise_tests[[var]] <- "wilcox.test" 
      }
    }
  }
  
  tbl_global <- tbl_data %>%
    tbl_summary(by = all_of(conf$group_by), statistic = gtsum_stats, missing = "ifany", missing_text = "Missing (NA)") %>%
    add_p(test = gtsum_global_tests) %>% modify_header(p.value = "**Global p-value**")
  
  if (isTRUE(conf$pairwise)) {
    valid_groups <- tbl_data %>% group_by(!!sym(conf$group_by)) %>% summarise(n = n(), .groups = "drop") %>% filter(n >= 2) %>% pull(!!sym(conf$group_by)) %>% as.character()
    all_comparisons <- combn(valid_groups, 2, simplify = FALSE)
    pairwise_tbls <- list()
    for (comp in all_comparisons) {
      comp_data <- tbl_data %>% filter(!!sym(conf$group_by) %in% comp) %>% mutate(!!sym(conf$group_by) := droplevels(factor(!!sym(conf$group_by))))
      ptbl <- comp_data %>% tbl_summary(by = all_of(conf$group_by), statistic = gtsum_stats, missing = "no") %>% add_p(test = gtsum_pairwise_tests) %>% modify_header(p.value = sprintf("**%s vs %s**", comp[1], comp[2])) %>% modify_column_hide(all_stat_cols()) 
      pairwise_tbls <- append(pairwise_tbls, list(ptbl))
    }
    gtsum_obj <- tbl_merge(tbls = append(list(tbl_global), pairwise_tbls), tab_spanner = FALSE)
  } else {
    gtsum_obj <- tbl_global
  }
  
  gtsum_obj <- gtsum_obj %>% bold_labels() %>% modify_header(label = "**Variable**")
  write_csv(as_tibble(gtsum_obj), file.path(output_dir_data, paste0(tbl_name, ".csv")))
  
  tbl_html <- as_kable_extra(gtsum_obj, format = "html") %>% kable_styling(bootstrap_options = c("hover", "condensed"), full_width = FALSE, position = "left") %>% row_spec(0, extra_css = "white-space: nowrap;") 
  print(tbl_html)
  cat("\n---\n\n")
  generated_tables[[tbl_name]] <- gtsum_obj
}

# Clean Memory
rm(tbl_data, gtsum_stats, gtsum_global_tests, gtsum_pairwise_tests, tbl_global, gtsum_obj, tbl_html, conf)
invisible(gc())
```

# Figures
`r what_this_does("Generated dynamic clinical visualizations (stacked histograms, violin plots) based on the configuration dictionary.", "Automatically applied appropriate parametric or non-parametric pairwise statistical comparisons to the plots.", "Assembled individual plots into comprehensive multi-panel figure layouts using patchwork.")`
```{r figure_engine}
group_levels <- levels(master_spine$cohort_group) 
plot_colors <- get_project_colors(group_levels) 

individual_plots <- list()
stitched_figures <- list()

generate_base_plot <- function(conf) {
  if (conf$type == "stacked_histogram") {
    hist_bins <- if (!is.null(conf$bins)) conf$bins else 30
    p <- ggplot(master_spine, aes(x = .data[[conf$x_var]], fill = .data[[conf$fill_var]])) + 
      geom_histogram(bins = hist_bins, color = "white", alpha = 0.8) +
      scale_fill_manual(values = conf$fill_palette) + 
      labs(title = conf$title, x = conf$x_var, y = "Count", fill = conf$fill_var)
    if (!is.null(conf$facet_by)) p <- p + facet_wrap(vars(!!sym(conf$facet_by)))
    
  } else if (conf$type == "violin") {
    plot_data <- master_spine %>% filter(!is.na(.data[[conf$y_var]]))
    actual_y_label <- if (!is.null(conf$y_label)) conf$y_label else conf$y_var
    
    p <- ggplot(plot_data, aes(x = .data[[conf$x_var]], y = .data[[conf$y_var]])) + 
      geom_violin(aes(fill = .data[[conf$x_var]]), alpha = 0.4, color = "transparent") +
      geom_boxplot(width = 0.2, fill = "white", color = "grey30", outlier.shape = NA) +
      geom_jitter(aes(color = .data[[conf$x_var]]), width = 0.15, alpha = 0.6) +
      scale_fill_manual(values = plot_colors) + scale_color_manual(values = plot_colors) +
      labs(title = conf$title, x = "Cohort Group", y = actual_y_label) + theme(legend.position = "none")
    
    if (!is.null(conf$y_transform) && conf$y_transform == "log10") {
      p <- p + scale_y_log10(expand = expansion(mult = c(0.05, 0.40)))
    } else {
      p <- p + scale_y_continuous(expand = expansion(mult = c(0.05, 0.40)))
    }
    
    if (isTRUE(conf$add_stat_test)) {
      is_parametric <- check_parametric(plot_data, conf$y_var, conf$x_var, stat_config, transform = conf$y_transform) 
      dyn_pairwise_method <- if (is_parametric) "t.test" else "wilcox.test"
      dyn_global_method <- if (is_parametric) "anova" else "kruskal.test"
      
      if (isTRUE(conf$pairwise)) {
        valid_groups <- plot_data %>% group_by(!!sym(conf$x_var)) %>% summarise(n = n(), .groups = "drop") %>% filter(n >= 2) %>% pull(!!sym(conf$x_var)) %>% as.character()
        if (length(valid_groups) >= 2) {
          all_comparisons <- combn(valid_groups, 2, simplify = FALSE)
          sig_comparisons <- list()
          for (comp in all_comparisons) {
            g1 <- plot_data[[conf$y_var]][plot_data[[conf$x_var]] == comp[1]]
            g2 <- plot_data[[conf$y_var]][plot_data[[conf$x_var]] == comp[2]]
            if (length(g1) >= 2 && length(g2) >= 2) {
              if (!is.null(conf$y_transform) && conf$y_transform == "log10") {
                g1 <- log10(g1[g1 > 0]); g2 <- log10(g2[g2 > 0])
              }
              p_val <- tryCatch({
                if (dyn_pairwise_method == "t.test") t.test(g1, g2)$p.value else wilcox.test(g1, g2, exact = FALSE)$p.value 
              }, error = function(e) 1) 
              if (isTRUE(p_val < stat_config$p_cutoff)) sig_comparisons <- append(sig_comparisons, list(comp))
            }
          }
          if (length(sig_comparisons) > 0) p <- p + stat_compare_means(comparisons = sig_comparisons, method = dyn_pairwise_method, label = "p.signif", hide.ns = TRUE, step.increase = 0.1)
        }
      } else {
        p <- p + stat_compare_means(method = dyn_global_method, label.y.npc = "top")
      }
    }
  }
  p <- p + theme_project_base()
  return(p)
}

for (fig_name in names(figure_configs)) {
  fig_conf <- figure_configs[[fig_name]]
  patchwork_ready_plots <- list()
  plot_names <- names(fig_conf$plots)
  for (i in seq_along(plot_names)) {
    p_name <- plot_names[i]
    p_conf <- fig_conf$plots[[p_name]]
    pure_plot <- generate_base_plot(p_conf)
    individual_plots[[p_name]] <- pure_plot
    abc_title <- sprintf("%s. %s", LETTERS[i], p_conf$title)
    patchwork_ready_plots[[i]] <- pure_plot + labs(title = abc_title)
  }
  stitched_figures[[fig_name]] <- wrap_plots(patchwork_ready_plots, ncol = 2) + plot_annotation(title = fig_conf$figure_title, theme = theme(plot.title = element_text(size = 16, face = "bold")))
}

# Clean Memory
rm(pure_plot, patchwork_ready_plots, p_name, p_conf, abc_title, fig_conf, plot_names, i)
invisible(gc())
```

## Merged Figures
`r what_this_does("Rendered the assembled multi-panel figures for the HTML report.", "Exported high-resolution PNG copies of the merged figures to the output directory.")`
```{r print_figures, fig.width=12, fig.height=8}
for (fig_name in names(stitched_figures)) {
  print(stitched_figures[[fig_name]])
  ggsave(filename = file.path(output_dir_data, paste0(fig_name, ".png")), plot = stitched_figures[[fig_name]], width = 12, height = 8, dpi = 300, bg = "white")
}
```

## Individual Plots {.tabset}
`r what_this_does("Generated tabbed sections in the HTML report for easy navigation of individual plots.", "Exported standalone high-resolution PNGs for each individual plot.")`
```{r print_individual_plots, results='asis', fig.width=7, fig.height=5.5}
for (p_name in names(individual_plots)) {
  tab_title <- individual_plots[[p_name]]$labels$title
  if (is.null(tab_title)) tab_title <- p_name
  
  cat(sprintf("\n\n### %s\n\n", tab_title))
  print(individual_plots[[p_name]])
  cat("\n\n")
  ggsave(filename = file.path(output_dir_data, paste0(p_name, ".png")), plot = individual_plots[[p_name]], width = 7, height = 5.5, dpi = 300, bg = "white")
}
```

# Matrix Correlation
## Factor Exploration
`r what_this_does("Scanned all factor variables to identify their category distributions and assess variance.", "Filtered out cleanly binarized variables to reduce noise.", "Generated an audit table grouping remaining factors by level counts (>2 levels, exactly 2 non-standard levels, or 0 variance) to guide manual binarization.")`
```{r factor_exploration}
get_level_pct <- function(vec) {
  vec <- vec[!is.na(vec)]
  if(length(vec) == 0) return("All NA")
  tbl <- prop.table(table(vec)) * 100
  level_strings <- paste0(names(tbl), ": ", sprintf("%.1f%%", tbl))
  return(paste(level_strings, collapse = " <br> ")) 
}

exploration_df <- data.frame(Variable = character(), Category = character(), Distribution = character(), stringsAsFactors = FALSE)

for (col in names(master_spine)) {
  vec <- master_spine[[col]]
  if (is.factor(vec)) {
    clean_vec <- vec[!is.na(vec)]
    unique_levels <- unique(as.character(clean_vec))
    n_levels <- length(unique_levels)
    dist_str <- get_level_pct(vec)
    cat_name <- NA
    
    if (n_levels <= 1) {
      cat_name <- "3: Zero Variance (<= 1 Level)"
    } else if (n_levels > 2) {
      cat_name <- "1: > 2 Levels (Needs Bucketing/Exclusion)"
    } else if (n_levels == 2) {
      lower_levels <- tolower(unique_levels)
      is_strict_binary <- all(lower_levels %in% c("yes", "no", "1", "0"))
      if (!is_strict_binary) cat_name <- "2: Exactly 2 Levels (Needs Manual 1/0 Mapping)"
    }
    
    if (!is.na(cat_name)) {
      exploration_df <- rbind(exploration_df, data.frame(Variable = col, Category = cat_name, Distribution = dist_str, stringsAsFactors = FALSE))
    }
  }
}

if(nrow(exploration_df) > 0) {
  exploration_df %>% arrange(Category, Variable) %>% kableExtra::kbl(format = "html", escape = FALSE, caption = "Variables Requiring Binarization Attention") %>% kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left") %>% kableExtra::column_spec(1, bold = TRUE) %>% kableExtra::collapse_rows(columns = 2, valign = "top") 
} else {
  cat("**Result:** All factor variables are cleanly binary (1/0) and ready for the correlation matrix!\n\n")
}

# Clean Memory
rm(exploration_df, vec, clean_vec, unique_levels, dist_str)
invisible(gc())
```


## Data Prep and Matrix Calculation
`r analysis_decision("Filtered out sparse variables with < 10 observations to ensure robust correlation metrics.", "Standardized categorical data into ordinal ranks.", "Applied a global NA-to-Zero fill on the correlation matrix to stabilize hierarchical clustering.")`

`r what_this_does("Standardized clinical factors into ordinal ranks and calculated the master Spearman correlation matrix.", "Implemented a global NA-to-Zero fill to ensure mathematical stability during hierarchical clustering.", "Generated transparent audit tables documenting numeric mapping and variable exclusion for clinical validation.", "Exported the finalized correlation matrix to CSV and cleared intermediate variables from RAM.")`

```{r mat_corr_calc, warning=FALSE}
# 1. Audit Table: Binarization & Ordinal Mapping -------------------------------
mapping_audit_df <- data.frame(
  Source_Category = c("Sex", "Cohort Group", "Smoking/History", "Severity Ranks", "Clinical Visits"),
  Original_Levels = c("Female / Male", "Control / Non-Active / Active", "Never / Previously / Current", "None / Rare / Moderate / Extensive", "Visit 0 / FU 1 / FU 2"),
  Numeric_Value = c("1 / 0", "0 / 1 / 2", "0 / 1 / 2", "0 / 1 / 2 / 3", "0 / 1 / 2"),
  stringsAsFactors = FALSE
)

# 2. Execution: Translation & Filtering ----------------------------------------
heatmap_dir <- file.path(output_dir_data, "heatmaps")
if (!dir.exists(heatmap_dir)) dir.create(heatmap_dir, recursive = TRUE)

cor_data_prep <- master_spine %>%
  mutate(
    Subset_Diffuse = if_else(`Subsets of SSc according to LeRoy (1988):` == "diffuse cutaneous", 1, 0),
    Subset_Limited = if_else(`Subsets of SSc according to LeRoy (1988):` == "limited cutaneous", 1, 0),
    Subset_Sine    = if_else(`Subsets of SSc according to LeRoy (1988):` == "sine scleroderma", 1, 0)
  ) %>%
  select(-`Subsets of SSc according to LeRoy (1988):`) %>%
  mutate(across(where(is.factor), ~ {
    val <- tolower(as.character(.x))
    case_when(
      val == "1" ~ 1, val == "0" ~ 0,
      val %in% c("weiblich", "female") ~ 1, val %in% c("männlich", "male") ~ 0,
      val == tolower(GRP_CONTROL)     ~ 0, 
      val == tolower(GRP_NON_ACTIVE)  ~ 1, 
      val == tolower(GRP_ACTIVE)      ~ 2,
      val == "never" ~ 0, val == "previously" ~ 1, val == "current" ~ 2,
      val == "none" ~ 0, val == "rare" ~ 1, val == "moderate" ~ 2, val == "extensive" ~ 3,
      val %in% c("1", "i") ~ 1, val == "ii" ~ 2, val == "iii" ~ 3,
      TRUE ~ NA_real_ 
    )
  })) %>%
  select(where(is.numeric))

# 3. Audit Table: Exclusions ---------------------------------------------------
valid_counts <- colSums(!is.na(cor_data_prep))
dropped_cols <- names(valid_counts)[valid_counts < corr_config$min_required_obs]

exclusion_audit_df <- data.frame(
  Variable = names(valid_counts),
  Observations = as.numeric(valid_counts),
  Status = if_else(names(valid_counts) %in% dropped_cols, "DROPPED", "KEPT"),
  Reason = if_else(names(valid_counts) %in% dropped_cols, 
                   sprintf("Sparse data (< %d obs)", corr_config$min_required_obs), 
                   "Passed Sparsity Filter"),
  stringsAsFactors = FALSE
) %>% filter(Status == "DROPPED")

cor_mat_final <- cor_data_prep %>% select(all_of(names(valid_counts)[valid_counts >= corr_config$min_required_obs]))

# SANITY CHECK: Ensure we have enough columns to run a correlation matrix
stopifnot("FATAL: Insufficient columns remaining for correlation matrix." = ncol(cor_mat_final) >= 2)

# 4. Math Engine ---------------------------------------------------------------
cor_matrix <- cor(cor_mat_final, use = "pairwise.complete.obs", method = "spearman")
cor_matrix[is.na(cor_matrix)] <- 0 

p_matrix <- matrix(1, ncol(cor_mat_final), ncol(cor_mat_final), 
                   dimnames = list(colnames(cor_mat_final), colnames(cor_mat_final)))

for (i in 1:(ncol(cor_mat_final) - 1)) {
  for (j in (i + 1):ncol(cor_mat_final)) {
    if(sum(complete.cases(cor_mat_final[[i]], cor_mat_final[[j]])) >= 3) {
      p_matrix[i, j] <- p_matrix[j, i] <- tryCatch({ cor.test(cor_mat_final[[i]], cor_mat_final[[j]], method="spearman")$p.value }, error=function(e) 1)
    }
  }
}

write_csv(as.data.frame(cor_matrix), file.path(output_dir_data, "clinical_spearman_rho.csv"))

# 5. Render Audit Tables (Aggregated at end of chunk) --------------------------
cat("### Data Transformation Audit Report\n\n")

# Table 1: Mappings
print(mapping_audit_df %>% 
  kableExtra::kbl(caption = "1. Binarization and Ordinal Mapping Dictionary", format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::scroll_box(width = "100%", height = "300px"))

cat("\n\n")

# Table 2: Exclusions
if(nrow(exclusion_audit_df) > 0) {
  print(exclusion_audit_df %>% 
    kableExtra::kbl(caption = "2. Variable Exclusion Log (Sparsity Filter)", format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
    kableExtra::scroll_box(width = "100%", height = "300px"))
} else {
  cat("#### 2. Variable Exclusion Log\n\n- No variables were dropped due to sparsity.\n")
}

# Clear Memory
rm(cor_data_prep, valid_counts, dropped_cols, mapping_audit_df, exclusion_audit_df)
invisible(gc())
```


### How to Interpret These Heatmaps
These networks visualize the **Spearman Rank Correlation** between clinical features. Because clinical data is often non-linear and ordinal (e.g., severity scores), Spearman accurately captures whether two variables move together, regardless of their exact numeric scale.

* **Color (Direction):** <span style="color:#d73027;">**Red**</span> indicates a positive correlation (as Variable A increases, Variable B increases). <span style="color:#4575b4;">**Blue**</span> indicates a negative correlation (as Variable A increases, Variable B decreases).
* **Intensity (Strength):** Darker, more saturated colors represent stronger mathematical relationships (closer to +1.0 or -1.0).
* **Blank Spaces (Significance):** If a square is entirely blank, the correlation failed to reach statistical significance (p ≥ 0.05). Only statistically robust relationships are displayed.
* **Clustering:** In the Intra-Domain plots (Self vs. Self), variables are automatically ordered using hierarchical clustering (`ward.D2`). Variables that behave similarly across the cohort will group tightly together.

## Intra Domain Correlation MAtrix {.tabset}

`r what_this_does("Visualized local correlations within clinical domains using hierarchical clustering.", "Optimized text scaling and margins for maximum legibility.", "Dynamically saved high-resolution PNGs to the heatmap output subfolder.")`
```{r mat_corr_plot, results='asis'}
# 1. Shared Aesthetics
col_palette <- colorRampPalette(c(HM_Z_LOW, HM_Z_MID, HM_Z_HIGH))(200)

# 2. Filter domains for plotting
for (d_name in names(clinical_domains)) {
  
  raw_vars <- clinical_domains[[d_name]]
  existing_vars <- intersect(raw_vars, colnames(cor_matrix))
  
  if (length(existing_vars) < 2) next

  v_vals <- sapply(cor_mat_final[, existing_vars, with=FALSE], function(x) var(x, na.rm=TRUE))
  valid_vars <- names(v_vals)[!is.na(v_vals) & v_vals > 0]
  
  if (length(valid_vars) < 2) {
    cat(sprintf("### %s\n", gsub("_", " ", d_name)))
    cat("\n> **Note:** Domain skipped. Insufficient variance detected.\n\n")
    next
  }

  sub_cor <- cor_matrix[valid_vars, valid_vars, drop = FALSE]
  sub_p   <- p_matrix[valid_vars, valid_vars, drop = FALSE]
  
  if (!all(dim(sub_cor) == dim(sub_p))) next

  cat(sprintf("### %s\n", gsub("_", " ", d_name)))
  
  dyn_size <- 4 + (length(valid_vars) * 0.5)
  png(file.path(output_dir_data, "heatmaps", paste0("Intra_", d_name, ".png")), 
      width=dyn_size, height=dyn_size, units="in", res=300)
  
  try({
    corrplot(sub_cor, method="color", type="upper", order="hclust", 
             col=col_palette, tl.col="black", tl.srt=45,
             tl.cex=corr_config$label_font_size, 
             p.mat=sub_p, sig.level=corr_config$p_cutoff, insig="blank", 
             diag=FALSE, title=gsub("_", " ", d_name), 
             mar=c(0, 0, 2, 0))
  }, silent = TRUE)
  dev.off()
  
  # Render to HTML
  try({
    corrplot(sub_cor, method="color", type="upper", order="hclust", 
             col=col_palette, tl.col="black", tl.cex=corr_config$label_font_size, tl.srt=45, 
             p.mat=sub_p, sig.level=corr_config$p_cutoff, insig="blank", 
             diag=FALSE, mar=c(0, 0, 1, 0))
  }, silent = TRUE)
  
  cat("\n\n")
}

# Cleanup
rm(sub_cor, sub_p, valid_vars, existing_vars, v_vals)
invisible(gc())
```
## Interdomain Combos {.tabset}

`r what_this_does("Visualized correlations between different biological systems using non-square matrices.", "Dynamically saved high-resolution PNGs to the heatmap output subfolder.")`
```{r inter_domain, results='asis'} 
# 1. Regenerate the safe plot list (ensuring names exist in the matrix)
plotting_domains <- list()
for (n in names(clinical_domains)) {
  v <- intersect(clinical_domains[[n]], colnames(cor_matrix))
  if (length(v) >= 1) plotting_domains[[n]] <- v
}

domain_combos <- combn(names(plotting_domains), 2, simplify = FALSE)
col_palette <- colorRampPalette(c(HM_Z_LOW, HM_Z_MID, HM_Z_HIGH))(200)

for (combo in domain_combos) {
  d1_name <- combo[1]; d2_name <- combo[2]
  v1_raw <- plotting_domains[[d1_name]]; v2_raw <- plotting_domains[[d2_name]]
  
  # Safety Filters
  v1_vals <- sapply(cor_mat_final[, v1_raw, with=FALSE], function(x) var(x, na.rm=TRUE))
  v1_valid <- names(v1_vals)[!is.na(v1_vals) & v1_vals > 0]
  v2_vals <- sapply(cor_mat_final[, v2_raw, with=FALSE], function(x) var(x, na.rm=TRUE))
  v2_valid <- names(v2_vals)[!is.na(v2_vals) & v2_vals > 0]
  
  if (length(v1_valid) < 1 || length(v2_valid) < 1) next 
  
  sub_cor <- cor_matrix[v1_valid, v2_valid, drop = FALSE]
  sub_p   <- p_matrix[v1_valid, v2_valid, drop = FALSE]
  
  if (!all(dim(sub_cor) == dim(sub_p))) next

  # Tracking for Domain Labels
  lab_y <- d1_name; lab_x <- d2_name

  # Always display larger axis as X 
  if (nrow(sub_cor) > ncol(sub_cor)) {
    sub_cor <- t(sub_cor)
    sub_p   <- t(sub_p)
    lab_y <- d2_name; lab_x <- d1_name
  }

  cat(sprintf("### %s vs %s\n", gsub("_", " ", d1_name), gsub("_", " ", d2_name)))
  
  dynamic_width  <- 4 + (ncol(sub_cor) * 0.7)
  dynamic_height <- 4 + (nrow(sub_cor) * 0.7)
  
  # Export PNG
  png(file.path(output_dir_data, "heatmaps", paste0("Inter_", d1_name, "_vs_", d2_name, ".png")), 
      width = dynamic_width, height = dynamic_height, units = "in", res = 300)
  
  try({
    corrplot(sub_cor, method="color", type="full", col=col_palette, 
             tl.col="black", tl.srt=45, tl.cex=corr_config$label_font_size,
             p.mat=sub_p, sig.level=corr_config$p_cutoff, insig="blank", 
             mar=c(2, 4, 4, 2)) 
    
    mtext(gsub("_", " ", lab_x), side = 3, line = 2, cex = 1.6, font = 2)
    mtext(gsub("_", " ", lab_y), side = 2, line = 2, cex = 1.6, font = 2)
  }, silent = TRUE)
  dev.off()
  
  # Render to HTML
  try({
    corrplot(sub_cor, method="color", type="full", col=col_palette, 
             tl.col="black", tl.srt=45, tl.cex=corr_config$label_font_size,
             p.mat=sub_p, sig.level=corr_config$p_cutoff, insig="blank", 
             mar=c(2, 4, 2, 2))
    
    mtext(gsub("_", " ", lab_x), side = 3, line = 0.5, cex = 1.2, font = 2)
    mtext(gsub("_", " ", lab_y), side = 2, line = 2, cex = 1.2, font = 2)
  }, silent = TRUE)
  
  cat("\n\n")
}

# Final Clean up
graphics.off()
rm(sub_cor, sub_p, plotting_domains, col_palette, v1_valid, v2_valid, d1_name, d2_name)
invisible(gc())
```

```{r finish_message}
cat("Script 02 Complete!")
```

