---
title: "Clinical Feature Clustering (Sobanski Replication)"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction
This script performs unsupervised clustering on clinical baseline characteristics to discover distinct clinical phenotypes within our SSc cohort, drawing methodology from Sobanski et al. (2019). Because clinical data features a mix of continuous and categorical variables, we employ Random Forest imputation followed by two parallel dimensionality reduction strategies: Gower's Distance and Factor Analysis of Mixed Data (FAMD).

# Setup and Import
`r what_this_does("Loaded FactoMineR for mixed-data dimensionality reduction and missForest for Random Forest imputation.", "Imported the pre-cleaned clinical master spine.", "Sourced the custom omics clustering engines.")`

```{r setup_and_import}
library(tidyverse)
library(patchwork)
library(here)
library(missForest) # Random Forest Imputation
library(FactoMineR) # FAMD
library(factoextra) # FAMD Visuals
library(cluster)    # Gower & PAM
library(gtsummary)  # Clinical Tables
library(msigdbr)

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "02b_clinical_clustering"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Configuration & Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
data_dictionary <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
olink_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

# Fetch multiple GSEA databases into a named list
pathway_dbs <- list(
  "Hallmark" = msigdbr(species = "Homo sapiens", category = "H") %>% select(gs_name, gene_symbol),
  "GO_BP" = msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% select(gs_name, gene_symbol),
  "Immunologic_Signature" = msigdbr(species = "Homo sapiens", category = "C7") %>% select(gs_name, gene_symbol),
  "Reactome" = msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME") %>% select(gs_name, gene_symbol)
)
```

# Configuration
`r what_this_does("Defined the exact 24 clinical variables mandated by  Sobanski's publication.", "Configured the FAMD parameters to extract the top components representing the mixed-data variance.", "Established dual clustering jobs")`

## Clustering

```{r cluster_config}
# 1. Variable Selection (Matching the 24 vars from Sobanski 2019)
sobanski_vars <- c(

  # Symptoms / organ involvement
  "Raynaud phenomenon",
  "Esophageal symptom at the time of visit",
  "Stomach Symptoms",
  "Intestinal Symptoms",
  "Digital ulcer",
  "Joint involvement",                  # APPROXIMATION Sobanksi used "Joint Synovitis"
  "Joint Contractures",
  "Tendon friction rubs",
  "Proximal muscle weakness",
  "Muscle atrophy",
  # Not available: Arterial hypertension, palpitations, renal crisis

  # Laboratory values
  "Anti-centromere (ACA)",
  "Anti-Scl70",                         # Corresponds to anti-topoisomerase I
  "Anti-RNA polymerase III",            # APPROXIMATION Sobanski used "Anti Nuclear Antibody"
  "CK value in serum (U/L)",
  "Proteinuria (>300mg/d)",

  # Other tests / organ assessments
  "ILD diagnosed via HRCT",             # APPROXIMATION Sobanski used "Lung fibrosis on plain radiography" maybe honeycombing better?
  "FVC %pred",                          # APPROXIMATION Sobanski used "Restrictive defect on Pulmonary Function Test"
  "PH/ILD",                             # APPROXIMATION Sobanski used "suspected PH on cardiac echography"
  # Not available: Conduction blocks, abnormal diastolic function

  # Skin involvement
  "Modified rodnan skin score",

  # Demographics / covariates
  "Age",
  "Sex",
  "cohort_group"
)

# Filter variables to ensure they actually exist in your master_spine
valid_clustering_vars <- intersect(sobanski_vars, colnames(master_spine))

# 2. FAMD Configs
famd_configs <- list(
  famd_clinical = list(
    title = "FAMD: Clinical Baseline Sobanski Variables",
    max_dims = 10,  # Calculates up to 10 dims for the scree plot context
    top_dim = 5,    # NEW: Strictly subsets the matrix to 5 dims for clustering
    color_vars = c("cohort_group", "Sex", "Age")
  )
)

# 3. Optimization Sweeps
k_opt_configs <- list(
  opt_gower = list(
    title = "Optimization: Gower Distance Matrix",
    input_matrix = "mat_gower", 
    k_range = 2:8,
    is_dist_matrix = TRUE, # <--- Triggers the new Silhouette engine
    algorithms = list(
      pam = list(run=TRUE),
      hclust = list(run=TRUE, method="ward.D2")
    )
  ),
  opt_famd = list(
    title = "Optimization: FAMD Dimension Matrix",
    input_matrix = "mat_famd", 
    k_range = 2:8,
    gap_bootstraps = 0,
    is_dist_matrix = FALSE, # <--- Triggers classic fviz_nbclust
    algorithms = list(
      kmeans = list(run=TRUE, nstart=25),
      pam = list(run=TRUE, metric="euclidean"),
      hclust = list(run=TRUE, distance="euclidean", method="ward.D2")
    )
  )
)

# 4. Clustering Configs
clustering_configs <- list(

  job_famd = list(
    title = "FAMD Coordinate Clustering",
    input_matrix = "mat_famd", # Defined below
    k = 2, 
    n_neighbors = 10,
    is_dist_matrix = FALSE,
    algorithms = list(
      hclust = list(run=TRUE, distance="euclidean", method="ward.D2") 
    )
  )
)

# 5. Diagnostic Configs
diagnostic_configs <- list(
  evaluate_jobs = c("job_famd"),
  correlate_vars = c("cohort_group", "Age", "Sex")
)

```

## Phenotyping
```{r phenotype_config}
# 6. Clinical Phenotyping Targets
clin_vars_to_highlight <- list(
  "Sobanski Clustering Parameters" = valid_clustering_vars # Highlights the exact 24 vars used for clustering
)

clin_vars_to_include <- list(
  "Exploratory" = c(clinical_domains$Demographics_and_Scores, clinical_domains$Pulmonary, 
                    clinical_domains$Cardiac, clinical_domains$Labs_and_Biomarkers, 
                    clinical_domains$Cutaneous_and_Vascular, clinical_domains$Musculoskeletal, 
                    clinical_domains$Gastrointestinal) 
)

pheno_configs <- list(
  pheno_famd = list(
    target_job = "job_famd",
    title = "Clinical Phenotypes: FAMD Clustering",
    target_algorithms = c("hclust", "pam", "kmeans"), # You can evaluate multiple algorithms at once!
    subset_filter = list(column = "cohort_group", value = c("Active", "Non-Active")),
    variables = clin_vars_to_include,
    highlight_variables = clin_vars_to_highlight,
    p_cutoff = 0.05,
    must_include_vars = c("cohort_group", "Sex", "Age")
  )
)


# 7. Proteomics Feature Phenotyping (DEA)
feature_configs <- list(
  
  feat_famd_k2 = list(
    target_job = "job_famd",
    title = "Feature Phenotypes: FAMD Clustering",
    input_matrix = "raw", 
    covariates = c("Age"),
    # Algorithms are now the primary key to enable contrast-per-algorithm
    algorithms = list(
      hclust = list(
        contrasts = list(
          c1_vs_c2 = list(title = "Cluster 1 vs Cluster 2 Ssc only", desc = "", 
                          target=c("Cluster 1"), ref=c("Cluster 2"),
                          subset_filter = list(column = "cohort_group", value = c("Active", "Non-Active")))
        )
      )
    ),
    dea_p_cutoff = 0.05,
    dea_fc_cutoff = 0.58
  )
)

# 8. DEA UI Configs (Required for wrap_dea_pipeline)
heatmap_config <- list(top_n = 30, split_by_group = TRUE, row_font_size = 8, annotation_vars = c("cohort_group", "Age", "Sex"))
omics_config <- list(gsea_p_cutoff = 0.20, gsea_min_size = 3, network_p_cutoff = 0.1)
go_genesets <- list() # Assuming no explicit GSEA in this clinical script, but required by the wrapper
```



# Preprocessing & Imputation
`r analysis_decision("Restricted the dataset strictly to the SSc cohort, removing healthy controls to prevent the clustering algorithm from trivially splitting Sick vs. Healthy.", "Applied Random Forest imputation (missForest) to handle missing values. Unlike simple mean-imputation, RF captures non-linear clinical interactions to predict missing values for both continuous and categorical variables.")`

`r what_this_does("Subsets the master spine to the active and non-active SSc patients.", "Executes the wrap_rf_imputation engine to generate a gapless clinical matrix.", "Outputs a clear audit table displaying the degree of missingness resolved by the algorithm.")`

```{r preprocessing_and_imputation}
# 1. Filter Cohort & Select Variables
# We only want to cluster the SSc patients, so we drop controls.
ssc_only <- master_spine %>% 
  filter(cohort_group %in% c(GRP_ACTIVE, GRP_NON_ACTIVE)) %>%
  select(Subject_ID, all_of(valid_clustering_vars))

# 2. Execute Imputation Engine
imp_res <- wrap_rf_imputation(ssc_only, id_col = "Subject_ID", maxiter = 10, ntree = 100)

if (imp_res$status == "success") {
  clean_clin_df <- imp_res$imputed_df
  
  # Audit Report
  if (!is.null(imp_res$audit) && nrow(imp_res$audit) > 0) {
    cat("### Imputation Audit\n\n")
    print(
      imp_res$audit %>%
        kableExtra::kbl(format = "html", caption = "Missing Values Resolved via Random Forest") %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
    )
    cat(sprintf("\n\n**OOB Imputation Error:**\n- Continuous (NRMSE): %.3f\n- Categorical (PFC): %.3f\n\n", 
                imp_res$error_metrics[1], imp_res$error_metrics[2]))
  } else {
    cat("### Imputation Audit\n\n> Dataset was completely gapless. No imputation required.\n\n")
  }
} else {
  stop("Imputation failed.")
}

# Clean workspace
rm(ssc_only, imp_res)
invisible(gc())
```


# Dimensionality Reduction
`r analysis_decision("Calculated two distinct mathematical coordinate spaces. Gower distance preserves the raw scaling and binary nature of categorical variables.", "Factor Analysis of Mixed Data (FAMD) extracts continuous dimensions of variance. We strictly subsetted FAMD to the top 5 dimensions to aggressively strip statistical noise before downstream clustering.")`

`r what_this_does("Executes wrap_gower_distance to generate a patient-to-patient similarity matrix.", "Executes wrap_famd_pipeline to generate PCA-like continuous coordinates.", "Generates a Scree Plot and a ComplexHeatmap detailing the top clinical drivers of each selected dimension.")`

```{r dimensionality_reduction, fig.height=8, fig.width=16}
# 1. Gower Distance Matrix
prepared_matrices <- list()
gower_res <- wrap_gower_distance(clean_clin_df, id_col = "Subject_ID")
prepared_matrices[["mat_gower"]] <- gower_res$matrix

# 2. Factor Analysis of Mixed Data (FAMD)
famd_res <- wrap_famd_pipeline(
  df = clean_clin_df, 
  id_col = "Subject_ID", 
  famd_conf = famd_configs$famd_clinical, 
  famd_id = "Baseline_Clinical", 
  base_output_dir = output_dir_data, 
  project_colors_func = get_project_colors
)

prepared_matrices[["mat_famd"]] <- famd_res$matrix

# Print FAMD Diagnostics
cat("### FAMD Diagnostics\n\n")
print(famd_res$plots$scree | famd_res$plots$var_contrib)
cat("\n\n")

# Print the new Top Drivers Heatmap
if (!is.null(famd_res$plots$drivers_heatmap)) {
  cat("#### Top Dimension Drivers (FAMD)\n\n")
  ComplexHeatmap::draw(famd_res$plots$drivers_heatmap)
  cat("\n\n")
}

# Print Individual Overlays
for (c_var in famd_configs$famd_clinical$color_vars) {
  plot_key <- paste0("overlay_", c_var)
  
  if (!is.null(famd_res$plots[[plot_key]])) {
    cat(sprintf("#### Overlay: %s\n\n", c_var))
    print(famd_res$plots[[plot_key]])
    cat("\n\n")
  }
}

# Clean Workspace
rm(gower_res, famd_res)
invisible(gc())
```

# Algorithm Optimization
`r what_this_does("Executes algorithm-aware hyperparameter sweeps to evaluate multiple numbers of clusters.", "Evaluates Silhouette boundaries manually for Gower distance and via factored coordinate variance for FAMD.")`

```{r algorithm_optimization, results='asis', fig.width=10, fig.height=5}
for (opt_id in names(k_opt_configs)) {
  conf <- k_opt_configs[[opt_id]]
  cat(sprintf("\n\n### %s\n\n", conf$title))
  
  target_matrix <- prepared_matrices[[conf$input_matrix]]
  if (is.null(target_matrix)) next
  
  res <- wrap_k_optimization_pipeline(target_matrix, conf, opt_id, output_dir_data)
  
  if (res$status == "success") {
    for (algo in names(res$plots)) {
      print(res$plots[[algo]])
      cat("\n\n")
    }
  }
}
```


# Unsupervised Clustering {.tabset}
`r what_this_does("Ingests the matrices.", "Runs the specified algorithms at the requested number of clusters.", "Generates 2D UMAPs to visualize the discovered clusters.")`

```{r unsupervised_clustering, fig.width=10, fig.height=8}
global_cluster_payloads <- list()

for (c_id in names(clustering_configs)) {
  conf <- clustering_configs[[c_id]]
  cat(sprintf("\n\n## %s {.tabset}\n\n", conf$title))
  
  target_matrix <- prepared_matrices[[conf$input_matrix]]
  if (is.null(target_matrix)) next
  
  res <- wrap_clustering_pipeline(target_matrix, master_spine, conf, c_id, output_dir_data, get_project_colors)
  global_cluster_payloads[[c_id]] <- res
  
  if (res$status == "success") {
    for (algo in names(res$plots$umaps)) {
      cat(sprintf("\n\n### %s\n\n", toupper(algo)))
      
      # NEW: Print Dendrogram if it exists for this algorithm
      if (!is.null(res$plots$dendrograms[[algo]])) {
        cat("#### Hierarchical Dendrogram\n\n")
        print(res$plots$dendrograms[[algo]])
        cat("\n\n")
      }
      
      cat("#### UMAP Projection\n\n")
      print(res$plots$umaps[[algo]])
      cat("\n\n")
    }
  } 
}
cat("\n\n")
```

# Clustering Diagnostic {.tabset}
`r what_this_does("Calculates Silhouette scores to determine which dimensionality approach (Gower vs FAMD) and which algorithm generated the most mathematically distinct boundaries.", "Plots prediction confidence against key clinical variables to ensure the algorithm isn't biased by covariates like Age or Sex.")`

```{r clustering_diagnostic, results='asis', fig.width=10, fig.height=5}
global_diagnostic_payloads <- list()

for (job_id in diagnostic_configs$evaluate_jobs) {
  payload <- global_cluster_payloads[[job_id]]
  
  if (!is.null(payload) && payload$status == "success") {
    
    # Extract the exact matrix and config used for this job
    target_matrix <- prepared_matrices[[clustering_configs[[job_id]]$input_matrix]]
    cluster_conf <- clustering_configs[[job_id]]
    
    diag_res <- run_clustering_diagnostics(
      payload, target_matrix, master_spine, data_dictionary, 
      diagnostic_configs, cluster_conf, job_id, output_dir_data
    )
    
    global_diagnostic_payloads[[job_id]] <- diag_res
    
    if (!is.null(diag_res)) {
      # Inject Markdown Header for the Job Tabset
      cat(sprintf("\n\n## %s\n\n", cluster_conf$title))
      
      # 1. Print the Global HTML Leaderboard Table
      print(diag_res$plots[["leaderboard"]])
      cat("\n\n")
      
      # 2. Print Variable Correlation/Confidence Plots
      for (v in names(diag_res$plots)) {
        if (v == "leaderboard") next
        
        # Print the ggplot object
        print(diag_res$plots[[v]])
        cat("\n\n") # Crucial empty lines for pandoc rendering
      }
    }
  }
}
```

# Clinical Phenotype
`r analysis_decision("Muted non-significant phenotypes (p >= 0.05) to zero on the heatmap, ensuring visual emphasis is strictly placed on true biological drivers rather than random noise.")`

`r what_this_does("Calculates Log-Odds Ratios and Z-Scores to map the clinical fingerprint of the discovered endotypes.", "Generates statistical summary tables, contrast heatmaps, and highlight grids for streamlined visual profiling.")`

```{r phenotyping, fig.height=16, fig.width=14, warning=FALSE, results='asis'}
for (p_id in names(pheno_configs)) {
  conf <- pheno_configs[[p_id]]
  cat(sprintf("\n\n## %s {.tabset}\n\n", conf$title))
  
  payload <- global_cluster_payloads[[conf$target_job]]
  if (is.null(payload)) next
  
  res <- wrap_clinical_phenotyping(payload, master_spine, data_dictionary, conf, p_id, output_dir_data)
  
  if (res$status == "success") {
    for (algo in names(res$plots)) {
      # Inject Markdown Header for the Algorithm Tabset
      cat(sprintf("\n\n### %s {.tabset}\n\n", toupper(algo)))
      
      # Print Master Table
      cat("\n\n#### Summary Table\n\n")
      if (!is.null(res$tables[[algo]])) {
        tbl <- res$tables[[algo]] %>% 
          gtsummary::as_kable_extra(caption = sprintf("Clinical Summary: %s", toupper(algo))) %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
        print(tbl)
        cat("\n\n")
      }
      
      
      
      # --- Loop through nested Heatmaps ---
      if (length(res$plots[[algo]]$heatmap) > 0) {
        for (h_name in names(res$plots[[algo]]$heatmap)) {
          cat(sprintf("\n\n#### Heatmap: %s\n\n", gsub("_", " ", h_name)))
          ComplexHeatmap::draw(res$plots[[algo]]$heatmap[[h_name]])
          cat("\n\n")
        }
      }
      
     # --- Loop through nested Highlight Tables & Grids ---
      if (length(res$plots[[algo]]$highlights) > 0) {
        for (g_name in names(res$plots[[algo]]$highlights)) {
          cat(sprintf("\n\n#### Highlight: %s\n\n", gsub("_", " ", g_name)))
          
          # NEW: Print the targeted Highlight Table
          if (!is.null(res$tables$highlights[[algo]][[g_name]])) {
            print(res$tables$highlights[[algo]][[g_name]])
            cat("\n\n")
          }
          
          # Print the Highlight Grid Plot
          print(res$plots[[algo]]$highlights[[g_name]])
          cat("\n\n")
        }
      }
    }
  }
}
```

# Feature Phenotyping {.tabset}
`r analysis_decision("Projected the mathematically discovered clinical endotypes onto the Olink proteomics matrix to identify underlying molecular drivers.", "Applied Limma modeling controlling for Age and Sex to ensure the protein fold-changes are strictly tied to the phenotype.")`

`r what_this_does("Filters the clinical spine to the targeted clinical clusters.", "Executes rigorous Differential Expression Analysis (DEA) between Cluster 1 and Cluster 2.", "Generates volcano plots, top protein tables, and supervised heatmaps to visualize the molecular footprint of the clinical phenotypes.")`

```{r feature_phenotyping, fig.width=12, fig.height=10, results='asis'}
for (f_id in names(feature_configs)) {
  conf <- feature_configs[[f_id]]
  payload <- global_cluster_payloads[[conf$target_job]]
  target_matrix <- if(conf$input_matrix == "raw") olink_mat else prepared_matrices[[conf$input_matrix]]
  
  if (is.null(payload) || is.null(target_matrix)) next

  cat(sprintf("\n\n## %s {.tabset}\n\n", conf$title))

  for (algo_name in names(conf$algorithms)) {
    cat(sprintf("\n\n### %s {.tabset}\n\n", toupper(algo_name)))
    
    # Inject cluster labels
    temp_spine <- master_spine
    temp_spine$Dynamic_Cluster <- as.character(payload$data$assignments[[algo_name]][match(temp_spine$Subject_ID, names(payload$data$assignments[[algo_name]]))])
    temp_spine <- temp_spine %>% filter(!is.na(Dynamic_Cluster), Dynamic_Cluster != "Noise")

    algo_conf <- conf$algorithms[[algo_name]]
    
    for (contrast_name in names(algo_conf$contrasts)) {
      c_conf <- algo_conf$contrasts[[contrast_name]]
      
      cat(sprintf("\n\n#### %s\n\n", c_conf$title))
      cat(sprintf("**Description:** %s\n\n", c_conf$desc))

      contrast_df <- temp_spine
      
      dea_conf <- list(
        title = sprintf("%s | Alg: %s", c_conf$title, toupper(algo_name)),
        group_col = "Dynamic_Cluster",
        target_groups = c_conf$target, 
        ref_groups = c_conf$ref,        
        covariates = conf$covariates,
        dea_p_cutoff = conf$dea_p_cutoff,
        dea_fc_cutoff = conf$dea_fc_cutoff
      )
      
      # Explicitly define the DEA ID string so it can be reused locally
      current_dea_id <- sprintf("%s_%s_%s", f_id, algo_name, contrast_name)
      
      # Execute Limma Wrapper using the defined variable
      res <- wrap_dea_pipeline(target_matrix, contrast_df, dea_conf, current_dea_id, output_dir_data, go_genesets, omics_config, heatmap_config, get_project_colors)
      
      if (res$status == "success") {
        
        # A. Top 10 Table
        if (nrow(res$tables$significant_proteins) > 0) {
          tbl <- head(res$tables$significant_proteins, 10) %>% 
            mutate(logFC = round(logFC, 2), adj.P.Val = formatC(adj.P.Val, format = "e", digits = 2)) %>%
            kableExtra::kbl(format = "html", caption = "Top 10 Significant Proteins") %>% 
            kableExtra::kable_styling(bootstrap_options = "striped")
          print(tbl)
          cat("\n\n") 
        } else {
          cat("> **Note:** No significant proteins found matching the FDR and FC criteria.\n\n")
        }
        
        # B. Volcano Plot
        if (!is.null(res$plots$volcano)) {
          print(res$plots$volcano)
          cat("\n\n") 
        }
        
        # C. Heatmap (Direct Draw for Interactive Viewing)
        if (!is.null(res$plots$heatmap)) {
          ComplexHeatmap::draw(res$plots$heatmap, newpage = TRUE)
          cat("\n\n")
        }
        
        # D. Pathway Analysis (GSEA)
        if (length(res$plots$gsea) > 0) {
          cat("\n\n##### Pathway Analysis {.tabset}\n\n")
          
          for (db_name in names(res$plots$gsea)) {
            cat(sprintf("\n\n###### %s\n\n", db_name))
            
            p_obj <- res$plots$gsea[[db_name]]
            
            if (!is.null(p_obj$dotplot)) {
              print(p_obj$dotplot)
              cat("\n\n")
            }
            
            if (!is.null(p_obj$ridgeplot)) {
              print(p_obj$ridgeplot)
              cat("\n\n")
            }
            
            if (!is.null(p_obj$emap)) {
              print(p_obj$emap)
              cat("\n\n")
            }
          }
        } else {
          cat("> **Note:** No pathways passed criteria for visualization.\n\n")
        }
        
      } else {
        cat(sprintf("> **DEA Failed:** %s\n\n", res$error_msg))
      }
    }
  }
}
```


# Appendix
`r what_this_does("Appended the raw underlying R code from the custom predictive engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 02b Complete! \n")
```

```{r legacy}
  # job_gower_k2 = list(
  #   title = "Gower Distance Clustering (k=2)",
  #   input_matrix = "mat_gower", # Defined below in the execution chunk
  #   k = 2, 
  #   n_neighbors = 10,
  #   is_dist_matrix = TRUE, # <--- NEW FLAG FOR ENGINE
  #   algorithms = list(
  #     pam = list(run=TRUE), 
  #     hclust = list(run=TRUE, method="ward.D2")
  #   )
  # )

  # pheno_gower = list(
  #   target_job = "job_gower_k2",
  #   title = "Clinical Phenotypes: Gower Distance",
  #   target_algorithms = c("hclust", "pam"),
  #   subset_filter = list(column = "cohort_group", value = c("Active", "Non-Active")),
  #   variables = clin_vars_to_include,
  #   highlight_variables = clin_vars_to_highlight,
  #   p_cutoff = 0.05,
  #   must_include_vars = c("cohort_group", "Sex", "Age")
  # ),
  # 
  # feat_gower_k2 = list(
  #   target_job = "job_gower_k2",
  #   title = "Gower dist matrix k2",
  #   input_matrix = "raw", 
  #   covariates = c("Age"),
  #   # Algorithms are now the primary key to enable contrast-per-algorithm
  #   algorithms = list(
  #     hclust = list(
  #       contrasts = list(
  #         c1_vs_c2 = list(title = "Cluster 1 vs Cluster 2 Ssc only", desc = "", 
  #                         target=c("Cluster 1"), ref=c("Cluster 2"),
  #                         subset_filter = list(column = "cohort_group", value = c("Active", "Non-Active")))
  #       )
  #     )
  #   ),
  #   dea_p_cutoff = 0.05,
  #   dea_fc_cutoff = 0.58
  # ),
```

