---
title: "Proteomics Variance"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction

Goals:
-   To explore which variables have an influence on the omics data (by PCA, VPA and exploratory VPA)
-   To determine which variables correlate with each other
-   To determine which variables should be included in the downstream Differential Expression Analysis

## DEA Variable Inclusion Logic
| Scenario | Action | Explanation |
|:---|:---|:---|
| High VPA Variance & Low Correlation with Contrast | Include | Noise |
| High VPA VAriance & High Correlation with Contrast | Exclude | Confounder: Correcting for this could erase the signal |
| High VPA Variance & High correlation with Covariate | Pick the VPA winner | Redundancy/Collinearity |
| Low VPA VAriance | Exclude | Unnecessarily reduces degrees of freedom |

Further criteria:
-   Limit number of variables to abeout 1 variable per 10 patients.

# Setup and Import

`r what_this_does("Loaded analytical packages (FactoMineR, factoextra) and parallel processing tools.", "Sourced custom omics engines to handle repetitive matrix operations.", "Imported the twin clinical spines, proteomics matrix, and the Data Dictionary.", "Initialized parallel processing to accelerate computationally intensive variance partition calculations.")`

```{r setup_import, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(patchwork)
library(here)
library(future)
library(future.apply)
library(vsn) 
library(hexbin)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "03a_proteomics_variance"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "01_general_engines.R"))
source(here::here("code", "02_proteomics_engines.R"))

# Load Pre-Cleaned Data from Scripts 01a, 01b, and 02
master_spine          <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
numeric_spine         <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine_numeric.rds"))
data_dictionary       <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))

olink_mat             <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))
clinical_correlations <- readRDS(here::here("output", "02a_cohort_characterization", "clinical_correlation_matrices.rds"))

# Initialize multi-core use
plan(multisession, workers = 4)
```

# Configuration

`r what_this_does("Dynamically engaged the Data Dictionary shield to automatically block zero-variance, heavily skewed, outlier-heavy, or highly imbalanced variables from crashing the models.", "Defined mapping parameters for Principal Component Analysis (PCA).", "Configured Exploratory VPA rules and automated DEA selection thresholds.", "Established Targeted VPA models to evaluate explanatory power.")`

```{r config}
# 0. Data Dictionary Shield ----------------------------------------------------
unstable_vars <- data_dictionary %>%
  filter(grepl("ZERO VARIANCE|IMBALANCED|SKEWED|OUTLIER", toupper(QC_Flag))) %>%
  pull(Variable)

base_excludes <- c(
  "inCentraxx", "SELECTED", "Subject_ID", "Birth_Year", "Raynaud VAS Unknown", 
  "Face", "Chest", "Abdomen", "Left_Upper Arms", "Right_Upper Arms", 
  "Left_Forearms", "Right_Forearms", "Left_Hands", "Right_Hands", 
  "Left_Fingers", "Right_Fingers", "Left_Thighs", "Right_Thighs", 
  "Left_Legs", "Right_Legs", "Left_Feet", "Right_Feet", "Active_our",
  "Geschlecht", "Entnahmedatum", "Entnahmedatum_Clean", "Geburtsdatum", "Distance in m"
)

# 1. Exploratory  VPA ----------------------------------------------------------
exploratory_vpa_configs <- list(
  discovery_scan_patients = list(
    title              = "Clinical Discovery: Patients Only",
    subset_to_groups   = c(GRP_ACTIVE, GRP_NON_ACTIVE),
    exclude_vars       = unique(c(base_excludes, unstable_vars)), 
    exclude_patterns   = c("Date", "\\?", "Datum"), 
    include_vars       = c(), 
    missingness_cutoff = 0.10,
    top_n_plot         = 60,
    dist_plot_limit    = 30, 
    corr_plot_limit    = 30, 
    heatmap_add_vars   = c("cohort_group", "Sex", "Age", "Inflamm_active", "fibrotic_active", "vascular_active") 
  )
)

# 2. Automated DEA Variable Selector Configurations ----------------------------
dea_selector_configs <- list(
  active_vs_nonactive_1 = list(
    title = "Active vs Non-Active",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),    # Pure Text
    ref_groups = c(GRP_NON_ACTIVE),   # Pure Text
    vpa_source_id = "discovery_scan_patients", 
    must_include = c(),          
    must_exclude = c("log_Creatinine", "Uric ACid (mg/dl)", "mRSS_a", "Modified rodnan skin score", "eustarAI", "SSc Antibodies", "Delta_ECHO_Years", "Delta_ECG_Years"), 
    min_samples_per_param = 15,
    max_cor_rho           = 0.25, 
    max_cor_p             = 0.05,
    pre_selection_log     = "Some variables excluded due to non-normal distribution (mostly bimodal). eustar AI excluded due to high correlation with contrast. SSC Antibodies excluded due to skewed distribution."
  )
)

# 3. Manual Model Adjustment and Creations  ------------------------------------
manual_model_adjustments <- list(
  active_vs_nonactive_1 = list(
    type = "adjustment", 
    drop = c("Right ventricular area (cm2)","LVEF %"), 
    add  = c(), 
    log  = "Removed cardiac parameters since they did not explain a large enough portion of variance in the multivariate VPA. Could not include Sex or Age due to correlation with log_Creatinine."
  ),
  legacy_model = list(
    type = "custom",     
    title = "Legacy Model",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),    
    ref_groups = c(GRP_NON_ACTIVE),   
    covariates = c("Age", "Sex"),
    log  = "Legacy model, created to compare the new models to."
  ),
  inflamm_vs_non_inflamm = list(
    type = "custom",      
    title = "Inflammatory Active vs Non-Inflammatory",
    group_col = "Inflamm_active",
    target_groups = c("yes"), # Standardized to Pure Text
    ref_groups = c("No"),     # Standardized to Pure Text
    covariates = c("log_Creatinine", "log_NT_proBNP__pg_ml"),
    log  = "Custom model contrasting inflammatory active (yes) vs non-inflammatory (No) patients."
  ),
  vascular_vs_non_vascular = list(
    type = "custom",      
    title = "Vascular Active vs Non-Vascular",
    group_col = "vascular_active",
    target_groups = c("yes"), # Standardized to Pure Text
    ref_groups = c("No"),     # Standardized to Pure Text
    covariates = c("log_Creatinine", "log_NT_proBNP__pg_ml"),
    log  = "Custom model contrasting vascular active (yes) vs non-vascular (No) patients."
  ),
  fibrotic_vs_non_fibrotic = list(
    type = "custom",      
    title = "Fibrotic Active vs Non-Fibrotic",
    group_col = "fibrotic_active",
    target_groups = c("yes"), # Standardized to Pure Text
    ref_groups = c("No"),     # Standardized to Pure Text
    covariates = c("log_Creatinine", "log_NT_proBNP__pg_ml"),
    log  = "Custom model contrasting fibrotic active (yes) vs non-fibrotic (No) patients."
  )
)

# 4. PCA Configs----------------------------------------------------------------
pca_configs <- list(
  pca_global = list(
    title = "PCA: Global Cohort Structure",
    subset_to_contrast = NULL, 
    color_vars = c("cohort_group", "Sex", "Age", "Sample_Age") 
  ),
  pca_candidate_vars = list(
    title = "PCA: Candidate Variables",
    subset_to_contrast = list(
      group_col = "cohort_group",
      target_groups = c(GRP_ACTIVE, GRP_NON_ACTIVE)
    ), 
    color_vars = c("cohort_group", "Sex", "Age", "log_NT_proBNP__pg_ml", "log_Creatinine") 
  )
)

# 5. Final Validation ----------------------------------------------------------
dea_validation_rules <- list(
  min_samples_per_param = 10, 
  min_group_n           = 2,
  max_cor_rho           = 0.25, 
  max_cor_p             = 0.05  
)

# 6. Global Omics Filtering ----------------------------------------------------
global_var_filter <- list(
  title = "Protein Variance Distribution",
  omics_type = "Protein",
  top_n_features = 200
)
```

# Exploratory Variance Partition Analysis

`r analysis_decision("Restricted the exploratory scan to patient cohorts only (excluding healthy controls) because clinical data is only available for patients not for healthy controls", "Removed variables with a >10% missingness ratio to prevent modeling artifacts caused by highly sparse clinical data.", "Removed variables with zero or near-zero variance (e.g., factors with only one level remaining after NA-drops) to ensure mathematical convergence of the mixed-effects models.", "Removed administrative IDs and raw date columns, preventing spurious correlations with procedural metadata.", "Removed components of the mRSS (e.g. right hand) and other duplicate variables to prevent redundancy")`

`r what_this_does("Ran univariate Variance Partition Analysis across all eligible clinical variables to independently rank their impact on the proteomics matrix.", "Automatically filtered out sparse variables based on the missingness cutoff and generated an exclusion log.", "Rendered summary tables and bar charts highlighting the top clinical drivers of proteomic variance.")`

```{r exploratory_vpa_execution, results='asis', fig.height=15}
for (vpa_id in names(exploratory_vpa_configs)) {
  conf <- exploratory_vpa_configs[[vpa_id]]
  cat(sprintf("#### Scan: %s\n\n", conf$title))
  
  res <- wrap_vpa_univariate_pipeline(
    omics_mat       = olink_mat, 
    clin_df         = master_spine, 
    vpa_conf        = conf, 
    vpa_id          = vpa_id, 
    base_output_dir = output_dir_data
  )
  
  if(!is.null(res)) {
    
    # INTELLIGENT RENDER ENGINE (Interactive vs Knitting)
    if (interactive()) {
      print(res$tables$exclusion)
      print(res$univariate_bar)
      print(res$tables$summary)
    } else {
      cat(as.character(res$tables$exclusion))
      cat("\n\n<br>\n\n") 
      print(res$univariate_bar)
      cat("\n\n<br>\n\n")
      cat(as.character(res$tables$summary))
      cat("\n\n<br>\n\n")
    }
  }
  cat("\n\n---\n\n")
}

# Clean memory
rm(vpa_id, conf, res)
invisible(gc())
```

# Top Hit Quality Control

`r what_this_does("Generates a dynamic visual quality control checkpoint for the top Univariate VPA hits.", "Plots a multi-panel grid of distributions to identify extreme outliers or skew.", "Renders a clustered correlation heatmap to instantly reveal blocks of collinear variables before automated selection begins.")`

```{r vpa_qc_checkpoint, results='asis', fig.width=14, fig.height=10}
# Initialize output buffer to prevent code chunk fragmentation
for (vpa_id in names(exploratory_vpa_configs)) {
  conf <- exploratory_vpa_configs[[vpa_id]]
  html_out <- paste0(sprintf("#### QC: %s\n\n", conf$title))
  
  vpa_csv_path <- file.path(output_dir_data, "VPA_Univariate", vpa_id, paste0(vpa_id, "_Univariate_Results.csv"))
  if (!file.exists(vpa_csv_path)) {
    html_out <- paste0(html_out, "> **SKIPPED:** VPA Results not found.\n\n")
    next
  }
  
  vpa_df <- read_csv(vpa_csv_path, show_col_types = FALSE)
  
  # Call the DRY Engine Wrapper
  qc_res <- wrap_vpa_qc_pipeline(vpa_df, master_spine, clinical_correlations, conf, vpa_id, output_dir_data)
  
  # Append any missing variable warnings to the buffer
  if (length(qc_res$html_warnings) > 0) {
    html_out <- paste0(html_out, paste(qc_res$html_warnings, collapse = "\n\n"), "\n\n")
  }
  
  # Dump HTML string so far, then print plots sequentially
  cat(html_out)
  html_out <- "" # Reset buffer
  
  if (!is.null(qc_res$plots[["distribution_grid"]])) {
    print(qc_res$plots[["distribution_grid"]])
    cat("\n\n<br><hr>\n\n")
  }
  
  if (!is.null(qc_res$plots[["correlation_heatmap"]])) {
    print(qc_res$plots[["correlation_heatmap"]])
    cat("\n\n<br>\n\n")
  }
}

# Clean memory
rm(html_out, vpa_id, conf, vpa_csv_path, vpa_df, qc_res)
invisible(gc())
```


# Variance Filtering

`r what_this_does("Calculated the variance of all measured features across the entire patient cohort.", "Generated a density histogram to visualize the global distribution of variance.", "Extracted and saved the highest-variance features to an RDS file to strip background noise for downstream unsupervised discovery scripts.")`

`r why_is_this_done("Features that do not change across the cohort (low variance) are biological background noise. Removing them here, at the end of the script, prevents dilution of true signals in downstream algorithms like UMAP or hierarchical clustering, while ensuring we do not prematurely drop them from earlier targeted VPA or DEA steps.")`

```{r global_variance_filter, warning=FALSE}
var_res <- wrap_global_variance_filter(olink_mat, global_var_filter, output_dir_data)
print(var_res$plot)

# Clean memory
rm(var_res)
invisible(gc())
```

#Scedasticity Plot
`r what_this_does("Generates collapsible visual evidence (Before-State Multi-VPA and PCA) for any variables that were manually rejected from the automated draft models.")`

```{r scedasticity_plot, fig.width=9, fig.height=8}
# Plot A: Ranked mean
msd_npx_ranked <- meanSdPlot(as.matrix(olink_mat), plot = FALSE)
p_msd_npx_ranked <- msd_npx_ranked$gg + 
  ggtitle("NPX Values Ranked") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
p_msd_npx_ranked

# Plot B: Mean
msd_npx_scale <- meanSdPlot(as.matrix(olink_mat), plot = FALSE, ranks = FALSE)
p_msd_npx_scale <- msd_npx_scale$gg + 
  ggtitle("NPX Values original scale") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
p_msd_npx_scale
```


# Appendix

`r what_this_does("Generates collapsible visual evidence (Before-State Multi-VPA and PCA) for any variables that were manually rejected from the automated draft models.")`

```{r diagnostic_graveyard, results='asis'}
for (id in names(final_locked_models)) {
  m <- final_locked_models[[id]]
  
  if (length(m$dropped_vars) > 0) {
    cat(sprintf("### Rejected Diagnostics: %s\n\n", m$title))
    
    # Render Dropdown wrapper
    cat(sprintf("<details><summary style='cursor: pointer; font-weight: bold; color: #2c3e50; font-size: 1.2em;'>▶ Click to view 'Before' State Visuals (Dropped: %s)</summary>\n<div style='padding: 15px; border-left: 3px solid #c0392b; margin-top: 10px; background-color: #fcfcfc;'>\n\n", paste(m$dropped_vars, collapse=", ")))
    
    cat("> **Note:** The plots below show the *Automated Draft* model before manual exclusion. They demonstrate how the dropped variables behaved in the context of the full draft model.\n\n")
    
    # 1. Run 'Before' VPA
    cat("#### Multi-VPA (Draft State)\n\n")
    draft_vpa_conf <- list(title = sprintf("Draft VPA: %s", m$title), variables = c(m$group_col, m$auto_vars))
    vpa_res <- wrap_vpa_pipeline(olink_mat, master_spine, draft_vpa_conf, paste0("Draft_", id), output_dir_data)
    if(!is.null(vpa_res) && vpa_res$status == "success") { print(vpa_res$plots$bar); cat("\n\n") }
    
    # 2. Run 'Before' PCA
    cat("#### PCA (Draft State)\n\n")
    draft_pca_conf <- list(title = sprintf("Draft PCA: %s", m$title), subset_to_contrast = list(group_col = m$group_col, target_groups = c(m$target_groups, m$ref_groups)), color_vars = c(m$group_col, m$auto_vars))
    pca_res <- wrap_pca_pipeline(olink_mat, master_spine, draft_pca_conf, paste0("Draft_", id), output_dir_data, get_project_colors)
    if (!interactive()) {
      cat(sprintf("![](%s)\n\n![](%s)\n\n![](%s)\n\n", pca_res$paths[["scree"]], pca_res$paths[["loadings"]], pca_res$paths[["merged_drivers"]]))
      if (!is.null(pca_res$paths[["merged_overlays"]])) cat(sprintf("![](%s)\n\n", pca_res$paths[["merged_overlays"]]))
    } else {
      print(pca_res$plots[["scree"]]); print(pca_res$plots[["loadings"]]); print(pca_res$plots[["merged_drivers"]])
      for (c_var in draft_pca_conf$color_vars) { if (!is.null(pca_res$plots[[paste0("PCA_Plot_", c_var)]])) print(pca_res$plots[[paste0("PCA_Plot_", c_var)]]) }
    }
    
    # Close Dropdown
    cat("\n</div>\n</details>\n\n<hr>\n\n")
  }
}
```

`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=c(readLines(here::here("code", "01_general_engines.R")), readLines(here::here("code", "02_proteomics_engines.R"))), eval=FALSE}

```

```{r complete}
cat("Script 03a Complete! \n")
```
