---
title: "03a_proteomics_variance"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction

-   To explore which variables have an influence on the omics data (by PCA, VPA and exploratory VPA)
-   To determine which variables correlate with each other
-   To determine which variables should be included in the downstream Differential Expression Analysis

## DEA VAriable Inclusion Logic
| Scenario | Action | Explanation |
|:---|:---|:---|
| High VPA Variance & Low Correlation with Contrast | Include | Noise |
| High VPA VAriance & High Correlation with Contrast | Compare Include/Exclude | Confounder: Correcting for this could erase the signal |
| High VPA Variance & High correlation with Covariate | Pick the VPA winner | Redundancy/Collinearity |
| Low VPA VAriance | Exclude | Unnecessarily reduces degrees of freedom |

Limit umber of variables to abeout 1 variable per 10 patients.


# Setup and Import

`r what_this_does("Loaded analytical packages (FactoMineR, factoextra) and parallel processing tools (future, progressr).", "Sourced custom omics engines to handle repetitive matrix operations.", "Imported the pre-cleaned clinical spine and Olink proteomics matrix.", "Initialized parallel processing to accelerate computationally intensive variance partition calculations.")`

```{r setup_import, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(patchwork)
library(here)
library(future)
library(future.apply)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "03a_proteomics_variance"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data from Script 01
master_spine <- readRDS(here::here("output", "01_data_import_and_qc", "clinical_spine.rds"))
olink_mat <- readRDS(here::here("output", "01_data_import_and_qc", "clean_olink_matrix.rds"))
clinical_correlations <- readRDS(here::here("output", "02_cohort_characterization", "clinical_correlation_matrices.rds"))

# Initialize multi-core use
plan(multisession, workers = 4)
```

# Configuration

`r what_this_does("Defined mapping parameters for Principal Component Analysis (PCA) to overlay specific clinical metadata on global cohorts.", "Configured Exploratory VPA rules, setting exclusion patterns to prevent IDs and raw dates from skewing the variance scan.", "Established Targeted VPA models to evaluate the combined explanatory power of standard covariates versus newly discovered clinical hits.")`

```{r config}
# 1. Exploratory  VPA ----------------------------------------------------------
exploratory_vpa_configs <- list(
  
  discovery_scan_patients = list(
    title              = "Clinical Discovery: Patients Only",
    subset_to_groups   = c(GRP_ACTIVE, GRP_NON_ACTIVE),
    
    # Filtering Rules
    exclude_vars       = c("inCentraxx", "SELECTED", "Subject-ID (EUSTAR)", "Geburtsdatum", "Raynaud VAS Unknown", "Birth_Year"),
    exclude_patterns   = c("Date", "?", "Datum"), 
    include_vars       = c(), 
    missingness_cutoff = 0.10,
    
    # Execution & Visualization Rules
    top_n_plot         = 50 
  )
)

# 2. Candidates for correcting variables
# Variables are selected based on the Exploratory VPA in this script
# Selected variables will be assessed using a PCA and a Multivariate VPA

candidate_vars <- list(
  set1 = c("Creatinine", "NT-proBNP (pg/ml)", "Age", "DLCO (SB) (% predicted)")
)

# 3. PCA Master Layouts (Independent grids) ------------------------------------
pca_configs <- list(
  pca_global = list(
    title = "PCA: Global Cohort Structure",
    subset_to_contrast = NULL, 
    color_vars = c("cohort_group", "Sex", "Age", "Sample_Age") 
  ),
  
  pca_DEA_candidate_vars = list(
    title = "PCA: Proposed DEA Variables (Active/Non-Active Only)",
    subset_to_contrast = list(
      group_col = "cohort_group",
      target_groups = c(GRP_ACTIVE, GRP_NON_ACTIVE)
    ), 
    color_vars = c("cohort_group", "Inflamm_active", "fibrotic_active", "vascular_active", candidate_vars$set1) 
  )
  
)

# 4. Targeted VPA --------------------------------------------------------------

targeted_vpa_configs <- list(
  
  active_nonactive = list(
    title     = "Potential DEA Model for: Active vs. Non-Active",
    variables = c("cohort_group", candidate_vars$set1)  
  ),
  
  inflammation = list(
    title     = "Potential DEA Model for: Inflammed vs. Non-Inflammed SSc Patients",
    variables = c("Inflamm_active", candidate_vars$set1)  
  ),
  
  fibrotic = list(
    title     = "Potential DEA Model for: Fibrotic vs. Non-Fibrotic SSc Patients",
    variables = c("fibrotic_active", candidate_vars$set1)  
  ),
  
   vascular = list(
    title     = "Potential DEA Model for: Vascular vs. Non-Vascular SSc Patients",
    variables = c("vascular_active", candidate_vars$set1)  
  ),
   SSC_vs_Ctrl = list(
    title     = "Potential DEA Model for: SSc vs. Control",
    variables = c("cohort_group", "Age", "Sex")  
  )
)

# 5. Pre-DEA Validation Configurations -----------------------------------------
dea_validation_rules <- list(
  min_samples_per_param = 10, 
  min_group_n           = 2,
  max_cor_rho           = 0.70, # Threshold for Variance Inflation
  max_cor_p             = 0.05
)

pre_dea_configs <- list(
  active_vs_nonactive = list(
    title         = "Active vs Non-Active",
    group_col     = "cohort_group",
    target_groups = c(GRP_ACTIVE),
    ref_groups    = c(GRP_NON_ACTIVE),
    covariates    = candidate_vars$set1
  ),
  inflammed = list(
    title         = "Inflammed vs. Non-Inflammed",
    group_col     = "Inflamm_active",
    target_groups = c(1),
    ref_groups    = c(0),
    covariates    = candidate_vars$set1
  ),
  fibrotic = list(
    title         = "Fibrotic vs. Non-Fibrotic",
    group_col     = "fibrotic_active",
    target_groups = c(1),
    ref_groups    = c(0),
    covariates    = candidate_vars$set1
  ),
  vascular = list(
    title         = "Vascular vs. Non-Vascular",
    group_col     = "vascular_active",
    target_groups = c(1),
    ref_groups    = c(0),
    covariates    = candidate_vars$set1
  ),
  ssc_vs_ctrl = list(
    title         = "SSc vs Healthy Control",
    group_col     = "cohort_group",
    target_groups = c(GRP_ACTIVE, GRP_NON_ACTIVE),
    ref_groups    = c(GRP_CONTROL),
    covariates    = c("Age", "Sex") 
  )
)
```

# Exploratory Variance Partition Analysis

`r analysis_decision("Restricted the exploratory scan to patient cohorts only (excluding healthy controls) because clinical data is only available for patients not for healthy controls", "Enforced a strict missingness cutoff (max 10% missing data per variable) to prevent modeling artifacts caused by highly sparse clinical data.", "Automatically excluded variables with zero or near-zero variance (e.g., factors with only one level remaining after NA-drops) to ensure mathematical convergence of the mixed-effects models.", "Applied blacklist filtering and regex pattern matching to remove administrative IDs and raw date columns, preventing spurious correlations with procedural metadata.")`

`r what_this_does("Ran univariate Variance Partition Analysis across all eligible clinical variables to independently rank their impact on the proteomics matrix.", "Automatically filtered out sparse variables based on the missingness cutoff and generated an exclusion log.", "Rendered summary tables and bar charts highlighting the top clinical drivers of proteomic variance.")`

```{r exploratory_vpa_execution, results='asis', fig.height=12}
cat("### Executing Exploratory Variance Partition Analysis...\n\n")

for (vpa_id in names(exploratory_vpa_configs)) {
  conf <- exploratory_vpa_configs[[vpa_id]]
  cat(sprintf("#### Scan: %s\n\n", conf$title))
  
  res <- wrap_vpa_univariate_pipeline(
    omics_mat       = olink_mat, 
    clin_df         = master_spine, 
    vpa_conf        = conf, 
    vpa_id          = vpa_id, 
    base_output_dir = output_dir_data
  )
  
  if(!is.null(res)) {
    
    # INTELLIGENT RENDER ENGINE (Interactive vs Knitting)
    if (interactive()) {
      print(res$tables$exclusion)
      print(res$univariate_bar)
      print(res$tables$summary)
    } else {
      cat(as.character(res$tables$exclusion))
      cat("\n\n<br>\n\n") 
      print(res$univariate_bar)
      cat("\n\n<br>\n\n")
      cat(as.character(res$tables$summary))
      cat("\n\n<br>\n\n")
    }
  }
  cat("\n\n---\n\n")
}

# Clean memory
rm(vpa_id, conf, res)
invisible(gc())
```

# PCA

`r what_this_does("Executed the PCA pipeline to assess global cohort structure and identify potential outliers or batch effects.", "Generated scree plots to evaluate dimensional variance capture.", "Rendered and exported feature loadings and metadata-colored PCA overlays (e.g., by Cohort, Age, Sex).")`

```{r execution_engine_PCA, results='asis'}
cat("### Executing Principal Component Analysis...\n\n")

# Process each PCA configuration through the Master Pipeline
for (pca_name in names(pca_configs)) {
  
  pca_conf <- pca_configs[[pca_name]]
  
  # CALL THE MASTER PIPELINE
  pca_results <- wrap_pca_pipeline(
    omics_mat           = olink_mat, 
    clin_df             = master_spine, 
    pca_conf            = pca_conf, 
    pca_id              = pca_name, 
    base_output_dir     = output_dir_data, 
    project_colors_func = get_project_colors
  )
  
  # PRINT TO HTML REPORT
  if (!interactive()) {
    cat(sprintf("#### %s\n\n", pca_conf$title))
    
    cat(sprintf("![](%s)\n\n", pca_results$paths[["scree"]]))
    cat(sprintf("![](%s)\n\n", pca_results$paths[["loadings"]]))
    cat(sprintf("![](%s)\n\n", pca_results$paths[["merged_drivers"]]))

    if (!is.null(pca_results$paths[["merged_overlays"]])) {
      cat("#### Metadata Overlays (Side-by-Side Comparison Grid)\n\n")
      
      subtitle_desc <- case_when(
        is.null(pca_conf$subset_to_contrast) ~ "Showing: Global Cohort",
        TRUE ~ pca_conf$title 
      )
      
      cat(sprintf("> **Grid View for:** %s\n\n", subtitle_desc))
      
      # Insert the already created master PNG
      cat(sprintf("![](%s)\n\n", pca_results$paths[["merged_overlays"]]))
    }
  } else {
    # INTERACTIVE MODE: Print individual plots sequentially so they don't squash!
    cat(sprintf("#### Interactive Viewer: %s\n\n", pca_conf$title))
    print(pca_results$plots[["scree"]])
    print(pca_results$plots[["loadings"]])
    print(pca_results$plots[["merged_drivers"]])
    
    # Loop through and print each overlay plot on its own
    for (c_var in pca_conf$color_vars) {
      plot_key <- paste0("PCA_Plot_", c_var)
      if (!is.null(pca_results$plots[[plot_key]])) {
        print(pca_results$plots[[plot_key]])
      }
    }
  }
  
  cat("---\n\n")
}

# Clean memory
rm(pca_name, pca_conf, pca_results, subtitle_desc, plot_key, c_var)
invisible(gc())
```

# Targeted Variance Partition Analysis

`r what_this_does("Executed multivariate Variance Partition Analysis to model the combined influence of specific clinical variable sets.", "Calculated the unique percentage of proteomics variance explained by each variable while adjusting for the others.", "Outputted bar and violin plots to visualize the distribution of explained variance across all measured proteins.")`

```{r targeted_vpa_engine, results='asis', fig.height=10, fig.width=8}
for (vpa_id in names(targeted_vpa_configs)) {
  conf <- targeted_vpa_configs[[vpa_id]]
  
  cat(sprintf("#### Analysis: %s\n\n", conf$title))
  
  res <- wrap_vpa_pipeline(
    omics_mat       = olink_mat,  
    clin_df         = master_spine, 
    vpa_conf        = conf, 
    vpa_id          = vpa_id, 
    base_output_dir = output_dir_data
  )
  
  if(!is.null(res) && res$status == "success") {
    print(res$plots$bar)
    cat("\n\n")
    print(res$plots$violin)
  } else if (!is.null(res) && res$status == "failed") {
    # Prints the exact reason the model was skipped
    cat(sprintf("> <span style='color:red;'>**SKIPPED:** %s</span>\n", res$error_msg))
  }
}

# Clean memory
rm(vpa_id, conf, res)
invisible(gc())
```

# Pre-DEA Design Validation

`r analysis_decision("Enforced the clinical 'Rule of 10' (minimum 10 samples per parameter) for all DEA designs to mathematically prevent overfitting.", "Mandated a strict minimum of 2 surviving patients per contrast group to ensure valid standard error estimation in downstream limma models.","Implemented automated exclusion for variables that collapse to zero variance or a single level after complete-case filtration.","Evaluated linear independence via QR matrix decomposition to actively block collinear variables from entering the limma pipeline.", "Cross-referenced all covariates against the cohort Spearman correlation matrix to block Variance Inflation (rho >= 0.70, p <= 0.05).")`

`r what_this_does("Tested proposed Differential Expression Analysis configurations against 5 mathematical pillars: Complete Case Survival, Variance Stability, Matrix Collinearity, Variance Inflation, and Statistical Power.", "Generated an integrated, variable-by-variable mathematical audit table for each proposed model to guide downstream DEA script configurations.")`
```{r pre_dea_validation, results='asis'}
for (model_id in names(pre_dea_configs)) {
  
  # Call the new wrapper engine
  res <- wrap_pre_dea_validation(
    clin_df      = master_spine,
    conf         = pre_dea_configs[[model_id]],
    rules        = dea_validation_rules,
    cor_matrices = clinical_correlations
  )
  
  # Intelligent Render Engine
  if (interactive()) {
    print(res$html_table)
  } else {
    cat(as.character(res$html_table))
  }
  
  cat("\n\n") 
}

# Clean memory
rm(model_id, res)
invisible(gc())
```

# Appendix

`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 03a Complete! \n")
```
