---
title: "Proteomics Variance"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction

Goals:
-   To explore which variables have an influence on the omics data (by PCA, VPA and exploratory VPA)
-   To determine which variables correlate with each other
-   To determine which variables should be included in the downstream Differential Expression Analysis

## DEA Variable Inclusion Logic
| Scenario | Action | Explanation |
|:---|:---|:---|
| High VPA Variance & Low Correlation with Contrast | Include | Noise |
| High VPA VAriance & High Correlation with Contrast | Exclude | Confounder: Correcting for this could erase the signal |
| High VPA Variance & High correlation with Covariate | Pick the VPA winner | Redundancy/Collinearity |
| Low VPA VAriance | Exclude | Unnecessarily reduces degrees of freedom |

Further criteria:
-   Limit number of variables to abeout 1 variable per 10 patients.

# Setup and Import

`r what_this_does("Loaded analytical packages (FactoMineR, factoextra) and parallel processing tools (future, progressr).", "Sourced custom omics engines to handle repetitive matrix operations.", "Imported the pre-cleaned clinical spine and Olink proteomics matrix.", "Initialized parallel processing to accelerate computationally intensive variance partition calculations.")`

```{r setup_import, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(patchwork)
library(here)
library(future)
library(future.apply)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "03a_proteomics_variance"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data from Script 01
master_spine <- readRDS(here::here("output", "01_data_import_and_qc", "clinical_spine.rds"))
olink_mat <- readRDS(here::here("output", "01_data_import_and_qc", "clean_olink_matrix.rds"))
clinical_correlations <- readRDS(here::here("output", "02_cohort_characterization", "clinical_correlation_matrices.rds"))

# Initialize multi-core use
plan(multisession, workers = 4)
```

# Configuration

`r what_this_does("Defined mapping parameters for Principal Component Analysis (PCA) to overlay specific clinical metadata on global cohorts.", "Configured Exploratory VPA rules, setting exclusion patterns to prevent IDs and raw dates from skewing the variance scan.", "Established Targeted VPA models to evaluate the combined explanatory power of standard covariates versus newly discovered clinical hits.")`

```{r config}
# 1. Exploratory  VPA ----------------------------------------------------------
exploratory_vpa_configs <- list(
  discovery_scan_patients = list(
    title              = "Clinical Discovery: Patients Only",
    subset_to_groups   = c(GRP_ACTIVE, GRP_NON_ACTIVE),
    exclude_vars       = c("inCentraxx", "SELECTED", "Subject-ID (EUSTAR)", "Geburtsdatum", "Raynaud VAS Unknown", "Birth_Year", "Face", "Chest", "Abdomen", "Left_Upper Arms", "Right_Upper Arms", "Left_Forearms", "Right_Forearms", "Left_Hands", "Right_Hands", "Left_Fingers", "Right_Fingers", "Left_Thighs", "Right_Thighs", "Left_Legs", "Right_Legs", "Left_Feet", "Right_Feet", "Geschlecht", "Active_our"),
    exclude_patterns   = c("Date", "?", "Datum"), 
    include_vars       = c(), 
    missingness_cutoff = 0.10,
    top_n_plot         = 60,
    
    # QC Checkpoint Parameters -------------------------------------------------
    dist_plot_limit    = 30, # Max variables to draw in the histogram grid
    corr_plot_limit    = 30, # Max VPA hits to feed into the heatmap
    heatmap_add_vars   = c("cohort_group", "Sex", "Age", "Inflamm_active", "fibrotic_active", "vascular_active") # Extra vars to force into heatmap
  )
)

# 2. Automated DEA Variable Selector Configurations ----------------------------
dea_selector_configs <- list(
  active_vs_nonactive_1 = list(
    title = "Active vs Non-Active",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),
    ref_groups = c(GRP_NON_ACTIVE),
    vpa_source_id = "discovery_scan_patients", 
    must_include = c(),          
    must_exclude = c("Creatinine", "NT-proBNP (pg/ml)", "Distance in m", "log_Distance_in_m", "Uric ACid (mg/dl)", "mRSS_a", "Modified rodnan skin score", "eustarAI", "SSc Antibodies", "Delta_ECHO_Years", "Deöta_ECG_Years"), # Excluded based on QC
    min_samples_per_param = 15,
    max_cor_rho           = 0.25, 
    max_cor_p             = 0.05,
    # --- NEW: Pre-Selection Log ---
    pre_selection_log     = "Some variables excluded due to non-normal distribution (mostly bimodal). eustar AI excluded due to high correlation with contrast. SSC Antibodies excluded due to skewed distribution (11% vs 89%)."
  )
)

# 3. Manual Model Adjustment and Creations  ------------------------------------
manual_model_adjustments <- list(
  
  active_vs_nonactive_1 = list(
    type = "adjustment", 
    drop = c("Right ventricular area (cm2)","LVEF %"), 
    add  = c(), 
    log  = "Removed cardiac parameters since theyx did not explain a large enough portion of variance in the multivariate VPA. Could not include Sex or Age due to correlation with log_creatinine"
  ),
  
  legacy_model = list(
    type = "custom",     
    title = "Custom: Legacy Model",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),
    ref_groups = c(GRP_NON_ACTIVE),
    covariates = c("Age", "Sex"),
    log  = "Legacy model, created to compare the new models to."
  )
  
)

# 4. PCA Configs----------------------------------------------------------------
pca_configs <- list(
  pca_global = list(
    title = "PCA: Global Cohort Structure",
    subset_to_contrast = NULL, 
    color_vars = c("cohort_group", "Sex", "Age", "Sample_Age") 
  ),
  
  pca_candidate_vars = list(
    title = "PCA: Candidate Variables",
    subset_to_contrast = list(
      group_col = "cohort_group",
      target_groups = c(GRP_ACTIVE, GRP_NON_ACTIVE)
    ), 
    color_vars = c("cohort_group", "Sex", "Age", "log_NT_ProBNP", "log_creatinine") 
  )
)

# 5. Final Validation ----------------------------------------------------------
dea_validation_rules <- list(
  min_samples_per_param = 10, 
  min_group_n           = 2,
  max_cor_rho           = 0.25, 
  max_cor_p             = 0.05  
)
```

# Exploratory Variance Partition Analysis

`r analysis_decision("Restricted the exploratory scan to patient cohorts only (excluding healthy controls) because clinical data is only available for patients not for healthy controls", "Removed variables with a >10% missingness ratio to prevent modeling artifacts caused by highly sparse clinical data.", "Removed variables with zero or near-zero variance (e.g., factors with only one level remaining after NA-drops) to ensure mathematical convergence of the mixed-effects models.", "Removed administrative IDs and raw date columns, preventing spurious correlations with procedural metadata.", "Removed components of the mRSS (e.g. right hand) and other duplicate variables to prevent redundancy")`

`r what_this_does("Ran univariate Variance Partition Analysis across all eligible clinical variables to independently rank their impact on the proteomics matrix.", "Automatically filtered out sparse variables based on the missingness cutoff and generated an exclusion log.", "Rendered summary tables and bar charts highlighting the top clinical drivers of proteomic variance.")`

```{r exploratory_vpa_execution, results='asis', fig.height=15}
for (vpa_id in names(exploratory_vpa_configs)) {
  conf <- exploratory_vpa_configs[[vpa_id]]
  cat(sprintf("#### Scan: %s\n\n", conf$title))
  
  res <- wrap_vpa_univariate_pipeline(
    omics_mat       = olink_mat, 
    clin_df         = master_spine, 
    vpa_conf        = conf, 
    vpa_id          = vpa_id, 
    base_output_dir = output_dir_data
  )
  
  if(!is.null(res)) {
    
    # INTELLIGENT RENDER ENGINE (Interactive vs Knitting)
    if (interactive()) {
      print(res$tables$exclusion)
      print(res$univariate_bar)
      print(res$tables$summary)
    } else {
      cat(as.character(res$tables$exclusion))
      cat("\n\n<br>\n\n") 
      print(res$univariate_bar)
      cat("\n\n<br>\n\n")
      cat(as.character(res$tables$summary))
      cat("\n\n<br>\n\n")
    }
  }
  cat("\n\n---\n\n")
}

# Clean memory
rm(vpa_id, conf, res)
invisible(gc())
```

# Top Hit Quality Control

`r what_this_does("Generates a dynamic visual quality control checkpoint for the top Univariate VPA hits.", "Plots a multi-panel grid of distributions to identify extreme outliers or skew.", "Renders a clustered correlation heatmap to instantly reveal blocks of collinear variables before automated selection begins.")`

```{r vpa_qc_checkpoint, results='asis', fig.width=14, fig.height=10}
# Initialize output buffer to prevent code chunk fragmentation
for (vpa_id in names(exploratory_vpa_configs)) {
  conf <- exploratory_vpa_configs[[vpa_id]]
  html_out <- paste0(sprintf("#### QC: %s\n\n", conf$title))
  
  vpa_csv_path <- file.path(output_dir_data, "VPA_Univariate", vpa_id, paste0(vpa_id, "_Univariate_Results.csv"))
  if (!file.exists(vpa_csv_path)) {
    html_out <- paste0(html_out, "> **SKIPPED:** VPA Results not found.\n\n")
    next
  }
  
  vpa_df <- read_csv(vpa_csv_path, show_col_types = FALSE)
  
  # Call the DRY Engine Wrapper
  qc_res <- wrap_vpa_qc_pipeline(vpa_df, master_spine, clinical_correlations, conf, vpa_id, output_dir_data)
  
  # Append any missing variable warnings to the buffer
  if (length(qc_res$html_warnings) > 0) {
    html_out <- paste0(html_out, paste(qc_res$html_warnings, collapse = "\n\n"), "\n\n")
  }
  
  # Dump HTML string so far, then print plots sequentially
  cat(html_out)
  html_out <- "" # Reset buffer
  
  if (!is.null(qc_res$plots[["distribution_grid"]])) {
    print(qc_res$plots[["distribution_grid"]])
    cat("\n\n<br><hr>\n\n")
  }
  
  if (!is.null(qc_res$plots[["correlation_heatmap"]])) {
    print(qc_res$plots[["correlation_heatmap"]])
    cat("\n\n<br>\n\n")
  }
}

# Clean memory
rm(html_out, vpa_id, conf, vpa_csv_path, vpa_df, qc_res)
invisible(gc())
```

# Automated DEA Variable Selector {.tabset}

`r what_this_does("Executes a greedy top-down selection algorithm to automatically construct the optimal Limma covariates.", "Forces must-include variables into the model, checking for confounders and variance inflation.", "Evaluates remaining variables by univariate VPA strength, rejecting those that collinearize with existing selections or break statistical power.", "Computes final Variance Inflation Factors (VIF) and outputs a custom verification heatmap of the proposed model.")`

```{r dea_selector_engine, results='asis'}
# Store the machine's drafts in memory for the Appendix
auto_draft_models <- list()

for (sel_id in names(dea_selector_configs)) {
  conf <- dea_selector_configs[[sel_id]]
  cat(sprintf("### %s\n\n", conf$title))
  
  # Print the pre-selection log if it exists!
  if (!is.null(conf$pre_selection_log)) {
    cat(sprintf("> **Pre-Selection Note:** *%s*\n\n", conf$pre_selection_log))
  }
  
  vpa_csv_path <- file.path(output_dir_data, "VPA_Univariate", conf$vpa_source_id, paste0(conf$vpa_source_id, "_Univariate_Results.csv"))
  if (!file.exists(vpa_csv_path)) next
  
  vpa_df <- read_csv(vpa_csv_path, show_col_types = FALSE)
  res <- wrap_dea_selector(clin_df = master_spine, conf = conf, vpa_df = vpa_df, cor_matrices = clinical_correlations)
  
  if (res$status == "success") {
    auto_draft_models[[sel_id]] <- res$selected_vars
    
    if (interactive()) print(res$html_table) else cat(as.character(res$html_table))
    cat("\n\n<br>\n\n")
    if (!is.null(res$heatmap)) { print(res$heatmap); cat("\n\n") }
  } else {
    cat(sprintf("> **SKIPPED:** %s\n\n", res$msg))
  }
}
```

# Manual Adjustments
```{r model_adjustments, results='asis'}
final_locked_models <- list()

for (id in names(manual_model_adjustments)) {
  adj <- manual_model_adjustments[[id]]
  
  if (is.null(adj$type) || adj$type == "adjustment") {
    if (!(id %in% names(auto_draft_models))) next 
    
    auto_vars <- auto_draft_models[[id]]
    conf <- dea_selector_configs[[id]]
    final_vars <- setdiff(unique(c(auto_vars, adj$add)), adj$drop)
    
    final_locked_models[[id]] <- list(
      id = id, title = conf$title, group_col = conf$group_col,
      target_groups = conf$target_groups, ref_groups = conf$ref_groups,
      covariates = final_vars, auto_vars = auto_vars, dropped_vars = adj$drop
    )
    
    # FIX 1: Stripped leading spaces
    cat(sprintf(
"<div style='border: 1px solid #ddd; padding: 15px; border-radius: 5px; background-color: #f9f9f9; margin-bottom: 20px;'>
<h4 style='margin-top: 0;'>%s</h4>
<b>Automated Draft:</b> %s<br>
<b>Final Locked Model:</b> %s<br>
<b>Biostatistician Note:</b> <i>%s</i>
</div>\n\n", 
    conf$title, 
    ifelse(length(auto_vars)>0, paste(auto_vars, collapse=", "), "None"), 
    ifelse(length(final_vars)>0, paste(final_vars, collapse=", "), "None"), 
    adj$log))
    
  } else if (adj$type == "custom") {
    final_vars <- adj$covariates
    
    final_locked_models[[id]] <- list(
      id = id, title = adj$title, group_col = adj$group_col,
      target_groups = adj$target_groups, ref_groups = adj$ref_groups,
      covariates = final_vars, auto_vars = character(0), dropped_vars = character(0)
    )
    
    # FIX 1: Stripped leading spaces
    cat(sprintf(
"<div style='border: 1px solid #3498db; padding: 15px; border-radius: 5px; background-color: #ebf5fb; margin-bottom: 20px;'>
<h4 style='margin-top: 0;'>%s <span style='color:#2980b9; font-size:0.8em;'>(Manual Custom Model)</span></h4>
<b>Final Locked Model:</b> %s<br>
<b>Biostatistician Note:</b> <i>%s</i>
</div>\n\n", 
    adj$title, 
    ifelse(length(final_vars)>0, paste(final_vars, collapse=", "), "None"), 
    adj$log))
  }
}

rm(id, adj, auto_vars, conf, final_vars)
invisible(gc())
```


# PCA {.tabset}

`r what_this_does("Executed the PCA pipeline to assess global cohort structure and identify potential outliers or batch effects.", "Generated scree plots to evaluate dimensional variance capture.", "Rendered and exported feature loadings and metadata-colored PCA overlays (e.g., by Cohort, Age, Sex).")`

`r why_is_this_done("The PCA is used to validate the Variable Selector.", "The PCA can show more complex relationships than mere correlation, that might warrant variable exclusion.")`

```{r execution_engine_PCA, results='asis'}
# Run Manual / Standalone PCAs defined in the Configuration Chunk
for (pca_name in names(pca_configs)) {
  pca_conf <- pca_configs[[pca_name]]
  pca_results <- wrap_pca_pipeline(olink_mat, master_spine, pca_conf, pca_name, output_dir_data, get_project_colors)
  
 cat(sprintf("\n\n## %s\n\n", pca_conf$title))
  
  # FIX 2: Print plots directly so knitr handles the HTML embedding
  print(pca_results$plots[["scree"]])
  cat("\n\n")
  print(pca_results$plots[["loadings"]])
  cat("\n\n")
  print(pca_results$plots[["merged_drivers"]])
  cat("\n\n")
  
  for (c_var in pca_conf$color_vars) { 
    plot_key <- paste0("PCA_Plot_", c_var)
    if (!is.null(pca_results$plots[[plot_key]])) {
      print(pca_results$plots[[plot_key]]) 
      cat("\n\n")
    }
  }
}

cat("\n")

# Clean memory
rm(pca_name, pca_conf, pca_results)
if (exists("plot_key")) rm(plot_key)
if (exists("c_var")) rm(c_var)
invisible(gc())
```

# Targeted Variance Partition Analysis {.tabset}

`r what_this_does("Executed multivariate Variance Partition Analysis to model the combined influence of specific clinical variable sets.", "Calculated the unique percentage of proteomics variance explained by each variable while adjusting for the others.", "Outputted bar and violin plots to visualize the distribution of explained variance across all measured proteins.")`

`r why_is_this_done("The multivariate VPA is used to validate the Variable Selector (which is based o the univariate VPA).", "The multivariate VPA shows a more accurate estimation of explained variance thatn the univariate", "If the explained variance of a variable drops markedly, this warrants variable exclusion")`

```{r targeted_vpa_engine, results='asis', fig.height=10, fig.width=8}
for (id in names(final_locked_models)) {
  m <- final_locked_models[[id]]
  cat(sprintf("\n\n## Analysis: %s (Locked Model)\n\n", m$title))
  
  dyn_conf <- list(title = m$title, variables = c(m$group_col, m$covariates))
  res <- wrap_vpa_pipeline(olink_mat, master_spine, dyn_conf, paste0("Locked_", id), output_dir_data)
  
  if(!is.null(res) && res$status == "success") {
    print(res$plots$bar); cat("\n\n"); print(res$plots$violin)
  } else {
    cat(sprintf("> <span style='color:red;'>**SKIPPED:** %s</span>\n", res$error_msg))
  }
}

# Clean memory
rm(id, m, dyn_conf, res)
invisible(gc())
```



# Pre-DEA Design Validation

`r analysis_decision("Only models that passed these criteria were accepted:", "Minimum 10 samples per parameter to prevent overfitting.", "Minimum of 2 surviving patients per contrast group to ensure valid standard error estimation in downstream limma models.","Variance > 0 after complete case filtratiion","No collinearity (checked by QR matrix)", "No correlation (rho >= 0.25 and p <= 0.05)")`

`r what_this_does("Tested proposed Differential Expression Analysis configurations against 5 mathematical pillars: Complete Case Survival, Variance Stability, Matrix Collinearity, Variance Inflation, and Statistical Power.", "Generated an integrated, variable-by-variable mathematical audit table for each proposed model to guide downstream DEA script configurations.")`

`r why_is_this_done("The Pre-DEA Design Validation is used to validate the candidate variables selected using the Automated Variable Selector, PCA and multivariate VPA.", "If the model passes this validation it is ready to be used in a DEA")`

```{r pre_dea_validation, results='asis'}
for (id in names(final_locked_models)) {
  m <- final_locked_models[[id]]
  
  dyn_conf <- list(title = sprintf("%s (Locked)", m$title), group_col = m$group_col, target_groups = m$target_groups, ref_groups = m$ref_groups, covariates = m$covariates)
  res <- wrap_pre_dea_validation(master_spine, dyn_conf, dea_validation_rules, clinical_correlations)
  
  if (interactive()) {
    print(res$html_table)
  } else {
    cat(as.character(res$html_table))
  }
  cat("\n\n") 
  
  # Print the metadata box BELOW the table
  cat(res$meta_html)
  cat("\n<br><hr>\n\n") 
}

# Clean memory
rm(id, m, dyn_conf, res)
invisible(gc())
```

# Appendix

`r what_this_does("Generates collapsible visual evidence (Before-State Multi-VPA and PCA) for any variables that were manually rejected from the automated draft models.")`

```{r diagnostic_graveyard, results='asis'}
for (id in names(final_locked_models)) {
  m <- final_locked_models[[id]]
  
  if (length(m$dropped_vars) > 0) {
    cat(sprintf("### Rejected Diagnostics: %s\n\n", m$title))
    
    # Render Dropdown wrapper
    cat(sprintf("<details><summary style='cursor: pointer; font-weight: bold; color: #2c3e50; font-size: 1.2em;'>▶ Click to view 'Before' State Visuals (Dropped: %s)</summary>\n<div style='padding: 15px; border-left: 3px solid #c0392b; margin-top: 10px; background-color: #fcfcfc;'>\n\n", paste(m$dropped_vars, collapse=", ")))
    
    cat("> **Note:** The plots below show the *Automated Draft* model before manual exclusion. They demonstrate how the dropped variables behaved in the context of the full draft model.\n\n")
    
    # 1. Run 'Before' VPA
    cat("#### Multi-VPA (Draft State)\n\n")
    draft_vpa_conf <- list(title = sprintf("Draft VPA: %s", m$title), variables = c(m$group_col, m$auto_vars))
    vpa_res <- wrap_vpa_pipeline(olink_mat, master_spine, draft_vpa_conf, paste0("Draft_", id), output_dir_data)
    if(!is.null(vpa_res) && vpa_res$status == "success") { print(vpa_res$plots$bar); cat("\n\n") }
    
    # 2. Run 'Before' PCA
    cat("#### PCA (Draft State)\n\n")
    draft_pca_conf <- list(title = sprintf("Draft PCA: %s", m$title), subset_to_contrast = list(group_col = m$group_col, target_groups = c(m$target_groups, m$ref_groups)), color_vars = c(m$group_col, m$auto_vars))
    pca_res <- wrap_pca_pipeline(olink_mat, master_spine, draft_pca_conf, paste0("Draft_", id), output_dir_data, get_project_colors)
    if (!interactive()) {
      cat(sprintf("![](%s)\n\n![](%s)\n\n![](%s)\n\n", pca_res$paths[["scree"]], pca_res$paths[["loadings"]], pca_res$paths[["merged_drivers"]]))
      if (!is.null(pca_res$paths[["merged_overlays"]])) cat(sprintf("![](%s)\n\n", pca_res$paths[["merged_overlays"]]))
    } else {
      print(pca_res$plots[["scree"]]); print(pca_res$plots[["loadings"]]); print(pca_res$plots[["merged_drivers"]])
      for (c_var in draft_pca_conf$color_vars) { if (!is.null(pca_res$plots[[paste0("PCA_Plot_", c_var)]])) print(pca_res$plots[[paste0("PCA_Plot_", c_var)]]) }
    }
    
    # Close Dropdown
    cat("\n</div>\n</details>\n\n<hr>\n\n")
  }
}
```

`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 03a Complete! \n")
```
