---
title: "03b_proteomics_differential"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and import
`r what_this_does("Loaded required bioinformatics packages (limma, clusterProfiler, msigdbr).", "Imported the pre-cleaned clinical spine and Olink proteomics matrix.", "Fetched the Human Gene Ontology (GO) Biological Processes database for downstream pathway enrichment analysis.")`
```{r setup_import, echo=FALSE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(limma)
library(clusterProfiler)
library(msigdbr)
library(patchwork)
library(here)
library(kableExtra)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "03b_proteomics_differential"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Configuration & Engines
source(here::here("code", "00_config.R"))
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data from Script 01
master_spine <- readRDS(here::here("output", "01_data_import_and_qc", "clinical_spine.rds"))
omics_mat <- readRDS(here::here("output", "01_data_import_and_qc", "clean_olink_matrix.rds"))

# Fetch GO Biological Processes Database for GSEA
go_genesets <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% select(gs_name, gene_symbol)
```

# Analysis Configurator
`r what_this_does("Defined global mathematical thresholds for Gene Set Enrichment Analysis (GSEA).", "Configured targeted Differential Expression Analysis (DEA) contrasts, specifying reference groups, target groups, and required covariates (e.g., Age, Sex, Sample_Age).", "Established Trajectory mapping rules to compare protein and pathway shifts across multiple distinct contrasts.")`
```{r analysis_config}
# 1. Global Mathematical Thresholds
omics_config <- list(
  gsea_p_cutoff  = 0.20, 
  gsea_min_size  = 3     
)


# 3. Differential Expression Contrasts (Target vs Reference)
contrast_configs <- list(
  active_vs_control = list(
    title = "Active vs Control",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),
    ref_groups = c(GRP_CONTROL),
    is_continuous = FALSE,
    covariates = c("Age", "Sex"),
    dea_p_cutoff = 0.05,  # <-- INDIVIDUAL CUTOFFS
    dea_fc_cutoff = 0.5
  ),
  nonactive_vs_control = list(
    title = "Nonactive vs Control",
    group_col = "cohort_group",
    target_groups = c(GRP_NON_ACTIVE),
    ref_groups = c(GRP_CONTROL),
    is_continuous = FALSE,
    covariates = c("Age", "Sex"),
    dea_p_cutoff = 0.05,
    dea_fc_cutoff = 0.5
  ),
  active_vs_nonactive = list(
    title = "Active vs Nonactive",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),
    ref_groups = c(GRP_NON_ACTIVE),
    is_continuous = FALSE,
    covariates = c("Age", "Sex"),
    dea_p_cutoff = 0.05,
    dea_fc_cutoff = 0.2
  ),
  active_vs_nonactive_sample_age_corr = list(
    title = "Active vs Nonactive (Sample Age Corrected)",
    group_col = "cohort_group",
    target_groups = c(GRP_ACTIVE),
    ref_groups = c(GRP_NON_ACTIVE),
    is_continuous = FALSE,
    covariates = c("Age", "Sex", "Sample_Age"), 
    dea_p_cutoff = 0.05,
    dea_fc_cutoff = 0.5
  )
)

## 4. Cross-Contrast Trajectory Plots
cross_contrasts <- list(
  active_vs_nonactive_trajectory = list(
    title = "Trajectory: Active vs Nonactive",
    x_axis_contrast = "active_vs_control",      
    y_axis_contrast = "nonactive_vs_control",   
    top_n_specific  = 5,  
    top_n_shared    = 5,
    top_n_pathways  = 10,       # Per Quadrant
    pathway_rank_var = "rank_score" # Internal rank score uses lowest FDR sum
  )
)

# 5. Heatmap Configuration
heatmap_config <- list(
  top_n = 30,
  split_by_group = FALSE, 
  row_font_size = 8
)

# 6. Comparative Settings
cross_contrasts$active_vs_nonactive_trajectory$top_n_upset <- 10      
cross_contrasts$active_vs_nonactive_trajectory$top_n_pathways <- 20   


cat("Pipelines and Mathematical Engines fully configured.\n")
```

# Execution Engines

## Differential Expression Analysis & GSEA
`r what_this_does("Executed linear modeling (limma) for each defined contrast to identify differentially expressed proteins while adjusting for covariates.", "Generated Volcano plots and static Heatmaps to visualize significant fold changes and expression clusters.", "Performed Gene Set Enrichment Analysis (GSEA) on full fold-change ranked lists and rendered Dot and Ridge plots for biological interpretation.")`
```{r execution_engine_DEA_GSEA, fig.width=8,fig.height=8}
cat("### Executing Differential Expression & Contrast-Specific Discovery...\n\n")

master_dea_results <- list()
master_gsea_results <- list() 

for (c_name in names(contrast_configs)) {
  
  # Run Master Pipeline
  res <- wrap_dea_pipeline(
    omics_mat           = omics_mat, 
    clin_df             = master_spine, 
    dea_conf            = contrast_configs[[c_name]], 
    dea_id              = c_name, 
    base_output_dir     = output_dir_data, 
    go_db               = go_genesets, 
    omics_config        = omics_config, 
    heatmap_config      = heatmap_config, 
    project_colors_func = get_project_colors
  )
  
  # Report Status
  cat(sprintf("#### Contrast: %s\n", contrast_configs[[c_name]]$title))
  cat(sprintf("*Diagnostic: Samples used: %d Target, %d Reference*\n\n", 
              res$diagnostics$n_target, res$diagnostics$n_ref))
  
  if(res$status == "failed") {
    cat(sprintf("<span style='color:red;'>**SKIPPED:** %s</span>\n\n", res$error_msg))
    next
  }
  
  master_dea_results[[c_name]] <- res$data$dea
  master_gsea_results[[c_name]] <- res$data$gsea

  # Rendering
  print(res$plots$volcano); cat("\n\n")
  
  if(nrow(res$tables$significant_proteins) > 0) {
    cat(sprintf("##### Significant Proteins (%d total)\n\n", nrow(res$tables$significant_proteins)))
    print(kbl(res$tables$significant_proteins, format = "html", digits = 4, align = "c", 
              caption = sprintf("Significant Results: %s", contrast_configs[[c_name]]$title)) %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left"))
    cat("\n\n")
  }

  draw(res$plots$heatmap); cat("\n\n")

  if(!is.null(res$plots$gsea_dotplot)) print(res$plots$gsea_dotplot); cat("\n\n")
  if(!is.null(res$plots$gsea_ridge)) print(res$plots$gsea_ridge); cat("\n\n")
  
  
  cat("--- \n\n")
}
```

## Trajectory Mapping
`r what_this_does("Mapped the fold changes of proteins across two distinct clinical contrasts to identify shared and diverging biological trajectories (Quadrant Plots).", "Generated Venn diagrams to visualize the intersection of significant features between contrasts.", "Calculated and plotted cross-contrast pathway trajectories to reveal higher-order biological mechanisms driving cohort differences.")`
```{r execution_engine_trajectory, fig.width=8, fig.height=8}
cat("### Executing Comparative Analysis (Trajectories)...\n\n")

for (cross_name in names(cross_contrasts)) {
  conf <- cross_contrasts[[cross_name]]
  
  # CALL THE MASTER PIPELINE
  # FIX: Added 'omics_config' to the arguments below
  traj_res <- wrap_trajectory_pipeline(
    cross_conf          = conf, 
    cross_id            = cross_name, 
    master_dea          = master_dea_results, 
    master_gsea         = master_gsea_results, 
    contrast_configs    = contrast_configs,
    base_output_dir     = output_dir_data, 
    project_colors_func = get_project_colors, 
    highlight_color     = COLOR_HIGHLIGHT,
    omics_config        = omics_config      # <--- ADD THIS LINE
  )
  
  cat(sprintf("#### Comparison: %s\n", conf$title))
  
  if(traj_res$status == "failed") {
    cat(sprintf("<span style='color:red;'>**SKIPPED:** %s</span>\n\n", traj_res$error_msg))
    next
  }

  # Inline HTML Rendering
  if(!is.null(traj_res$plots$protein_trajectory)) {
    print(traj_res$plots$protein_trajectory); cat("\n\n")
  }
  
  if(!is.null(traj_res$plots$venn)) {
    cat("##### Feature Intersection (Venn Diagram)\n")
    print(traj_res$plots$venn); cat("\n\n")
  }

  if(!is.null(traj_res$plots$pathway_trajectory)) {
    cat("##### Pathway Trajectory\n")
    print(traj_res$plots$pathway_trajectory); cat("\n\n")
  }
  
  cat("--- \n\n")
}
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 03b Complete! \n")
```
