---
title: "Clustering"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded required packages including UMAP for non-linear dimensionality reduction and gtsummary for clinical cross-referencing.", "Imported the pre-cleaned clinical spine and Olink proteomics matrix.", "Sourced custom omics engines")`

```{r setup_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(umap)       # Non-linear reduction
library(cluster)    # Diagnostics
library(dbscan)     # HDBSCAN Density clustering
library(igraph)     # Louvain Graph clustering
library(gtsummary)  # Clinical tables$
library(msigdbr)

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "03c_proteomics_clustering"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Global DEA Path
proteomics_dea_path <- here::here("output", "03b_proteomics_differential")

# Source Configuration & Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
data_dictionary <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
olink_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

# Fetch GO Biological Processes Database for GSEA
go_genesets <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% select(gs_name, gene_symbol)
```

# Configurations
`r what_this_does("Configured the Matrix Factory, Algorithm Optimization, Clustering, and dual Phenotyping (Clinical & Feature) parameters.")`

```{r config}
# 0. Data Dictionary Shield & Base Excludes
unstable_vars <- data_dictionary %>% filter(grepl("ZERO VARIANCE", toupper(QC_Flag))) %>% pull(Variable)
base_excludes <- c("inCentraxx", "SELECTED", "Subject_ID", "Birth_Year", "Raynaud VAS Unknown", "Face", "Chest", "Abdomen", "Left_Upper Arms", "Right_Upper Arms", "Left_Forearms", "Right_Forearms", "Left_Hands", "Right_Hands", "Left_Fingers", "Right_Fingers", "Left_Thighs", "Right_Thighs", "Left_Legs", "Right_Legs", "Left_Feet", "Right_Feet", "Active_our", "Geschlecht", "Entnahmedatum", "Entnahmedatum_Clean", "Distance_in_m")
final_excludes <- unique(c(base_excludes, unstable_vars))

# 1. Matrix Factory Configurations
matrix_prep_configs <- list(
  mat_avc_age_residualized = list(
    subset_var = NULL,
    base_features = list(
      method = "dea", dea_path = proteomics_dea_path, source_file = "dea_active_vs_control.rds", 
      fdr_cutoff = 0.05, fc_cutoff = 0.58, min_features = 30, top_n = 200 
    ),
    transform = list(method = "residualize", covariates = c("Age"))
  )
)

# 2. Pre-Flight Algorithm Optimization Configs
k_opt_configs <- list(
  k_opt_multi_topology = list(
    title = "Optimization: Age Regressed",
    input_matrix = "mat_avc_age_residualized",
    k_range = 2:8,
    gap_bootstraps = 0, 
    algorithms = list(
      pam = list(run=TRUE, metric="euclidean"),
      kmeans = list(run=TRUE, nstart=25),
      hclust = list(run=TRUE, distance="euclidean", method="ward.D2"),
      hdbscan = list(run=TRUE, minPts_range = 2:15),
      louvain = list(run=TRUE, k_neighbors = 4, resolution_range = seq(0.1, 2.0, by=0.1))
    )
  )
)

# 3. Clustering Configs 
clustering_configs <- list(
  job_avc = list(
    title = "Age Regressed Clustering",
    input_matrix = "mat_avc_age_residualized", 
    k = 3, 
    n_neighbors = 5,
    algorithms = list(
      kmeans = list(run=TRUE, nstart=25), 
      pam = list(run=TRUE, metric="euclidean"), 
      hclust = list(run=TRUE, distance="euclidean", method="ward.D2"), 
      hdbscan = list(run=TRUE, minPts=3),          
      louvain = list(run=TRUE, k_neighbors=4, resolution=0.5) 
    )
  )
)

# 4. Diagnostic Targets
diagnostic_configs <- list(
  evaluate_jobs = c("job_avc"),
  correlate_vars = c("cohort_group", "Age", "Sex")
)

# 5. Clinical Phenotyping Targets
pheno_configs <- list(
  pheno_avc = list(
    target_job = "job_avc",
    title = "Age Regressed Clustering",
    target_algorithms = c("kmeans", "hclust","louvain"),
    variables = unique(c(ssc_clinical_parameters, "Age", "Sex", "cohort_group")),
    highlight_variables = c("Age", "Sex", "cohort_group") ,
    p_cutoff = 0.05,
    must_include_vars = c("Age", "Sex", "cohort_group")
  )
)

# 6. Feature Phenotyping Targets (DEA)
feature_configs <- list(
  feat_avc = list(
    target_job = "job_avc",
    title = "Age Regressed Clustering",
    input_matrix = "raw", 
    covariates = c("Age"),
    # Algorithms are now the primary key to enable contrast-per-algorithm
    algorithms = list(
      kmeans = list(
        contrasts = list(
          c1_vs_c2 = list(title = "Cluster 1 vs Cluster 2", desc = "Active-like (high mRSS & high anti-RNA-Pol-III) vs Control-like", target=c("Cluster 1"), ref=c("Cluster 2")),
          c1_vs_c3 = list(title = "Cluster 1 vs Cluster 3", desc = "Active-like (high mRSS & high anti-RNA-Pol-III) vs Nonactive-Active mixed", target=c("Cluster 1"), ref=c("Cluster 3"))
        )
      ),
      hclust = list(
        contrasts = list(
          c2_vs_c1 = list(title = "Cluster 2 vs Cluster 1", desc = "Active-like (high mRSS & high anti-RNA-Pol-III) vs Control-like", target=c("Cluster 2"), ref=c("Cluster 1")),
          c2_vs_c3 = list(title = "Cluster 2 vs Cluster 3", desc = "Active-like (high mRSS & high anti-RNA-Pol-III) vs Nonactive-Active mixed", target=c("Cluster 2"), ref=c("Cluster 3"))
        )
      ),
      louvain = list(
        contrasts = list(
          c2_vs_c1 = list(title = "Cluster 2 vs Cluster 1", desc = "Active-like (high mRSS) vs Control-like", target=c("Cluster 2"), ref=c("Cluster 1")),
          c2_vs_c3 = list(title = "Cluster 2 vs Cluster 3", desc = "Active-like (high mRSS) vs Nonactive-Active mixed", target=c("Cluster 2"), ref=c("Cluster 3"))
        )
      )
    ),
    dea_p_cutoff = 0.05,
    dea_fc_cutoff = 0.58
  )
)

# 6.b DEA Heatmap Configs
heatmap_config <- list(
  top_n = 30,
  split_by_group = TRUE, 
  row_font_size = 8
)

# 6.c GSEA configs (called omics_configs to keep things confusing)
omics_config <- list(
  gsea_p_cutoff  = 0.20, 
  gsea_min_size  = 3     
)
```

# Matrix Factory
`r analysis_decision("Employed three distinct matrix transformations to isolate the disease signature from the Age confounder. The baseline matrix remains unmodified. The pruned matrix surgically removes features correlating heavily with Age (Spearman rho > 0.4). The residualized matrix uses linear modeling to mathematically regress Age out of the coordinate space.")` 

`r what_this_does("Matrix Factory: Dynamically generates custom feature matrices by subsetting the Olink proteomics landscape.", "Transformation: Applies mathematical corrections (Residualization vs. Pruning) to isolate the clinical signature from the Age confounder.", "Audit: Tracks the inclusion/exclusion status of every feature against defined biological thresholds (FDR/FC) to ensure transparency in the dimensionality of your input space.")`

```{r matrix_factory, results='asis'}
# Global storage for the prepared mathematical matrices
prepared_matrices <- list()

for (prep_id in names(matrix_prep_configs)) {
  
  conf <- matrix_prep_configs[[prep_id]]
  res <- wrap_matrix_factory(olink_mat, master_spine, conf, prep_id)
  
  if (res$status == "success") {
    prepared_matrices[[prep_id]] <- res$matrix
    
    # 1. Print the raw text log generated by the engine
    cat(res$log_text, "\n\n") 
    
    # 2. Format and Print the Audit Table
    if (!is.null(res$audit_tbl) && nrow(res$audit_tbl) > 0) {
      formatted_tbl <- res$audit_tbl %>%
        mutate(
          # Clean up the numbers
          logFC = round(logFC, 3),
          FDR = formatC(FDR, format = "e", digits = 2),
          
          # Inject HTML color tags for the status
          Final_Status = case_when(
            grepl("Included", Final_Status) ~ sprintf("<span style='color:#27ae60; font-weight:bold;'>%s</span>", Final_Status),
            grepl("Excluded", Final_Status) ~ sprintf("<span style='color:#c0392b; font-weight:bold;'>%s</span>", Final_Status),
            TRUE ~ Final_Status
          )
        ) %>%
        kableExtra::kbl(format = "html", escape = FALSE, caption = sprintf("Matrix Audit: %s", prep_id)) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
        kableExtra::scroll_box(height = "400px") 
        
      print(formatted_tbl)
    }
    
  } else {
    cat(sprintf("FAILED: %s\n", res$error_msg))
  }
}
```
# Algorithm Optimization
`r what_this_does("Executes algorithm-aware hyperparameter sweeps.", "Evaluates Silhouette boundaries for K-Means, PAM, and H-Clust.", "Traces the Noise-to-Cluster tradeoff for HDBSCAN density, and measures Modularity peaks for Louvain network communities.")`

```{r algorithm_optimization}
for (opt_id in names(k_opt_configs)) {
  conf <- k_opt_configs[[opt_id]]
  
  target_matrix <- prepared_matrices[[conf$input_matrix]]
  if (is.null(target_matrix)) next
  
  res <- wrap_k_optimization_pipeline(target_matrix, conf, opt_id, output_dir_data)
  
  if (res$status == "success") {
    for (algo in names(res$plots)) {
      print(res$plots[[algo]])
    }
  }
}
```


# Unsupervised Clustering {.tabset}
`r what_this_does("Ingested the dynamically pre-processed matrices from the Matrix Factory.", "Calculated mathematical distances and assigned endotypes using true high-dimensional protein matrices.", "Calculated 2D UMAP coordinates exclusively for visualization, painting high-dimensional labels onto the squashed axes.")`

`r why_is_this_done("Clustering on projected 2D UMAP space introduces manifold distortion. By calculating groupings mathematically across all protein dimensions, we ensure biological robustness, even if the 2D visual projection appears slightly overlapping.")`
```{r unsupervised_clustering, results='asis'}
# Store the payload globally so the diagnostic chunk can access it later!
global_cluster_payloads <- list()

for (c_id in names(clustering_configs)) {
  conf <- clustering_configs[[c_id]]
  # Inject the Job Title
  cat(sprintf("\n\n## %s {.tabset}\n\n", conf$title))
  
  target_matrix <- prepared_matrices[[conf$input_matrix]]
  if (is.null(target_matrix)) next
  
  res <- wrap_clustering_pipeline(target_matrix, master_spine, conf, c_id, output_dir_data, get_project_colors)
  global_cluster_payloads[[c_id]] <- res
  
  if (res$status == "success") {
    for (algo in names(res$plots$umaps)) {
      # Inject the Algorithm Tab
     cat(sprintf("\n\n### %s\n\n", toupper(algo)))
     print(res$plots$umaps[[algo]])
      cat("\n\n") # Crucial empty lines for pandoc rendering
    }
  } 
}
```
# Clustering Diagnostic {.tabset}

`r analysis_decision("Employed the Silhouette Score to mathematically quantify algorithm confidence. The algorithm with the highest average width indicates the most robust, well-separated endotypes.")`

`r what_this_does("Extracted the underlying distance matrices and specific patient assignments from the upstream multi-algorithm pipeline.", "Calculated the global Silhouette width for K-Means, PAM, and Hierarchical clustering.", "Generated comparative bar charts to guide algorithmic selection.")`

```{r clustering_diagnostic,results='asis'}
# Store diagnostic results to extract the "best_algo" for Consensus later
global_diagnostic_payloads <- list()

for (job_id in diagnostic_configs$evaluate_jobs) {
  payload <- global_cluster_payloads[[job_id]]
  
  if (!is.null(payload) && payload$status == "success") {
    
    # Extract the exact matrix and config used for this job
    target_matrix <- prepared_matrices[[clustering_configs[[job_id]]$input_matrix]]
    cluster_conf <- clustering_configs[[job_id]]
    
    diag_res <- run_clustering_diagnostics(
      payload, target_matrix, master_spine, data_dictionary, 
      diagnostic_configs, cluster_conf, job_id, output_dir_data
    )
    
    global_diagnostic_payloads[[job_id]] <- diag_res
    
    if (!is.null(diag_res)) {
      # 1. Print the New Global HTML Table
      print(diag_res$plots[["leaderboard"]])
      
      # 2. Print Variable Correlation Plots
      for (v in names(diag_res$plots)) {
        if (v == "leaderboard") next
        print(diag_res$plots[[v]])
      }
    }
  }
}
```

# Clinical Phenotyping {.tabset}
`r analysis_decision("Muted non-significant phenotypes (p >= 0.05) to zero on the heatmap, ensuring visual emphasis is strictly placed on true biological drivers rather than random noise.")`

`r what_this_does("Calculates Log-Odds Ratios and Z-Scores to map the clinical fingerprint of the discovered endotypes.")`

```{r clinical_phenotyping, fig.height=6, fig.width=12, warning=FALSE, results='asis'}
for (p_id in names(pheno_configs)) {
  conf <- pheno_configs[[p_id]]
  # Assuming you have the job name mapped, or just print conf$target_job
  cat(sprintf("\n\n## %s {.tabset}\n\n", conf$title))
  payload <- global_cluster_payloads[[conf$target_job]]
  if (is.null(payload)) next
  
  res <- wrap_clinical_phenotyping(payload, master_spine, data_dictionary, conf, p_id, output_dir_data)
  
  if (res$status == "success") {
    for (algo in names(res$plots)) {
      # Inject Markdown Header for the Tabset
      cat(sprintf("\n\n### %s\n\n", toupper(algo)))
      
      # Print Table (Converted to kableExtra for consistent HTML styling)
      if (!is.null(res$tables[[algo]])) {
        tbl <- res$tables[[algo]] %>% 
          gtsummary::as_kable_extra(caption = sprintf("Clinical Summary: %s", toupper(algo))) %>%
          kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
        print(tbl)
        cat("\n\n")
      }
      
      # Print Heatmap (Using ComplexHeatmap::draw)
      if (!is.null(res$plots[[algo]]$heatmap)) {
        ComplexHeatmap::draw(res$plots[[algo]]$heatmap)
        cat("\n\n")
      }
      
      # Print Grid
      if (!is.null(res$plots[[algo]]$highlights)) {
        print(res$plots[[algo]]$highlights)
        cat("\n\n")
      }
    }
  }
}
```

# Feature Phenotyping {.tabset}
`r what_this_does("Runs rigorous Differential Expression Analysis directly between the new, algorithmically discovered endotypes.")`

```{r feature_phenotyping, fig.width=8, fig.height=8, results='asis'}
for (f_id in names(feature_configs)) {
  conf <- feature_configs[[f_id]]
  payload <- global_cluster_payloads[[conf$target_job]]
  target_matrix <- if(conf$input_matrix == "raw") olink_mat else prepared_matrices[[conf$input_matrix]]
  
  if (is.null(payload) || is.null(target_matrix)) next

  # Tier 1: Job Title
  cat(sprintf("\n\n## %s {.tabset}\n\n", conf$title))

  for (algo_name in names(conf$algorithms)) {
    # Tier 2: Algorithm Tab
    cat(sprintf("\n\n### %s {.tabset}\n\n", toupper(algo_name)))
    
    # Inject cluster labels for this algorithm
    temp_spine <- master_spine
    temp_spine$Dynamic_Cluster <- as.character(payload$data$assignments[[algo_name]][match(temp_spine$Subject_ID, names(payload$data$assignments[[algo_name]]))])
    temp_spine <- temp_spine %>% filter(!is.na(Dynamic_Cluster), Dynamic_Cluster != "Noise")

    # Tier 3: Contrast Loop
    algo_conf <- conf$algorithms[[algo_name]]
    for (contrast_name in names(algo_conf$contrasts)) {
      c_conf <- algo_conf$contrasts[[contrast_name]]
      
      # The \n\n padding here is crucial for pandoc
      cat(sprintf("\n\n#### %s\n\n", c_conf$title))
      cat(sprintf("**Description:** %s\n\n", c_conf$desc))
      
      dea_conf <- list(
        title = sprintf("%s | Alg: %s", c_conf$title, toupper(algo_name)),
        group_col = "Dynamic_Cluster",
        target_groups = c_conf$target,
        ref_groups = c_conf$ref,
        covariates = conf$covariates,
        dea_p_cutoff = conf$dea_p_cutoff,
        dea_fc_cutoff = conf$dea_fc_cutoff
      )
      
      res <- wrap_dea_pipeline(target_matrix, temp_spine, dea_conf, sprintf("%s_%s_%s", f_id, algo_name, contrast_name), output_dir_data, go_genesets, omics_config, heatmap_config, get_project_colors)
      
      if (res$status == "success") {
        
        # 1. TABLE
        if (nrow(res$tables$significant_proteins) > 0) {
          tbl <- head(res$tables$significant_proteins, 10) %>% 
            mutate(logFC = round(logFC, 2), adj.P.Val = formatC(adj.P.Val, format = "e", digits = 2)) %>%
            kableExtra::kbl(format = "html", caption = "Top 10 Significant Proteins") %>% 
            kableExtra::kable_styling(bootstrap_options = "striped")
          print(tbl)
          cat("\n\n") # Reset markdown parser
        }
        
        # 2. VOLCANO PLOT
        if (!is.null(res$plots$volcano)) {
          print(res$plots$volcano)
          cat("\n\n") # Reset markdown parser
        }
        
        # 3. GSEA DOTPLOT
        if (!is.null(res$plots$gsea_dotplot)) {
          print(res$plots$gsea_dotplot)
          cat("\n\n")
        } else {
          cat("*Note: No GSEA Dotplot generated (insufficient pathways passed FDR).* \n\n")
        }
        
        # 4. GSEA RIDGEPLOT
        if (!is.null(res$plots$gsea_ridge)) {
          print(res$plots$gsea_ridge)
          cat("\n\n")
        } else {
          cat("*Note: No GSEA Ridgeplot generated (requires >= 2 significant pathways).* \n\n")
        }
        
        # 5. COMPLEXHEATMAP (Moved to the very end to prevent tabset hijacking)
        if (!is.null(res$plots$heatmap)) {
          ComplexHeatmap::draw(res$plots$heatmap)
          cat("\n\n") # The ultimate safety buffer
        }
      }
    }
  }
}
```

<!-- # Consensus Export -->

<!-- `r analysis_decision("Utilized a Reference-Based Overlap algorithm to translate arbitrary cluster numbers (e.g. Cluster 1 vs Cluster 3) into a unified coordinate space.", "Automatically anchored the alignment matrix to the algorithm with the highest average Silhouette confidence.")` -->

<!-- ` r what_this_does("Evaluated row-wise mode to determine the Final Consensus Endotype for every patient.", "Assigned patients failing the stability threshold into a 'Mixed/Transitional' category.", "Exported a comprehensive .rds file containing translated sub-clusters, stability scores, and individual silhouette scores for downstream multi-omics integration.")` -->

<!-- ```{r consensus_export} -->
<!-- for (export_id in names(export_configs)) { -->
<!--   conf <- export_configs[[export_id]] -->

<!--   target_payload <- global_cluster_payloads[[conf$target_job]] -->
<!--   target_diag <- global_diagnostic_payloads[[conf$target_job]] -->

<!--   if (is.null(target_payload) || is.null(target_diag)) next -->

<!--   # Auto-select anchor if requested -->
<!--   anchor <- if (conf$anchor_algorithm == "auto") target_diag$best_algo else conf$anchor_algorithm -->

<!--   res <- wrap_consensus_export(target_payload, anchor, conf, export_id, output_dir_data) -->

<!--   if (res$status == "success") { -->
<!--     # Print the raw text log natively -->
<!--     cat(res$text_log, "\n") -->

<!--     # Render the new diagnostics dynamically to the RStudio Viewer -->
<!--     print(res$plots$convergence_bar) -->
<!--     print(res$plots$consensus_umap) -->
<!--   } -->
<!-- } -->
<!-- ``` -->


# Appendix
`r what_this_does("Appended the raw underlying R code from the custom predictive engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 03c Complete! \n")
```
