---
title: "Clustering"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded required packages including UMAP for non-linear dimensionality reduction and gtsummary for clinical cross-referencing.", "Imported the pre-cleaned clinical spine and Olink proteomics matrix.", "Sourced custom omics engines tailored for unsupervised learning.")`
```{r setup_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(umap)       # Non-linear reduction
library(gtsummary)  # Clinical tables

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "03c_proteomics_clustering"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Configuration & Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data (Twin Spine Architecture)
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
data_dictionary <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
olink_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

# Load High Variance Proteins (generated at end of 03a)
high_var_path <- here::here("output", "03a_proteomics_variance", "high_var_features.rds")
high_var_prots <- if(file.exists(high_var_path)) readRDS(high_var_path) else NULL
```

# Configurations
`r what_this_does("Defined configuration parameters for unsupervised molecular clustering.", "Calculated a Non-Active specific high-variance protein subset using the global variance engine.", "Set K-means centers (k) to test different numbers of potential endotypes.", "Selected specific clinical variables to automatically profile against the newly discovered clusters.")`

```{r config}
# 0. Data Dictionary Shield & Base Excludes ------------------------------------
# Automatically identify variables that are mathematically unstable, skewed, or contain outliers
unstable_vars <- data_dictionary %>%
  filter(grepl("ZERO VARIANCE|IMBALANCED|SKEWED|OUTLIER", toupper(QC_Flag))) %>%
  pull(Variable)

# Base administrative or redundant variables to exclude
base_excludes <- c(
  "inCentraxx", "SELECTED", "Subject_ID", "Birth_Year", "Raynaud VAS Unknown", 
  "Face", "Chest", "Abdomen", "Left_Upper Arms", "Right_Upper Arms", 
  "Left_Forearms", "Right_Forearms", "Left_Hands", "Right_Hands", 
  "Left_Fingers", "Right_Fingers", "Left_Thighs", "Right_Thighs", 
  "Left_Legs", "Right_Legs", "Left_Feet", "Right_Feet", "Active_our",
  "Geschlecht", "Entnahmedatum", "Entnahmedatum_Clean", "Distance_in_m" 
)

final_excludes <- unique(c(base_excludes, unstable_vars))

# 1. Non-Active Specific High Variance Proteins --------------------------------
na_clin <- master_spine %>% filter(cohort_group == GRP_NON_ACTIVE)
na_mat <- olink_mat[, intersect(colnames(olink_mat), na_clin$Subject_ID)]

na_var_conf <- list(
  title = "Non-Active Variance Filter", 
  top_n_features = 260, 
  omics_type = "Protein"
)

# Run the variance filter engine specifically on the Non-Active matrix
na_var_res <- wrap_global_variance_filter(na_mat, na_var_conf, output_dir_data)
na_high_var_prots <- na_var_res$features

# 2. Molecular Clustering Configs ----------------------------------------------
clustering_configs <- list(
  
  # Test a 2-cluster split (Full Cohort)
  k2_split_filtered = list(
    title = "Molecular Endotypes (k=2) | Full Cohort High Variance",
    k = 2,
    n_neighbors = 15, 
    include_features = high_var_prots, 
    include_clinical_vars = unname(unlist(clinical_domains)), 
    exclude_clinical_vars = final_excludes 
  ),
  
  # Test a 3-cluster split (Full Cohort)
  k3_split_filtered = list(
    title = "Molecular Endotypes (k=3) | Full Cohort High Variance",
    k = 3,
    n_neighbors = 15, 
    include_features = high_var_prots, 
    include_clinical_vars = unname(unlist(clinical_domains)), 
    exclude_clinical_vars = final_excludes 
  ),
  
  # Non-Active Sub-Clustering
  nonactive_only_k2 = list(
    title = "Non-Active Sub-Clustering (k=2) | Internal High Variance",
    subset_groups = c(GRP_NON_ACTIVE), # Focus only on Non-Actives
    k = 2,
    n_neighbors = 5, # Lowered due to much smaller N-size of this subset
    include_features = na_high_var_prots, # Using the internally filtered top 260
    include_clinical_vars = unname(unlist(clinical_domains)),
    exclude_clinical_vars = final_excludes
  )
)
```

# Unsupervised Clustering {.tabset}
`r what_this_does("Calculated UMAP coordinates to map high-dimensional proteomics into 2D space based on biological similarity.", "Applied K-Means clustering to officially assign patients to new molecular groups.", "Cross-referenced the newly discovered groups against the clinical spine to identify defining clinical characteristics.")`

`r why_is_this_done("Since predefined clinical labels (like Inflammatory vs Non-Inflammatory) did not yield strong proteomic differences, the approach was inverted. By letting the proteins organically group the patients, we can discover hidden disease endotypes and retrospectively identify what clinical features they share.")`
```{r UMAP, results='asis'}
for (c_id in names(clustering_configs)) {
  conf <- clustering_configs[[c_id]]
  cat(sprintf("\n\n## %s\n\n", conf$title))
  
  res <- wrap_clustering_pipeline(olink_mat, master_spine, conf, c_id, output_dir_data, get_project_colors)
  
  if (res$status == "success") {
    
    # 1. Print the UMAP
    print(res$plots$umap)
    cat("\n\n<br>\n\n")
    
    # 2. Print Diagnostic Histograms (if any variables were dropped due to skew)
    if (!is.null(res$plots$excluded_hist)) {
      cat("#### Diagnostic: Skewed Variables Excluded\n\n")
      print(res$plots$excluded_hist)
      cat("\n\n<br>\n\n")
    }
    
    # 3. Print the Clinical Cross-Reference Table with Log
    if (!is.null(res$tables$clinical_profile)) {
      cat(sprintf("#### Clinical Profile of %s\n\n", conf$title))
      
      # Print Auto-Filter Log
      log_msg <- sprintf("> **Auto-QA/QC Filter Log:**\n> * **Dropped Categorical (<10%% frequency):** %s\n> * **Dropped Continuous (|Skew| > 2.5):** %s\n\n",
                         ifelse(length(res$log$dropped_cat) > 0, paste(res$log$dropped_cat, collapse = ", "), "None"),
                         ifelse(length(res$log$dropped_cont) > 0, paste(res$log$dropped_cont, collapse = ", "), "None"))
      cat(log_msg)
      
      # Extract HTML from gtsummary object for rendering
      tbl_html <- res$tables$clinical_profile %>% 
        as_kable_extra(format = "html") %>% 
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
        
      if (interactive()) {
        print(res$tables$clinical_profile)
      } else {
        cat(as.character(tbl_html))
      }
    }
  } else {
    cat(sprintf("> <span style='color:red;'>**SKIPPED:** %s</span>\n\n", res$error_msg))
  }
}

# Clean memory
rm(c_id, conf, res)
invisible(gc())
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom predictive engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 03c Complete! \n")
```
