---
title: "Clustering"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded required packages including UMAP for non-linear dimensionality reduction and gtsummary for clinical cross-referencing.", "Imported the pre-cleaned clinical spine and Olink proteomics matrix.", "Sourced custom omics engines")`

```{r setup_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(umap)       # Non-linear reduction
library(cluster)    # Diagnostics
library(dbscan)     # HDBSCAN Density clustering
library(igraph)     # Louvain Graph clustering
library(gtsummary)  # Clinical tables

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "03c_proteomics_clustering"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Global DEA Path
proteomics_dea_path <- here::here("output", "03b_proteomics_differential")

# Source Configuration & Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
data_dictionary <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
olink_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))
```

# Configurations
`r what_this_does("Configured the Matrix Factory, Algorithm Optimization, Clustering, and dual Phenotyping (Clinical & Feature) parameters.")`

```{r config}
# 0. Data Dictionary Shield & Base Excludes
unstable_vars <- data_dictionary %>% filter(grepl("ZERO VARIANCE", toupper(QC_Flag))) %>% pull(Variable)
base_excludes <- c("inCentraxx", "SELECTED", "Subject_ID", "Birth_Year", "Raynaud VAS Unknown", "Face", "Chest", "Abdomen", "Left_Upper Arms", "Right_Upper Arms", "Left_Forearms", "Right_Forearms", "Left_Hands", "Right_Hands", "Left_Fingers", "Right_Fingers", "Left_Thighs", "Right_Thighs", "Left_Legs", "Right_Legs", "Left_Feet", "Right_Feet", "Active_our", "Geschlecht", "Entnahmedatum", "Entnahmedatum_Clean", "Distance_in_m")
final_excludes <- unique(c(base_excludes, unstable_vars))

# 1. Matrix Factory Configurations
matrix_prep_configs <- list(
  mat_avc_age_residualized = list(
    subset_var = NULL,
    base_features = list(
      method = "dea", dea_path = proteomics_dea_path, source_file = "dea_active_vs_control.rds", 
      fdr_cutoff = 0.05, fc_cutoff = 0.58, min_features = 30, top_n = 200 
    ),
    transform = list(method = "residualize", covariates = c("Age"))
  )
)

# 2. Pre-Flight Algorithm Optimization Configs
k_opt_configs <- list(
  k_opt_multi_topology = list(
    title = "Optimization: Age Regressed",
    input_matrix = "mat_avc_age_residualized",
    k_range = 2:8,
    gap_bootstraps = 0, 
    algorithms = list(
      pam = list(run=TRUE, metric="euclidean"),
      kmeans = list(run=TRUE, nstart=25),
      hclust = list(run=TRUE, distance="euclidean", method="ward.D2"),
      hdbscan = list(run=TRUE, minPts_range = 2:15),
      louvain = list(run=TRUE, k_neighbors = 4, resolution_range = seq(0.1, 2.0, by=0.1))
    )
  )
)

# 3. Clustering Configs 
clustering_configs <- list(
  job_avc = list(
    title = "Age Regressed Clustering",
    input_matrix = "mat_avc_age_residualized", 
    k = 3, 
    n_neighbors = 5,
    algorithms = list(
      kmeans = list(run=TRUE, nstart=25), 
      pam = list(run=TRUE, metric="euclidean"), 
      hclust = list(run=TRUE, distance="euclidean", method="ward.D2"), 
      hdbscan = list(run=TRUE, minPts=4),          
      louvain = list(run=TRUE, k_neighbors=4, resolution=0.5) 
    )
  )
)

# 4. Diagnostic Targets
diagnostic_configs <- list(
  evaluate_jobs = c("job_avc"),
  correlate_vars = c("cohort_group", "Age", "Sex")
)

# 5. Clinical Phenotyping Targets
pheno_configs <- list(
  pheno_avc = list(
    target_job = "job_avc",
    target_algorithms = c("kmeans", "louvain"),
    variables = unique(c(ssc_clinical_parameters, "Age", "Sex", "cohort_group")),
    highlight_variables = c("Age", "Sex", "cohort_group") ,
    p_cutoff = 0.05,
    must_include_vars = c("Age", "Sex", "cohort_group")
  )
)

# 6. Feature Phenotyping Targets (DEA)
feature_configs <- list(
  feat_avc = list(
    target_job = "job_avc",
    target_algorithms = c("kmeans", "louvain"),
    input_matrix = "raw", # Use the pure unaltered Olink matrix for unbiased biological discovery
    covariates = c("Age"), 
    contrasts = list(
      c1_vs_c2 = list(title = "Cluster 1 vs Cluster 2", target_groups = c("Cluster 1"), ref_groups = c("Cluster 2")),
      c3_vs_rest = list(title = "Cluster 3 vs Rest", target_groups = c("Cluster 3"), ref_groups = c("Cluster 1", "Cluster 2"))
    ),
    dea_p_cutoff = 0.05,
    dea_fc_cutoff = 0.58
  )
)

# 6.b DEA Heatmap Configs
heatmap_config <- list(
  top_n = 30,
  split_by_group = FALSE, 
  row_font_size = 8
)
```

# Matrix Factory
`r analysis_decision("Employed three distinct matrix transformations to isolate the disease signature from the Age confounder. The baseline matrix remains unmodified. The pruned matrix surgically removes features correlating heavily with Age (Spearman rho > 0.4). The residualized matrix uses linear modeling to mathematically regress Age out of the coordinate space.")` 

```{r matrix_factory, results='asis'}
# Global storage for the prepared mathematical matrices
prepared_matrices <- list()

for (prep_id in names(matrix_prep_configs)) {
  
  conf <- matrix_prep_configs[[prep_id]]
  res <- wrap_matrix_factory(olink_mat, master_spine, conf, prep_id)
  
  if (res$status == "success") {
    prepared_matrices[[prep_id]] <- res$matrix
    
    # 1. Print the raw text log generated by the engine
    cat(res$log_text, "\n\n") 
    
    # 2. Format and Print the Audit Table
    if (!is.null(res$audit_tbl) && nrow(res$audit_tbl) > 0) {
      formatted_tbl <- res$audit_tbl %>%
        mutate(
          # Clean up the numbers
          logFC = round(logFC, 3),
          FDR = formatC(FDR, format = "e", digits = 2),
          
          # Inject HTML color tags for the status
          Final_Status = case_when(
            grepl("Included", Final_Status) ~ sprintf("<span style='color:#27ae60; font-weight:bold;'>%s</span>", Final_Status),
            grepl("Excluded", Final_Status) ~ sprintf("<span style='color:#c0392b; font-weight:bold;'>%s</span>", Final_Status),
            TRUE ~ Final_Status
          )
        ) %>%
        kableExtra::kbl(format = "html", escape = FALSE, caption = sprintf("Matrix Audit: %s", prep_id)) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
        kableExtra::scroll_box(height = "400px") 
        
      print(formatted_tbl)
    }
    
  } else {
    cat(sprintf("FAILED: %s\n", res$error_msg))
  }
}
```
# Algorithm Optimization
`r what_this_does("Executes algorithm-aware hyperparameter sweeps.", "Evaluates Silhouette boundaries for K-Means, PAM, and H-Clust.", "Traces the Noise-to-Cluster tradeoff for HDBSCAN density, and measures Modularity peaks for Louvain network communities.")`

```{r algorithm_optimization}
for (opt_id in names(k_opt_configs)) {
  conf <- k_opt_configs[[opt_id]]
  
  target_matrix <- prepared_matrices[[conf$input_matrix]]
  if (is.null(target_matrix)) next
  
  res <- wrap_k_optimization_pipeline(target_matrix, conf, opt_id, output_dir_data)
  
  if (res$status == "success") {
    for (algo in names(res$plots)) {
      print(res$plots[[algo]])
    }
  }
}
```


# Unsupervised Clustering {.tabset}
`r what_this_does("Ingested the dynamically pre-processed matrices from the Matrix Factory.", "Calculated mathematical distances and assigned endotypes using true high-dimensional protein matrices.", "Calculated 2D UMAP coordinates exclusively for visualization, painting high-dimensional labels onto the squashed axes.")`

`r why_is_this_done("Clustering on projected 2D UMAP space introduces manifold distortion. By calculating groupings mathematically across all protein dimensions, we ensure biological robustness, even if the 2D visual projection appears slightly overlapping.")`
```{r UMAP, results='asis'}
# Store the payload globally so the diagnostic chunk can access it later!
global_cluster_payloads <- list()

for (c_id in names(clustering_configs)) {
  conf <- clustering_configs[[c_id]]
  
  target_matrix <- prepared_matrices[[conf$input_matrix]]
  if (is.null(target_matrix)) next
  
  res <- wrap_clustering_pipeline(target_matrix, master_spine, conf, c_id, output_dir_data, get_project_colors)
  global_cluster_payloads[[c_id]] <- res
  
  if (res$status == "success") {
    for (algo in names(res$plots$umaps)) {
      # Explicitly print the UMAP so it appears in the RStudio Plots pane
      print(res$plots$umaps[[algo]])
    }
  } 
}
```
# Clustering Diagnostic {.tabset}

`r analysis_decision("Employed the Silhouette Score to mathematically quantify algorithm confidence. The algorithm with the highest average width indicates the most robust, well-separated endotypes.")`

`r what_this_does("Extracted the underlying distance matrices and specific patient assignments from the upstream multi-algorithm pipeline.", "Calculated the global Silhouette width for K-Means, PAM, and Hierarchical clustering.", "Generated comparative bar charts to guide algorithmic selection.")`

```{r clustering_diagnostic}
# Store diagnostic results to extract the "best_algo" for Consensus later
global_diagnostic_payloads <- list()

for (job_id in diagnostic_configs$evaluate_jobs) {
  payload <- global_cluster_payloads[[job_id]]
  
  if (!is.null(payload) && payload$status == "success") {
    
    # Extract the exact matrix and config used for this job
    target_matrix <- prepared_matrices[[clustering_configs[[job_id]]$input_matrix]]
    cluster_conf <- clustering_configs[[job_id]]
    
    diag_res <- run_clustering_diagnostics(
      payload, target_matrix, master_spine, data_dictionary, 
      diagnostic_configs, cluster_conf, job_id, output_dir_data
    )
    
    global_diagnostic_payloads[[job_id]] <- diag_res
    
    if (!is.null(diag_res)) {
      # 1. Print the New Global HTML Table
      print(diag_res$plots[["leaderboard"]])
      
      # 2. Print Variable Correlation Plots
      for (v in names(diag_res$plots)) {
        if (v == "leaderboard") next
        print(diag_res$plots[[v]])
      }
    }
  }
}
```

# Clinical Phenotyping
`r analysis_decision("Muted non-significant phenotypes (p >= 0.05) to zero on the heatmap, ensuring visual emphasis is strictly placed on true biological drivers rather than random noise.")`

`r what_this_does("Calculates Log-Odds Ratios and Z-Scores to map the clinical fingerprint of the discovered endotypes.")`

```{r clinical_phenotyping, fig.height=6, fig.width=12, warning=FALSE}
for (p_id in names(pheno_configs)) {
  conf <- pheno_configs[[p_id]]
  payload <- global_cluster_payloads[[conf$target_job]]
  if (is.null(payload)) next
  
  res <- wrap_clinical_phenotyping(payload, master_spine, data_dictionary, conf, p_id, output_dir_data)
  
  if (res$status == "success") {
    for (algo in names(res$plots)) {
      cat(sprintf("\n\n#### Algorithm: %s\n", toupper(algo)))
      
      # Print Table
      if (!is.null(res$tables[[algo]])) {
        print(res$tables[[algo]] %>% as_gt() %>% gt::tab_header(title=sprintf("Clinical Summary: %s", toupper(algo))))
      }
      # Print Heatmap (Using ComplexHeatmap::draw)
      if (!is.null(res$plots[[algo]]$heatmap)) {
        ComplexHeatmap::draw(res$plots[[algo]]$heatmap)
      }
      # Print Grid
      if (!is.null(res$plots[[algo]]$highlights)) print(res$plots[[algo]]$highlights)
    }
  }
}
```

# Feature Phenotyping {.tabset}
`r what_this_does("Runs rigorous Differential Expression Analysis directly between the new, algorithmically discovered endotypes.")`

```{r feature_phenotyping}
for (f_id in names(feature_configs)) {
  conf <- feature_configs[[f_id]]
  payload <- global_cluster_payloads[[conf$target_job]]
  target_matrix <- if(conf$input_matrix == "raw") olink_mat else prepared_matrices[[conf$input_matrix]]
  
  if (is.null(payload) || is.null(target_matrix)) next
  
  for (algo_name in conf$target_algorithms) {
    # 1. Dynamically update the spine with this specific algorithm's clusters
    temp_spine <- master_spine
    temp_spine$Dynamic_Cluster <- as.character(payload$data$assignments[[algo_name]][match(temp_spine$Subject_ID, names(payload$data$assignments[[algo_name]]))])
    
    # Filter out Noise if it exists (HDBSCAN artifacts)
    temp_spine <- temp_spine %>% filter(!is.na(Dynamic_Cluster), Dynamic_Cluster != "Noise")

    # 2. Iterate through contrasts and call the unified DEA engine
    for (contrast_name in names(conf$contrasts)) {
      c_conf <- conf$contrasts[[contrast_name]]
      
      # Build dynamic config
      dea_conf <- list(
        title = sprintf("%s | Alg: %s | Job: %s", c_conf$title, toupper(algo_name), f_id),
        group_col = "Dynamic_Cluster",
        target_groups = c_conf$target_groups,
        ref_groups = c_conf$ref_groups,
        covariates = conf$covariates,
        dea_p_cutoff = conf$dea_p_cutoff,
        dea_fc_cutoff = conf$dea_fc_cutoff
      )
      
      # Reuse the wrap_dea_pipeline 
      res <- wrap_dea_pipeline(
        omics_mat = target_matrix,
        clin_df = temp_spine,
        dea_conf = dea_conf,
        dea_id = sprintf("%s_%s_%s", f_id, algo_name, contrast_name),
        base_output_dir = output_dir_data,
        go_db = go_genesets, # Ensure go_genesets is loaded in this script
        omics_config = omics_config,
        heatmap_config = heatmap_config,
        project_colors_func = get_project_colors
      )
      
      if (res$status == "success") {
        cat(sprintf("\n\n#### %s: %s\n", toupper(algo_name), c_conf$title))
        print(res$plots$volcano)
        
        # Display Heatmap (ComplexHeatmap object)
        if (!is.null(res$plots$heatmap)) ComplexHeatmap::draw(res$plots$heatmap)
        
        # Print Table
        if (nrow(res$tables$significant_proteins) > 0) {
          print(head(res$tables$significant_proteins, 10) %>%
            mutate(logFC = round(logFC, 2), adj.P.Val = formatC(adj.P.Val, format = "e", digits = 2)) %>%
            kableExtra::kbl(format = "html") %>% kableExtra::kable_styling(bootstrap_options = "striped"))
        }
      }
    }
  }
}
```

<!-- # Consensus Export -->

<!-- `r analysis_decision("Utilized a Reference-Based Overlap algorithm to translate arbitrary cluster numbers (e.g. Cluster 1 vs Cluster 3) into a unified coordinate space.", "Automatically anchored the alignment matrix to the algorithm with the highest average Silhouette confidence.")` -->

<!-- ` r what_this_does("Evaluated row-wise mode to determine the Final Consensus Endotype for every patient.", "Assigned patients failing the stability threshold into a 'Mixed/Transitional' category.", "Exported a comprehensive .rds file containing translated sub-clusters, stability scores, and individual silhouette scores for downstream multi-omics integration.")` -->

<!-- ```{r consensus_export} -->
<!-- for (export_id in names(export_configs)) { -->
<!--   conf <- export_configs[[export_id]] -->

<!--   target_payload <- global_cluster_payloads[[conf$target_job]] -->
<!--   target_diag <- global_diagnostic_payloads[[conf$target_job]] -->

<!--   if (is.null(target_payload) || is.null(target_diag)) next -->

<!--   # Auto-select anchor if requested -->
<!--   anchor <- if (conf$anchor_algorithm == "auto") target_diag$best_algo else conf$anchor_algorithm -->

<!--   res <- wrap_consensus_export(target_payload, anchor, conf, export_id, output_dir_data) -->

<!--   if (res$status == "success") { -->
<!--     # Print the raw text log natively -->
<!--     cat(res$text_log, "\n") -->

<!--     # Render the new diagnostics dynamically to the RStudio Viewer -->
<!--     print(res$plots$convergence_bar) -->
<!--     print(res$plots$consensus_umap) -->
<!--   } -->
<!-- } -->
<!-- ``` -->


# Appendix
`r what_this_does("Appended the raw underlying R code from the custom predictive engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 03c Complete! \n")
```
