---
title: "Non-Active Subgroup Analysis"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

# Introduction
Goal: To define a robust molecular boundary differentiating "Active-like" from "Control-like" profiles within the clinical "Non-Active" SSc cohort. 

We generate a continuous polarity-aligned heatmap score based on the SSc disease signature. We then perform a sensitivity analysis by slicing this continuous score at three distinct thresholds (50/50, 40/60, and 60/40) and executing parallel downstream Clinical Phenotyping, Differential Expression Analysis (DEA), and Gene Set Enrichment Analysis (GSEA). Finally, cross-contrast trajectory mapping validates the biological stability of the molecular boundary.

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Sourced configuration and engine scripts. Imported clinical spines, data dictionary, proteomics matrices, and the master Active vs Control DEA signature.", "Defined pathway database targets for downstream GSEA.")`

```{r setup_and_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(ComplexHeatmap)
library(circlize)
library(kableExtra)
library(ggrepel)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Paths
current_file <- "03d_proteomics_subgroups"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "01_general_engines.R"))
source(here::here("code", "02_proteomics_engines.R"))

# Load Data
master_spine  <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
data_dict     <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds")) 
olink_mat     <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))
Svc_dea       <- readRDS(here::here("output", "03b_proteomics_differential", "dea_ssc_vs_control.rds"))

pathway_dbs <- list(
  "Hallmark" = msigdbr(species = "Homo sapiens", category = "H") %>% select(gs_name, gene_symbol),
  "GO_BP" = msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% select(gs_name, gene_symbol),
  "Immunologic_Signature" = msigdbr(species = "Homo sapiens", category = "C7") %>% select(gs_name, gene_symbol),
  "Reactome" = msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME") %>% select(gs_name, gene_symbol)
)
```

# Configuration
`r what_this_does("Established highly parameterized lists controlling heatmap extraction, subgroup quantiles, DEA model parameters, clinical phenotyping parameters, and cross-contrast trajectory mapping.")`

```{r configuration}
# 1. Subgroup Definitions
groups <- list(
  active = GRP_ACTIVE,
  control = GRP_CONTROL,
  target = GRP_NON_ACTIVE
)

# 2. Heatmap & Split Threshold Configs
hm_conf <- list(
  top_n_prot = 30
)

split_conf <- list(
  # Formatted as Name = Percentile Cutoff (Higher Score = Active-Like)
  # e.g., 40/60 means 40% Active-like, so we cut at the 60th percentile
  Split_50_50 = 0.50,
  Split_40_60 = 0.60, 
  Split_60_40 = 0.40  
)

# 3. DEA & Heatmap Engines Config
dea_base_conf <- list(
  target_groups = "Active-Like",
  ref_groups = "Control-Like",
  covariates = c("Age", "Sex")
)

omics_config <- list(
  logFC_cutoff = log2(1.2),
  p_val_cutoff = 0.05,
  gsea_p_cutoff = 0.05
)

heatmap_config <- list(
  top_n = 30,
  split_by_group = TRUE,
  annotation_vars = c("Age", "Sex")
)

# 4. Phenotyping Config
pheno_vars_to_include <- data_dict %>% filter(QC_Flag == "PASS") %>% pull(Variable)

clin_vars_to_highlight <- list(
  "SSc clinical parameters" = ssc_clinical_parameters,
  "Medication" = clinical_domains$Medications,
  "Demographics_and_Scores" = clinical_domains$Demographics_and_Scores
)

clin_vars_to_include <- list(
  "Exploratory" = c(clinical_domains$Pulmonary, clinical_domains$Cardiac, clinical_domains$Labs_and_Biomarkers, clinical_domains$Cutaneous_and_Vascular, clinical_domains$Musculoskeletal, clinical_domains$Gastrointestinal, clinical_domains$Medications) 
)

pheno_base_conf <- list(
  target_algorithms = c("SSc_Splits"),
  subset_filter = NULL,
  composite_group = NULL,
  variables = clin_vars_to_include,
  highlight_variables = clin_vars_to_highlight,
  p_cutoff = 0.05,
  must_include_vars = c() # Dynamically updated in loop
)

# 5. Trajectory Mapping Configs
traj_configs <- list(
  vs_40_60 = list(
    x_axis_contrast = "Split_50_50", 
    y_axis_contrast = "Split_40_60",
    pathway_rank_var = "p.adjust",
    top_n_pathways = 8
  ),
  vs_60_40 = list(
    x_axis_contrast = "Split_50_50", 
    y_axis_contrast = "Split_60_40",
    pathway_rank_var = "p.adjust",
    top_n_pathways = 8
  )
)

# Map dynamic variables to custom colors wrapper
custom_colors <- setNames(c("#c0392b", "#2980b9"), c("Active-Like", "Control-Like"))
```

# Scoring
`r analysis_decision("Signature Selection: Defined the molecular axis using the top 30 most significant proteins from the global SSc vs Control DEA.", "Polarity Validation: Generated a strict complete hierarchical clustering of all samples, dynamically aligning the left-to-right percentile score so that 100 always points toward the 'Active' side")`
`r what_this_does("Generated a complete hierarchical clustering dendrogram of all patients. Extracted physical leaf order to assign a 0-100 percentile rank score for every Non-Active patient.")`
```{r scoring, fig.width=14, fig.height=8}
# 1. Isolate Matrix to Master Signature
sig_prots <- avc_dea %>% slice_min(adj.P.Val, n = hm_conf$top_n_prot) %>% pull(Feature)
valid_sig <- intersect(sig_prots, rownames(olink_mat))

hm_clin <- master_spine %>% filter(!is.na(cohort_group)) %>% droplevels()
hm_mat <- t(scale(t(olink_mat[valid_sig, hm_clin$Subject_ID])))

# 2. Strict Complete Clustering
hm <- Heatmap(hm_mat, clustering_distance_columns = "euclidean", clustering_method_columns = "complete")
hm_drawn <- draw(hm, padding = unit(c(2, 2, 2, 2), "mm")) # Draw implicitly constructs the object

# 3. Extract Order and Map to 0-100 Percentile Rank
order_idx <- column_order(hm_drawn)
ordered_ids <- colnames(hm_mat)[order_idx]

hm_df <- data.frame(
  Subject_ID = ordered_ids,
  HM_Rank = 1:length(ordered_ids)
) %>%
  mutate(HM_Percentile = (HM_Rank / max(HM_Rank)) * 100) %>%
  inner_join(hm_clin %>% select(Subject_ID, cohort_group), by="Subject_ID")

# 4. Polarity Check & Correction
mean_active <- mean(hm_df$HM_Percentile[hm_df$cohort_group == groups$active])
mean_control <- mean(hm_df$HM_Percentile[hm_df$cohort_group == groups$control])

if (mean_active < mean_control) {
  hm_df$HM_Active_Score <- 100 - hm_df$HM_Percentile
} else {
  hm_df$HM_Active_Score <- hm_df$HM_Percentile
}

hm_final <- hm_df %>% filter(cohort_group == groups$target) %>% select(Subject_ID, HM_Active_Score)

# Memory Cleanup
rm(hm_mat, hm_drawn, hm, hm_df); invisible(gc())
```

# Sensitivity Subgroup Thresholding
`r analysis_decision("Data Isolation: Filtering exclusively to the Non-Active cohort before calculating quantiles ensures the 50/50, 40/60, and 60/40 splits accurately reflect distributions within the target group, untainted by actual Active/Control samples.")`
`r what_this_does("Computed exact quantile thresholds based on the Non-Active patient scores. Assigned strict binary labels ('Active-Like' vs 'Control-Like'). Exported the master splitting index as an RDS.")`

```{rsensitivity_subgroup_thresholding}
# 1. Calculate Cutoffs
cutoff_50 <- quantile(hm_final$HM_Active_Score, probs = split_conf$Split_50_50, na.rm=TRUE)
cutoff_40 <- quantile(hm_final$HM_Active_Score, probs = split_conf$Split_40_60, na.rm=TRUE)
cutoff_60 <- quantile(hm_final$HM_Active_Score, probs = split_conf$Split_60_40, na.rm=TRUE)

# 2. Append Binary Columns
split_spine <- hm_final %>%
  mutate(
    Split_50_50 = ifelse(HM_Active_Score > cutoff_50, "Active-Like", "Control-Like"),
    Split_40_60 = ifelse(HM_Active_Score > cutoff_40, "Active-Like", "Control-Like"),
    Split_60_40 = ifelse(HM_Active_Score > cutoff_60, "Active-Like", "Control-Like")
  )

# Merge back into clinical spine for DEA
eval_spine <- master_spine %>% inner_join(split_spine, by="Subject_ID")

# 3. Save Master Splits
saveRDS(split_spine, file.path(output_dir_data, "non_active_splits.rds"))

# Print Diagnostics
cat("=== Subgroup Sizes ===\n")
cat(sprintf("50/50 Split: %d Active-Like | %d Control-Like\n", sum(split_spine$Split_50_50=="Active-Like"), sum(split_spine$Split_50_50=="Control-Like")))
cat(sprintf("40/60 Split: %d Active-Like | %d Control-Like\n", sum(split_spine$Split_40_60=="Active-Like"), sum(split_spine$Split_40_60=="Control-Like")))
cat(sprintf("60/40 Split: %d Active-Like | %d Control-Like\n", sum(split_spine$Split_60_40=="Active-Like"), sum(split_spine$Split_60_40=="Control-Like")))
```

# Phenotyping & DEA
`r analysis_decision("Dynamic Covariate Validation: Before executing Limma for each split, the script dynamically verifies variance for requested covariates. If a covariate lacks variance within a specific binary split (e.g., all 'Active-Like' happen to be Female), it is automatically dropped from the statistical model to prevent algebraic crashes.", "Mock Clustering Payload: By wrapping our binary split column into a list structure mirroring standard clustering outputs, we can seamlessly leverage the existing wrap_clinical_phenotyping function without architectural changes.")`
`r what_this_does("Iterated through the three predefined splits. Injected binary labels into the clinical phenotyping engine to evaluate physical patient traits. Constructed custom Limma configurations, evaluated covariate stability, and deployed the DEA/GSEA engine. Stored all results in unified master lists.")`

```{r phenotyping_and_dea}
master_dea_res <- list()
master_gsea_res <- list()
master_pheno_res <- list()
split_names <- c("Split_50_50", "Split_40_60", "Split_60_40")

for (s_name in split_names) {
  
  cat(sprintf("\nProcessing %s...\n", s_name))
  
  # ---------------------------------------------------------
  # A. Clinical Phenotyping
  # ---------------------------------------------------------
  # Create Mock Clustering Payload
  mock_payload <- list(data = list(assignments = list()))
  mock_payload$data$assignments[["SSc_Splits"]] <- setNames(eval_spine[[s_name]], eval_spine$Subject_ID)
  
  # Configure Pheno
  s_pheno_conf <- pheno_base_conf
  s_pheno_conf$must_include_vars <- c(s_pheno_conf$must_include_vars, s_name)
  
  pheno_res <- wrap_clinical_phenotyping(
    clustering_payload = mock_payload,
    clin_df = eval_spine,
    dict_df = data_dict,
    conf = s_pheno_conf,
    job_id = s_name,
    base_output_dir = output_dir_data
  )
  master_pheno_res[[s_name]] <- pheno_res

  # ---------------------------------------------------------
  # B. Differential Expression (Limma + GSEA)
  # ---------------------------------------------------------
  s_dea_conf <- dea_base_conf
  s_dea_conf$group_col <- s_name
  s_dea_conf$title <- paste("Non-Active:", gsub("_", " ", s_name))
  s_dea_conf$custom_color_map <- custom_colors
  
  # 1. Variance-based Covariate Selection (Failsafe)
  valid_covs <- c()
  for (cov in s_dea_conf$covariates) {
    if (cov %in% colnames(eval_spine) && length(unique(na.omit(eval_spine[[cov]]))) > 1) {
      valid_covs <- c(valid_covs, cov)
    } else {
      message(sprintf("Warning: Covariate '%s' dropped for %s due to zero variance.", cov, s_name))
    }
  }
  s_dea_conf$covariates <- valid_covs
  
  # 2. Add annotation vars (must be valid in subset)
  s_heatmap_conf <- heatmap_config
  s_heatmap_conf$annotation_vars <- intersect(heatmap_config$annotation_vars, valid_covs)
  
  # 3. Execute Wrap
  dea_res <- wrap_dea_pipeline(
    omics_mat = olink_mat,
    clin_df = eval_spine,
    dea_conf = s_dea_conf,
    dea_id = s_name,
    base_output_dir = output_dir_data,
    go_db = NULL, # Handled directly by pathway_dbs loop in function
    omics_config = omics_config,
    heatmap_config = s_heatmap_conf,
    project_colors_func = function(x) get_project_colors(x, custom_colors)
  )
  
  # 4. Store
  if (dea_res$status == "success") {
    master_dea_res[[s_name]] <- dea_res$data$dea
    # Assumes a primary DB (e.g., GO_BP) is first in pathway_dbs list
    primary_db_name <- names(pathway_dbs)[1]
    master_gsea_res[[s_name]] <- dea_res$data$gsea[[primary_db_name]] 
  } else {
    warning("DEA Failed for ", s_name, ": ", dea_res$error_msg)
  }
}

# Cleanup RAM
rm(mock_payload, s_pheno_conf, s_dea_conf, s_heatmap_conf, dea_res, pheno_res); invisible(gc())
```

# Trajectories {.tabset}
`r what_this_does("Rendered multi-modal sensitivity metrics contrasting the foundational 50/50 split against both asymmetric splits. Includes Venn intersections of significant proteins, scatter trajectories mapping log Fold-Change stability, and quadrant analyses of Gene Set Enrichment Normalized Enrichment Scores (NES).")`

```{r trajectories}
contrast_dummy <- list(
  Split_50_50 = list(title = "50/50 Split", target_groups = "Active-Like"),
  Split_40_60 = list(title = "40/60 Split", target_groups = "Active-Like"),
  Split_60_40 = list(title = "60/40 Split", target_groups = "Active-Like")
)

html_out <- ""

for (traj_id in names(traj_configs)) {
  
  conf <- traj_configs[[traj_id]]
  title_str <- sprintf("%s vs %s", conf$x_axis_contrast, conf$y_axis_contrast)
  
  if (!interactive()) html_out <- paste0(html_out, sprintf("\n\n## %s\n\n", title_str))
  
  traj_res <- wrap_trajectory_pipeline(
    cross_conf = conf,
    cross_id = traj_id,
    master_dea = master_dea_res,
    master_gsea = master_gsea_res,
    contrast_configs = contrast_dummy,
    base_output_dir = output_dir_data,
    project_colors_func = function(x) get_project_colors(x, custom_colors),
    highlight_color = "#f39c12",
    omics_config = omics_config
  )
  
  if (traj_res$status == "success") {
    p_grid <- (traj_res$plots$venn | traj_res$plots$protein_trajectory) / traj_res$plots$pathway_trajectory
    p_grid <- p_grid + plot_annotation(title = sprintf("Sensitivity Analysis: %s", title_str))
    
    if (interactive()) {
      print(p_grid)
    } else {
      # Render plot directly to HTML doc
      print(p_grid) 
      cat("\n\n")
    }
  } else {
    warning("Trajectory failed: ", traj_res$error_msg)
  }
}

if (!interactive()) cat(html_out)
```

# Clinical Table {.tabset}
`r what_this_does("Rendered the HTML clinical statistical tables constructed during the evaluation loop to compare phenotypic outcomes generated by altering the molecular splitting threshold.")`

```{r clincal_table}
for (s_name in split_names) {
  cat(sprintf("\n\n## %s\n\n", gsub("_", " ", s_name)))
  
  tbl_obj <- master_pheno_res[[s_name]]$tables[["SSc_Splits"]]
  
  if (!is.null(tbl_obj)) {
    # Shield HTML Output via raw markdown tags
    cat("```{=html}\n")
    print(gtsummary::as_kable_extra(tbl_obj, format = "html") %>%
            kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE))
    cat("\n```\n\n")
  } else {
    cat("> No clinical table generated for this split.\n\n")
  }
}
```


# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=c(readLines(here::here("code", "01_general_engines.R")), readLines(here::here("code", "02_proteomics_engines.R"))), eval=FALSE}

```

```{r complete}
cat("Script 03d Complete! \n")
```

