---
title: "Non-Active Subgroup Analysis"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

# Introduction
Goal: To mathematically triangulate the molecular phenotype of "Non-Active" SSc patients. Because they exhibit extreme clinical heterogeneity, we hypothesize this group contains distinct molecular sub-populations (Active-like vs Control-like). 

We will derive three independent, continuous scores aligning Non-Active patients along the Control-to-Active molecular axis.

<div style='border: 2px solid #e74c3c; padding: 15px; border-radius: 5px; background-color: #fdf2e9; margin-bottom: 20px;'>
<h4 style='margin-top: 0 !important; color: #c0392b;'>⚠️ NOTE: Medication Confounders</h4>
The computational pipeline currently lacks patient medication histories. If a Non-Active patient presents with a highly "Control-like" molecular signature, consider that this may represent successful pharmacological immunosuppression rather than a naturally benign disease state.
</div>

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Imported the clinical spine, numeric spine, data dictionary, and proteomics matrix.", "Loaded the pre-calculated Active vs Control DEA results to extract the master disease signature.", "Sourced the updated omics engines, including the Smart Correlation Engine.")`

```{r setup_and_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(ComplexHeatmap)
library(circlize)
library(kableExtra)
library(ggpubr)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Paths
current_file <- "03d_proteomics_subgroups"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "00_omics_engines.R"))

# Load Data
master_spine  <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
numeric_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine_numeric.rds"))
data_dict     <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds")) 
olink_mat     <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

# Load DEA Signature (Active vs Control)
avc_dea <- readRDS(here::here("output", "03b_proteomics_differential", "dea_active_vs_control.rds"))
```

# Configuration
```{r configuration}
# 1. Subgroup Definitions
groups <- list(
  active = GRP_ACTIVE,
  control = GRP_CONTROL,
  target = GRP_NON_ACTIVE
)

# 2. Score Generation Configurations
score_conf <- list(
  top_n_proteins = 260, # Number of highly variable/significant proteins to define the axis
  pca_dimensions = 5    # Compute centroid distances across PC1-PC5
)

# 3. Discovery Correlation Thresholds
cor_conf <- list(
  max_rows = 30,
  prot_fdr = 0.05,
  prot_rho = 0.40,
  clin_fdr = 0.10,
  clin_rho = 0.30
)

# 4. Plot Configurations
plot_conf <- list(
  dist_bins = 20,
  hm_font = 8
)
```

# Scores
`r analysis_decision("Signature Definition: We defined the molecular axis using the Top 260 most significant proteins from the Active vs Control DEA. This guarantees all three scores measure specifically where the Non-Active patient falls on the SSc disease spectrum.", "Polarity Alignment: All scores are mathematically aligned so that 100 = Maximally Active-Like, and 0 = Maximally Control-Like.")`

## 5D PCA Centroid Ratio
`r what_this_does("Ran a Principal Component Analysis on the signature proteins.", "Calculated the exact geometric centroid (average coordinate) of the true Active patients and true Control patients across PC1-PC5.", "Calculated the Euclidean distance of each Non-Active patient to both centroids.", "Computed a ratio (Control Distance / Total Distance) scaled 0-100.")`
```{r score_1_pca, fig.width=14, fig.height=8}
# Extract Top 260 Proteins
sig_prots <- avc_dea %>% slice_min(adj.P.Val, n = score_conf$top_n_proteins) %>% pull(Feature)
valid_sig <- intersect(sig_prots, rownames(olink_mat))

# Restrict Matrix and deal with any NAs for PCA
sub_mat <- olink_mat[valid_sig, ]
sub_mat[is.na(sub_mat)] <- 0 
pca_mat <- t(scale(t(sub_mat)))

# Run PCA
pca_res <- prcomp(t(pca_mat), center = FALSE, scale. = FALSE)
pca_df <- as.data.frame(pca_res$x[, 1:score_conf$pca_dimensions]) %>% rownames_to_column("Subject_ID")
pca_df <- pca_df %>% inner_join(master_spine %>% select(Subject_ID, cohort_group), by="Subject_ID")

# Calculate Centroids
cent_active <- pca_df %>% filter(cohort_group == groups$active) %>% select(starts_with("PC")) %>% colMeans()
cent_control <- pca_df %>% filter(cohort_group == groups$control) %>% select(starts_with("PC")) %>% colMeans()

# Calculate Euclidean Distances for Non-Actives
na_pca <- pca_df %>% filter(cohort_group == groups$target)
dist_df <- data.frame(Subject_ID = na_pca$Subject_ID)

dist_df$Dist_Active <- apply(na_pca %>% select(starts_with("PC")), 1, function(x) sqrt(sum((x - cent_active)^2)))
dist_df$Dist_Control <- apply(na_pca %>% select(starts_with("PC")), 1, function(x) sqrt(sum((x - cent_control)^2)))

# Score: Ratio of distances (100 = sits exactly on Active Centroid)
dist_df <- dist_df %>% mutate(PCA_Active_Score = (Dist_Control / (Dist_Active + Dist_Control)) * 100)

# Plots: Scree
var_exp <- (pca_res$sdev^2 / sum(pca_res$sdev^2)) * 100
plot_df_scree <- data.frame(PC = 1:10, Var = var_exp[1:10])

p_scree <- ggplot(plot_df_scree, aes(x=factor(PC), y=Var)) +
  geom_bar(stat="identity", fill="steelblue", color="black") +
  geom_text(aes(label = sprintf("%.1f%%", Var)), vjust = -0.5, size = 5, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + # Expands Y-axis so text fits
  theme_project_base() + 
  labs(title="PCA Scree Plot", x="Principal Component", y="% Variance")

print(p_scree)
cat("\n\n<br>\n\n")

# Plots: PC1-PC5 Drivers
pc_plots <- list()
for(i in 1:5) {
  pc_name <- paste0("PC", i)
  pc_loads <- as.data.frame(pca_res$rotation) %>% select(all_of(pc_name)) %>% 
    arrange(desc(abs(.data[[pc_name]]))) %>% head(10) %>% rownames_to_column("Protein")
  
  pc_plots[[i]] <- ggplot(pc_loads, aes(x=reorder(Protein, .data[[pc_name]]), y=.data[[pc_name]], fill=.data[[pc_name]]>0)) +
    geom_bar(stat="identity", color="black") + coord_flip() + theme_project_base(10) + 
    scale_fill_manual(values = c("TRUE" = "#c0392b", "FALSE" = "#2980b9")) +
    theme(legend.position="none", plot.title = element_text(size = 12)) + 
    labs(title=sprintf("%s Drivers", pc_name), x=NULL, y="Loading")
}

p_load_grid <- wrap_plots(pc_plots, ncol=5)
print(p_load_grid)
```

## Signature Correlation
`r what_this_does("Extracted the log2 Fold-Change vector from the Active vs Control DEA.", "Ran a Spearman correlation between this master signature vector and the raw protein vector of each individual Non-Active patient.", "Normalized the Spearman Rho to a 0-100 scale.")`

```{r signature_correlation_score}
# Extract logFC Signature
logfc_sig <- avc_dea %>% filter(Feature %in% valid_sig) %>% select(Feature, logFC) %>% deframe()

# Calculate Spearman for each Non-Active Patient
cor_scores <- sapply(dist_df$Subject_ID, function(id) {
  pt_vec <- olink_mat[valid_sig, id]
  cor(pt_vec, logfc_sig[names(pt_vec)], method="spearman", use="complete.obs")
})

sig_df <- data.frame(
  Subject_ID = names(cor_scores),
  Raw_Rho = cor_scores
) %>%
  # Min-Max Scale to 0-100
  mutate(Sig_Active_Score = ((Raw_Rho - min(Raw_Rho)) / (max(Raw_Rho) - min(Raw_Rho))) * 100)
```

## Heatmap based Score
`r what_this_does("Generated a complete hierarchical clustering dendrogram of all patients using the signature proteins.", "Extracted the strict left-to-right physical ordering of the leaves.", "Aligned the polarity (ensuring the side where Actives cluster is 100) and assigned continuous percentile scores to the Non-Active patients based on their physical placement.")`

```{r score_3_heatmap, fig.width=10, fig.height=7}
# 1. Build Heatmap Data
hm_clin <- master_spine %>% 
  filter(Subject_ID %in% pca_df$Subject_ID, !is.na(cohort_group)) %>%
  droplevels()

hm_mat <- t(scale(t(olink_mat[valid_sig, hm_clin$Subject_ID])))

# 2. Fetch proper colors using the actual levels
group_colors <- get_project_colors(levels(hm_clin$cohort_group))

# 3. Build Heatmap
hm <- Heatmap(hm_mat, 
              name = "Z-Score",
              show_row_names = FALSE,
              show_column_names = FALSE,
              top_annotation = HeatmapAnnotation(Group = hm_clin$cohort_group, col = list(Group = group_colors)),
              clustering_distance_columns = "euclidean",
              clustering_method_columns = "complete")

# 4. Draw directly to the HTML
hm_drawn <- draw(hm)

# Extract order
order_idx <- column_order(hm_drawn)
ordered_ids <- colnames(hm_mat)[order_idx]

# 5. Map Order to 0-100 Percentile
hm_df <- data.frame(
  Subject_ID = ordered_ids,
  HM_Rank = 1:length(ordered_ids)
) %>%
  mutate(HM_Percentile = (HM_Rank / max(HM_Rank)) * 100) %>%
  inner_join(hm_clin %>% select(Subject_ID, cohort_group), by="Subject_ID")

# Check Polarity (Are Actives on the Left (0) or Right (100)?)
mean_active <- mean(hm_df$HM_Percentile[hm_df$cohort_group == groups$active])
mean_control <- mean(hm_df$HM_Percentile[hm_df$cohort_group == groups$control])

if (mean_active < mean_control) {
  hm_df$HM_Active_Score <- 100 - hm_df$HM_Percentile
} else {
  hm_df$HM_Active_Score <- hm_df$HM_Percentile
}

hm_final <- hm_df %>% filter(cohort_group == groups$target) %>% select(Subject_ID, HM_Active_Score)
```

## Score Characterization
`r what_this_does("Merged the three scores into a master Non-Active index.", "Plotted density histograms to visually evaluate if the Non-Active cohort is continuous (bell curve) or comprised of two distinct subgroups (bimodal).", "Plotted the pairwise correlations between the three independent mathematical approaches.")`

## Score Characterization
`r what_this_does("Merged the three scores into a master Non-Active index.", "Plotted density histograms to visually evaluate if the Non-Active cohort is continuous (bell curve) or comprised of two distinct subgroups (bimodal).", "Rendered a standalone overlaid density plot to compare the distributions of the three scoring methods.", "Plotted the pairwise correlations between the three independent mathematical approaches.")`

```{r score_characterization, fig.width=14, fig.height=6}
# Merge Master Score Dataframe
score_master <- dist_df %>% select(Subject_ID, PCA_Active_Score) %>%
  inner_join(sig_df %>% select(Subject_ID, Sig_Active_Score), by="Subject_ID") %>%
  inner_join(hm_final, by="Subject_ID")

# 1. Density Distributions (Faceted)
score_long <- score_master %>% pivot_longer(-Subject_ID, names_to="Score_Type", values_to="Score")

p_dens_facet <- ggplot(score_long, aes(x=Score, fill=Score_Type)) +
  geom_density(alpha=0.6) +
  facet_wrap(~Score_Type, scales="free_y") +
  theme_project_base() + theme(legend.position="none") +
  labs(title="Faceted Score Distributions", x="Score (100 = Active-Like, 0 = Control-Like)", y="Density")

print(p_dens_facet)
cat("\n\n<br>\n\n")

# 2. Density Distributions (Overlaid Standalone)
p_dens_overlay <- ggplot(score_long, aes(x=Score, fill=Score_Type, color=Score_Type)) +
  geom_density(alpha=0.4, linewidth=1) +
  theme_project_base() + theme(legend.position="bottom", legend.title=element_blank()) +
  labs(title="Overlaid Score Distributions", x="Score (100 = Active-Like, 0 = Control-Like)", y="Density") +
  theme(plot.margin = margin(t = 10, r = 150, b = 10, l = 150, unit = "pt"))
 

print(p_dens_overlay)
cat("\n\n<br>\n\n")

# 3. Pairwise Correlations with ggpubr Stats
p1 <- ggplot(score_master, aes(x=PCA_Active_Score, y=Sig_Active_Score)) + 
  geom_point(alpha=0.7) + geom_smooth(method="lm", color="red") + 
  stat_cor(label.y.npc = "top") + stat_regline_equation(label.y.npc = 0.9) +
  theme_project_base()

p2 <- ggplot(score_master, aes(x=PCA_Active_Score, y=HM_Active_Score)) + 
  geom_point(alpha=0.7) + geom_smooth(method="lm", color="red") + 
  stat_cor(label.y.npc = "top") + stat_regline_equation(label.y.npc = 0.9) +
  theme_project_base()

p3 <- ggplot(score_master, aes(x=Sig_Active_Score, y=HM_Active_Score)) + 
  geom_point(alpha=0.7) + geom_smooth(method="lm", color="red") + 
  stat_cor(label.y.npc = "top") + stat_regline_equation(label.y.npc = 0.9) +
  theme_project_base()

p_corrs <- (p1 | p2 | p3) + plot_annotation(title="Pairwise Agreement of Mathematical Scoring Methods")
print(p_corrs)
```

# Subgrouping
## Subgroup Drivers {.tabset}
`r what_this_does("Deployed the Smart Correlation Engine.", "Correlated all 3,000 proteins against the three scores to find the exact biological drivers pulling Non-Actives toward the Active phenotype.", "Mapped the numeric clinical spine (using Data Type logic) to find the physiological markers of 'Active-like' Non-Active patients.")`

```{r subgroup_drivers, results='asis', warning=FALSE}
# Prepare Matrices (Non-Actives Only)
na_ids <- score_master$Subject_ID
prot_mat_na <- olink_mat[, na_ids]

clin_df_na <- numeric_spine %>% filter(Subject_ID %in% na_ids)
clin_mat_na <- t(clin_df_na %>% column_to_rownames("Subject_ID") %>% select(where(is.numeric)))

score_targets <- c("PCA_Active_Score", "Sig_Active_Score", "HM_Active_Score")
modalities <- list(proteomics = prot_mat_na, clinical = clin_mat_na)

html_out <- ""

for (mod_name in names(modalities)) {
  if (!interactive()) html_out <- paste0(html_out, sprintf("\n\n## %s {.tabset}\n\n", tools::toTitleCase(mod_name)))
  
  for (score_name in score_targets) {
    clean_score <- gsub("_", " ", score_name)
    if (!interactive()) html_out <- paste0(html_out, sprintf("\n\n### %s\n\n", clean_score))
    
    # Calculate using Smart Engine
    score_vec <- setNames(score_master[[score_name]], score_master$Subject_ID)
    res_df <- run_smart_mass_cor(modalities[[mod_name]], score_vec, dict_df = data_dict)
    
    # Export full dataset
    export_name <- sprintf("Drivers_%s_vs_%s.csv", tools::toTitleCase(mod_name), score_name)
    write_csv(res_df, file.path(output_dir_data, export_name))
    
    # Filter
    thresh_fdr <- if(mod_name == "proteomics") cor_conf$prot_fdr else cor_conf$clin_fdr
    thresh_rho <- if(mod_name == "proteomics") cor_conf$prot_rho else cor_conf$clin_rho
    
    sig_df <- res_df %>% filter(FDR < thresh_fdr & abs(Effect) >= thresh_rho)
    
    # Comprehensive Descriptive Header
    caption <- sprintf("
<div style='background-color:#f9f9f9; padding: 10px; border-left: 4px solid #2980b9; margin-bottom: 15px;'>
  <b>Statistical Overview</b><br>
  <i>Cohort Size:</i> %d Non-Active patients<br>
  <i>Thresholds:</i> FDR < %s, |Effect| >= %s<br>
  <i>Note:</i> 'Effect' represents Spearman's Rho for continuous variables, and Wilcoxon Effect Size (r) for binary categoricals. Displaying top %d.
</div>", length(na_ids), thresh_fdr, thresh_rho, min(nrow(sig_df), cor_conf$max_rows))
    
    if (nrow(sig_df) > 0) {
      tbl_html <- sig_df %>% head(cor_conf$max_rows) %>%
        kbl(format = "html", digits = 3, align = "c") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
        scroll_box(height = "500px")
      
      if (interactive()) {
        cat(sprintf("\n=== %s vs %s ===\n", tools::toTitleCase(mod_name), clean_score))
        cat(caption, "\n")
        print(tbl_html)
      } else {
        html_out <- paste0(html_out, caption, "\n", tbl_html, "\n\n")
      }
    } else {
      empty_msg <- "> No features passed the significance thresholds.\n\n"
      if (interactive()) {
        cat(sprintf("\n=== %s vs %s ===\n", tools::toTitleCase(mod_name), clean_score))
        cat(caption, "\n", empty_msg)
      } else {
        html_out <- paste0(html_out, caption, "\n", empty_msg)
      }
    }
  }
}

if (!interactive()) cat(html_out)

# Cleanup RAM
rm(prot_mat_na, clin_mat_na, clin_df_na, na_ids, html_out)
```

```{r complete}
cat("Script 03d Complete! \n")
```

