---
title: "Lipidomics Topography & Missingness"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import

`r what_this_does("Loaded required bioinformatics and visualization packages (UpSetR for missingness intersections).", "Imported the pre-cleaned clinical spine , imputed and sparse lipidomics matrix.", "Sourced the new lipid_engines.R file designed specifically for sparse mass spectrometry data.")`
```{r setup_and_import, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(kableExtra)
library(UpSetR) # Highly recommended for missingness topography
library(readxl)      # Reading messy Excel files
library(data.table)
library(UpSetR)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "04a_lipidomics_topography"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines (Assuming we start the new file here!)
source(here::here("code", "03_lipidomics_engines.R"))

# Load Pre-Cleaned Data
# Assuming 01c or similar contains your raw/imputed lipid data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
lipid_mat_imputed <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_lipid_matrix.rds"))
lipid_mat_unimputed <- readRDS(here::here("output", "01a_data_import_and_qc", "raw_lipid_matrix.rds"))
clin_dict <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
lod_df <- read_excel(here::here("data", "Lipidomics_LOD.xlsx"), sheet = "Master")
```

# Configurations
`r what_this_does("Defined thresholds for categorizing lipids based on missingness to dynamically route them into different statistical pipelines (DEA vs. Binarization).", "Configured visualization parameters for the missingness overview and topographical histograms.")`

```{r conf}
# 1. Nested Visualization & Analysis Configuration
topography_config <- list(
  # Base Missingness Settings
  missingness = list(
    hist_bins = 20,
    log_transform_viz = TRUE
  ),
  
  # Phenotype Evaluation Settings (Fully Parameterized)
  phenotype = list(
    vars_to_test = c("cohort_group", "Sex", "Age"), # Add or remove spine columns here!
    continuous_bins = 3                             # e.g., 3 = Tertiles for numeric vars
  ),
  
  # Global Export Settings
  export = list(
    width = 10,
    height = 6,
    dpi = 300
  )
)
```

# Topography
`r what_this_does("Evaluated global matrix missingness and generated corresponding topographies (Barplots, Heatmaps, LOD scatter).", "Exported generated plots to the output directory to preserve visual artifacts.")`
```{r topography, fig.width=12, fig.height=10, results='asis'}
# 1. Execute Wrapper 
missingness_res <- wrap_missingness_topography(
  lipid_mat = lipid_mat_unimputed,
  config = topography_config,
  output_dir = output_dir_data,
  lod_df = lod_df
)

# 2. Print Diagnostics and Plots natively
cat("### Global Matrix Diagnostics\n\n")
cat(sprintf("* **Total Lipids Evaluated:** %d\n\n", missingness_res$diagnostics$total_lipids))

cat("### Global Missingness Distribution\n\n")
print(missingness_res$plots$missingness_bars)
cat("\n\n")

if(!is.null(missingness_res$plots$co_missingness_heatmap)) {
  cat("### Missingness Intersections (Heatmap)\n\n")
  cat("*Note: Co-Missingness evaluates how often lipid 'i' AND lipid 'j' are missing in the same patient.*\n\n")
  print(missingness_res$plots$co_missingness_heatmap)
  cat("\n\n")
}

if(!is.null(missingness_res$plots$lod_scatter)) {
  cat("### Limit of Detection (LOD) Topography\n\n")
  print(missingness_res$plots$lod_scatter)
  cat("\n\n")
}

# 3. RAM Cleanup
#rm(missingness_res)
```
Comissingess: How many times are lipid 'i' AND lipid 'j' missing in the same patient?
The diagonal displays the overall missingness of that lipid.

# Distribution Signatures
`r what_this_does("Generated and exported topographical histograms for all strictly measured lipids.", "Flagged ultra-sparse lipids (n=1) with text annotations.")`
`r analysis_decision("Zeros and NAs were explicitly excluded here to accurately evaluate the underlying continuous variance of successfully measured samples, necessary to assess structural validity for continuous models.")`
```{r execution_engine_histograms, fig.width=14, fig.height=10, results='asis'}
# 1. Execute Wrapper
hist_res <- wrap_distribution_signatures(
  lipid_mat = lipid_mat_unimputed,
  config = topography_config,
  output_dir = output_dir_data,
  project_colors_func = get_project_colors,
  lod_df = lod_df
)

# 2. Print Output natively
cat("\n\n*Note: Zeros/NAs are excluded here to evaluate the underlying continuous distribution of successfully measured samples.*\n\n")
print(hist_res$plots$hists)
cat("\n\n")

# 3. RAM Cleanup
rm(hist_res)
```
# Clinical Phenotype Missingness {.tabset}
`r what_this_does("Evaluated missingness percentages dynamically partitioned by parameterized clinical variables.", "Exported generated bar charts dynamically.")`
`r analysis_decision("Continuous clinical variables were dynamically identified via the data dictionary and binned into quantiles to assess categorical biological dropout trends, justifying Fisher's Exact binarization.")`
```{r execution_engine_phenotype, fig.width=15, fig.height=6, results='asis'}
# 1. Execute Wrapper
pheno_res <- wrap_phenotype_completeness(
  lipid_mat = lipid_mat_unimputed,
  clin_df = master_spine,
  clin_dict = clin_dict,
  config = topography_config,
  output_dir = output_dir_data
)

# 2. Dynamically Print Tabset Headers and Plots
if(length(pheno_res$plots) == 0) {
  cat("**WARNING:** No clinical variables successfully mapped. Check configuration and spine.\n\n")
} else {
  for (p_name in names(pheno_res$plots)) {
    cat(sprintf("\n\n## %s\n\n", p_name))
    print(pheno_res$plots[[p_name]])
    cat("\n\n")
  }
}

# 3. RAM Cleanup
rm(pheno_res)
```

# Lipid Groups
`r what_this_does("Creates an UpSet Plot of the different lipid groupings")`
```{r lipid_groups, fig.width=8, fig.height=10}
# 1. Define the columns we want to evaluate
cols_to_evaluate <- c("Class", "Precursor", "Function")

# 2. Clean and isolate the relevant data, keeping only UNIQUE mediators
df_clean <- lod_df %>%
  dplyr::select(Mediator, dplyr::all_of(cols_to_evaluate)) %>%
  dplyr::distinct(Mediator, .keep_all = TRUE) %>% # <-- NEW: Drop technical replicates
  tidyr::drop_na() # Remove any lipids missing this categorical data

# 3. Pivot LONG to aggregate all categories, then create a unique Set Name
df_long <- df_clean %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(cols_to_evaluate),
    names_to = "Category_Type",
    values_to = "Category_Value"
  ) %>%
  dplyr::mutate(
    Set_Name = paste0(Category_Type, "_", Category_Value),
    Presence = 1 
  )

# 4. Pivot WIDE to create the binary matrix UpSetR requires
df_wide <- df_long %>%
  dplyr::select(Mediator, Set_Name, Presence) %>%
  tidyr::pivot_wider(
    names_from = Set_Name,
    values_from = Presence,
    values_fill = 0 # <-- FIX: Simplified to scalar 0 to prevent type-conversion errors
  )

# 5. Extract the dynamically generated set names
# We exclude the 'Mediator' column, as UpSetR only wants the binary set columns
set_names <- setdiff(colnames(df_wide), "Mediator")

# 6. Generate the UpSet Plot
# Note: UpSetR strictly requires a standard data.frame, not a tibble
UpSetR::upset(
  as.data.frame(df_wide),
  sets = set_names,
  order.by = "freq",               # Order intersections by size
  nsets = length(set_names),       # Dynamically include all our dummy columns
  nintersects = 30,                # Show the top 30 combinations
  mainbar.y.label = "Lipid Count",
  sets.x.label = "Total Lipids per Category",
  text.scale = c(1.2, 1.2, 1, 1, 1.2, 1) # Adjust text sizing for readability
)

rm(df_clean, df_long, df_wide)
```


# Appendix
`r what_this_does("Appended the raw underlying R code from the custom lipid omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine,code=readLines(here::here("code", "03_lipidomics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 04a Complete! \n")
```

