---
title: "Sparse Lipidomics: Differential Expression & LSEA Prep"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded statistical packages including logistf for Firths penalized regression.", "Imported the pre-cleaned clinical spine, clinical data dictionary, and sparse lipidomics matrix.", "Sourced custom lipid omics engines.")`

```{r setup_and_import}
library(tidyverse)
library(patchwork)
library(here)
library(logistf)    # Firth's logistic regression
library(gtsummary)  # Clinical tables
library(kableExtra) # HTML tables

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "04b_lipidomics_differential"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Configuration & Engines
source(here::here("code", "00_lipid_engines.R"))

# Load Pre-Cleaned Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
data_dictionary <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
lipid_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "raw_lipid_matrix.rds"))

# Dynamic Feature Count
total_features <- nrow(lipid_mat)
```

# Configurations
`r what_this_does("Configured the Sparsity Triage thresholds and parameterized the Differential Expression routing logic, including the continuous binning toggle.")`

```{r configurations}
# 1. Sparsity Triage Configs ---------------------------
sparsity_configs <- list(
  global_triage = list(
    min_detect_rate = 0.10, 
    binarize_data = TRUE    # Convert to 1/0 (Present/Missing)
  )
)

# 2. DEA Routing Configs ---------------------------
# Note: Engine choice is automatic based on the clinical_data_dictionary Class.
dea_configs <- list(
  job_active_vs_nonactive = list(
    title = "ACtive vs. Non-Active",
    test_var = "cohort_group",
    target_groups = c("Active"),
    ref_groups = c("Non-Active"),
    bin_continuous = FALSE,  # N/A for categorical, but safely ignored
    p_cutoff = 0.05
  ),
  
  job_age_continuous = list(
    title = "Age (Firth's)",
    test_var = "Age",
    bin_continuous = FALSE,  # Routes to Firth's Penalized Regression
    p_cutoff = 0.05
  ),
  
  job_age_binned = list(
    title = "Age (Fisher's)",
    test_var = "Age",
    bin_continuous = TRUE,   # Bins at median, routes to Fisher's
    p_cutoff = 0.05
  )
)

# 3. Plot Configurations ---------------------------
plot_configs <- list(
  volcano = list(
    point_size = 2,
    alpha = 0.8,
    color_sig = "firebrick",
    color_ns = "grey70"
  )
)
```

# Sparsity Triage & Binarization
`r what_this_does("Evaluates global missingness across the matrix. Binarizes the data into Presence/Absence and quarantines ultra-sparse features to save statistical power, reserving them for downstream family-level aggregation.")`

```{r}
# Triage Math ---------------------------
conf <- sparsity_configs$global_triage

# 1. Binarize Matrix
if(conf$binarize_data) {
  binary_mat <- ifelse(is.na(lipid_mat) | lipid_mat == 0, 0, 1)
} else {
  binary_mat <- lipid_mat
}

# 2. Calculate global detection rates
detection_rates <- rowMeans(binary_mat, na.rm = TRUE)

# 3. Split features
testable_lipids <- names(detection_rates[detection_rates >= conf$min_detect_rate])
ultra_sparse_lipids <- names(detection_rates[detection_rates < conf$min_detect_rate])

# 4. Create final working matrix
mat_testable <- binary_mat[testable_lipids, , drop = FALSE]

# 5. Export ultra-sparse list for later Lipid Set Enrichment (LSEA)
saveRDS(ultra_sparse_lipids, file.path(output_dir_data, "ultra_sparse_lipids_for_LSEA.rds"))

# Cleanup ---------------------------
rm(detection_rates, conf)
```

# Differential Routing Engine {.tabset}
`r what_this_does("Dynamically queries the Data Dictionary to determine metadata classes. Routes categorical/binned variables to Fisher's Exact Test and continuous variables to Firth's Penalized Regression. Compiles unified Volcano Plots and saves independent stat modules.")`

```{r Differential}
# Global storage for LSEA ranking
ranked_results_list <- list()

for (job_id in names(dea_configs)) {
  job <- dea_configs[[job_id]]
  
  cat(sprintf("\n\n## %s\n\n", job$title))
  
  # 1. Dictionary Lookup ---------------------------
  var_class <- data_dictionary %>% 
    filter(Variable == job$test_var) %>% 
    pull(Class) %>% 
    tolower()
    
  if (length(var_class) == 0) var_class <- "unknown"
  
  # 2. Pre-processing & Engine Assignment ---------------------------
  temp_spine <- master_spine %>% filter(!is.na(.data[[job$test_var]]))
  target_engine <- "fisher"
  
  if (var_class %in% c("numeric", "integer", "double", "continuous")) {
    if (isTRUE(job$bin_continuous)) {
      # Binning Continuous -> Fisher
      median_val <- median(temp_spine[[job$test_var]], na.rm = TRUE)
      temp_spine <- temp_spine %>%
        mutate(!!job$test_var := ifelse(.data[[job$test_var]] > median_val, "High", "Low"))
      target_engine <- "fisher"
      cat(sprintf("*Note: %s was binned at median (%.2f) and routed to Fisher's Exact Test.*\n\n", job$test_var, median_val))
      
    } else {
      # Raw Continuous -> Firth
      target_engine <- "firth"
      cat(sprintf("*Note: %s is continuous and was routed to Firth's Penalized Logistic Regression.*\n\n", job$test_var))
    }
  } else {
      cat(sprintf("*Note: %s is categorical/binary and was routed to Fisher's Exact Test.*\n\n", job$test_var))
  }
  
  # 3. Execute Statistical Wrapper ---------------------------
  # (Assuming wrap_sparse_dea calls your worker_run_fishers_exact / worker_run_firth_logistic)
  res <- wrap_sparse_dea(
    mat = mat_testable, 
    clin_df = temp_spine, 
    config = job, 
    engine = target_engine,
    job_name = job_id,
    out_dir = output_dir_data
  )
  
  if (res$status == "success") {
    
    # Store for LSEA 
    ranked_results_list[[job_id]] <- res$tables$full_results
    saveRDS(res$tables$full_results, file.path(output_dir_data, sprintf("%s_full_stats.rds", job_id)))
    
    # Render Output Table
    if (nrow(res$tables$significant_features) > 0) {
      tbl <- head(res$tables$significant_features, 15) %>% 
        mutate(
          Log2OR = round(Log2OR, 2), 
          P_Value = formatC(P_Value, format = "e", digits = 2),
          FDR = formatC(FDR, format = "e", digits = 2)
        ) %>%
        kableExtra::kbl(format = "html", caption = sprintf("Top Significant Lipids: %s", job$title)) %>% 
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
      print(tbl)
      cat("\n\n") 
    } else {
      cat("*No significant features survived FDR correction.* \n\n")
    }
    
    # Render Volcano Plot (Dots)
    if (!is.null(res$plots$volcano)) {
      # Standard dot-based volcano plot returned from the worker
      print(res$plots$volcano)
      cat("\n\n") 
    }
    
  } else {
     cat(sprintf("**FAILED:** %s\n\n", res$error_msg))
  }
  
  # Cleanup temp spine to save memory
  rm(temp_spine, var_class, target_engine, res)
}
```

# Export 
`r what_this_does("Aggregates all continuous and categorical statistical outputs, combining them with the quarantined ultra-sparse features, and exports a unified ranked database for structural family-level enrichment analysis.")`

```{r export}
# LSEA Export Math ---------------------------
# Compile the full ranked lists into a single payload
lsea_payload <- list(
  testable_stats = ranked_results_list,
  ultra_sparse_background = ultra_sparse_lipids
)

# Save the unified payload
export_path <- file.path(output_dir_data, "LSEA_unified_payload.rds")
saveRDS(lsea_payload, export_path)

# Cleanup ---------------------------
rm(lsea_payload, ranked_results_list)
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom lipid omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 04a Complete! \n")
```


