---
title: "PLS-DA & VIP Biomarker Discovery"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

```{r init, include=FALSE}
rm(list = ls())
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded mixOmics for multivariate statistical modeling.", "Imported the pre-cleaned clinical spine and lipidomics matrix.", "Sourced custom lipid omics engines.")`

```{r setup_and_import, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(ropls)   # Core PLS-DA engine
library(kableExtra) # HTML tables

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "04d_lipidomics_plsda"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Asset paths (Public/GitHub accessible)
assets_dir <- here::here("docs", "assets", current_file)
if (!dir.exists(assets_dir)) dir.create(assets_dir, recursive = TRUE)

# Source Configuration & Engines
source(here::here("code", "03_lipidomics_engines.R"))

# Load Pre-Cleaned Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
lipid_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "raw_lipid_matrix.rds"))

# Dynamic Feature Count
total_features <- nrow(lipid_mat)
```


# Configurations
`r what_this_does("Configured the strict Sparsity Triage threshold (50%) to prevent modeling heavily fabricated data.", "Parameterized the PLS-DA contrast jobs.")`

```{r configurations}
# 1. Sparsity Triage Configs ---------------------------
# PLS-DA requires stricter missingness thresholds than standard DEA
sparsity_configs <- list(
  global_triage = list(
    min_detect_rate = 0.25  # Must be measured in at least 25% of the cohort
  )
)

# 2. PLS-DA Routing Configs ---------------------------
plsda_configs <- list(
  job_active_vs_nonactive = list(
    title = "Active vs. Non-Active",
    test_var = "cohort_group",
    target_groups = c("Active"),
    ref_groups = c("Non-Active"),
    bin_continuous = FALSE,
    vip_top_n = 20
  ),
  
  job_ssc_vs_ctrl = list(
    title = "SSc vs. Control",
    test_var = "cohort_group",
    target_groups = c("Active", "Non-Active"),
    ref_groups = c("Control"),
    bin_continuous = FALSE,
    vip_top_n = 20
  )
)
```

# Sparsity Triage & Global Imputation

`r what_this_does("Evaluates global missingness and filters out ultra-sparse features.", "Imputes the remaining gaps by substituting missing values with 1/5 of the lowest positive measurement observed for that specific lipid across the entire cohort.", "Outputs a detailed data manipulation audit trail.")`

```{r sparsity_triage_and_global_imputation, results='asis'}
# 1. Triage Math ---------------------------
conf <- sparsity_configs$global_triage

binary_mat <- ifelse(is.na(lipid_mat) | lipid_mat == 0, 0, 1)
detection_rates <- rowMeans(binary_mat, na.rm = TRUE)

testable_lipids <- names(detection_rates[detection_rates >= conf$min_detect_rate])
removed_lipids <- names(detection_rates[detection_rates < conf$min_detect_rate])
mat_triage <- lipid_mat[testable_lipids, , drop = FALSE]

# 2. Imputation Engine ---------------------------
imputation_res <- worker_impute_min_fifth(mat_triage)
mat_imputed <- imputation_res$mat

cat(c(sprintf("### Triage & Imputation Audit {.unlisted .unnumbered}\n\n"), 
    sprintf("* **Starting Lipids:** %d\n", nrow(lipid_mat)),
    sprintf("* **Removed by Triage (Failed %.0f%% threshold):** %d\n", conf$min_detect_rate * 100, length(removed_lipids)),
    sprintf("* **Surviving Lipids:** %d\n", nrow(mat_triage)),
    sprintf("* **Data Points Imputed:** %d (Using 1/5th MNAR minimum rule)\n\n", imputation_res$imputed_count)))

# Save Imputed Matrix for downstream use
saveRDS(mat_imputed, file.path(output_dir_data, "imputed_lipid_matrix.rds"))

# Cleanup ---------------------------
rm(binary_mat, detection_rates, conf, mat_triage, imputation_res, removed_lipids)
invisible(gc())
```

# Discriminant Analysis {.tabset}
`r what_this_does("Performs sub-setting and data wrangling for the specific contrast.", "Calculates variance to dynamically exclude constant features.", "Executes auto-scaled Partial Least Squares Discriminant Analysis (PLS-DA).", "Extracts and visualizes Variable Importance in Projection (VIP) scores to identify primary driving biomarkers.")`

```{r discriminant_analysis}
# Global storage for VIP lists
vip_results_list <- list()

for (job_id in names(plsda_configs)) {
  job <- plsda_configs[[job_id]]
  
  cat(sprintf("\n\n## %s\n\n", job$title))
  
  # Continuous variable handling (binning for classification model)
  temp_spine <- master_spine %>% filter(!is.na(.data[[job$test_var]]))
  if (isTRUE(job$bin_continuous)) {
    median_val <- median(temp_spine[[job$test_var]], na.rm = TRUE)
    temp_spine <- temp_spine %>% mutate(!!job$test_var := ifelse(.data[[job$test_var]] > median_val, "High", "Low"))
    cat(sprintf("*Note: Continuous variable %s was binned at median (%.2f) to permit classification modeling.*\n\n", job$test_var, median_val))
  }
  
  # Execute Wrapper
  res <- wrap_plsda_pipeline(
    mat = mat_imputed, 
    clin_df = temp_spine, 
    config = job, 
    job_name = job_id,
    out_dir = output_dir_data,
    assets_dir = assets_dir,
    project_colors_func = get_project_colors
  )
  
  # Print the Data Wrangling Audit Trail
  if (!is.null(res$audit)) {
    cat(c(sprintf("### Data Wrangling Audit {.unlisted .unnumbered}\n\n"),
          sprintf("* **Starting cohort:** %d samples\n", res$audit$start_samples),
          sprintf("* **Filtered to target groups:** %d samples\n", res$audit$filtered_samples),
          sprintf("* **Dropped due to missing clinical labels:** %d samples\n", res$audit$dropped_na_labels),
          sprintf("* **Variance Check:** %d lipids had zero variance in this specific subset and were excluded.\n", res$audit$zero_var_dropped),
          sprintf("* **Final input to PLS-DA:** %d samples x %d lipids\n\n", res$audit$final_samples, res$audit$final_lipids)))
  }
  
  if (res$status == "success") {
    
    vip_results_list[[job_id]] <- res$tables$vip
    saveRDS(res$tables$vip, file.path(output_dir_data, sprintf("%s_vip_stats.rds", job_id)))
    
    # Render Plots Individually
    cat(sprintf("### PLS-DA Scores: %s {.unlisted .unnumbered}\n\n", job$title))
    if (interactive() && !isTRUE(getOption('knitr.in.progress'))) { 
      print(res$plots$scores); cat("\n\n")
    } else { 
      print(res$plots$scores)
      cat("\n\n") 
    }
    
    cat(sprintf("### VIP Biomarkers Plot: %s {.unlisted .unnumbered}\n\n", job$title))
    if (interactive() && !isTRUE(getOption('knitr.in.progress'))) { 
      print(res$plots$vip); cat("\n\n")
    } else { 
      print(res$plots$vip)
      cat("\n\n") 
    }
    
    # Render Table
    cat(sprintf("### Top VIP Biomarkers Table {.unlisted .unnumbered}\n\n"))
    
    tbl <- head(res$tables$vip, 15) %>% 
      mutate(VIP = round(VIP, 3)) %>%
      kableExtra::kbl(format = "html", caption = sprintf("Top 15 VIP Features: %s", job$title)) %>% 
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
    
    print(tbl)
    cat("\n\n")
    
    # Safely clear the plotting objects now that they are inside the success loop
    rm(tbl)
    
  } else {
     cat(sprintf("**FAILED:** %s\n\n", res$error_msg))
  }
  
  rm(temp_spine, res)
}
```

# Appendix

`r what_this_does("Appended the raw underlying R code from the custom lipid omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=readLines(here::here("code", "03_lipidomics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script 04d Complete! \n")
```
