---
title: "Target Investigator"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded bioinformatics and graphics packages, including httr for API requests.",  "Imported the Olink proteomics, binary lipidomics, and clinical matrices, and sourced the required computational engines.")`

```{r setup_and_import, error=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(kableExtra)
library(ggraph)
library(igraph)
library(ggpubr)
library(grid)      # Required for ggraph arrows
library(httr)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "05a_investigator"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Legacy Engines
source(here::here("code", "01_general_engines.R"))
source(here::here("code", "02_proteomics_engines.R")) # really required??
source(here::here("code", "05_multiomics_engines.R"))

# Load Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
numeric_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine_numeric.rds"))
clin_dict <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
omics_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))
lipid_mat <- readRDS(here::here("output", "04b_lipidomics_differential", "lipid_binary_matrix.rds"))

# Load External Grouping Sources (Example: Clustering output)
# Note: Ensure this dataframe has a 'Subject_ID' column for joining
clustering_res <- readRDS(here::here("output", "03c_proteomics_clustering", "Clustering", "job_svc","job_svc_Master_Assignments.rds"))

```

# Omnipath Import
`r what_this_does("Imports latest OmniPath curated dataset", "If not available uses the last downloaded version")`
```{r omnipath_import}
# OmniPath Caching System
omnipath_cache_path <- here::here("output", "omnipath_full_db_cache.tsv")
omnipath_url <- "https://omnipathdb.org/interactions?datasets=omnipath&genesymbols=1"

api_res <- tryCatch(httr::GET(omnipath_url, httr::timeout(30)), error = function(e) NULL)

if (!is.null(api_res) && httr::status_code(api_res) == 200) {
  # Server is UP: Download, load into memory, and save a fresh cache
  raw_text <- httr::content(api_res, as = "text", encoding = "UTF-8")
  omnipath_full_db <- readr::read_tsv(raw_text, show_col_types = FALSE)
  readr::write_tsv(omnipath_full_db, omnipath_cache_path)
  message("OmniPath API successful: Downloaded and cached fresh database.")
} else {
  # Server is DOWN: Fallback to local cache
  if (file.exists(omnipath_cache_path)) {
    omnipath_full_db <- readr::read_tsv(omnipath_cache_path, show_col_types = FALSE)
    message("OmniPath API unavailable: Successfully loaded from local cache.")
  } else {
    omnipath_full_db <- NULL
    message("OmniPath API unavailable and no local cache found. Network graphs will be skipped.")
  }
}
```


# Analysis Configurator
`r what_this_does("Defined the investigative parameters, including sub-cohort subsetting, external groupings for visual highlights, and biological network thresholds.")`

```{r analysis_configurator}
# 1. Ignore me

# 2. Global Investigator Configurations
investigator_config <- list(
  cor_fdr_cutoff = 0.05,
  plot_theme = theme_minimal()
)

# Common highlight list
highlights <- list(
      list(var = "Age", source = "master_spine"),
      list(var = "Sex", source = "master_spine"),
      list(var = "Sample_Age", source = "master_spine"),
      list(var = "cohort_group", source = "master_spine"), 
      list(var = "kmeans_Cluster", source = "clustering_res")
      )

# 3. Job Definitions
investigation_jobs <- list(

  job_casp4 = list(
    title = "Investigate CASP4",
    targets = c("CASP4"),
    highlight_vars = highlights,
    clin_test_list = ssc_clinical_parameters,
    residualize_conf = NULL
  ),
  
  job_ido1 = list(
    title = "Investigate IDO1",
    targets = c("IDO1"),
    highlight_vars = highlights,
    clin_test_list = ssc_clinical_parameters,
    residualize_conf = NULL
  )
)

# Configuration for distributions plots
plot_configs <- list(
  violin = list(alpha = 0.7, width = 0.8),
  boxplot = list(width = 0.2, alpha = 0.9, outlier.shape = NA)
)
```

# Target Investigation {.tabset .ignore-toc}
`r what_this_does("Iterated through the defined targets to compute baseline distributions, categorical highlights, and cross-domain correlation statistics (Proteomics, Lipidomics, Clinical).", " Automatically aligned and subsetted data matrices based on configuration, and prepared tables displaying both statistically significant hits and top non-significant runners-up.")`

```{r target_investigator, results='asis'}
# Global list to store data intended for the Network Chunk
network_queue <- list()
master_investigation_results <- list()

for (j_name in names(investigation_jobs)) {
  conf <- investigation_jobs[[j_name]]
  cat(sprintf("\n\n## %s \n\n", conf$title))
  
  # 1. Base Matrices
  w_olink <- omics_mat
  w_lipid <- lipid_mat
  w_spine <- master_spine
  
  # 2. Apply Subsetting Filter (if configured)
  if (!is.null(conf$subset_filter)) {
    filter_var <- conf$subset_filter$var
    filter_keep <- conf$subset_filter$keep
    if(filter_var %in% colnames(w_spine)) {
      w_spine <- w_spine %>% filter(.data[[filter_var]] == filter_keep)
      cat(sprintf("> *Subsetting Applied: Retaining only '%s' == '%s' (N = %d remaining).*\n\n", filter_var, filter_keep, nrow(w_spine)))
    }
  }
  
  valid_subjects <- w_spine$Subject_ID
  
  # 3. Align Matrices (FIXED INTERSECTION LOGIC)
  w_clin_temp <- numeric_spine %>% select(Subject_ID, any_of(conf$clin_test_list)) %>% column_to_rownames("Subject_ID") %>% as.matrix() %>% t()
  
  common_subjects <- Reduce(intersect, list(colnames(w_olink), colnames(w_lipid), colnames(w_clin_temp), valid_subjects))
  
  w_olink <- w_olink[, common_subjects, drop = FALSE]
  w_lipid <- w_lipid[, common_subjects, drop = FALSE]
  w_clin  <- w_clin_temp[, common_subjects, drop = FALSE]
  
  cat(sprintf("> *Diagnostic: Matrices aligned. %d patients common across Olink, Lipids, and Clinical data.*\n\n", length(common_subjects)))
  
  all_valid_targets <- intersect(conf$targets, c(rownames(w_olink), rownames(w_lipid)))
  if (length(all_valid_targets) == 0) { 
    cat(sprintf("> <span style='color:red;'>**SKIPPED:** None of the targets found.</span>\n\n"))
    next 
  }
  
  for (target in all_valid_targets) {
    
    cat(sprintf("\n\n### Target: %s {.toc-ignore}\n\n", target))
    
    if (target %in% rownames(w_olink)) {
      t_vec <- w_olink[target, ]
      t_type <- "Continuous"
    } else {
      t_vec <- w_lipid[target, ]
      t_type <- "Categorical_Binary"
    }
    
    # --- TAB A: Overall Distribution ---
    cat(sprintf("\n\n#### Data Profile & Highlights {.tabset .tabset-pills}\n\n"))
    
    cat(sprintf("\n\n##### Overall Population Profile\n\n"))
    
    df_dist <- data.frame(Value = t_vec) %>% drop_na()
    n_total <- nrow(df_dist)
    
    if (t_type == "Continuous") {
      p_hist <- ggplot(df_dist, aes(x = Value)) +
        geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
        labs(title = sprintf("Global Distribution: %s", target), subtitle = sprintf("Total Valid N = %d", n_total), x = "Abundance", y = "Count") +
        investigator_config$plot_theme
    } else {
      p_hist <- ggplot(df_dist, aes(x = factor(Value))) +
        geom_bar(fill = "steelblue", alpha = 0.8) +
        labs(title = sprintf("Global Detection: %s", target), subtitle = sprintf("Total Valid N = %d", n_total), x = "Detected (0/1)", y = "Count") +
        investigator_config$plot_theme
    }
    
    print(p_hist)
    cat("\n\n")
    ggsave(file.path(output_dir_data, sprintf("%s_%s_histogram.png", j_name, target)), p_hist, width=5, height=4)
    
    # --- TAB B: Specific Highlights ---
    
    for (hl_conf in conf$highlight_vars) {
      # FIXED: Added \n\n to the start of the string
      cat(sprintf("\n\n##### %s\n\n", hl_conf$var))
      
      res_hl <- wrap_highlight_plot(
        t_vec = t_vec, target_name = target, t_type = t_type, 
        hl_conf = hl_conf, env = parent.frame(), 
        plot_theme = investigator_config$plot_theme, clin_dict = clin_dict
      )
      
      if (res_hl$status == "success") {
        print(res_hl$plot)
        cat("\n\n")
        ggsave(file.path(output_dir_data, sprintf("%s_%s.png", j_name, res_hl$p_name)), res_hl$plot, width=5, height=4)
      } else {
        cat(sprintf("> Highlight failed for '%s' (Source: %s): %s\n\n", hl_conf$var, hl_conf$source, res_hl$msg))
      }
    }
    
    # --- TAB C: Correlation Results (Table Split Logic) ---
    cat(sprintf("\n\n#### Correlates {.tabset .tabset-pills}\n\n"))
    
    cor_results <- wrap_investigator_correlation(target_vec = t_vec, target_type = t_type, olink_mat = w_olink, lipid_mat = w_lipid, clin_mat = w_clin, clin_dict = clin_dict)
    
    for (domain in names(cor_results)) {
      # FIXED: Added \n\n to the start of the string
      cat(sprintf("\n\n##### %s Domain\n\n", domain))
      
      domain_df_full <- cor_results[[domain]] %>% filter(Feature != target) %>% arrange(FDR)
      
      # Queue Proteomics Data for the Network Chunk
      if (domain == "Proteomics") {
        network_queue[[paste(j_name, target, sep="_")]] <- list(target = target, df = domain_df_full)
      }
      
      # Table Splitting
      df_sig <- domain_df_full %>% filter(FDR < investigator_config$cor_fdr_cutoff) %>% mutate(Status = "Significant")
      df_nonsig <- domain_df_full %>% filter(FDR >= investigator_config$cor_fdr_cutoff) %>% head(10) %>% mutate(Status = "Top Non-Sig")
      
      df_display <- bind_rows(df_sig, df_nonsig)
      master_investigation_results[[paste(j_name, target, domain, sep="_")]] <- domain_df_full
      
      if (nrow(df_display) > 0) {
        if (isTRUE(getOption('knitr.in.progress'))) {
          html_table <- kbl(df_display, format = "html", digits = 4, align = "c", caption = sprintf("%s Correlates: %s", domain, target)) %>%
              kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left") %>%
              row_spec(which(df_display$Status == "Top Non-Sig"), color = "grey50") %>%
              scroll_box(width = "100%", height = "400px")
          cat(as.character(html_table))
        } else {
          print(rmarkdown::paged_table(df_display))
        }
      } else {
        cat("> No features evaluated.\n")
      }
      cat("\n\n")
    }
  }
}
```

# Directed Biological Networks {.tabset}
`r what_this_does("Rendered high-resolution, directed biological network graphs leveraging a database-first API query to OmniPath, mapping dataset significance onto validated primary and secondary biological interactions.")`

```{r directed_biological_networks, fig.height=12, fig.width=12, results='asis'}
for (net_job in names(network_queue)) {
  meta <- network_queue[[net_job]]
  
  cat(sprintf("\n\n## Network: %s\n\n", meta$target))
  
  p_net <- wrap_directed_network(
    target = meta$target, 
    prot_cor_df = meta$df, 
    fdr_cutoff = investigator_config$cor_fdr_cutoff, 
    plot_theme = investigator_config$plot_theme
  )
  
  if (!is.null(p_net)) {
    print(p_net)
    ggsave(file.path(output_dir_data, sprintf("%s_directed_network.png", net_job)), p_net, width=10, height=10)
  } else {
    cat("> *API returned no high-confidence biological network for this target, or internet is disconnected.*\n\n")
  }
}
```


# Export
`r what_this_does("Exported the complete statistical correlation tables for each investigated target to the output directory as CSV files.")`

```{r export}
for (res_name in names(master_investigation_results)) {
  # Save to dynamic output directory
  target_file <- file.path(output_dir_data, paste0("investigator_", res_name, ".csv"))
  write.csv(master_investigation_results[[res_name]], target_file, row.names = FALSE)
}
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=c(readLines(here::here("code", "01_general_engines.R")), readLines(here::here("code", "02_proteomics_engines.R")), readLines(here::here("code", "05_multiomics_engines.R"))), eval=FALSE}

```

```{r complete}
cat("Script 05a Complete! \n")
```

  <!-- # ========================================================================= -->
  <!-- # KITCHEN SINK / EXPLANATORY TEMPLATE (Commented Out) -->
  <!-- # Use this as a reference for all available configuration options. -->
  <!-- # ========================================================================= -->
  <!-- # job_example_all_features = list( -->
  <!-- #   title = "Investigate TARGET_X (Comprehensive)", -->
  <!-- #   targets = c("TARGET_X"), # Can list multiple targets here -->
  <!-- # -->
  <!-- #   # SUBSETTING (Optional): Filters the initial spine data before analysis -->
  <!-- #   # var: Column in master_spine. keep: The exact string value to retain. -->
  <!-- #   subset_filter = list(var = "cohort_group", keep = "SSc"), -->
  <!-- # -->
  <!-- #   # HIGHLIGHTS (Optional): Generates detailed distribution plots -->
  <!-- #   # source: Must match the R environment variable name of the loaded dataframe -->
  <!-- #   highlight_vars = list( -->
  <!-- #     list(var = "Age", source = "master_spine"),  -->
  <!-- #     list(var = "kmeans_Cluster", source = "clustering_res") -->
  <!-- #   ), -->
  <!-- # -->
  <!-- #   # CLINICAL TESTS: List of features to correlate against in the Clinical Domain -->
  <!-- #   clin_test_list = ssc_clinical_parameters, -->
  <!-- # -->
  <!-- #   # RESIDUALIZATION (Optional): Corrects continuous omics matrices for covariates -->
  <!-- #   residualize_conf = list(formula = "~ Age + Sex") -->
  <!-- # ), -->
  <!-- # ========================================================================= -->

  

  
