---
title: "Target Investigator"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded required bioinformatics packages and graphics libraries.", "Imported the pre-cleaned clinical spine, clinical dictionary, continuous Olink proteomics matrix, and binary lipidomics matrix.", "Sourced the legacy engine files and the new investigator-specific engines.")`

```{r setup_and_import}
library(tidyverse)
library(patchwork)
library(here)
library(kableExtra)
library(ggraph)
library(igraph)
# library(limma) #really required??

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file <- "05a_investigator"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Legacy Engines
source(here::here("code", "01_general_engines.R"))
# source(here::here("code", "02_proteomics_engines.R")) # really required??
source(here::here("code", "05_multiomics_engines.R"))

# Load Data
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
clin_dict <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
omics_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))
lipid_mat <- readRDS(here::here("output", "04b_lipidomics_differential", "lipid_binary_matrix.rds"))
```


# Analysis Configurator
`r what_this_does("Defined specific groupings of clinical variables to streamline interpretations.", "Configured the investigative jobs, defining the targets (single or multiple), the background subgrouping strategy, and the statistical constraints (e.g., residualizing out age).")`
`r analysis_decision("Age and Sex are residualized in jobs 2 and 3 to ensure identified correlations represent independent biological signals rather than demographic confounders.")`

```{r analysis_configurator}
# 1. Clinical Parameter Groupings
clin_lists <- list(
  cardiac_vars = c("NTproBNP", "LVEF", "Heart_Failure_Status"),
  pulmonary_vars = c("FVC_percent", "DLCO_percent", "ILD_Status"),
  demographic_vars = c("Age", "BMI", "Sex")
)

# 2. Global Investigator Configurations
investigator_config <- list(
  cor_fdr_cutoff = 0.05,
  plot_theme = theme_minimal()
)

# 3. Job Definitions
investigation_jobs <- list(
  job1_single_protein = list(
    title = "Investigate CXCL10 (Single Protein)",
    targets = c("CXCL10"),
    subgroup_var = "cohort_group",   # Variable to split distribution plots by
    clin_test_list = c(clin_lists$cardiac_vars, clin_lists$pulmonary_vars),
    residualize_conf = NULL          # No covariate correction
  ),
  
  job2_single_lipid = list(
    title = "Investigate Cer(d18:1/24:0) (Single Lipid)",
    targets = c("Cer_d18_1_24_0"),
    subgroup_var = "Inflamm_active",
    clin_test_list = unlist(clin_lists), # Test all mapped clinical vars
    residualize_conf = list(method = "residualize", covariates = c("Age", "Sex"))
  ),
  
  job3_target_network = list(
    title = "Investigate Fibrosis Signature (Multi-Target Network)",
    targets = c("COL1A1", "MMP9", "TIMP1", "Cer_d18_1_24_0", "SM_d18_1_16_0"),
    subgroup_var = "fibrotic_active",
    clin_test_list = clin_lists$pulmonary_vars,
    residualize_conf = list(method = "residualize", covariates = c("Age"))
  )
)

# Configuration for distributions plots
plot_configs <- list(
  violin = list(alpha = 0.7, width = 0.8),
  boxplot = list(width = 0.2, alpha = 0.9, outlier.shape = NA)
)
```

# Target Investigation {.tabset}
`r what_this_does("Iterated through defined jobs to extract target profiles, correct for covariates if requested, and plot clinical subgroup distributions.", "Orchestrated targeted correlation networks across proteomics, lipidomics, and specific clinical parameters.", "Dynamically saved plots and exported full significant summary tables per job.")`

```{r target_investigator}
master_investigation_results <- list()

for (j_name in names(investigation_jobs)) {
  conf <- investigation_jobs[[j_name]]
  
  cat(sprintf("\n\n## %s\n\n", conf$title))
  
  # 1. Prep Specific Matrices & Residualize
  w_olink <- omics_mat
  w_lipid <- lipid_mat
  
  if (!is.null(conf$residualize_conf)) {
    # Call wrapper factory for Olink
    fac_olink <- wrap_matrix_factory(w_olink, master_spine, list(transform = conf$residualize_conf), prep_id = j_name)
    w_olink <- fac_olink$matrix
    
    # Note: Residualizing a purely binary lipid matrix via limma yields continuous residuals.
    # To maintain our binary-track logic, we do NOT residualize the lipid matrix here. 
    # Covariates for lipid targets must be handled upstream or accepted as unadjusted in this script.
    cat("> *Note: Covariate residualization applied to continuous proteomics matrix only. Binary lipid matrix retained in raw state to support Fisher's Exact routing.*\n\n")
  }
  
  # 2. Extract Valid Targets
  valid_olink_targets <- intersect(conf$targets, rownames(w_olink))
  valid_lipid_targets <- intersect(conf$targets, rownames(w_lipid))
  all_valid_targets <- c(valid_olink_targets, valid_lipid_targets)
  
  if (length(all_valid_targets) == 0) {
    cat(sprintf("> <span style='color:red;'>**SKIPPED:** None of the targets found in matrices.</span>\n\n"))
    next
  }
  
  # 3. Prep Clinical Matrix for testing
  w_clin <- master_spine %>%
    select(Subject_ID, any_of(conf$clin_test_list)) %>%
    column_to_rownames("Subject_ID") %>%
    t() # Transpose to match omics matrix format (features in rows)
  
  # 4. Multi-Target Network Assessment (If > 1 target)
  if (length(all_valid_targets) > 1) {
    cat("### Target Network Inter-correlations\n\n")
    
    # Combine target data into a single matrix for pairwise checking
    target_mat <- rbind(w_olink[valid_olink_targets, , drop=FALSE], 
                        w_lipid[valid_lipid_targets, , drop=FALSE])
    
    # Calculate pairwise correlation matrix (Spearman handles continuous vs continuous and cont vs binary)
    cor_mat <- cor(t(target_mat), method = "spearman", use = "pairwise.complete.obs")
    
    # Simple Network Visualization
    graph <- graph_from_adjacency_matrix(abs(cor_mat) > 0.3, mode="undirected", diag=FALSE)
    p_net <- ggraph(graph, layout = "fr") + 
      geom_edge_link(aes(edge_alpha = after_stat(index)), show.legend = FALSE) + 
      geom_node_point(size = 5, color = "steelblue") + 
      geom_node_text(aes(label = name), repel = TRUE, size = 4) +
      theme_void() +
      ggtitle("Target Correlation Network (|Rho| > 0.3)")
    
    print(p_net)
    cat("\n\n")
    
    # Save network
    ggsave(file.path(output_dir_data, sprintf("%s_network.png", j_name)), p_net, width=6, height=6)
  }
  
  # 5. Iterative Single Target Deep-Dive
  for (target in all_valid_targets) {
    cat(sprintf("### Target: %s {.tabset .tabset-pills .toc-ignore}\n\n", target))
    
    # Determine target source and extract vector
    if (target %in% valid_olink_targets) {
      t_vec <- w_olink[target, ]
      t_type <- "Continuous"
    } else {
      t_vec <- w_lipid[target, ]
      t_type <- "Categorical_Binary"
    }
    
    # --- TAB A: Distribution ---
    cat(sprintf("#### Distribution\n\n"))
    
    df_plot <- data.frame(
      Value = t_vec,
      Subgroup = master_spine[[conf$subgroup_var]][match(names(t_vec), master_spine$Subject_ID)]
    ) %>% drop_na()
    
    if (t_type == "Continuous") {
      p_dist <- ggplot(df_plot, aes(x = Subgroup, y = Value, fill = Subgroup)) +
        geom_violin(alpha = plot_configs$violin$alpha, width = plot_configs$violin$width) +
        geom_boxplot(width = plot_configs$boxplot$width, alpha = plot_configs$boxplot$alpha, outlier.shape = plot_configs$boxplot$outlier.shape) +
        labs(title = sprintf("%s Expression by %s", target, conf$subgroup_var), y = "Abundance (NPX/Residual)", x = "") +
        investigator_config$plot_theme +
        theme(legend.position = "none")
    } else {
      # Bar plot for binary lipid distributions
      p_dist <- ggplot(df_plot, aes(x = Subgroup, fill = factor(Value))) +
        geom_bar(position = "fill") +
        labs(title = sprintf("%s Detection Rate by %s", target, conf$subgroup_var), y = "Proportion", fill = "Detected (0/1)") +
        investigator_config$plot_theme
    }
    
    print(p_dist)
    cat("\n\n")
    ggsave(file.path(output_dir_data, sprintf("%s_%s_distribution.png", j_name, target)), p_dist, width=5, height=4)
    
    # --- TAB B: Correlations ---
    cat(sprintf("#### Significant Correlates\n\n"))
    
    cor_results <- orchestrate_correlations(
      target_vec = t_vec, target_type = t_type, 
      olink_mat = w_olink, lipid_mat = w_lipid, 
      clin_mat = w_clin, clin_dict = clin_dict
    )
    
    # Bind, filter for significance, and sort
    final_cor_df <- bind_rows(cor_results) %>%
      filter(FDR < investigator_config$cor_fdr_cutoff) %>%
      filter(Feature != target) %>% # Remove self-correlation
      arrange(FDR)
    
    # Store to master list for export
    master_investigation_results[[paste(j_name, target, sep="_")]] <- final_cor_df
    
    # Print Table dynamically to HTML
    if (nrow(final_cor_df) > 0) {
      html_table <- kbl(final_cor_df, format = "html", digits = 4, align = "c", 
                        caption = sprintf("Significant Correlates: %s (FDR < %s)", target, investigator_config$cor_fdr_cutoff)) %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left") %>%
          scroll_box(width = "100%", height = "400px")
      
      cat(as.character(html_table))
    } else {
      cat(sprintf("> No significant correlations found at FDR < %s.\n", investigator_config$cor_fdr_cutoff))
    }
    cat("\n\n")
    
    # Memory Management per target
    rm(t_vec, df_plot, p_dist, cor_results, final_cor_df)
  }
  
  # Memory Management per job
  rm(w_olink, w_lipid, w_clin)
  gc()
}
```

# Export
`r what_this_does("Exported the complete significant correlation tables for each investigated target as CSV files to the output directory.")`

```{r export}
for (res_name in names(master_investigation_results)) {
  # Save to dynamic output directory
  target_file <- file.path(output_dir_data, paste0("investigator_", res_name, ".csv"))
  write.csv(master_investigation_results[[res_name]], target_file, row.names = FALSE)
}
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r print_engine, code=c(readLines(here::here("code", "01_general_engines.R")), readLines(here::here("code", "02_proteomics_engines.R")), readLines(here::here("code", "05_multiomics_engines.R"))), eval=FALSE}

```

```{r complete}
cat("Script 05a Complete! \n")
```
