---
title: "05a: miRNA-Seq QC and Preprocessing"
author: "Alexander Gysin(alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded essential libraries and sourced custom command-line tool engines.", "Established storage separation: heavy processing files map to the scratch drive, while lightweight web assets map to the local project.", "Located the raw R1 FastQ files on the scratch drive.")`

```{r setup_and_import, message=FALSE, error=FALSE}
library(tidyverse)
library(here)
library(pbmcapply)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file    <- "05a_qc_and_preprocessing"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Web Assets Path (Lightweight HTML reports only)
docs_asset_dir  <- here::here("docs", "assets", current_file, "qc_reports")

# Scratch Directory (Heavy I/O: FastQ files and raw FastQC data)
scratch_out_dir <- file.path("~/scratch/miRNA_project", current_file)

# Source Engines
source(here::here("code", "05_sRNA_engines.R"))

# Locate raw FastQ files on Scratch
raw_fastq_dir <- "~/scratch/o42057_NovaSeq_260623_X506"
raw_files <- list.files(raw_fastq_dir, pattern = "R1.*\\.fastq\\.gz$", full.names = TRUE)

if(length(raw_files) == 0) {
  warning("No raw fastq files found. Please verify the scratch raw_fastq_dir path.")
}
```

# Analysis Configurator
`r what_this_does("Introduced a test-run toggle to safely evaluate the pipeline on a subset of samples.", "Defined system resource limits to dynamically harness the supercomputer without causing I/O bottlenecks.", "Configured UMI-tools with a PCRE regex pattern to extract the 12bp UMI following the exact 19bp QIAseq 3' adapter.", "Set Cutadapt parameters to trim the adapter and filter degraded reads (<18nt).")`
```{r analysis_config}
preprocessing_config <- list(
  
  # 0. Pipeline Run Mode
  run_mode = list(
    test_run = FALSE,    # Toggle to TRUE to only process a small subset
    test_samples = 2    # Number of samples to process if test_run is TRUE
  ),
  
  # 1. HPC System Limits
  system = list(
    max_cores  = 128,   # This is currently configured for the HPC
    core_ratio = 0.75   # Use 75% of available cores up to the max_cap
  ),
  
  # 2. UMI Extraction (QIAseq Kit Regex)
  umi = list(
    pattern = ".*AACTGTAGGCACCATCAAT(?P<umi_1>.{12}).*"
  ),
  
  # 3. Adapter Trimming
  cutadapt = list(
    adapter    = "AACTGTAGGCACCATCAAT",
    min_length = 18 # Discard reads smaller than typical mature miRNAs
  )
)

# Estimation per sample: 20min
```

# Preprocessing Execution 
`r what_this_does("Evaluated the test-run toggle to dynamically route outputs and subset input files if active.", "Executed the parallelized preprocessing pipeline.", "Ran FastQC and aggregated via MultiQC for raw reads.", "Extracted UMIs to the header and trimmed the 3' adapters.", "Ran FastQC and aggregated via MultiQC for the final processed reads.")`
```{r preprocessing_execution, results='asis', eval=FALSE}
# 1. Dynamically route folders based on test mode (FIXED path logic)
folder_suffix <- ifelse(preprocessing_config$run_mode$test_run, "_test", "")
target_folder <- paste0(current_file, folder_suffix)

scratch_out_dir <- file.path("~/scratch/miRNA_project", target_folder)
docs_asset_dir  <- here::here("docs", "assets", target_folder, "qc_reports")

# 2. Subset files if testing
if(preprocessing_config$run_mode$test_run) {
  cat("> **Notice:** TEST RUN ACTIVE. Processing only", preprocessing_config$run_mode$test_samples, "samples.\n\n")
  raw_files <- head(raw_files, preprocessing_config$run_mode$test_samples)
}

# 3. Execute
if(length(raw_files) > 0) {
  
  pipeline_res <- wrap_preprocessing_pipeline(
    fastq_files  = raw_files, 
    config       = preprocessing_config, 
    scratch_dir  = scratch_out_dir,
    docs_dir     = docs_asset_dir
  )
  
  if(pipeline_res$status == "complete") {
    cat(sprintf("> **Success:** Processed %d/%d samples successfully.\n\n", 
                pipeline_res$diagnostics$successful_processed, 
                pipeline_res$diagnostics$total_samples))
    
    cat("Processed reads are stored on the scratch drive at: `", pipeline_res$paths$processed_fastq_dir, "`\n\n")
    
    # Print trapped errors if any failed
    if(length(pipeline_res$diagnostics$errors) > 0) {
      cat("### Error Logs from Failed Samples\n\n```text\n")
      cat(head(pipeline_res$diagnostics$errors, 3), sep = "\n\n-----------------\n\n")
      cat("\n```\n\n")
    }
  }
} else {
  cat("> **Notice:** Execution skipped. No raw fastq files loaded.\n\n")
}
```

# Quality Control Dashboards {.tabset}
`r what_this_does("Embedded the interactive MultiQC HTML reports directly into the workflowr page using iframe shielding, isolating their complex CSS/JS from Pandoc's DOM parser.")`

```{r qc_dashboard, results='asis'}
# Dynamically map relative paths based on test mode
folder_suffix <- ifelse(preprocessing_config$run_mode$test_run, "_test", "")
target_folder <- paste0(current_file, folder_suffix)

raw_qc_rel_path <- file.path("assets", target_folder, "qc_reports", "raw_multiqc_report.html")
pro_qc_rel_path <- file.path("assets", target_folder, "qc_reports", "processed_multiqc_report.html")

cat("## Raw Reads QC\n\n")
if(file.exists(here::here("docs", raw_qc_rel_path))) {
  cat(sprintf('<iframe src="%s" width="100%%" height="800px" style="border:none;"></iframe>\n\n', raw_qc_rel_path))
} else {
  cat("*Raw QC report not yet generated.*\n\n")
}

cat("## Processed Reads QC\n\n")
if(file.exists(here::here("docs", pro_qc_rel_path))) {
  cat(sprintf('<iframe src="%s" width="100%%" height="800px" style="border:none;"></iframe>\n\n', pro_qc_rel_path))
} else {
  cat("*Processed QC report not yet generated.*\n\n")
}
```

# Export and Cleanup
`r what_this_does("Exported the processed data paths to RDS so Script 05b can seamlessly load them from the scratch drive, accounting for test mode names.", "Purged unneeded variables and enforced garbage collection to free RAM.")`

```{r export_and_cleanup}
if(exists("pipeline_res") && pipeline_res$status == "complete") {
  # Save ONLY the pointer path locally to the home directory (append _test if active)
  file_suffix <- ifelse(preprocessing_config$run_mode$test_run, "_test", "")
  export_name <- paste0("processed_fastq_dir_path", file_suffix, ".rds")
  
  saveRDS(
    pipeline_res$paths$processed_fastq_dir, 
    file.path(output_dir_data, export_name)
  )
}

# Strict Memory Optimization
rm(list = ls(pattern = "^raw_|^pipeline_"))
gc()
```

Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r appendix, code=c(readLines(here::here("code", "0_config.R")), readLines(here::here("code", "05_sRNA_engines.R"))), eval=FALSE}

```


```{r complete}
cat("Script 05a Complete! \n")
```

