---
title: "sRNA Alignment and Counting"
author: "Alexander Gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup_boxes, include=FALSE}
rm(list = ls())
source(here::here("code", "00_config.R"))
```

# Setup and Import
`r what_this_does("Loaded essential libraries and sourced custom engines.", "Dynamically imported the scratch drive path containing the processed FastQ files from Script 05a.")`

```{r setup_and_import, message=FALSE}
library(tidyverse)
library(here)
library(pbmcapply)
library(Rsubread)
library(Rsamtools)

# Knitr Config
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Output paths
current_file    <- "05b_alignment_and_counting"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "05_sRNA_engines.R"))

# Always read the pointer saved by 05a
rds_path <- here::here("output", "05a_sRNA_qc_and_preprocessing", "processed_fastq_dir_path.rds")
# Fallback for test mode inheritance if run separately
test_rds_path <- here::here("output", "05a_sRNA_qc_and_preprocessing", "processed_fastq_dir_path_test.rds")

if(file.exists(rds_path)) {
  processed_fastq_dir <- readRDS(rds_path)
} else if(file.exists(test_rds_path)) {
  processed_fastq_dir <- readRDS(test_rds_path)
} else {
  stop("Could not find RDS output from 05a. Please run 05a_sRNA_qc_and_preprocessing.Rmd first.")
}

processed_files <- list.files(processed_fastq_dir, pattern = "\\.fastq\\.gz$", full.names = TRUE)

if(length(processed_files) == 0) {
  warning("No processed fastq files found in the inherited scratch directory.")
}
```

# Analysis Configurator
`r what_this_does("Shifted reference index to strictly target the miRNAome using miRBase annotations.", "Utilized Bowtie 1 parameters tailored for short reads to retain optimal mappings and allow miRDeep2 weighting.", "Automated deduplication and Fastx formatting prior to cohort-level miRDeep2 quantification.")`

```{r analysis_configurator}
align_count_config <- list(
  
  # 0. Pipeline Run Mode
  run_mode = list(
    test_run = FALSE,        
    test_samples = 2         
  ),
  
  # 1. HPC System Limits
  system = list(
    max_cores  = 128,  
    core_ratio = 0.75  
  ),
  
  # 2. Reference Genomes & Annotations (miRNAome specifically)
  reference = list(
    hairpin_url = "[https://www.mirbase.org/ftp/CURRENT/hairpin.fa](https://www.mirbase.org/ftp/CURRENT/hairpin.fa)",
    mature_url  = "[https://www.mirbase.org/ftp/CURRENT/mature.fa](https://www.mirbase.org/ftp/CURRENT/mature.fa)",
    
    hairpin_path = "~/scratch/miRNA_project/resources/genomes/miRBase/hairpin.fa",
    mature_path  = "~/scratch/miRNA_project/resources/genomes/miRBase/mature.fa",
    index_path   = "~/scratch/miRNA_project/resources/genomes/miRBase/bowtie_index/hairpin_index"
  ),
  
  # 3. Alignment Parameters (Bowtie 1)
  align = list(
    # -v 0: 0 mismatches allowed (strict)
    # -a: report all valid alignments (required for miRDeep2 multimapping weight)
    # --best --strata: group best hits
    # --norc: no reverse complement (highly strand specific)
    bowtie_params = "-v 0 -a --best --strata --norc"
  )
)

# Apply Test Subset
if (align_count_config$run_mode$test_run) {
  cat("> **Notice:** TEST RUN ACTIVE. Processing only", align_count_config$run_mode$test_samples, "samples.\n\n")
  processed_files <- head(processed_files, align_count_config$run_mode$test_samples)
}
```

# Download Reference and Build Bowtie Index
`r what_this_does("Downloaded the hairpin and mature FASTA sequences directly to the scratch drive.", "Generated the specialized Bowtie 1 index required for short-read alignment.")`

```{r reference}
# 1. Download
worker_download_reference(align_count_config$reference$hairpin_url, align_count_config$reference$hairpin_path)
worker_download_reference(align_count_config$reference$mature_url, align_count_config$reference$mature_path)

# 2. Build Bowtie 1 Index
# Bowtie 1 indices typically end with .1.ebwt
if (!file.exists(paste0(align_count_config$reference$index_path, ".1.ebwt"))) {
  message("Bowtie 1 index not found. Building index for hairpin sequences...")
  
  dir.create(dirname(align_count_config$reference$index_path), recursive = TRUE, showWarnings = FALSE)
  
  conda_bin  <- path.expand("~/miniconda3/envs/mirna_env/bin")
  cmd_build  <- file.path(conda_bin, "bowtie-build")
  
  # Execute bowtie-build
  system2(cmd_build, args = c(align_count_config$reference$hairpin_path, align_count_config$reference$index_path))
  message("Bowtie index built successfully.")
} else {
  message("Bowtie index found. Skipping build step.")
}
```

# Align, Format and Quantify
`r what_this_does("Executed the parallelized orchestrator to map reads via Bowtie, sort via Samtools, and deduplicate UMIs via UMICollapse.", "Converted BAMs back to collapsed FastQ files.", "Passed the cohort to miRDeep2 for weighted quantification.")`

```{r align_format_quantify, eval = FALSE}
folder_suffix   <- ifelse(align_count_config$run_mode$test_run, "_test", "")
target_folder   <- paste0(current_file, folder_suffix)
scratch_out_dir <- file.path("~/scratch/miRNA_project", target_folder)

if(length(processed_files) > 0) {
  
  # This wrapper now runs the pipeline end-to-end and returns the miRDeep2 count matrix natively
  quant_res <- wrap_alignment_dedup(
    fastq_files = processed_files, 
    config      = align_count_config, 
    scratch_dir = scratch_out_dir
  )
  
  if(quant_res$status == "complete") {
    cat(sprintf("> **Success:** Aligned, deduplicated, and quantified %d/%d samples successfully.\n\n", 
                quant_res$diagnostics$successful_processed, 
                quant_res$diagnostics$total_samples))
    
    cat("Quantification files stored at: `", quant_res$paths$quant_dir, "`\n\n")
    
    if(length(quant_res$diagnostics$errors) > 0) {
      cat("### Error Logs from Failed Samples\n\n```text\n")
      cat(head(quant_res$diagnostics$errors, 3), sep = "\n\n-----------------\n\n")
      cat("\n```\n\n")
    }
  }
} else {
  cat("> **Notice:** Execution skipped. No preprocessed fastq files loaded.\n\n")
}
```


# Export 
`r what_this_does("Extracted the clean raw count matrix generated by miRDeep2 and exported it for downstream Differential Expression analysis.")`

```{r export}
# The matrix is already loaded into R memory by the wrapper function
if(exists("quant_res") && quant_res$status == "complete" && !is.null(quant_res$counts)) {
  
  file_suffix <- ifelse(align_count_config$run_mode$test_run, "_test", "")
  
  # Export Count Matrix (RDS for data integrity in downstream R scripts)
  matrix_path <- file.path(output_dir_data, paste0("raw_count_matrix", file_suffix, ".rds"))
  saveRDS(quant_res$counts, matrix_path)
  
  # Export as CSV for quick viewing/sharing
  csv_path <- file.path(output_dir_data, paste0("raw_count_matrix", file_suffix, ".csv"))
  write.csv(quant_res$counts, csv_path, row.names = FALSE)
  
  cat("> **Export Complete:** Raw count matrix saved to `", matrix_path, "`\n")
}
```

Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`

```{r appendix, code=c(readLines(here::here("code", "0_config.R")), readLines(here::here("code", "05_sRNA_engines.R"))), eval=FALSE}

```


```{r complete}
cat("Script 05b Complete! \n")
```
