---
title: "Side Project - Ellen (Extreme Vascular Phenotypes)"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---
```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Introduction
This is a side-project focused on comparing extreme vascular phenotypes (5 vs 5). It executes a covariate-free Differential Expression Analysis (DEA) due to the low sample size, and visualizes the top drivers from the full cohort alongside specific proteins of interest using Z-scored heatmaps.

# Setup and Import
`r what_this_does("Loaded standard analysis packages.", "Imported the pre-cleaned spine, the Olink matrix, and the full vascular DEA results from script 03b.")`

```{r setup_and_import, message=FALSE, warning=FALSE}
# Packages
library(tidyverse)
library(data.table)
library(limma)
library(ComplexHeatmap)
library(circlize)
library(here)
library(kableExtra)
library(msigdbr)

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "z01_sideproject_ellen"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "00_omics_engines.R"))

# Data Import
master_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
omics_mat <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

full_dea_path <- here::here("output", "03b_proteomics_differential", "full_vascular_dea_results.rds")
if(file.exists(full_dea_path)) {
  full_vascular_dea <- readRDS(full_dea_path)
} else {
  warning("full_vascular_dea_results.rds not found! Run 03b first.")
  full_vascular_dea <- NULL
}

# Fetch GO Biological Processes Database for GSEA (Engine requirement)
go_genesets <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% select(gs_name, gene_symbol)
```

# Configuration
`r what_this_does("Configured the specific Subject IDs for the extreme vascular phenotypes.", "Mapped human-readable protein names to standard Olink symbols.", "Defined the covariate-free DEA parameters with a shortened title to prevent heatmap rendering clipping.", "Dynamically overwrites the global color dictionary to fix the engine's hardcoded heatmap annotations.")`

```{r configuration}
# 1. Patient Selection (ACTION REQUIRED: Fill these in!)
manual_patient_selection <- list(
  vascular_active = c("Act_2", "Act_34", "Act_47", "Act_29", "Act_12"),   
  vascular_inactive = c("Act_33", "Nonact_20", "Act_16", "Nonact_38", "Nonact_35") 
)

# 2. Proteins of Interest Translation Map
proteins_of_interest <- c(
  "ECE1",    # Endothelin-converting enzyme 1
  "HIF1A",   # Hypoxia-inducible factor 1-alpha
  "PGF",     # Placental growth factor
  "FGF2",    # Fibroblast growth factor 2
  "FGF21",   # Fibroblast growth factor 21
  "TNF",     # Tumor necrosis factor
  "VEGFA",   # Vascular endothelial growth factor A
  "THBS1",   # Thrombospondin 1
  "CXCL10"   # CXCL10
)

# 3. DEA Configs
contrast_5v5 <- list(
  title = "Vascular 5v5", # FIX: Shortened title prevents left-clipping over the narrow 10-patient heatmap
  group_col = "phenotype_group",
  target_groups = c("Vascular_Active"),
  ref_groups = c("Vascular_Inactive"),
  is_continuous = FALSE,
  covariates = c(), # INTENTIONALLY EMPTY due to N=10
  dea_p_cutoff = 0.05,
  dea_fc_cutoff = 0.5
)

# 4. Engine Configs
omics_config <- list(
  gsea_p_cutoff  = 0.20, 
  gsea_min_size  = 3    
)

heatmap_config <- list(
  top_n = 30,
  split_by_group = FALSE, 
  row_font_size = 8
)

# 5. FIX: Global Color Override
# By overwriting the function globally in this script, the internal engine is tricked 
# into pulling our custom Red/Blue colors for the phenotype_group annotations!
original_get_project_colors <- get_project_colors
get_project_colors <- function(requested_levels) {
  cols <- original_get_project_colors(requested_levels)
  cols[names(cols) == "Vascular_Active"] <- "#e74c3c"   # Red
  cols[names(cols) == "Vascular_Inactive"] <- "#3498db" # Blue
  return(cols)
}
```

# Data Preparation
`r what_this_does("Subsets the massive clinical spine and proteomics matrix down to just the 10 selected patients.", "Decodes the binary 'Sex' variable back to 'Male'/'Female' to match the global config color dictionary.", "Creates a localized 'phenotype_group' variable strictly for this 5v5 contrast and intercepts the engine's hardcoded cohort annotations.")`
```{r data_preparation}
# Subset clinical spine
clin_10 <- master_spine[Subject_ID %in% c(manual_patient_selection$vascular_active, manual_patient_selection$vascular_inactive)]

if(nrow(clin_10) != 10) {
  cat(sprintf("<b style='color:red;'>WARNING: Only found %d of the 10 specified patients in the spine! Please check your ID list.</b>\n\n", nrow(clin_10)))
}

# Decode Sex
if ("Sex" %in% names(clin_10)) {
  clin_10[, Sex := fcase(
    Sex == 1 | Sex == "1", SEX_FEMALE,
    Sex == 0 | Sex == "0", SEX_MALE,
    default = NA_character_
  )]
}

# Create strict 5v5 grouping column
clin_10[, phenotype_group := fcase(
  Subject_ID %in% manual_patient_selection$vascular_active, "Vascular_Active",
  Subject_ID %in% manual_patient_selection$vascular_inactive, "Vascular_Inactive"
)]
clin_10[, phenotype_group := factor(phenotype_group, levels = c("Vascular_Active", "Vascular_Inactive"))]

# FIX: Overwrite cohort_group so the engine draws the vascular names!
clin_10[, cohort_group := phenotype_group]

# Subset Omics Matrix
omics_mat_10 <- omics_mat[, clin_10$Subject_ID, drop = FALSE]

cat(sprintf("Successfully isolated %d patients (%d Active vs %d Inactive) and %d proteins.\n", 
            nrow(clin_10), 
            sum(clin_10$phenotype_group == "Vascular_Active"),
            sum(clin_10$phenotype_group == "Vascular_Inactive"),
            nrow(omics_mat_10)))
```

# 5v5 Differential Expression Analysis
`r analysis_decision("Omitted clinical covariates from the DEA to preserve degrees of freedom given the N=10 constraint.")`

`r what_this_does("Constructs a covariate-free linear model to compare the 5 Vascular Active vs 5 Vascular Inactive patients.", "Computes empirical Bayes moderated t-statistics (limma) to calculate Fold Changes and False Discovery Rates (FDR).","Generates a Volcano plot highlighting the top 20 most extreme proteins.","Ranks the entire proteome using the Pi-score formula (Pi = |logFC| * -log10(FDR)) to balance magnitude and significance.","Extracts the top 30 Pi-score proteins and renders an automated Z-scored heatmap to visualize the separation between the 10 patients.")`

```{r 5v5_DEA, results='asis', fig.width=8, fig.height=8}
# Run Engine
res_5v5 <- wrap_dea_pipeline(
  omics_mat           = omics_mat_10, 
  clin_df             = clin_10, 
  dea_conf            = contrast_5v5, 
  dea_id              = "extreme_5v5", 
  base_output_dir     = output_dir_data, 
  go_db               = go_genesets, 
  omics_config        = omics_config, 
  heatmap_config      = heatmap_config, 
  project_colors_func = get_project_colors# <--- CHANGED THIS LINE
)

if(res_5v5$status == "success") {
  print(res_5v5$plots$volcano); cat("\n\n")
  
  if(nrow(res_5v5$tables$significant_proteins) > 0) {
    cat(sprintf("##### Significant Proteins (%d total)\n\n", nrow(res_5v5$tables$significant_proteins)))
    print(kbl(res_5v5$tables$significant_proteins, format = "html", digits = 4, align = "c") %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left"))
    cat("\n\n")
  } else {
  }
  
  draw(res_5v5$plots$heatmap); cat("\n\n")
} else {
  cat(sprintf("**DEA Failed:** %s\n", res_5v5$error_msg))
}
```
`r key_insight("No significantly differentially expressed proteins", "Despite this the heatmap separates the vascular and non-vascular patients in the clustering", "There is only a hint of a pattern in the heatmap", "The top left and bottom right tend to be up and the top right and bottom left tend to be downregulated")`

# Full-cohort Heatmap
`r analysis_decision("Extracted the Top 30 proteins from the FULL cohort vascular DEA ranked by lowest FDR.", "Z-scored these exact 30 proteins across just the 10 extreme patients to maximize visual contrast of the systemic signals.")`

`r what_this_does("Subsets the 10-patient omics matrix to the top 30 proteins identified in the full-cohort DEA.", "Calculated row-wise Z-scores to normalize protein abundance.", "Rendered an annotated ComplexHeatmap clustering the patients based on these systemic drivers.")`

```{r full_cohort_heatmap, results='asis', fig.width=8, fig.height=7}
if (!is.null(full_vascular_dea)) {
  
  # Extract Top 30 by FDR (adj.P.Val)
  top_30_features <- full_vascular_dea %>%
    arrange(adj.P.Val) %>%
    slice_head(n = 30) %>%
    pull(Feature)
  
  # Subset 10-patient matrix to these 30 features
  mat_top30 <- omics_mat_10[rownames(omics_mat_10) %in% top_30_features, , drop = FALSE]
  
  # Z-Score row-wise
  mat_top30_z <- t(scale(t(mat_top30)))
  
  # Annotations & Colors
  col_fun <- colorRamp2(c(-2, 0, 2), c("#2980b9", "white", "#c0392b"))
  group_colors <- c("Vascular_Active" = "#e74c3c", "Vascular_Inactive" = "#3498db")
  ha <- HeatmapAnnotation(Group = clin_10$phenotype_group, col = list(Group = group_colors))
  
  ht2 <- Heatmap(
    mat_top30_z,
    name = "Z-Score",
    top_annotation = ha,
    col = col_fun,
    column_title = "Top 30 Full-Cohort Vascular Proteins (Subsetted to N=10)",
    row_names_gp = gpar(fontsize = 8),
    column_names_gp = gpar(fontsize = 10),
    cluster_columns = TRUE, 
    show_row_dend = FALSE
  )
  
  cat("> **Note:** The proteins displayed below are the top 30 ranked by lowest FDR in the *full* cohort DEA. None of these proteins have passed the FDR < 0.05 threshold.\n\n")
  
  draw(ht2)
  cat("\n\n")
  
} else {
  cat("> Skipped: Full cohort DEA results not found.\n")
}
```
`r key_insight("No visible pattern")`

# Proteins of Interest Heatmap
`r analysis_decision("Filtered the 10-patient matrix for the predefined list of human-translated proteins of interest.", "Applied Z-scoring to highlight relative abundance shifts strictly between the extreme phenotypes.")`

`r what_this_does("Matched the predefined proteins of interest against the available Olink assays, generating a warning for any missing targets.", "Calculated row-wise Z-scores for the surviving proteins.", "Rendered an annotated ComplexHeatmap, intentionally preserving the custom row order defined in the configuration.")`

```{r poi_heatmap, results='asis', fig.width=8, fig.height=4}
# Find which proteins actually exist in the matrix
available_poi <- intersect(proteins_of_interest, rownames(omics_mat_10))
missing_poi <- setdiff(proteins_of_interest, rownames(omics_mat_10))

if (length(missing_poi) > 0) {
  cat(sprintf("<b style='color:#d35400;'>Note: The following proteins were not found in the Olink matrix and will be excluded: %s</b>\n\n", paste(missing_poi, collapse = ", ")))
}

if (length(available_poi) > 0) {
  
  # Subset and Z-Score
  mat_poi <- omics_mat_10[available_poi, , drop = FALSE]
  mat_poi_z <- t(scale(t(mat_poi)))
  
  ht3 <- Heatmap(
    mat_poi_z,
    name = "Z-Score",
    top_annotation = ha, # Reuse annotation from Goal 2
    col = col_fun,       # Reuse color scale from Goal 2
    column_title = "Proteins of Interest (Z-Scored on Extreme N=10)",
    row_names_gp = gpar(fontsize = 10, fontface = "bold"),
    column_names_gp = gpar(fontsize = 10),
    cluster_columns = TRUE,
    cluster_rows = FALSE, # Keep user's predefined list order
    show_row_dend = FALSE
  )
  
  draw(ht3)
  cat("\n\n")
  
} else {
  cat("> Skipped: No proteins of interest found in the dataset.\n")
}
```
`r key_insight("No visible pattern")`

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script z01 Complete! \n")
```
