---
title: "Sideproject: Vascular 2"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

# Introduction
Goal: To identify SSc patients with the most extreme "vascular disease" proteomic signatures to select their serum for functional in vitro organoid stimulation.

<div style='border: 2px solid #e74c3c; padding: 15px; border-radius: 5px; background-color: #fdf2e9; margin-bottom: 20px;'>
<h4 style='margin-top: 0; color: #c0392b;'>⚠️ NOTE: Medication Confounders</h4>
The computational pipeline currently lacks patient medication histories . If organoids exhibit divergent behavior, consider that they may be responding to circulating pharmaceutical agents in the serum rather than the endogenous vascular disease signature.
</div>

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

`r what_this_does("Loaded analytical packages for scoring, correlation, and clinical profiling.", "Imported the pre-cleaned clinical master spine, numeric spine, data dictionary, and Olink proteomics matrix.", "Sourced the master omics engines for downstream DEA and Heatmap rendering.")`

```{r setup_and_import, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(gtsummary)
library(kableExtra)
library(ComplexHeatmap)
library(circlize)
library(writexl)
library(ggrepel) # Added for non-overlapping scatterplot labels

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "z02_sideproject_vascular_2"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data & Global DEA Results
master_spine  <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
numeric_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine_numeric.rds"))
data_dict     <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds")) # Adjust to .csv and read_csv() if your dict is a csv!
olink_mat     <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

# Load Global Vascular DEA (from Script 03b)
global_vascular_dea <- readRDS(here::here("output", "03b_proteomics_differential", "full_vascular_dea_results.rds"))
```

# Configurations
`r what_this_does("Defined the exact sets of proteins that comprise the Vascular Scoring system.", "Set mathematical thresholds for the Direction Count Score (+1/-1 logic).", "Parameterised the sample sizes for the Differential Expresion Analysis (Extreme N) and the Biobank Shopping List.", "Configured strict FDR and Rho thresholds for the Proteomic and Clinical correlation discovery engines.")`

```{r configurations}
# 1. Protein Signatures
vascular_up_proteins   <- c("VEGFA", "PGF", "CCL2", "CXCL10", "EDNRB") 
vascular_down_proteins <- c("NOS3", "CALCA", "ACE2", "HGF")

# Proteins of Interest for Endothelial/Myocardial specific heatmaps
proteins_of_interest   <- unique(c(vascular_up_proteins, vascular_down_proteins, "VWF", "TIE1", "TNNT2"))

# 2. Scoring Parameters
score_config <- list(
  z_thresh   = 1.0,  # Threshold for Direction Count Score (+1 or -1)
  extreme_n  = 5,    # Number of patients to select for Top and Bottom groups (for DEA)
  shopping_n = 10    # Number of patients to export for Biobank retrieval
)

# 3. Clinical Audit & Visualization Parameters
plot_color_vars <- c("vascular_active", "Inflamm_active", "cohort_group")
audit_vars      <- c("Age", "Sex", "log_Creatinine", "SSc Antibodies", "disease_duration")

# 4. Global Omics Thresholds (Required for DEA wrapper)
omics_config <- list(gsea_p_cutoff = 0.20, gsea_min_size = 3)
heatmap_config <- list(top_n = 30, split_by_group = TRUE, row_font_size = 8)

# 5. NEW: Correlation Discovery Thresholds
cor_config <- list(
  max_table_rows = 30, # How many top hits to display in the HTML tables
  proteomics = list(
    fdr_thresh = 0.05,
    rho_thresh = 0.4
  ),
  clinical = list(
    fdr_thresh = 0.10, # Slightly looser FDR for clinical traits due to lower N and higher variance
    rho_thresh = 0.3
  )
)
```

# Scores
`r analysis_decision("Dual-Metric Stringency: The Magnitude Score captures total biological intensity, while the Count Score captures systemic breadth (preventing a single massive outlier protein from hijacking the selection).", "Patients are ranked in both metrics.", "Ranks are summed and highest and lowest concordance ranks are selected.")`

## Vascular Signature Scoring
`r what_this_does("Calculated two independent vascular metrics.", "Magnitude Score: The sum of aligned Z-scores (Downregulated proteins are inverted so positive always means pro-disease).", "Direction Count Score: Assigns +1 if a protein crosses the biological threshold in the 'disease' direction, and -1 if it crosses in the 'healthy' direction.")`

```{r scoring}
# Clean lists to only include available proteins
valid_up   <- intersect(vascular_up_proteins, rownames(olink_mat))
valid_down <- intersect(vascular_down_proteins, rownames(olink_mat))

# Restrict to SSc patients only
ssc_patients <- master_spine %>% filter(cohort_group %in% c(GRP_ACTIVE, GRP_NON_ACTIVE)) %>% pull(Subject_ID)
valid_ssc_ids <- intersect(ssc_patients, colnames(olink_mat))

# 1. Z-Score Scaling (Scaled across SSc patients only)
z_mat <- t(scale(t(olink_mat[, valid_ssc_ids])))

# 2. Magnitude Score Calculation (Aligned: Up - Down)
mag_up   <- colSums(z_mat[valid_up, , drop=FALSE], na.rm=TRUE)
mag_down <- colSums(z_mat[valid_down, , drop=FALSE], na.rm=TRUE)
mag_score <- mag_up - mag_down

# 3. Direction Count Score (+1/-1 logic)
# UP list logic
up_plus  <- colSums(z_mat[valid_up, , drop=FALSE] > score_config$z_thresh, na.rm=TRUE)
up_minus <- colSums(z_mat[valid_up, , drop=FALSE] < -score_config$z_thresh, na.rm=TRUE)

# DOWN list logic (Inverted biologically)
down_plus  <- colSums(z_mat[valid_down, , drop=FALSE] < -score_config$z_thresh, na.rm=TRUE) # Low is bad
down_minus <- colSums(z_mat[valid_down, , drop=FALSE] > score_config$z_thresh, na.rm=TRUE)  # High is good

count_score <- (up_plus - up_minus) + (down_plus - down_minus)

# 4. Master Score DataFrame
score_df <- data.frame(
  Subject_ID = valid_ssc_ids,
  Magnitude_Score = mag_score,
  Count_Score = count_score
) %>%
  mutate(
    # Concordance Ranking (Lower rank = more extreme disease)
    Rank_Mag   = rank(-Magnitude_Score),
    Rank_Count = rank(-Count_Score),
    Concordance_Rank = Rank_Mag + Rank_Count
  ) %>%
  arrange(Concordance_Rank)

# Merge back with clinical data for plotting
plot_spine <- master_spine %>% inner_join(score_df, by = "Subject_ID")
```

## Score Distribution {.tabset}
`r what_this_does("Plotted the density distributions of the newly calculated Vascular Scores across the entire cohort.", "Color-coded the histograms by clinical metadata to visually confirm if the proteomic score maps to real-world clinical status.")`

```{r score_dist_and_val, results='asis'}
for (c_var in plot_color_vars) {
  if (!(c_var %in% colnames(plot_spine))) next
  
  cat(sprintf("\n\n### Colored by: %s\n\n", c_var))
  
  p_mag <- ggplot(plot_spine %>% filter(!is.na(!!sym(c_var))), aes(x = Magnitude_Score, fill = !!sym(c_var))) +
    geom_density(alpha = 0.5) +
    theme_project_base() +
    labs(title = "Magnitude Score", x = "Magnitude (Sum of aligned Z-scores)") +
    theme(plot.title = element_text(size = 11, face = "italic"))
  
  p_cnt <- ggplot(plot_spine %>% filter(!is.na(!!sym(c_var))), aes(x = Count_Score, fill = !!sym(c_var))) +
    geom_histogram(alpha = 1, position = "stack", bins = 15, color = "black") +
    theme_project_base() +
    labs(title = "Direction Count Score", x = "Count Score") +
    theme(plot.title = element_text(size = 11, face = "italic"))
  
  # Use patchwork's plot_annotation to create a master title for the whole figure
  master_plot <- (p_mag | p_cnt) + 
    plot_annotation(
      title = sprintf("Vascular Score Distributions (Colored by: %s)", c_var),
      theme = theme(plot.title = element_text(size = 15, face = "bold"))
    )
  
  print(master_plot)
  cat("\n\n")
}
```

## Score Drivers
`r what_this_does("Calculated the Spearman correlation between each scoring protein's raw NPX and both the final derived Magnitude Score and Count Score.", "Identifies which specific proteins are contributing the strongest biological signal to the overall patient ranking across both methodologies.")`

```{r primary_score_drivers}
all_scoring_prots <- unique(c(valid_up, valid_down))

# Helper function to extract Rho, P-value, and assign significance stars
get_cor_stats <- function(p, score_col) {
  res <- cor.test(olink_mat[p, plot_spine$Subject_ID], plot_spine[[score_col]], method = "spearman", exact = FALSE)
  stars <- case_when(
    res$p.value < 0.001 ~ "***",
    res$p.value < 0.01  ~ "**",
    res$p.value < 0.05  ~ "*",
    TRUE ~ "ns"
  )
  return(data.frame(Rho = res$estimate, P_Val = res$p.value, Stars = stars))
}

# Calculate for both scores
mag_res <- lapply(all_scoring_prots, function(p) get_cor_stats(p, "Magnitude_Score")) %>% bind_rows() %>% mutate(Protein = all_scoring_prots, Score_Type = "Magnitude Score")
cnt_res <- lapply(all_scoring_prots, function(p) get_cor_stats(p, "Count_Score")) %>% bind_rows() %>% mutate(Protein = all_scoring_prots, Score_Type = "Count Score")

driver_df <- bind_rows(mag_res, cnt_res) %>%
  mutate(Direction = ifelse(Protein %in% valid_up, "Upregulated List", "Downregulated List"))

order_prots <- driver_df %>% filter(Score_Type == "Magnitude Score") %>% arrange(Rho) %>% pull(Protein)
driver_df$Protein <- factor(driver_df$Protein, levels = order_prots)
driver_df$Score_Type <- factor(driver_df$Score_Type, levels = c("Magnitude Score", "Count Score"))

ggplot(driver_df, aes(x = Protein, y = Rho, fill = Direction)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = Stars, y = Rho + sign(Rho)*0.05), size = 4, vjust = 0.75) +
  coord_flip() +
  facet_wrap(~ Score_Type) +
  scale_fill_manual(values = c("Upregulated List" = "#c0392b", "Downregulated List" = "#2980b9")) +
  theme_project_base() +
  labs(title = "Primary Score Drivers", subtitle = "Spearman Correlation: Individual Protein NPX vs Final Scores", x = NULL, y = "Spearman Rho")
```

## Score Correlations
`r what_this_does("Extracted clinical variables labeled 'PASS' from the data dictionary.", "Computed Spearman correlations mapping every valid protein and clinical trait against the Magnitude Score, Count Score, and Inverted Rank.", "Applied Benjamini-Hochberg FDR correction.", "Generated scrollable HTML tables of the top drivers and exported full results to CSV.")`

`r why_is_this_done("Because the sickest patients have the lowest numerical Concordance Rank (e.g., #1), disease-promoting proteins would normally show a negative mathematical correlation. By computationally inverting the rank (multiplying by -1), a positive Rho uniformly signifies an association with worse vascular disease across all three metrics.")`

```{r score_correlations_tables, results='asis', warning=FALSE}
# 1. Prepare Data Matrices -----------------------------------------------------
valid_ssc_ids <- plot_spine$Subject_ID

# Prepare the Score DataFrame with the Inverted Rank
score_df_sub <- plot_spine %>% 
  select(Subject_ID, Magnitude_Score, Count_Score, Concordance_Rank) %>%
  mutate(Inverted_Rank = -1 * Concordance_Rank)

# Prepare Clinical Matrix (Filtered by Dictionary 'PASS')
pass_vars <- data_dict %>% filter(QC_Flag == "PASS") %>% pull(Variable)
clin_vars <- intersect(colnames(numeric_spine), pass_vars)
clin_vars <- setdiff(clin_vars, "Subject_ID")

clin_df <- numeric_spine %>% filter(Subject_ID %in% valid_ssc_ids) %>% select(Subject_ID, all_of(clin_vars))

# FIX: Safely move ID to row names, then purge any accidental text columns before transposing
clin_mat <- t(clin_df %>% 
                column_to_rownames("Subject_ID") %>% 
                select(where(is.numeric)))

# Prepare Proteomics Matrix
prot_mat <- olink_mat[, valid_ssc_ids]

# 2. Correlation Engine --------------------------------------------------------
run_mass_cor <- function(data_mat, score_vec) {
  res <- apply(data_mat, 1, function(x) {
    if(sum(!is.na(x)) < 10) return(c(Rho=NA, P_Val=NA)) # Skip features with < 10 valid data points
    ct <- cor.test(x, score_vec, method="spearman", exact=FALSE)
    # unname() strips the hidden 'rho' label so dplyr can find the columns
    c(Rho=unname(ct$estimate), P_Val=unname(ct$p.value))
  })
  
  df <- as.data.frame(t(res)) %>%
    rownames_to_column("Feature") %>%
    filter(!is.na(Rho)) %>%
    mutate(FDR = p.adjust(P_Val, method="BH")) %>%
    arrange(desc(abs(Rho)))
  return(df)
}

# 3. Execution & HTML Construction Loop ----------------------------------------
# We store results in a master list for the next chunk
cor_results <- list(proteomics = list(), clinical = list())
score_targets <- c("Magnitude_Score", "Count_Score", "Inverted_Rank")
modalities <- list(proteomics = prot_mat, clinical = clin_mat)

# HTML string accumulator to prevent fragmenting the output box (Knitting only)
html_out <- ""

for (mod_name in names(modalities)) {
  if (!interactive()) {
    html_out <- paste0(html_out, sprintf("\n\n### %s {.tabset}\n\n", tools::toTitleCase(mod_name)))
  }
  mod_conf <- cor_config[[mod_name]]
  
  for (score_name in score_targets) {
    clean_score_name <- gsub("_", " ", score_name)
    if (!interactive()) {
      html_out <- paste0(html_out, sprintf("\n\n#### %s\n\n", clean_score_name))
    }
    
    # Calculate
    score_vec <- setNames(score_df_sub[[score_name]], score_df_sub$Subject_ID)
    res_df <- run_mass_cor(modalities[[mod_name]], score_vec)
    cor_results[[mod_name]][[score_name]] <- res_df
    
    # Export full dataset
    export_name <- sprintf("Correlations_%s_vs_%s.csv", tools::toTitleCase(mod_name), score_name)
    write_csv(res_df, file.path(output_dir_data, export_name))
    
    # Filter for Significant Top Hits for Display
    sig_df <- res_df %>% filter(FDR < mod_conf$fdr_thresh & abs(Rho) >= mod_conf$rho_thresh)
    
    caption_text <- sprintf("<br><i>Showing top %d significant hits (FDR < %s, |Rho| >= %s). Full list exported to output folder.</i>\n\n", 
                            min(nrow(sig_df), cor_config$max_table_rows), mod_conf$fdr_thresh, mod_conf$rho_thresh)
    
    if (nrow(sig_df) > 0) {
      tbl_html <- sig_df %>%
        head(cor_config$max_table_rows) %>%
        kbl(format = "html", digits = 3, align = "c") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>% # FIX: Set full_width to TRUE
        scroll_box(height = "400px")
      
      # The Dual-Path Rendering Engine
      if (interactive()) {
        cat(sprintf("\n=== %s vs %s ===\n", tools::toTitleCase(mod_name), clean_score_name))
        print(tbl_html) # Forces RStudio Viewer to render
        cat(caption_text, "\n")
      } else {
        html_out <- paste0(html_out, tbl_html, "\n\n", caption_text)
      }
      
    } else {
      empty_msg <- "> No features passed the significance thresholds.\n\n"
      
      if (interactive()) {
        cat(sprintf("\n=== %s vs %s ===\n", tools::toTitleCase(mod_name), clean_score_name))
        cat(empty_msg)
      } else {
        html_out <- paste0(html_out, empty_msg)
      }
    }
  }
}

# Print the entire compiled HTML structure at once (Knitting only)
if (!interactive()) {
  cat(html_out)
}
```

## Driver Concordance
`r what_this_does("Cross-referenced the Magnitude and Count score correlation profiles for both omics modalities.", "Categorized drivers as 'Common' (significant in both), 'Unique to Magnitude', or 'Unique to Count' based on configuration thresholds.", "Generated labeled quadrant scatterplots to visualize the core biological signals shared across the scoring architectures.")`

```{r score_correlations_scatter, fig.width=14, fig.height=7}
# 1. Quadrant Plotting Engine --------------------------------------------------
plot_quadrant <- function(mag_df, count_df, conf, title_prefix) {
  
  # Join dataframes safely
  df <- mag_df %>%
    inner_join(count_df, by = "Feature", suffix = c("_Mag", "_Count")) %>%
    mutate(
      Sig_Mag = FDR_Mag < conf$fdr_thresh & abs(Rho_Mag) >= conf$rho_thresh,
      Sig_Count = FDR_Count < conf$fdr_thresh & abs(Rho_Count) >= conf$rho_thresh,
      Category = case_when(
        Sig_Mag & Sig_Count ~ "Common Driver",
        Sig_Mag ~ "Unique to Magnitude",
        Sig_Count ~ "Unique to Count",
        TRUE ~ "Not Significant"
      ),
      Label = ifelse(Category != "Not Significant", Feature, "")
    )
  
  cat_colors <- setNames(
    c("#c0392b", "#E69F00", "#0072B2", "grey80"),
    c("Common Driver", "Unique to Magnitude", "Unique to Count", "Not Significant")
  )
  
  p <- ggplot(df, aes(x = Rho_Mag, y = Rho_Count, fill = Category, label = Label)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(shape = 21, size = 3, alpha = 0.8, color = "black") +
    geom_text_repel(size = 3, max.overlaps = 15, box.padding = 0.5, segment.color = "grey50") +
    scale_fill_manual(values = cat_colors) +
    theme_project_base() +
    labs(
      title = sprintf("%s: Driver Concordance", title_prefix),
      subtitle = sprintf("Thresholds: FDR < %s, |Rho| >= %s", conf$fdr_thresh, conf$rho_thresh),
      x = "Magnitude Score (Spearman Rho)",
      y = "Direction Count Score (Spearman Rho)"
    )
  
  return(p)
}

# 2. Generate and Save Plots ---------------------------------------------------
p_prot <- plot_quadrant(
  mag_df = cor_results$proteomics$Magnitude_Score, 
  count_df = cor_results$proteomics$Count_Score, 
  conf = cor_config$proteomics, 
  title_prefix = "Proteomics"
)

p_clin <- plot_quadrant(
  mag_df = cor_results$clinical$Magnitude_Score, 
  count_df = cor_results$clinical$Count_Score, 
  conf = cor_config$clinical, 
  title_prefix = "Clinical Parameters"
)

# Print side-by-side
print(p_prot | p_clin)

# Save to disk
ggsave(file.path(output_dir_data, "Quadrant_Proteomics.png"), p_prot, width = 8, height = 7, bg = "white")
ggsave(file.path(output_dir_data, "Quadrant_Clinical.png"), p_clin, width = 8, height = 7, bg = "white")

# Memory Cleanup (Delete variables no longer needed outside these chunks)
rm(clin_mat, prot_mat, score_df_sub, clin_df, cor_results, html_out)
```

# Patients
## Patient Concordance & Selection
`r what_this_does("Ranked patients based on their aggregate performance across both the Magnitude and Direction Count scoring systems.", "Designated the Top N and Bottom N patients as the extreme vascular phenotypes.", "Generated a scatterplot with overlaid linear regression to visually highlight the selected patients in the extreme corners of the concordance distribution.")`

```{r patient_concordance_and_selection, results='asis'}
# Define the Extreme Groups based on Concordance Rank
top_patients <- plot_spine %>% slice_min(Concordance_Rank, n = score_config$extreme_n) %>% pull(Subject_ID)
bot_patients <- plot_spine %>% slice_max(Concordance_Rank, n = score_config$extreme_n) %>% pull(Subject_ID)

lbl_unselected <- "Unselected"
lbl_top <- sprintf("Top %d (Vascular High)", score_config$extreme_n)
lbl_bot <- sprintf("Bottom %d (Vascular Low)", score_config$extreme_n)

plot_spine <- plot_spine %>%
  mutate(Selection = case_when(
    Subject_ID %in% top_patients ~ lbl_top,
    Subject_ID %in% bot_patients ~ lbl_bot,
    TRUE ~ lbl_unselected
  ))

selection_colors <- setNames(c("grey80", "#c0392b", "#2980b9"), c(lbl_unselected, lbl_top, lbl_bot))

p_scatter <- ggplot(plot_spine, aes(x = Magnitude_Score, y = Count_Score, fill = Selection)) +
  geom_point(shape = 21, size = 3, alpha = 0.8, color = "black") +
  scale_fill_manual(values = selection_colors) + 
  theme_project_base() +
  labs(title = "Score Concordance & Sera Selection", 
       subtitle = sprintf("Selecting top and bottom %d patients based on combined ranking", score_config$extreme_n),
       x = "Magnitude Score (Z-Score Sum)", y = "Count Score (+1/-1)")

print(p_scatter)
ggsave(file.path(output_dir_data, "Patient_Selection_Concordance.png"), p_scatter, width=8, height=6, bg="white")


tbl_data <- plot_spine %>% 
  filter(Selection != lbl_unselected) %>%
  select(Subject_ID, Selection, Magnitude_Score, Count_Score, Concordance_Rank) %>%
  arrange(Concordance_Rank)

# Print cleanly for both inline RStudio and Knitted HTML
tbl_data %>% 
  kbl(format = "html", digits = 2, align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left") %>%
  row_spec(0, bold = TRUE)
```


## Clinical Confounder Audit
`r what_this_does("Generates a cross-reference table of the selected extreme patients.", "Allows wet-lab experimenters to check if the 'High Vascular' serum happens to be systematically older, uremic (Creatinine), or tied to a specific autoantibody, which could artificially influence the organoids.")`

```{r clinical_confounder_audit}
audit_cols <- intersect(audit_vars, colnames(plot_spine))

audit_df <- plot_spine %>%
  filter(Selection != "Unselected") %>%
  select(Selection, all_of(audit_cols))

continuous_force_vars <- intersect(c("Age", "log_Creatinine", "disease_duration"), colnames(audit_df))

suppressMessages(suppressWarnings({
  tbl_audit <- audit_df %>%
    tbl_summary(
      by = Selection, 
      missing = "no",
      type = list(any_of(continuous_force_vars) ~ "continuous")
    ) %>%
    add_p() %>%
    bold_labels() %>%
    modify_header(label = "**Clinical Confounder**")
}))

tbl_audit %>% as_kable_extra(format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```


## Differential Expression Analysis {.tabset}
`r what_this_does("Runs the master Limma Differential Expression Analysis specifically on the Top N vs Bottom N selected patients.", "Generates standard Volcano plots and heatmaps to visually validate the biological extreme phenotype of the chosen sera.")`

```{r DEA, fig.height=8, fig.width=10, results='asis'}
# 1. Setup Custom DEA Config for the Wrapper
extreme_dea_conf <- list(
  title = sprintf("Organoid Candidates: Top %d vs Bottom %d", score_config$extreme_n, score_config$extreme_n),
  group_col = "Selection",
  target_groups = c(sprintf("Top %d (Vascular High)", score_config$extreme_n)),
  ref_groups = c(sprintf("Bottom %d (Vascular Low)", score_config$extreme_n)),
  is_continuous = FALSE,
  covariates = c(),
  dea_p_cutoff = 0.05,
  dea_fc_cutoff = 0.5,
  custom_color_map = selection_colors # <--- Pass custom colors safely
)

# 2. Run DEA Engine
dea_res <- wrap_dea_pipeline(
    omics_mat           = olink_mat, 
    clin_df             = plot_spine, 
    dea_conf            = extreme_dea_conf, 
    dea_id              = "extreme_vascular", 
    base_output_dir     = output_dir_data, 
    go_db               = NULL, 
    omics_config        = omics_config, 
    heatmap_config      = heatmap_config, 
    project_colors_func = get_project_colors
)

clin_extreme <- plot_spine %>% filter(Selection != "Unselected")

# HEATMAP 1: Local DEA (Extreme Phenotype)
cat("## Local DEA Heatmap\n\n")
if (dea_res$status == "success") {
  print(dea_res$plots$volcano)
  cat("\n\n<br>\n\n")
  
  hm_local <- run_heatmap_engine(
    mat = olink_mat, 
    clin = clin_extreme, 
    dea_res = dea_res$data$dea,
    title = "Local DEA (5v5)",
    n_top = 30, 
    split_by_group = FALSE, # Unsupervised clustering
    split_col = "Selection",
    custom_color_map = selection_colors
  )
  draw(hm_local)
  cat("\n\n")
} else {
  cat(sprintf("> <span style='color:red;'>**DEA Failed:** %s</span>\n\n", dea_res$error_msg))
}

# HEATMAP 2: Global DEA Signature
cat("## Global DEA Heatmap\n\n")
if (exists("global_vascular_dea") && !is.null(global_vascular_dea)) {
  hm_global <- run_heatmap_engine(
    mat = olink_mat, 
    clin = clin_extreme, 
    dea_res = global_vascular_dea,
    title = "Global DEA (full cohort)",
    n_top = 30, 
    split_by_group = FALSE,
    split_col = "Selection",
    custom_color_map = selection_colors
  )
  draw(hm_global)
  cat("\n\n")
} else {
  cat("> Global DEA results not found. Skipping Global Heatmap.\n\n")
}

# HEATMAP 3: Custom Proteins of Interest
cat("## Custom POI Heatmap\n\n")
valid_poi <- intersect(proteins_of_interest, rownames(olink_mat))
if (length(valid_poi) > 1 && dea_res$status == "success") {
  
  poi_limma_df <- dea_res$data$dea %>% filter(Feature %in% valid_poi)
  
  poi_heatmap <- run_heatmap_engine(
    mat = olink_mat, 
    clin = clin_extreme, 
    dea_res = poi_limma_df,
    title = "Selected Proteins of Interest (5v5 DEA)",
    n_top = length(valid_poi), 
    split_by_group = FALSE,
    split_col = "Selection",
    show_top_n_title = FALSE, # Suppress the "Top 9" text
    custom_color_map = selection_colors
  )
  draw(poi_heatmap)
  cat("\n\n")
} else {
  cat("> Insufficient Valid POIs to generate targeted heatmap.\n\n")
}
```

## Biobank Shopping List Export
`r what_this_does("Exports an extended priority list of extreme patients to CSV, providing the biobank with backup options in case primary aliquots are depleted.", "Prints the shopping list to the HTML report for quick reference.")`

```{r biobank_shopping_list_export}
# Select only the needed columns FIRST to prevent duplicate column warnings
shopping_top <- plot_spine %>% 
  slice_min(Concordance_Rank, n = score_config$shopping_n) %>% 
  mutate(Target_Group = "High Vascular Serum") %>%
  select(Target_Group, Subject_ID, Concordance_Rank, Magnitude_Score, Count_Score)

shopping_bot <- plot_spine %>% 
  slice_max(Concordance_Rank, n = score_config$shopping_n) %>% 
  mutate(Target_Group = "Low Vascular Serum") %>%
  select(Target_Group, Subject_ID, Concordance_Rank, Magnitude_Score, Count_Score)

# Bind clean rows
shopping_list <- bind_rows(shopping_top, shopping_bot) %>%
  arrange(Target_Group, Concordance_Rank)

export_path <- file.path(output_dir_data, "Biobank_Organoid_Shopping_List.xlsx")
write_xlsx(shopping_list, export_path)

# Print the Table to HTML
shopping_list %>% 
  kbl(format = "html", digits = 2, align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  row_spec(0, bold = TRUE)
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script z02 Complete! \n")
```

