---
title: "Sideproject: Vascular 2"
author: "alexander-gysin (alexander.gysin@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: inline
---

# Introduction
Goal: To identify SSc patients with the most extreme "vascular disease" proteomic signatures to select their serum for functional in vitro organoid stimulation.

<div style='border: 2px solid #e74c3c; padding: 15px; border-radius: 5px; background-color: #fdf2e9; margin-bottom: 20px;'>
<h4 style='margin-top: 0; color: #c0392b;'>⚠️ NOTE: Medication Confounders</h4>
The computational pipeline currently lacks patient medication histories . If organoids exhibit divergent behavior, consider that they may be responding to circulating pharmaceutical agents in the serum rather than the endogenous vascular disease signature.
</div>

```{r setup_boxes, include=FALSE}
source(here::here("code", "00_config.R"))
```

# Setup and Import

`r what_this_does("Loaded analytical packages for scoring, correlation, and clinical profiling.", "Imported the pre-cleaned clinical master spine, numeric spine, data dictionary, and Olink proteomics matrix.", "Sourced the master omics engines for downstream DEA and Heatmap rendering.")`

```{r setup_and_import, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(gtsummary)
library(kableExtra)
library(ComplexHeatmap)
library(circlize)
library(writexl)
library(ggrepel)
library(msigdbr)

# Knitr Config
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Output paths
current_file <- "z02_sideproject_vascular_2"
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data, recursive = TRUE)

# Source Engines
source(here::here("code", "00_omics_engines.R"))

# Load Pre-Cleaned Data & Global DEA Results
master_spine  <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine.rds"))
numeric_spine <- readRDS(here::here("output", "01b_clinical_feature_engineering", "curated_clinical_spine_numeric.rds"))
data_dict     <- readRDS(here::here("output", "01b_clinical_feature_engineering", "clinical_data_dictionary.rds"))
olink_mat     <- readRDS(here::here("output", "01a_data_import_and_qc", "clean_olink_matrix.rds"))

# Load Global Vascular DEA (from Script 03b)
global_vascular_dea <- readRDS(here::here("output", "03b_proteomics_differential", "full_vascular_dea_results.rds"))

# GSEA Datasets
pathway_dbs <- list(
  "Hallmark" = msigdbr(species = "Homo sapiens", category = "H") %>% select(gs_name, gene_symbol),
  "GO_BP" = msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP") %>% select(gs_name, gene_symbol),
  "Immunologic_Signature" = msigdbr(species = "Homo sapiens", category = "C7") %>% select(gs_name, gene_symbol),
  "Reactome" = msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME") %>% select(gs_name, gene_symbol)
)
```

# Configurations
`r what_this_does("Defined the exact sets of proteins that comprise the Vascular Scoring system.", "Set mathematical thresholds for the Direction Count Score (+1/-1 logic).", "Parameterised the sample sizes for the Differential Expresion Analysis (Extreme N) and the Biobank Shopping List.", "Configured strict FDR and Rho thresholds for the Proteomic and Clinical correlation discovery engines.")`

```{r configurations}
# 1. Protein Signatures
vascular_up_proteins   <- c("VEGFA", "PGF", "CCL2", "CXCL10", "EDNRB") 
vascular_down_proteins <- c("NOS3", "CALCA", "ACE2", "HGF")

# Proteins of Interest for Endothelial/Myocardial specific heatmaps
proteins_of_interest   <- unique(c(vascular_up_proteins, vascular_down_proteins, "VWF", "TIE1", "TNNT2"))

# 2. Scoring Parameters
score_config <- list(
  z_thresh = 1.0  # Threshold for Direction Count Score (+1 or -1)
)

# 3. MANUAL COHORT SELECTION
manual_cohort <- list(
  pool_20 = c("Act_7", "Act_33", "Act_44", "Nonact_24", "Nonact_15", 
              "Nonact_20", "Act_16", "Nonact_48", "Act_24", "Nonact_45", 
              "Nonact_22", "Nonact_34", "Nonact_31", "Nonact_30", "Nonact_35", 
              "Nonact_37", "Nonact_46", "Act_35", "Nonact_2", "Nonact_36", "Act_46", "Act_8", "Act_37", "Act_6", "Act_33", "Nonact_20", "Act_16", "Nonact_36"),
              
  high_4  = c("Act_46", "Act_8", "Act_37", "Act_6"), 
  low_4   = c("Act_33", "Nonact_20", "Act_16", "Nonact_36")      
)

# 4. Clinical Audit & Visualization Parameters
plot_color_vars <- c("vascular_active", "Inflamm_active", "cohort_group")
audit_vars      <- c("Age", "Sex", "log_Creatinine", "SSc Antibodies")

# 5. Global Omics Thresholds
omics_config <- list(
  gsea_p_cutoff    = 0.20, 
  gsea_min_size    = 3,
  network_p_cutoff = 0.10  
)
heatmap_config <- list(top_n = 30, split_by_group = TRUE, row_font_size = 8)

# 6. Correlation Discovery Thresholds
cor_config <- list(
  max_table_rows = 30,
  proteomics = list(fdr_thresh = 0.05, rho_thresh = 0.4),
  clinical = list(fdr_thresh = 0.10, rho_thresh = 0.3)
)
```

# Scores
`r analysis_decision("Dual-Metric Stringency: The Magnitude Score captures total biological intensity, while the Count Score captures systemic breadth (preventing a single massive outlier protein from hijacking the selection).", "Patients are ranked in both metrics.", "Ranks are summed and highest and lowest concordance ranks are selected.")`

## Vascular Signature Scoring
`r what_this_does("Calculated two independent vascular metrics.", "Magnitude Score: The sum of aligned Z-scores (Downregulated proteins are inverted so positive always means pro-disease).", "Direction Count Score: Assigns +1 if a protein crosses the biological threshold in the 'disease' direction, and -1 if it crosses in the 'healthy' direction.")`

```{r scoring}
# Clean lists to only include available proteins
valid_up   <- intersect(vascular_up_proteins, rownames(olink_mat))
valid_down <- intersect(vascular_down_proteins, rownames(olink_mat))

# Restrict to SSc patients only
ssc_patients <- master_spine %>% filter(cohort_group %in% c(GRP_ACTIVE, GRP_NON_ACTIVE)) %>% pull(Subject_ID)
valid_ssc_ids <- intersect(ssc_patients, colnames(olink_mat))

# 1. Z-Score Scaling (Scaled across SSc patients only)
z_mat <- t(scale(t(olink_mat[, valid_ssc_ids])))

# 2. Magnitude Score Calculation (Aligned: Up - Down)
mag_up   <- colSums(z_mat[valid_up, , drop=FALSE], na.rm=TRUE)
mag_down <- colSums(z_mat[valid_down, , drop=FALSE], na.rm=TRUE)
mag_score <- mag_up - mag_down

# 3. Direction Count Score (+1/-1 logic)
# UP list logic
up_plus  <- colSums(z_mat[valid_up, , drop=FALSE] > score_config$z_thresh, na.rm=TRUE)
up_minus <- colSums(z_mat[valid_up, , drop=FALSE] < -score_config$z_thresh, na.rm=TRUE)

# DOWN list logic (Inverted biologically)
down_plus  <- colSums(z_mat[valid_down, , drop=FALSE] < -score_config$z_thresh, na.rm=TRUE) # Low is bad
down_minus <- colSums(z_mat[valid_down, , drop=FALSE] > score_config$z_thresh, na.rm=TRUE)  # High is good

count_score <- (up_plus - up_minus) + (down_plus - down_minus)

# 4. Master Score DataFrame
score_df <- data.frame(
  Subject_ID = valid_ssc_ids,
  Magnitude_Score = mag_score,
  Count_Score = count_score
) %>%
  mutate(
    # Concordance Ranking (Lower rank = more extreme disease)
    Rank_Mag   = rank(-Magnitude_Score),
    Rank_Count = rank(-Count_Score),
    Concordance_Rank = Rank_Mag + Rank_Count
  ) %>%
  arrange(Concordance_Rank)

# Merge back with clinical data for plotting
plot_spine <- master_spine %>% inner_join(score_df, by = "Subject_ID")
```

## Score Distribution {.tabset}
`r what_this_does("Plotted the density distributions of the newly calculated Vascular Scores across the entire cohort.", "Color-coded the histograms by clinical metadata to visually confirm if the proteomic score maps to real-world clinical status.")`

```{r score_dist_and_val, results='asis'}
for (c_var in plot_color_vars) {
  if (!(c_var %in% colnames(plot_spine))) next
  
  cat(sprintf("\n\n### Colored by: %s\n\n", c_var))
  
  p_mag <- ggplot(plot_spine %>% filter(!is.na(!!sym(c_var))), aes(x = Magnitude_Score, fill = !!sym(c_var))) +
    geom_density(alpha = 0.5) +
    theme_project_base() +
    labs(title = "Magnitude Score", x = "Magnitude (Sum of aligned Z-scores)") +
    theme(plot.title = element_text(size = 11, face = "italic"))
  
  p_cnt <- ggplot(plot_spine %>% filter(!is.na(!!sym(c_var))), aes(x = Count_Score, fill = !!sym(c_var))) +
    geom_histogram(alpha = 1, position = "stack", bins = 15, color = "black") +
    theme_project_base() +
    labs(title = "Direction Count Score", x = "Count Score") +
    theme(plot.title = element_text(size = 11, face = "italic"))
  
  # Use patchwork's plot_annotation to create a master title for the whole figure
  master_plot <- (p_mag | p_cnt) + 
    plot_annotation(
      title = sprintf("Vascular Score Distributions (Colored by: %s)", c_var),
      theme = theme(plot.title = element_text(size = 15, face = "bold"))
    )
  
  print(master_plot)
  cat("\n\n")
}
```

## Score Drivers
`r what_this_does("Calculated the Spearman correlation between each scoring protein's raw NPX and both the final derived Magnitude Score and Count Score.", "Identifies which specific proteins are contributing the strongest biological signal to the overall patient ranking across both methodologies.")`

```{r primary_score_drivers}
all_scoring_prots <- unique(c(valid_up, valid_down))

# Helper function to extract Rho, P-value, and assign significance stars
get_cor_stats <- function(p, score_col) {
  res <- cor.test(olink_mat[p, plot_spine$Subject_ID], plot_spine[[score_col]], method = "spearman", exact = FALSE)
  stars <- case_when(
    res$p.value < 0.001 ~ "***",
    res$p.value < 0.01  ~ "**",
    res$p.value < 0.05  ~ "*",
    TRUE ~ "ns"
  )
  return(data.frame(Rho = res$estimate, P_Val = res$p.value, Stars = stars))
}

# Calculate for both scores
mag_res <- lapply(all_scoring_prots, function(p) get_cor_stats(p, "Magnitude_Score")) %>% bind_rows() %>% mutate(Protein = all_scoring_prots, Score_Type = "Magnitude Score")
cnt_res <- lapply(all_scoring_prots, function(p) get_cor_stats(p, "Count_Score")) %>% bind_rows() %>% mutate(Protein = all_scoring_prots, Score_Type = "Count Score")

driver_df <- bind_rows(mag_res, cnt_res) %>%
  mutate(Direction = ifelse(Protein %in% valid_up, "Upregulated List", "Downregulated List"))

order_prots <- driver_df %>% filter(Score_Type == "Magnitude Score") %>% arrange(Rho) %>% pull(Protein)
driver_df$Protein <- factor(driver_df$Protein, levels = order_prots)
driver_df$Score_Type <- factor(driver_df$Score_Type, levels = c("Magnitude Score", "Count Score"))

ggplot(driver_df, aes(x = Protein, y = Rho, fill = Direction)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = Stars, y = Rho + sign(Rho)*0.05), size = 4, vjust = 0.75) +
  coord_flip() +
  facet_wrap(~ Score_Type) +
  scale_fill_manual(values = c("Upregulated List" = "#c0392b", "Downregulated List" = "#2980b9")) +
  theme_project_base() +
  labs(title = "Primary Score Drivers", subtitle = "Spearman Correlation: Individual Protein NPX vs Final Scores", x = NULL, y = "Spearman Rho")
```

## Score Correlations
`r what_this_does("Extracted clinical variables labeled 'PASS' from the data dictionary.", "Computed Spearman correlations mapping every valid protein and clinical trait against the Magnitude Score, Count Score, and Inverted Rank.", "Applied Benjamini-Hochberg FDR correction.", "Generated scrollable HTML tables of the top drivers and exported full results to CSV.")`

`r why_is_this_done("Because the sickest patients have the lowest numerical Concordance Rank (e.g., #1), disease-promoting proteins would normally show a negative mathematical correlation. By computationally inverting the rank (multiplying by -1), a positive Rho uniformly signifies an association with worse vascular disease across all three metrics.")`

```{r score_correlations_tables, results='asis', warning=FALSE}
# 1. Prepare Data Matrices -----------------------------------------------------
valid_ssc_ids <- plot_spine$Subject_ID

# Prepare the Score DataFrame with the Inverted Rank
score_df_sub <- plot_spine %>% 
  select(Subject_ID, Magnitude_Score, Count_Score, Concordance_Rank) %>%
  mutate(Inverted_Rank = -1 * Concordance_Rank)

# Prepare Clinical Matrix (Filtered by Dictionary 'PASS')
pass_vars <- data_dict %>% filter(QC_Flag == "PASS") %>% pull(Variable)
clin_vars <- intersect(colnames(numeric_spine), pass_vars)
clin_vars <- setdiff(clin_vars, "Subject_ID")

clin_df <- numeric_spine %>% filter(Subject_ID %in% valid_ssc_ids) %>% select(Subject_ID, all_of(clin_vars))

# Safely move ID to row names, then purge any accidental text columns before transposing
clin_mat <- t(clin_df %>% 
                column_to_rownames("Subject_ID") %>% 
                select(where(is.numeric)))

# Prepare Proteomics Matrix
prot_mat <- olink_mat[, valid_ssc_ids]

# 2. Correlation Engine --------------------------------------------------------
run_mass_cor <- function(data_mat, score_vec) {
  res <- apply(data_mat, 1, function(x) {
    if(sum(!is.na(x)) < 10) return(c(Rho=NA, P_Val=NA)) # Skip features with < 10 valid data points
    ct <- cor.test(x, score_vec, method="spearman", exact=FALSE)
    # unname() strips the hidden 'rho' label so dplyr can find the columns
    c(Rho=unname(ct$estimate), P_Val=unname(ct$p.value))
  })
  
  df <- as.data.frame(t(res)) %>%
    rownames_to_column("Feature") %>%
    filter(!is.na(Rho)) %>%
    mutate(FDR = p.adjust(P_Val, method="BH")) %>%
    arrange(desc(abs(Rho)))
  return(df)
}

# 3. Execution & HTML Construction Loop ----------------------------------------
# We store results in a master list for the next chunk
cor_results <- list(proteomics = list(), clinical = list())
score_targets <- c("Magnitude_Score", "Count_Score", "Inverted_Rank")
modalities <- list(proteomics = prot_mat, clinical = clin_mat)

# HTML string accumulator to prevent fragmenting the output box (Knitting only)
html_out <- ""

for (mod_name in names(modalities)) {
  if (!interactive()) {
    html_out <- paste0(html_out, sprintf("\n\n### %s {.tabset}\n\n", tools::toTitleCase(mod_name)))
  }
  mod_conf <- cor_config[[mod_name]]
  
  for (score_name in score_targets) {
    clean_score_name <- gsub("_", " ", score_name)
    if (!interactive()) {
      html_out <- paste0(html_out, sprintf("\n\n#### %s\n\n", clean_score_name))
    }
    
    # Calculate
    score_vec <- setNames(score_df_sub[[score_name]], score_df_sub$Subject_ID)
    res_df <- run_mass_cor(modalities[[mod_name]], score_vec)
    cor_results[[mod_name]][[score_name]] <- res_df
    
    # Export full dataset
    export_name <- sprintf("Correlations_%s_vs_%s.csv", tools::toTitleCase(mod_name), score_name)
    write_csv(res_df, file.path(output_dir_data, export_name))
    
    # Filter for Significant Top Hits for Display
    sig_df <- res_df %>% filter(FDR < mod_conf$fdr_thresh & abs(Rho) >= mod_conf$rho_thresh)
    
    caption_text <- sprintf("<br><i>Showing top %d significant hits (FDR < %s, |Rho| >= %s). Full list exported to output folder.</i>\n\n", 
                            min(nrow(sig_df), cor_config$max_table_rows), mod_conf$fdr_thresh, mod_conf$rho_thresh)
    
    if (nrow(sig_df) > 0) {
      tbl_html <- sig_df %>%
        head(cor_config$max_table_rows) %>%
        kbl(format = "html", digits = 3, align = "c") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>% # FIX: Set full_width to TRUE
        scroll_box(height = "400px")
      
      # The Dual-Path Rendering Engine
      if (interactive()) {
        cat(sprintf("\n=== %s vs %s ===\n", tools::toTitleCase(mod_name), clean_score_name))
        print(tbl_html) # Forces RStudio Viewer to render
        cat(caption_text, "\n")
      } else {
        html_out <- paste0(html_out, tbl_html, "\n\n", caption_text)
      }
      
    } else {
      empty_msg <- "> No features passed the significance thresholds.\n\n"
      
      if (interactive()) {
        cat(sprintf("\n=== %s vs %s ===\n", tools::toTitleCase(mod_name), clean_score_name))
        cat(empty_msg)
      } else {
        html_out <- paste0(html_out, empty_msg)
      }
    }
  }
}

# Print the entire compiled HTML structure at once (Knitting only)
if (!interactive()) {
  cat(html_out)
}
```

## Driver Concordance
`r what_this_does("Cross-referenced the Magnitude and Count score correlation profiles for both omics modalities.", "Categorized drivers as 'Common' (significant in both), 'Unique to Magnitude', or 'Unique to Count' based on configuration thresholds.", "Generated labeled quadrant scatterplots to visualize the core biological signals shared across the scoring architectures.")`

```{r score_correlations_scatter, fig.width=14, fig.height=7}
# 1. Quadrant Plotting Engine --------------------------------------------------
plot_quadrant <- function(mag_df, count_df, conf, title_prefix) {
  
  # Join dataframes safely
  df <- mag_df %>%
    inner_join(count_df, by = "Feature", suffix = c("_Mag", "_Count")) %>%
    mutate(
      Sig_Mag = FDR_Mag < conf$fdr_thresh & abs(Rho_Mag) >= conf$rho_thresh,
      Sig_Count = FDR_Count < conf$fdr_thresh & abs(Rho_Count) >= conf$rho_thresh,
      Category = case_when(
        Sig_Mag & Sig_Count ~ "Common Driver",
        Sig_Mag ~ "Unique to Magnitude",
        Sig_Count ~ "Unique to Count",
        TRUE ~ "Not Significant"
      ),
      Label = ifelse(Category != "Not Significant", Feature, "")
    )
  
  cat_colors <- setNames(
    c("#c0392b", "#E69F00", "#0072B2", "grey80"),
    c("Common Driver", "Unique to Magnitude", "Unique to Count", "Not Significant")
  )
  
  p <- ggplot(df, aes(x = Rho_Mag, y = Rho_Count, fill = Category, label = Label)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(shape = 21, size = 3, alpha = 0.8, color = "black") +
    geom_text_repel(size = 3, max.overlaps = 15, box.padding = 0.5, segment.color = "grey50") +
    scale_fill_manual(values = cat_colors) +
    theme_project_base() +
    labs(
      title = sprintf("%s: Driver Concordance", title_prefix),
      subtitle = sprintf("Thresholds: FDR < %s, |Rho| >= %s", conf$fdr_thresh, conf$rho_thresh),
      x = "Magnitude Score (Spearman Rho)",
      y = "Direction Count Score (Spearman Rho)"
    )
  
  return(p)
}

# 2. Generate and Save Plots ---------------------------------------------------
p_prot <- plot_quadrant(
  mag_df = cor_results$proteomics$Magnitude_Score, 
  count_df = cor_results$proteomics$Count_Score, 
  conf = cor_config$proteomics, 
  title_prefix = "Proteomics"
)

p_clin <- plot_quadrant(
  mag_df = cor_results$clinical$Magnitude_Score, 
  count_df = cor_results$clinical$Count_Score, 
  conf = cor_config$clinical, 
  title_prefix = "Clinical Parameters"
)

# Print side-by-side
print(p_prot | p_clin)

# Save to disk
ggsave(file.path(output_dir_data, "Quadrant_Proteomics.png"), p_prot, width = 8, height = 7, bg = "white")
ggsave(file.path(output_dir_data, "Quadrant_Clinical.png"), p_clin, width = 8, height = 7, bg = "white")

# Memory Cleanup (Delete variables no longer needed outside these chunks)
rm(clin_mat, prot_mat, score_df_sub, clin_df, cor_results, html_out)
```

# Patients
## Patient Concordance & Selection
`r what_this_does("Tagged patients based on the predefined manual cohort lists (Pool of 20, High 4, Low 4).", "Generated a scatterplot to visually confirm where the hand-picked patients sit relative to the global mathematical extremes of the cohort.", "Printed a formatted table of the 20 selected patients displaying their global vascular scores and ranks.")`

```{r patient_concordance_and_selection, results='asis'}
# Define Labels
lbl_unselected <- "Global SSc Cohort"
lbl_pool       <- "Selected Pool (n=20)"
lbl_top        <- "Target: Vascular High (n=4)"
lbl_bot        <- "Target: Vascular Low (n=4)"

# Tag Patients
plot_spine <- plot_spine %>%
  mutate(Selection = case_when(
    Subject_ID %in% manual_cohort$high_4 ~ lbl_top,
    Subject_ID %in% manual_cohort$low_4 ~ lbl_bot,
    Subject_ID %in% manual_cohort$pool_20 ~ lbl_pool,
    TRUE ~ lbl_unselected
  )) %>%
  # Factor to control plotting order (so bold dots plot on top of grey ones)
  mutate(Selection = factor(Selection, levels = c(lbl_unselected, lbl_pool, lbl_top, lbl_bot)))

# Assign Colors
selection_colors <- setNames(c("grey90", "#F39C12", "#c0392b", "#2980b9"), 
                             c(lbl_unselected, lbl_pool, lbl_top, lbl_bot))

# 1. Scatterplot
p_scatter <- ggplot(plot_spine %>% arrange(Selection), aes(x = Magnitude_Score, y = Count_Score, fill = Selection)) +
  geom_point(shape = 21, size = 3, alpha = 0.8, color = "black") +
  scale_fill_manual(values = selection_colors) + 
  theme_project_base() +
  labs(title = "Score Concordance: Manual Sera Selection", 
       subtitle = "Visualizing the manually selected cohort within the global SSc distribution",
       x = "Magnitude Score (Z-Score Sum)", y = "Count Score (+1/-1)")

print(p_scatter)
ggsave(file.path(output_dir_data, "Manual_Selection_Concordance.png"), p_scatter, width=8, height=6, bg="white")
cat("\n\n")

# 2. Print Table for the 20 Pool
tbl_data <- plot_spine %>% 
  filter(Subject_ID %in% manual_cohort$pool_20) %>%
  select(Subject_ID, Selection, Magnitude_Score, Count_Score, Concordance_Rank) %>%
  arrange(Concordance_Rank)

# Render clean HTML table
tbl_data %>% 
  kbl(format = "html", digits = 2, align = "c", caption = "Global Scores for the 20 Manually Selected Patients") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left") %>%
  row_spec(0, bold = TRUE)
```

## Clinical Confounder Audit
`r what_this_does("Filtered the cohort strictly to the 8 extreme patients (4 High vs 4 Low).", "Generated a cross-reference table to audit potential clinical confounders (Age, Sex, Renal Function) prior to in vitro experiments.")`

```{r clinical_confounder_audit}
audit_cols <- intersect(audit_vars, colnames(plot_spine))

# Strictly filter to the final 8
audit_df <- plot_spine %>%
  filter(Selection %in% c("Target: Vascular High (n=4)", "Target: Vascular Low (n=4)")) %>%
  select(Selection, all_of(audit_cols)) %>%
  droplevels()

continuous_force_vars <- intersect(c("Age", "log_Creatinine", "disease_duration"), colnames(audit_df))

suppressMessages(suppressWarnings({
  tbl_audit <- audit_df %>%
    tbl_summary(
      by = Selection, 
      missing = "no",
      type = list(any_of(continuous_force_vars) ~ "continuous")
    ) %>%
    add_p() %>%
    bold_labels() %>%
    modify_header(label = "**Clinical Confounder**")
}))

tbl_audit %>% as_kable_extra(format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, position = "left")
```


## Differential Expression Analysis {.tabset}
`r what_this_does("Ran the master Limma DEA strictly comparing the 4 Manual High vs 4 Manual Low patients.", "Generated standard Volcano plots and heatmaps restricted to the 8 targeted patients to visually validate the biological extreme phenotype of the chosen sera.")`

```{r DEA, fig.height=8, fig.width=10, results='asis'}
# 1. Restrict data to the final 8 patients
clin_extreme <- plot_spine %>% filter(Selection %in% c("Target: Vascular High (n=4)", "Target: Vascular Low (n=4)")) %>% droplevels()

# 2. Setup Custom DEA Config for the Wrapper
extreme_dea_conf <- list(
  title = "Organoid Candidates: Manual High vs Manual Low",
  group_col = "Selection",
  target_groups = c("Target: Vascular High (n=4)"),
  ref_groups = c("Target: Vascular Low (n=4)"),
  is_continuous = FALSE,
  covariates = c(),
  dea_p_cutoff = 0.05,
  dea_fc_cutoff = 0.5,
  custom_color_map = selection_colors
)

# 3. Run DEA Engine
dea_res <- wrap_dea_pipeline(
    omics_mat           = olink_mat, 
    clin_df             = clin_extreme, 
    dea_conf            = extreme_dea_conf, 
    dea_id              = "extreme_vascular", 
    base_output_dir     = output_dir_data, 
    go_db               = NULL, 
    omics_config        = omics_config, 
    heatmap_config      = heatmap_config, 
    project_colors_func = get_project_colors
)

# HEATMAP 1: Local DEA (Extreme Phenotype)
cat("### Local DEA Heatmap\n\n")
if (dea_res$status == "success") {
  print(dea_res$plots$volcano)
  cat("\n\n<br>\n\n")
  
  hm_local <- run_heatmap_engine(
    mat = olink_mat, 
    clin = clin_extreme, 
    dea_res = dea_res$data$dea,
    title = "Local DEA (4v4 Manual)",
    n_top = 30, 
    split_by_group = FALSE,
    split_col = "Selection",
    custom_color_map = selection_colors
  )
  draw(hm_local)
  cat("\n\n")
} else {
  cat(sprintf("> <span style='color:red;'>**DEA Failed:** %s</span>\n\n", dea_res$error_msg))
}

# HEATMAP 2: Global DEA Signature
cat("### Global DEA Heatmap\n\n")
if (exists("global_vascular_dea") && !is.null(global_vascular_dea)) {
  hm_global <- run_heatmap_engine(
    mat = olink_mat, 
    clin = clin_extreme, 
    dea_res = global_vascular_dea,
    title = "Global Signature (Filtered to 8 Manual)",
    n_top = 30, 
    split_by_group = FALSE,
    split_col = "Selection",
    custom_color_map = selection_colors
  )
  draw(hm_global)
  cat("\n\n")
} else {
  cat("> Global DEA results not found. Skipping Global Heatmap.\n\n")
}

# HEATMAP 3: Custom Proteins of Interest
cat("### Custom POI Heatmap\n\n")
valid_poi <- intersect(proteins_of_interest, rownames(olink_mat))
if (length(valid_poi) > 1 && dea_res$status == "success") {
  
  poi_limma_df <- dea_res$data$dea %>% filter(Feature %in% valid_poi)
  
  poi_heatmap <- run_heatmap_engine(
    mat = olink_mat, 
    clin = clin_extreme, 
    dea_res = poi_limma_df,
    title = "Selected Proteins of Interest (4v4 Manual)",
    n_top = length(valid_poi), 
    split_by_group = FALSE,
    split_col = "Selection",
    show_top_n_title = FALSE,
    custom_color_map = selection_colors
  )
  draw(poi_heatmap)
  cat("\n\n")
} else {
  cat("> Insufficient Valid POIs to generate targeted heatmap.\n\n")
}
```

# Appendix
`r what_this_does("Appended the raw underlying R code from the custom omics engines to the HTML report's appendix to ensure complete analytical transparency and reproducibility.")`
```{r print_engine, code=readLines(here::here("code", "00_omics_engines.R")), eval=FALSE}

```

```{r complete}
cat("Script z02 Complete! \n")
```

